### 2015-08-24 Schloss_Cluster_PeerJ_2015.Rmd
Initial commit

---
title: "**The difference between precision and accuracy in assigning 16S rRNA gene sequences to operational taxonomic units**"
author: Patrick D. Schloss
date: '`r format(Sys.Date(), format="%B %d, %Y")`'
bibliography: references.bib
output:
  html_document:
    includes:
      in_header: header.tex
csl: peerj.csl
fontsize: 11pt
geometry: margin=1.0in
---

### 2015-09-26 code/run_vagc_clust.sh
Add vsearch AGC method

@@ -1,0 +1,14 @@
+#!bash
+
+FASTA=$1
+ROOT=$(echo $FASTA | sed 's/fasta/vagc/' | sed 's/.ng//')
+
+code/vsearch/vsearch --sizeout --derep_fulllength $FASTA --minseqlength 30 --threads 1 --uc $ROOT.sorted.uc --output $ROOT.sorted.fna --strand both --log $ROOT.sorted.log 
+
+code/vsearch/vsearch --maxaccepts 16 --usersort --id 0.97 --minseqlength 30 --wordlength 8 --uc $ROOT.clustered.uc --cluster_smallmem $ROOT.sorted.fna --maxrejects 64 --strand both --log $ROOT.clustered.log --sizeorder
+
+R -e "source('code/uc_to_list.R'); uc_to_list('$ROOT.sorted.uc', '$ROOT.clustered.uc')"
+
+rm $ROOT.sorted.uc $ROOT.sorted.fna $ROOT.sorted.log $ROOT.clustered.uc $ROOT.clustered.log
+
+

### 2015-09-28 code/summarize_mcc_ref.R
Add code to summarize mcc_ref data

@@ -1,0 +1,16 @@
+summarize_mcc_ref <- function(dataset){
+	methods <- c("an", "fn", "nn", "agc", "dgc", "closed", "open", "swarm", "vagc", "vdgc")
+
+	output_file_name <- paste0("data/process/", dataset, ".mcc_ref.summary")
+	write(x="mean\tlci\tuci\tfraction\tmethod", file=output_file_name)
+
+	for(m in methods){
+		file_name <- paste0("data/", dataset, "/", dataset, ".", m, ".ref_mcc")
+		ref_mcc <-read.table(file=file_name, header=T)
+		stats_mcc <- data.frame(t(apply(ref_mcc[,1:5], 2, function(x)c(mean(x), quantile(x, probs=c(0.025, 0.975))))))
+		stats_mcc$fraction <- gsub("X", "", rownames(stats_mcc))
+		stats_mcc$method <- rep(m, nrow(stats_mcc))
+		write.table(file=output_file_name, stats_mcc, col.names=F, row.names=F, append=T, quote=F)
+	}
+}
+

### 2015-09-28 Schloss_Cluster_PeerJ_2015.Rmd

@@
 ---
+
+## Introduction
+
+The ability to affordably generate millions of 16S rRNA gene sequences has allowed microbial ecologists to thoroughly characterize the microbial community composition of hundreds of samples.
+To simplify the complexity of these large datasets, it is helpful to cluster sequences into meaningful bins.
+These bins, known as operational taxonomic units (OTUs), are commonly used to compare the biodiversity contained within and between different samples.
+Such comparisons have enabled researchers to characterize the microbiota associated with the human body (REF), soil (REF), aquatic ecosystems (REF), and numerous other environments.
+Within the field of microbial ecology a convention has emerged where sequences are clustered into OTUs using a threshold of 3% dissimilarity or 97% similarity.
+One advantage of the OTU-based approach is that the definition of the bins is operational and can be changed to suit the needs of the particular project.
+However, with the dissemination of clustering algorithms within packages such as mothur and QIIME and stand-alone tools (REFS), it is important to understand how different clustering methods implement this conventional OTU threshold.
+Furthermore, it is necessary to understand how algorithm choice affects the precision and accuracy of assigning sequences to OTUs.
+Broadly speaking, three approaches have been developed to assign sequences to OTUs.
+
+The first approach has been referred to as phylotyping or closed reference clustering.
+This approach involves comparing sequences to a curated database and then clustering sequences together that are similar to the same reference sequence.
+Reference-based clustering methods suffer when the reference sequences do not reflect the composition of the community.
+If a large fraction of sequences are novel, then they cannot be assigned to an OTU.
+In addition, the reference sequences used in this application are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene.
+Thus, a sequence representing a fragment of the gene may be more than 97% similar to multiple reference sequences.
+Because of this, defining OTUs in the closed-reference approach is complicated because two sequences might be 97% similar to the same reference sequence; however, they may only be 94% similar to each other.
+The strength of the reference based approach is that the methods are generally fast, scaling linearly with the number of sequences being clustered.
+A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level in the Linnean taxonomic hierarchy.
+Ultimately, this approach is an example of supervised classification.
+
+The second approach has been referred to as *de novo* clustering.
+In this method, the similarity between sequences is used to cluster sequence rather than the similarity to a reference database.
+In contrast to the efficiency of closed-reference clustering, the speed of hierarchical *de novo* clustering algorithms scale quadratically with the number of unique sequences.
+The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences.
+If error rates can be reduced through stringent quality control measures then these problems can be overcome (REF).
+As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods.
+Previous comparisons of hierarchical and heuristic algorithms have shown that the average neighbor algorithm, a hierarchical algorithm outperforms the other *de novo* clustering algorithm.
+An early analysis using 16S rRNA gene sequence data generated using the 454 sequencing platform found that if these methods were applied to the same dataset using a subset of the sequences then a different number of OTUs were observed compared to clustering the full dataset and rarefying to the same number of sequences.
+Because the initial subsampling of the dataset was not repeated in the earlier analysis, it is unclear whether the result was product of the algorithm or sampling.
+One explanation could be that the clusters are getting better with additional sampling.
+Another observed problem with *de novo* clustering is that because it is necessary to break ties when assigning sequences to OTUs, clustering the same data multiple times may result in different clustering assignments.
+Whether the differences in assignments is meaningful is unclear; however the variation in results could represent equally valid clustering of the data.
+The strength of *de novo* clustering is its independence of references for carrying out the clustering step.
+After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU (ref).
+For this reason, *de novo* clustering has been preferred across the field.
+In contrast to closed-reference clustering, *de novo* clustering is an example of unsupervised clustering.
+
+The third approach is a hybrid of the closed-reference and *de novo* approaches.
+This approach has been called open-reference clustering.
+This method involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference.
+This method should exploit the strengths of both closed-referene and *de novo* clustering; however, the different OTU definitions generated by both approaches poses a possible problem when the methods are combined.
+An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels (REF).
+For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average neighbor algorithm.
+Those sequences that did not classify to a known family would also be clustered using the average neighbor algorithm.
+An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group (e.g.
+each family) is seen as being independent and can be processed separately.
+Such an approach would overcome the difficulty of mixing OTU definitions between the closed-reference and *de novo* approaches.
+
+The growth in options for assigning sequences using each of these three broad approaches has been considerable.
+It has been difficult to objectively assess the quality of OTU assignments.
+Some have focused on the time and memory required to process a dataset.
+These are valid parameters to assess when judging a clustering algorithm, but have little to say about the quality of the clustering.
+Others have attempted to judge the quality of an algorithm by its ability to generate data that parallels classification data.
+This approach is problematic because bacterial taxonomy often reflects the biases amongst bacterial systematicists and the rates of evolution across lineages are not the same.
+We recently proposed an approach for evaluating OTU assignments using the similarity between sequences (REF).
+Those sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives.
+Meanwhile, those sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives.
+Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and ratios using the Matthew's correlation coefficient (MCC; REF).
+That analysis focused on assessing *de novo* clustering algorithms and found that the average neighbor algorithm outperformed the other hierarchical and heuristic algorithms.
+
+A recent analysis by He and colleagues (REF) attempted to characterize the three general clustering approaches by focusing on what they called stability.
+They defined stability as the ability of an algorithm to provide the same clustering on a subset of the data as was found in the full dataset.
+Related to this, the authors expressed concern that because some algorithms use a random number generator to break ties there may be further instability between executions of algorithm.
+Their concept of stability fails to quantify for the accuracy of the clustering and instead focuses on the precision of the clustering.
+A method may be very precise, but low in accuracy.
+In the current analysis, we attempted to assess the accuracy and precision of the various clustering algorithms.
+Building on our previous analysis of clustering algorithms, our hypothesis was that the algorithms praised by the He study for their stability actually suffered a lack of accuracy.
+In addition, we assess these parameters in light of sequence quality using the original 454 dataset and a larger and more modern dataset generated using the MiSeq platform.
+
+## Results
+
+***Summary of He study.*** We sought to reproduce the more critical analyses performed in the He study.
+Similar to their study, we obtained the Canadian soil dataset from Roesch et al.
+(REF) and processed the sequences as described in their analysis.
+The first result we sought to reproduce was whether the clustering of the Roesch dataset using a subset of the data corresponded to the clustering observed with the full dataset.
+The He study found that when they used the open and closed-reference methods the subsampled clusters most closely resembled those of the full dataset when using the MCC.
+Among the *de novo* clustering methods they observed that the abundance-based greedy clustering (AGC) algorithm followed by single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) algorithms.
+They observed a broad range of MCC values among their AL replicates, which was not explained by the authors.
+The second result we sought to reproduce was the observation, originally made in the Roesch study, that rarefaction curves calculated using clusters obtained using a portion of the dataset did not overlap with rarefaction curves generated using clusters generated from the full dataset.
+In the He study, this appeared to be a more pronounced problem when using the CL, SL, and DGC algorithms.
+
+***Clustering approaches vary in their precision.*** We first calculated the MCC for for each of the clustering algorithms using 20, 40, 60, and 80% relative to the clusters formed by the algorithms using the full dataset.
+Because a random number generator is used in some of algorithms to break ties where pairs of sequences have the same distance between them, similar to the He study, we replicated each algorithm and subsample 30 times.
+
+***Effect of sampling depth on observed number of OTUs.***
+
+***Methods vary in their accuracy.***
+
+***Accuracy and precision affected by data quality.***
+
+***Deep sampling of 16S rRNA genes.***
+- V4 region not well represented in closed reference database
+
+***Closed-reference clustering also has a precision problem.***
+
+## Methods
+
+***Roesch dataset processed according to He study***
+After obtaining the 16S rRNA gene fragments from GenBank (accessions EF308591-EF361836), we followed the methods outlined by the He study by removing any sequence that contained an ambiguous base, was identified as being a chimera, and fell outside a defined sequence length.
+Although they reported observing a total of 50,542 sequences that were represented by 13,293 unique sequences, we obtained a total of `r length(scan("data/he/canada_soil.good.unique.pick.redundant.fasta", what="", sep="\n", quiet=T))/2` sequences that were represented by `r length(scan("data/he/canada_soil.good.unique.pick.fasta", what="", sep="\n", quiet=T))/2` unique sequences.
+Similar to the He study, we randomly sampled, without replacement, 20, 40, 60, and 80% of the sequences from the full data set.
+The random sampling was repeated 30 times.
+The order of the sequences in the full dataset was randomly permuted without replacement to generate an additional 30 datasets.
+For the hierarchical clustering algorithms and to generate a distance matrix, the pairwise distances between sequences were calculated in mothur using the pairwise.seqs command with the default Needleman-Wunsch alignment algorithm and parameters.
+Execution of the clustering algorithms was performed as described in the original He study using mothur and QIIME as appropriate.

### 2015-09-29 code/summarize_rarefaction.R
Make miseq analysis parallel others; summarize rarefaction data

@@ -1,0 +1,41 @@
+summarize_rarefaction <- function(folder){
+	methods <- c("an", "fn", "nn", "agc", "dgc", "closed", "open", "swarm")
+
+	output_file_name <- paste0("data/process/", folder, ".rarefaction.summary")
+	write(x="method\tfraction\tsobs\tsobs_lci\tsobs_uci\trarefied\trare_lci\trare_uci\tp_value", file=output_file_name)
+
+	for(m in methods){
+		print(m)
+		file_name <- paste0("data/", folder, "/", folder, ".", m, ".rarefaction")
+		rarefy <-read.table(file=file_name, header=T)
+		fractions <- unique(rarefy$fraction)
+
+		p <- numeric()
+		rare_mean <- numeric()
+		rare_lci <- numeric()
+		rare_uci <- numeric()
+		sobs_mean <- numeric()
+		sobs_lci <- numeric()
+		sobs_uci <- numeric()
+
+		for(f in fractions){
+			rarefy_sub <- rarefy[rarefy$f==f,]
+
+			sobs_mean[as.character(f)] <- mean(rarefy_sub$sobs)
+			sobs_lci[as.character(f)] <- quantile(rarefy_sub$sobs, prob=0.025)
+			sobs_uci[as.character(f)] <- quantile(rarefy_sub$sobs, prob=0.975)
+
+			rare_mean[as.character(f)] <- mean(rarefy_sub$rarefied)
+			rare_lci[as.character(f)] <- quantile(rarefy_sub$rarefied, prob=0.025)
+			rare_uci[as.character(f)] <- quantile(rarefy_sub$rarefied, prob=0.975)
+
+			if(f != 1 && m != "closed"){
+				p[as.character(f)] <- t.test(rarefy_sub$rarefied, rarefy_sub$sobs)$p.value
+			} else {
+				p[as.character(f)] <- NA
+			}
+		}
+		output <- cbind(rep(m, length(fractions)), fractions, sobs_mean, sobs_lci, sobs_uci, rare_mean, rare_lci, rare_uci, p)
+		write.table(file=output_file_name, output, col.names=F, row.names=F, append=T, quote=F, sep='\t')
+	}
+}

### 2015-09-29 Schloss_Cluster_PeerJ_2015.Rmd
Build figures 1 and 2

@@
 It has been difficult to objectively assess the quality of OTU assignments.
-Some have focused on the time and memory required to process a dataset.
+Some have focused on the time and memory required to process a dataset (REFS).
 These are valid parameters to assess when judging a clustering algorithm, but have little to say about the quality of the clustering.
@@
 Related to this, the authors expressed concern that because some algorithms use a random number generator to break ties there may be further instability between executions of algorithm.
-Their concept of stability fails to quantify for the accuracy of the clustering and instead focuses on the precision of the clustering.
+Their concept of stability does not account for the accuracy of the clustering and instead focuses on the precision of the clustering.
 A method may be very precise, but low in accuracy.
@@
 
-***Summary of He study.*** We sought to reproduce the more critical analyses performed in the He study.
+***Summary of He study.*** We sought to identify the more critical analyses performed in the He study.
 Similar to their study, we obtained the Canadian soil dataset from Roesch et al.
 (REF) and processed the sequences as described in their analysis.
-The first result we sought to reproduce was whether the clustering of the Roesch dataset using a subset of the data corresponded to the clustering observed with the full dataset.
-The He study found that when they used the open and closed-reference methods the subsampled clusters most closely resembled those of the full dataset when using the MCC.
+In our opinion there were three important tests.
+First, was to quantify whether the clustering observed for a subset of the data represented the same clustering that was found with the full dataset.
+The He study found that when they used the open and closed-reference methods clusters formed using the subset data most closely resembled those of the full dataset.
 Among the *de novo* clustering methods they observed that the abundance-based greedy clustering (AGC) algorithm followed by single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) algorithms.
 They observed a broad range of MCC values among their AL replicates, which was not explained by the authors.
-The second result we sought to reproduce was the observation, originally made in the Roesch study, that rarefaction curves calculated using clusters obtained using a portion of the dataset did not overlap with rarefaction curves generated using clusters generated from the full dataset.
-In the He study, this appeared to be a more pronounced problem when using the CL, SL, and DGC algorithms.
+Second, rarefaction curves calculated using clusters obtained using a portion of the dataset did not overlap with rarefaction curves generated using clusters generated from the full dataset.
+This result was originally observed in the Roesch study using the complete linkage algorithm and was reproduced in the He study where this result was more pronounced problem when using the CL, SL, and DGC algorithms relative to the other algorithms.
+Third, the authors attempted to describe the effects of the clustering instability on comparisons of communities.
+Such an analysis does not make sense for the *de novo* or open-reference approaches since the labels of the OTUs are not consistent between executions of the clustering algorithms.
+To overcome this, the authors selected representative sequences from OTUs and merged OTUs across different sized libraries to link OTUs.
+Regardless, the justification for this analysis is specious as the OTU assignments are based on the data available in the dataset when the sequences are clustered and comparing assignments in this manner are irreconcilable.
+We will attempt to replicate the first two observations.
 
-***Clustering approaches vary in their precision.*** We first calculated the MCC for for each of the clustering algorithms using 20, 40, 60, and 80% relative to the clusters formed by the algorithms using the full dataset.
+***Clustering approaches vary in their stability.*** We first sought to assess calculated the MCC for for each of the clustering algorithms using 20, 40, 60, and 80% relative to the clusters formed by the algorithms using the full dataset (Figure 1A).
 Because a random number generator is used in some of algorithms to break ties where pairs of sequences have the same distance between them, similar to the He study, we replicated each algorithm and subsample 30 times.
+Across these sequencing depths, we observed that the stability of the SL and CL algorithms were highly sensitive to sampling effort relative to the AL, AGC, and DGC algorithms (Figure 1A).
+Our results (Figure 1B) largely confirmed those of Figure 4C of He study with one notable exception.
+In the He study, when comparing the OTUs formed using the AL algorithm with 60% of the data, they observed a mean MCC value of approximately 0.63 (95% confidence interval between approximately 0.15 and 0.75).
+In contrast, we observed a mean value of `r round(z[z$method=="an" & z$fraction==0.6, "mean"], 2)` (95% confidence interval between `r round(z[z$method=="an" & z$fraction==0.6, "lci"], 2)` and `r round(z[z$method=="an" & z$fraction==0.6, "uci"], 2)`).
+This result suggests that the AL algorithm was far more stabile than indicated in the He study.
 
-***Effect of sampling depth on observed number of OTUs.***
+***Effect of sampling depth on observed number of OTUs.*** Next, we sought to determine whether the number of OTUs observed when clustering a subset of the data was the same as that observed when rarefying the full dataset back to the number of sequences in the subset.
+The analyses from the He and Roesch studies both compared the number of observed OTUs from the subset to the average number of expected OTUs from the full dataset instead of to the 95% confidence interval of what would be expected from the full dataset.
 
@@
 
-***Accuracy and precision affected by data quality.***
+***Stability, accuracy, and precision affected by data quality.***
 

### 2015-09-29 code/build_figure1.R

@@ -1,0 +1,62 @@
+build_ref_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
+
+	data <- read.table(file=paste0("data/process/", dataset, ".mcc_ref.summary"), header=T, stringsAsFactors=F)
+
+	methods <- c("fn", "nn", "an", "dgc", "agc", "closed", "open")
+	n_methods <- length(methods)
+
+	orig_methods <- data[data$method %in% methods,]
+
+	pdf(file=output_file_name, width=7.5, height=3.5)
+	layout(matrix(c(1,2), nrow=1), widths=c(4, 3.5))
+
+	par(mar=c(5.5, 5, 0.5, 0.5))
+
+	plot(NA, ylim=c(0.5,1), xlim=c(0.1,1), axes=F, ylab="", xlab="")
+	clrs <- rainbow(n_methods)
+	abline(v=threshold, col="gray", lwd=2)
+
+	for(m in 1:n_methods){
+		subset <- orig_methods[orig_methods$method==methods[m],]
+		points(subset$mean~subset$fraction, type="l", col=clrs[m], lwd=2)
+	}
+	axis(1)
+	axis(2, las=1)
+	box()
+
+	text(x=0.12, y=0.98, label="A", cex=1.5, font=2)
+
+	title(ylab="MCC value relative to\nfull dataset")
+
+
+
+
+
+	par(mar=c(5.5, 0.5, 0.5, 0.5))
+
+	sixty <- orig_methods[orig_methods$fraction==threshold,]
+	sixty$method <- factor(sixty$method, levels=methods)
+	sixty[sixty$method=="closed", c("lci", "uci")] <- c(0.999, 1.001)
+	plot(NA, ylim=c(0.5,1), xlim=c(0.75,7.25), axes=F, ylab="", xlab="")
+
+	for(m in 1:n_methods){
+		points(x=m, y=sixty[sixty$method==methods[m],"mean"], pch=19, col=clrs[m])
+
+		arrows(x0=m, y0=sixty[sixty$method==methods[m],"mean"], x1=m, y1=sixty[sixty$method==methods[m],"lci"], lwd=2, angle=90, length=0.1, col=clrs[m])
+		arrows(x0=m, y0=sixty[sixty$method==methods[m],"mean"], x1=m, y1=sixty[sixty$method==methods[m],"uci"], lwd=2, angle=90, length=0.1, col=clrs[m])
+	}
+
[…]

### 2015-09-29 code/build_figure2.R

@@ -1,0 +1,48 @@
+build_figure2 <- function(dataset, output_file_name){
+	rare <- read.table(file=paste0("data/process/", dataset, ".rarefaction.summary"), header=T, stringsAsFactors=T)
+
+	offset <- 0.1
+
+	n_fractions <- length(levels(factor(rare$fraction)))
+
+	methods <- c("fn", "nn", "an", "dgc", "agc", "closed", "open")
+	n_methods <- length(methods)
+
+	clrs <- rainbow(n_fractions)
+
+	orig_par <- par()
+
+	pdf(file=output_file_name, width=4, height=4)
+	par(mar=c(5,5,0.5,0.5))
+	plot(NA, xlim=c(0.8,length(methods)+0.2), ylim=c(0,max(rare$rare_uci)), axes=F, xlab="", ylab="")
+	par(cex=0.8)
+	for(m in 1:n_methods){
+
+		subset <- rare[rare$method==methods[m],]
+
+	#	the 95% confidence intervals were smaller or similar in size to the cex=1 plotting symbol
+	#	arrows(x0=m-offset, y0=subset[,"sobs"], y1=subset[,"sobs_lci"], col=clrs, length=0.05, angle=90)
+	#	arrows(x0=m-offset, y0=subset[,"sobs"], y1=subset[,"sobs_uci"], col=clrs, length=0.05, angle=90)
+		points(rep(m-offset, n_fractions), subset[,"sobs"], col=clrs, pch=19, cex=1.0)
+
+	#	the 95% confidence intervals were smaller or similar in size to the cex=1 plotting symbol
+	#	arrows(x0=m+offset, y0=subset[,"rarefied"], y1=subset[,"rare_lci"], col=clrs, length=0.05, angle=90)
+	#	arrows(x0=m+offset, y0=subset[,"rarefied"], y1=subset[,"rare_uci"], col=clrs, length=0.05, angle=90)
+		points(rep(m+offset, n_fractions), subset[,"rarefied"], col=clrs, pch=21, cex=1.0)
+	}
+
+	method_labels <- c("CL", "SL", "AL", "DGC", "AGC", "Closed-ref", "Open-ref")
+	axis(1, at=1:n_methods, label=method_labels, las=2)
+	axis(2, las=2, at=seq(0,6000,1000), label=c("0", "1,000", "2,000", "3,000", "4,000", "5,000", "6,000"))
+	box()
+
[…]

### 2015-09-29 code/summarize_rarefaction.R
Add vsearch methods

@@ -1,5 +1,5 @@
-summarize_rarefaction <- function(folder){
-	methods <- c("an", "fn", "nn", "agc", "dgc", "closed", "open", "swarm")
+summarize_rarefaction <- function(dataset){
+	methods <- c("an", "fn", "nn", "agc", "dgc", "closed", "open", "swarm", "vagc", "vdgc")
 
 	output_file_name <- paste0("data/process/", folder, ".rarefaction.summary")
 	write(x="method\tfraction\tsobs\tsobs_lci\tsobs_uci\trarefied\trare_lci\trare_uci\tp_value", file=output_file_name)

### 2015-09-29 code/summarize_mcc.R
Summarize MCC methods

@@ -1,0 +1,17 @@
+summarize_mcc <- function(dataset){
+
+	methods <- c("an", "fn", "nn", "agc", "dgc", "closed", "open", "swarm", "vagc", "vdgc")
+
+	output_file_name <- paste0("data/process/", dataset, ".mcc.summary")
+	write(x="mean\tlci\tuci\tfraction\tmethod", file=output_file_name)
+
+	for(m in methods){
+		file_name <- paste0("data/", dataset, "/", dataset, ".", m, ".pool_sensspec")
+		mcc_data <-read.table(file=file_name, header=T)
+		mcc_stats_table <- aggregate(mcc_data$mcc, by=list(mcc_data$fraction), function(x)c(mean=mean(x), quantile(x, probs=c(0.025, 0.975))))
+
+		mcc_summary <- cbind(mcc_stats_table$x, mcc_stats_table$Group, rep(m, nrow(mcc_stats_table)))
+		write.table(file=output_file_name, mcc_summary, col.names=F, row.names=F, append=T, quote=F)
+	}
+
+}

### 2015-09-29 code/build_figure3.R
Create figure 3 describing MCC of the clusters

@@ -1,6 +1,6 @@
-build_ref_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
+build_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
 
-	data <- read.table(file=paste0("data/process/", dataset, ".mcc_ref.summary"), header=T, stringsAsFactors=F)
+	data <- read.table(file=paste0("data/process/", dataset, ".mcc.summary"), header=T, stringsAsFactors=F)
 
 	methods <- c("fn", "nn", "an", "dgc", "agc", "closed", "open")
 	n_methods <- length(methods)
@@ -12,7 +12,7 @@ build_ref_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
 
 	par(mar=c(5.5, 5, 0.5, 0.5))
 
-	plot(NA, ylim=c(0.5,1), xlim=c(0.1,1), axes=F, ylab="", xlab="")
+	plot(NA, ylim=c(0,1), xlim=c(0.1,1), axes=F, ylab="", xlab="")
 	clrs <- rainbow(n_methods)
 	abline(v=threshold, col="gray", lwd=2)
 
@@ -36,8 +36,8 @@ build_ref_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
 
 	sixty <- orig_methods[orig_methods$fraction==threshold,]
 	sixty$method <- factor(sixty$method, levels=methods)
-	sixty[sixty$method=="closed", c("lci", "uci")] <- c(0.999, 1.001)
-	plot(NA, ylim=c(0.5,1), xlim=c(0.75,7.25), axes=F, ylab="", xlab="")
+
+	plot(NA, ylim=c(0.0,1), xlim=c(0.75,7.25), axes=F, ylab="", xlab="")
 
 	for(m in 1:n_methods){
 		points(x=m, y=sixty[sixty$method==methods[m],"mean"], pch=19, col=clrs[m])
@@ -49,8 +49,6 @@ build_ref_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
 	axis(1, at=1:7, label=c("CL", "SL", "AL", "DGC", "AGC", "Closed-ref", "Open-ref"), las=2)
 	#axis(2, las=1)
 	box()
-
-#	schloss <- read.table(file="data/process/schloss.mcc_ref.summary", header=T, stringsAsFactors=F)
 
 	text(x=0.87, y=0.98, label="B", cex=1.5, font=2)
 

### 2015-09-29 code/summarize_rarefaction.R
Build first three figures

@@ -1,12 +1,12 @@
 summarize_rarefaction <- function(dataset){
 	methods <- c("an", "fn", "nn", "agc", "dgc", "closed", "open", "swarm", "vagc", "vdgc")
 
-	output_file_name <- paste0("data/process/", folder, ".rarefaction.summary")
+	output_file_name <- paste0("data/process/", dataset, ".rarefaction.summary")
 	write(x="method\tfraction\tsobs\tsobs_lci\tsobs_uci\trarefied\trare_lci\trare_uci\tp_value", file=output_file_name)
 
 	for(m in methods){
 		print(m)
-		file_name <- paste0("data/", folder, "/", folder, ".", m, ".rarefaction")
+		file_name <- paste0("data/", dataset, "/", dataset, ".", m, ".rarefaction")
 		rarefy <-read.table(file=file_name, header=T)
 		fractions <- unique(rarefy$fraction)
 

### 2015-09-29 Schloss_Cluster_PeerJ_2015.Rmd
Working on R&D section

@@
 ---
-title: "**The difference between precision and accuracy in assigning 16S rRNA gene sequences to operational taxonomic units**"
+title: "**Accuracy and precision of 16S rRNA gene sequence assignments to OTUs is more important than the stability of OTUs**"
 author: Patrick D. Schloss
@@
 
-## Results
+## Results and Discussion
 
-***Summary of He study.*** We sought to identify the more critical analyses performed in the He study.
+***Summary and replication of He study.*** We sought to identify the more critical analyses performed in the He study.
 Similar to their study, we obtained the Canadian soil dataset from Roesch et al.
@@
 In our opinion there were three important tests.
-First, was to quantify whether the clustering observed for a subset of the data represented the same clustering that was found with the full dataset.
+
+First, we sought to quantify whether the clustering observed for a subset of the data represented the same clustering that was found with the full dataset.
 The He study found that when they used the open and closed-reference methods clusters formed using the subset data most closely resembled those of the full dataset.
@@
 They observed a broad range of MCC values among their AL replicates, which was not explained by the authors.
-Second, rarefaction curves calculated using clusters obtained using a portion of the dataset did not overlap with rarefaction curves generated using clusters generated from the full dataset.
-This result was originally observed in the Roesch study using the complete linkage algorithm and was reproduced in the He study where this result was more pronounced problem when using the CL, SL, and DGC algorithms relative to the other algorithms.
-Third, the authors attempted to describe the effects of the clustering instability on comparisons of communities.
-Such an analysis does not make sense for the *de novo* or open-reference approaches since the labels of the OTUs are not consistent between executions of the clustering algorithms.
-To overcome this, the authors selected representative sequences from OTUs and merged OTUs across different sized libraries to link OTUs.
-Regardless, the justification for this analysis is specious as the OTU assignments are based on the data available in the dataset when the sequences are clustered and comparing assignments in this manner are irreconcilable.
-We will attempt to replicate the first two observations.
-
-***Clustering approaches vary in their stability.*** We first sought to assess calculated the MCC for for each of the clustering algorithms using 20, 40, 60, and 80% relative to the clusters formed by the algorithms using the full dataset (Figure 1A).
+We first sought to assess calculated the MCC for for each of the clustering algorithms using 20, 40, 60, and 80% relative to the clusters formed by the algorithms using the full dataset (Figure 1A).
 Because a random number generator is used in some of algorithms to break ties where pairs of sequences have the same distance between them, similar to the He study, we replicated each algorithm and subsample 30 times.
@@
 This result suggests that the AL algorithm was far more stabile than indicated in the He study.
+Regardless, it supports the assertion that the clustering observed for the subset of the data does not represent the same clustering that was found with the full dataset; however, the significance of these differences is unclear.
 
-***Effect of sampling depth on observed number of OTUs.*** Next, we sought to determine whether the number of OTUs observed when clustering a subset of the data was the same as that observed when rarefying the full dataset back to the number of sequences in the subset.
-The analyses from the He and Roesch studies both compared the number of observed OTUs from the subset to the average number of expected OTUs from the full dataset instead of to the 95% confidence interval of what would be expected from the full dataset.
+Second, rarefaction curves calculated using clusters obtained using a portion of the dataset did not overlap with rarefaction curves generated using clusters generated from the full dataset.
+This result was originally observed in the Roesch study using the complete linkage algorithm and was reproduced in the He study where this result was more pronounced problem when using the CL, SL, and DGC algorithms relative to the other algorithms.
+The He and Roesch studies both found that the CL algorithm produced fewer OTUs in the subset than in the rarefied data.
+Expanding the analysis to other algorithms, the He study found that the SL algorithm produced more OTUs, the AGC produced fewer, and the other algorithms produced similar numbers of OTUs than expected when comparing the subsetted data to the rarefied data.
+Our results support those of these previous studies (Figure 2).
+It was clear that inter-method differences were generally more pronounced than the differences observed between rarefying from the full dataset and from clustering the subsetted data.
+The number of OTUs observed was largest using the CL algorithm, followed by the open-reference algorithm.
+The AL, AGC, and DGC algorithms all provided comparable numbers of OTUs.
+Finally, the closed-referene and SL algorithms generated the fewest number of OTUs.
 
-***Methods vary in their accuracy.***
+Third, the authors attempted to describe the effects of the clustering instability on comparisons of communities.
+They used Adonis to test whether the community structure represented in subsetted communities resembled that of the full dataset.
+Inspection of these results indicates that although they were able to detect significant p-values, they were were of marginal biological significance.
+Adonis R statistics close to zero indicate the community structures from the full and subsetted datasets overlapped while values of one indicate the communities are completely different.
+The He study observed adonis R statistics of 0.02 (closed-reference), 0.03 (open-reference), 0.07 (CL, AGC, DGC), and 0.16 (SL and AL).
+Regardless of the statistical or biological significance of these results, the analysis does not make sense since the *de novo* and open-reference approaches do not consistently label the OTUs that sequences belong to when the clustering algorithms are run multiple times with different random number seeds.
+To overcome this, the authors selected representative sequences from OTUs and merged OTUs across different sized libraries to link OTUs.
+The justification for this analysis is specious as the OTU assignments are based on the data available in the dataset when the sequences are clustered and comparing assignments in this manner are irreconcilable.
+It is not surprising that the only analysis that did not provide a significant p-value was for the closed-reference analysis, which is the only analysis that provides consistent OTU labels.
+Because this analysis was so poorly designed, we did not seek to reproduce it.
 
-***Stability, accuracy, and precision affected by data quality.***
+It is worth noting that the entire design of the He study is artificial.
+First, their analysis was based on a single sample.
+Researchers generally have dozens or hundreds of samples that are pooled and clustered together to enable comparison across samples.
+Second, all of the sequence data from these datasets is pooled for a single analysis.
+No one would ever perform an analysis based on a subset of their data.
+Because of these points, the value of identifying stable OTUs is unclear.
 
-***Deep sampling of 16S rRNA genes.***
-- V4 region not well represented in closed reference database
+***Methods vary in their accuracy.*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs.
+A method can generate highly stable OTUs, but the OTUs may be meaningless by poorly representing a specified cutoff in the assignment of sequences to those OTUs.
+To assess the accuracy and precision of the various methods, we made use of the pairwise distance between the unique sequences to count the number of true positives and negatives and the number of false positives and negatives for each method and sampling depth so that we could calculate the average MCC value as a measure of a method's accuracy and its variation as a measure of its precision.
+We made three important observations.
+First, each of the *de novo* methods varied in how sensitive their MCC values were to additional sequences (Figure 3A).
+The SL and CL algorithms were the most sensitive; however, the MCC values using 80 and 100% of the data did not vary meaningfully when using *de novo* algorithms.
+Second, the AL algorithm out-performed the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL.
+Third, with the possible exception of the CL algorithm, the MCC values for each of the only demonstrated a small amount of variation between runs of the algorithm with a different ordering of the input sequences.
+This indicates that although there may be variation between executions of the same algorithm, they produce OTU assignments that are equally good.
+Revisiting the concept of stability, we question the value of obtaining stable OTUs when the the full dataset is not optimally assigned to OTUs.
+Our analysis indicates that the best algorithm for assigning the Canadian soils sequences to OTUs using a 97% threshold is the AL algorithm.
+
+***Deep sampling of 16S rRNA genes.*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering algorithms.
+First, the Canadian soil dataset that the He study used to analyze the stability of OTUs is one of the earliest 16S rRNA gene sequence datasets published using the 454 FLX platform.
+Developments in sequencing technology now permits the sequencing of millions of sequences for a study.
+In addition, because the original phred quality scores and flowgram data are not available, it was not possible to adequately remove sequencing errors (REFS).
+The large number of sequences that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering algorithms.
+Second, the dataset used in the He study covered the V9 region of the 16S rRNA gene.
+For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference algorithms.
+Of the XXXXXX sequences in the default QIIME database, only XXXXXX fully cover the V9 region.
+Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene.
+With these points in mind, we compared the clustering assignment for each of these algorithms using a time series experiment that was obtained using mouse stool (REFS).
+The MiSeq platform was used to generate XX,XXX,XXX sequences from the V4 region of the 16S rRNA gene of XXX samples.
+Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02%.
+Although no dataset is perfect for exhaustively testing these clustering algorithms, this dataset was useful for demonstrating several points.
 
 ***Closed-reference clustering also has a precision problem.***
+- V9 region not well represented in closed reference database
+
+***Evalution of an open-source alternative to USEARCH.***
 

### 2015-09-30 Schloss_Cluster_PeerJ_2015.Rmd
Working on R&D section

@@
 
+```{r, cache=TRUE}
+n_miseq_seqs <- length(scan("data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.redundant.fix.fasta", what="", sep="\n", quiet=T))/2
+```
+
 ***Deep sampling of 16S rRNA genes.*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering algorithms.
@@
 With these points in mind, we compared the clustering assignment for each of these algorithms using a time series experiment that was obtained using mouse stool (REFS).
-The MiSeq platform was used to generate XX,XXX,XXX sequences from the V4 region of the 16S rRNA gene of XXX samples.
+The MiSeq platform was used to generate `r ` sequences from the V4 region of the 16S rRNA gene of XXX samples.
 Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02%.
@@
 
-***Closed-reference clustering also has a precision problem.***
-- V9 region not well represented in closed reference database
+[   Something, something, something.    ]
+
+```{r rand_ref_analysis}
+rand_ref_u <- read.table(file="data/rand_ref/hits.uclosed.summary", row.names=1)
+rand_ref_v <- read.table(file="data/rand_ref/hits.vclosed.summary", row.names=1)
+
+```
+
+***Additional problems with reference-based clustering.*** The He study indicated that the closed-reference algorithm generated perfectly stable OTUs.
+This was unsurprising since, by definition, the algorithm represents a one-to-one mapping of reads to a reference and treats the input sequences independently.
+An important test that was not performed in the He study was to determine whether the clustering was sensitive to the order of the sequences in the database.
+We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference algorithm with the full MiSeq dataset.
+Surprisingly, we observed that the number of OTUs generated was not the same in each of the randomizations.
+On average there were `r round(rand_ref_u["mean", 1], 2)` OTUs per randomization (range from `r round(rand_ref["min", 1], 2)` to `r round(rand_ref_u["max", 1], 2)`).
+The unrandomized reference resulted in `r rand_ref_u["original", 1]` OTUs, below the minimum observed number of OTUs when the references were randomized.
+This surprising result is likely because a heuristic is employed by USEARCH to determine when to stop searching the database such that the best possible match is not necessarily returned.
+If the order of the reference sequences differed between runs, then the heuristic criteria would be triggered at different points.
+To test this, we substituted VSEARCH for USEARCH in the closed-reference algorithm because VSEARCH does not utilize the heuristic but is otherwise modeled after USEARCH.
+In fact, all randomizations of the reference sequences and the original ordering of the reference sequences generated `r rand_ref_v["original", 1]` OTUs.
+
+```{r, cache=TRUE}
+v4_dist <- read.table(file="data/gg_13_8/gg_13_8_97.v4_ref.unique.dist")
+n_small_v4_dist <- sum(v4_dist$V3<=0.03)
+
+v19_dist <- read.table(file="data/gg_13_8/gg_13_8_97.v19_ref.unique.dist")
+n_small_v19_dist <- sum(v19_dist$V3<=0.03)
+
+v19_summary <- read.table(file="data/gg_13_8/gg_13_8_97.v19.summary", header=T)
+n_v19_good <- sum(v19_summary$start <= 3967 & v19_summary$end >= 6116)
+n_v4_good <- length(scan("data/gg_13_8/gg_13_8_97.v4_ref.names", what="", sep="\n", quiet=T))
+
+v_overlap <- read.table(file="data/rand_ref/overlap.vclosed.summary")
+```
+
+We noticed that even with VSEARCH, the reference OTU label that was assigned to each OTU differed even though the composition of the OTUs were the same across the randomized references.
+The default QIIME database that was used in the He study and by others contains full-length sequences that are at most 97% similar to each other.
+We suspect that the reference database was designed this way to overcome the limitations of the heuristic employed in USEARCH.
+This curated database is used for mapping sequences to a reference phylogenetic tree for carrying out UniFrac-based analyses (REFS) and to reference genomes for performing analyses such as picrust (REFS).
+Because the downstream applications depend on the correct and unique labeling of the OTUs, the lack of stability of the labeling was a concern.
+We hypothesized that within a given region  there would be sequences that were more similar to each other and possibly identical to each other.
+There were a total of `r nrow(v19)` sequences in the reference database and `r n_v19_good` of those sequences fully overlap the V4 region.
+Among those sequences that fully overlapped the V4 region, there were `r n_small_v19_dist` pairs of sequences that were more than 97% similar to each other over the full length of the 16S rRNA gene.
+When the extracted V4 sequences were dereplicated, we identified `r n_v19_good-n_v4_good` sequences that were identical to each other.
+Among the dereplicated V4 sequences there were `r n_small_v4_dist` pairs of sequences that were more than 97% similar to each other.
+The presence of duplicate V4 reference sequences explains the lack of labeling stability when using either USEARCH or VSEARCH to carry out the closed-reference algorithm.
+Using VSEARCH to perform the closed-reference clustering, an average of `r round(100*v_overlap["orig", "mean"], 1)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*v_overlap["orig", "min"], 1)` to `r round(100*v_overlap["orig", "max"], 1)`).
+
+```{r duplicate_analysis}
+dups <- read.table(file="data/gg_13_8/duplicate.analysis")
+max_taxa <- dups[dups$V3 == max(dups$V3),]
+```
+
+We next asked whether the lack of consistency in labeling would affect the taxonomic identification of the sequences.
+The `r sum(dups$V2)-nrow(dups)` duplicated V4 sequences represented `r nrow(dups)` unique sequences.
+Among the `r sum(dups$V2==2)` reference sequences that had one duplicate, `r sum(dups$V2==2 & dups$V3==2)` had discordant taxonomies.
+Among those `r sum(dups$V2 > 2)` sequences with two or more duplicates, `r sum(dups$V2>2 & dups$V3>=2)` had discordant taxonomies.
+Two sequences mapped to `r max_taxa[1,2]-1` and `r max_taxa[2,2]-1` duplicate sequences and both contained `r max_taxa[1,3]` different taxonomies.
+There was also a sequence had `r max(dups[,2]])-1` duplicates and contained `r dups[which.max(dups[,2]),3]` different taxonomies.
 

### 2015-10-01 Schloss_Cluster_PeerJ_2015.Rmd
Editing

@@
 output:
-  html_document:
+  pdf_document:
     includes:
@@
 ---
+
+```{r, echo=F}
+options(scipen=999)
+```
 
@@
 ## Results and Discussion
+```{r}
+z <- read.table(file="data/process/he.mcc_ref.summary", header=T)
+```
 
@@
 Surprisingly, we observed that the number of OTUs generated was not the same in each of the randomizations.
-On average there were `r round(rand_ref_u["mean", 1], 2)` OTUs per randomization (range from `r round(rand_ref["min", 1], 2)` to `r round(rand_ref_u["max", 1], 2)`).
+On average there were `r round(rand_ref_u["mean", 1], 2)` OTUs per randomization (range from `r round(rand_ref_u["min", 1], 2)` to `r round(rand_ref_u["max", 1], 2)`).
 The unrandomized reference resulted in `r rand_ref_u["original", 1]` OTUs, below the minimum observed number of OTUs when the references were randomized.
@@
 We hypothesized that within a given region  there would be sequences that were more similar to each other and possibly identical to each other.
-There were a total of `r nrow(v19)` sequences in the reference database and `r n_v19_good` of those sequences fully overlap the V4 region.
+There were a total of `r nrow(v19_summary)` sequences in the reference database and `r n_v19_good` of those sequences fully overlap the V4 region.
 Among those sequences that fully overlapped the V4 region, there were `r n_small_v19_dist` pairs of sequences that were more than 97% similar to each other over the full length of the 16S rRNA gene.
@@
 Two sequences mapped to `r max_taxa[1,2]-1` and `r max_taxa[2,2]-1` duplicate sequences and both contained `r max_taxa[1,3]` different taxonomies.
-There was also a sequence had `r max(dups[,2]])-1` duplicates and contained `r dups[which.max(dups[,2]),3]` different taxonomies.
+There was also a sequence had `r max(dups[,2])-1` duplicates and contained `r dups[which.max(dups[,2]),3]` different taxonomies.
 

### 2015-10-01 Schloss_Cluster_PeerJ_2015.Rmd
Editing

--- 2015-09-30 Schloss_Cluster_PeerJ_2015.Rmd
@@
 ---
+
+```{r knitr.settings, eval=TRUE, echo=TRUE, cache=TRUE}
+opts_chunk$set("fig.path"="figure/")
+opts_chunk$set("fig.align"="center")
+opts_chunk$set("dev" = c("png", "pdf"))
+
+opts_chunk$set("tidy" = TRUE)
+
+opts_chunk$set("echo" = TRUE)
+opts_chunk$set("eval" = TRUE)
+opts_chunk$set("warning" = FALSE)
+opts_chunk$set("cache" = FALSE)
+```
 
@@
 Regardless of the statistical or biological significance of these results, the analysis does not make sense since the *de novo* and open-reference approaches do not consistently label the OTUs that sequences belong to when the clustering algorithms are run multiple times with different random number seeds.
-To overcome this, the authors selected representative sequences from OTUs and merged OTUs across different sized libraries to link OTUs.
+To overcome this, the authors selected representative sequences from each OTU and used those representative sequences to link OTU assignments between the different sized sequence sets.
 The justification for this analysis is specious as the OTU assignments are based on the data available in the dataset when the sequences are clustered and comparing assignments in this manner are irreconcilable.
@@
 rand_ref_v <- read.table(file="data/rand_ref/hits.vclosed.summary", row.names=1)
-
 ```
 
-***Additional problems with reference-based clustering.*** The He study indicated that the closed-reference algorithm generated perfectly stable OTUs.
+***; reference-based clustering.*** The He study indicated that the closed-reference algorithm generated perfectly stable OTUs.
 This was unsurprising since, by definition, the algorithm represents a one-to-one mapping of reads to a reference and treats the input sequences independently.
@@
 
-***Roesch dataset processed according to He study***
+***Roesch Canadian soil dataset***
 After obtaining the 16S rRNA gene fragments from GenBank (accessions EF308591-EF361836), we followed the methods outlined by the He study by removing any sequence that contained an ambiguous base, was identified as being a chimera, and fell outside a defined sequence length.
@@
 For the hierarchical clustering algorithms and to generate a distance matrix, the pairwise distances between sequences were calculated in mothur using the pairwise.seqs command with the default Needleman-Wunsch alignment algorithm and parameters.
-Execution of the clustering algorithms was performed as described in the original He study using mothur and QIIME as appropriate.
+Execution of the clustering algorithms was performed as described in the original He study using mothur (v.1.37) and QIIME (v.1.XX) as appropriate.

### 2015-10-01 Schloss_Cluster_PeerJ_2015.Rmd

--- 2015-10-01 Schloss_Cluster_PeerJ_2015.Rmd
@@
 
-```{r, echo=F}
+```{r knitr.settings, eval=TRUE, echo=FALSE, cache=TRUE}
+opts_chunk$set("tidy" = TRUE)
+opts_chunk$set("echo" = TRUE)
+opts_chunk$set("eval" = TRUE)
+opts_chunk$set("warning" = FALSE)
+opts_chunk$set("cache" = FALSE)
+
 options(scipen=999)
@@
 rand_ref_v <- read.table(file="data/rand_ref/hits.vclosed.summary", row.names=1)
-
 ```

### 2015-10-01 Schloss_Cluster_PeerJ_2015.Rmd
Working on R&D section

@@
 ```{r, cache=TRUE}
-n_miseq_seqs <- length(scan("data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.redundant.fix.fasta", what="", sep="\n", quiet=T))/2
+gg_summary <- read.table(file="data/gg_13_8/gg_13_8_97.v19.summary", header=T)
+n_v4_overlap <- sum(gg_summary$start <= 3967 & gg_summary$end >= 6116)
+n_v9_overlap <- sum(gg_summary$start <= 10179 & gg_summary$end >= 10683)
+
+n_miseq_seqs <- nrow(read.table(file="data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.redundant.summary"))
+
+n_unique_miseq_seqs <- nrow(read.table(file="data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.summary"))
+
+n_groups <- ncol(read.table(file="data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.denovo.uchime.pick.pick.count_table")) - 2 #subtract 2 to remove the sequence name and total number of sequences
 ```
@@
 For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference algorithms.
-Of the XXXXXX sequences in the default QIIME database, only XXXXXX fully cover the V9 region.
+Of the `r nrow(gg_summary)` sequences in the default QIIME database, only `r n_v4_overlap` fully cover the V9 region.
+In contrast, `r n_v4_overlap` of the sequences fully covered the V4 region.
+Inadequate coverage of the V9 region would adversely affect the ability of the reference-based methods to assign sequences to OTUs.
 Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene.
 With these points in mind, we compared the clustering assignment for each of these algorithms using a time series experiment that was obtained using mouse stool (REFS).
-The MiSeq platform was used to generate `r ` sequences from the V4 region of the 16S rRNA gene of XXX samples.
-Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02%.
+The MiSeq platform was used to generate `r n_miseq_seqs` sequences from the V4 region of the 16S rRNA gene of `r n_groups` samples.
+Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02% (REFS).
 Although no dataset is perfect for exhaustively testing these clustering algorithms, this dataset was useful for demonstrating several points.
 
+[   Something, something, something.    ]
+
+***Evalution of an open-source alternative to USEARCH.*** The AGC and DGC algorithms appear to perform as well and possibly than the hierarchical clustering algorithms for some datasets.
+These algorithms utilize the USEARCH program (REF), which is not available as open source code and is only available for free to academic users as a 32-bit program.
+Access for non-academic users and those needing the 64-bit version is available commercially from the developer.
+An alternative to USEARCH is VSEARCH, which is being developed in parallel to USEARCH as a open-source alternative.
+One subtle difference between the two programs is that USEARCH employs a heuristic to determine when to stop searching a database.
+This means that the best possible match is not necessarily returned once a highly similar hit is obtained.
+In contrast, VSEARCH does not utilize the heuristic, which the VSEARCH developers claim enhances the sensitivity relative to USEARCH.
+Using the two datasets, we determined whether the AGC and DGC algorithms as implemented by the two methods yielded sequencing assignments of similar quality.
+
+ This also explains the variation observed in the AGC and DGC algorithms, which rely upon the USEARCH algorithm.
 [   Something, something, something.    ]
@@
 
-***Additional problems with reference-based clustering.*** The He study indicated that the closed-reference algorithm generated perfectly stable OTUs.
+***Problems with reference-based clustering.*** The He study indicated that the closed-reference algorithm generated perfectly stable OTUs.
 This was unsurprising since, by definition, the algorithm represents a one-to-one mapping of reads to a reference and treats the input sequences independently.
 An important test that was not performed in the He study was to determine whether the clustering was sensitive to the order of the sequences in the database.
-We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference algorithm with the full MiSeq dataset.
+The default QIIME database, which was used in the He study and by others, contains full-length sequences that are at most 97% similar to each other.
+We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference algorithm with the full MiSeq dataset, which contained `r n_unique_miseq_seqs` unique sequences.
 Surprisingly, we observed that the number of OTUs generated was not the same in each of the randomizations.
-On average there were `r round(rand_ref_u["mean", 1], 2)` OTUs per randomization (range from `r round(rand_ref_u["min", 1], 2)` to `r round(rand_ref_u["max", 1], 2)`).
-The unrandomized reference resulted in `r rand_ref_u["original", 1]` OTUs, below the minimum observed number of OTUs when the references were randomized.
-This surprising result is likely because a heuristic is employed by USEARCH to determine when to stop searching the database such that the best possible match is not necessarily returned.
-If the order of the reference sequences differed between runs, then the heuristic criteria would be triggered at different points.
-To test this, we substituted VSEARCH for USEARCH in the closed-reference algorithm because VSEARCH does not utilize the heuristic but is otherwise modeled after USEARCH.
-In fact, all randomizations of the reference sequences and the original ordering of the reference sequences generated `r rand_ref_v["original", 1]` OTUs.
+On average there were `r round(rand_ref_u["mean", 1], 2)` sequences that mapped to a reference OTU per randomization (range from `r round(rand_ref_u["min", 1], 2)` to `r round(rand_ref_u["max", 1], 2)`).
+The original ordering of the reference resulted in `r rand_ref_u["original", 1]` sequences being mapped, less than the minimum observed number of mapped sequences when the references were randomized.
+This surprising result was likely due to the performance of the USEARCH heuristic.
+As with the AGC and DGC algorithms, if the order of the reference sequences differed between runs, then the heuristic criteria would be triggered at different points.
+To test this further, we substituted VSEARCH for USEARCH in the closed-reference algorithm.
+When we used VSEARCH the original ordering of the reference sequences and all randomizations were able to map `r rand_ref_v["original", 1]` sequences to reference OTUs.
+That VSEARCH mapped more sequences than even the maximum number of sequences found across the USEARCH randomizations supports the VSEARCH developers' assertion that it has greater sensitivity than USEARCH.
 
@@
 v19_summary <- read.table(file="data/gg_13_8/gg_13_8_97.v19.summary", header=T)
-n_v19_good <- sum(v19_summary$start <= 3967 & v19_summary$end >= 6116)
-n_v4_good <- length(scan("data/gg_13_8/gg_13_8_97.v4_ref.names", what="", sep="\n", quiet=T))
+n_v4_overlap
+n_v4_unique <- length(scan("data/gg_13_8/gg_13_8_97.v4_ref.names", what="", sep="\n", quiet=T))
 
 v_overlap <- read.table(file="data/rand_ref/overlap.vclosed.summary")
-```
 
-We noticed that even with VSEARCH, the reference OTU label that was assigned to each OTU differed even though the composition of the OTUs were the same across the randomized references.
-The default QIIME database that was used in the He study and by others contains full-length sequences that are at most 97% similar to each other.
-We suspect that the reference database was designed this way to overcome the limitations of the heuristic employed in USEARCH.
-This curated database is used for mapping sequences to a reference phylogenetic tree for carrying out UniFrac-based analyses (REFS) and to reference genomes for performing analyses such as picrust (REFS).
-Because the downstream applications depend on the correct and unique labeling of the OTUs, the lack of stability of the labeling was a concern.
-We hypothesized that within a given region  there would be sequences that were more similar to each other and possibly identical to each other.
-There were a total of `r nrow(v19_summary)` sequences in the reference database and `r n_v19_good` of those sequences fully overlap the V4 region.
-Among those sequences that fully overlapped the V4 region, there were `r n_small_v19_dist` pairs of sequences that were more than 97% similar to each other over the full length of the 16S rRNA gene.
-When the extracted V4 sequences were dereplicated, we identified `r n_v19_good-n_v4_good` sequences that were identical to each other.
-Among the dereplicated V4 sequences there were `r n_small_v4_dist` pairs of sequences that were more than 97% similar to each other.
-The presence of duplicate V4 reference sequences explains the lack of labeling stability when using either USEARCH or VSEARCH to carry out the closed-reference algorithm.
-Using VSEARCH to perform the closed-reference clustering, an average of `r round(100*v_overlap["orig", "mean"], 1)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*v_overlap["orig", "min"], 1)` to `r round(100*v_overlap["orig", "max"], 1)`).
-
-```{r duplicate_analysis}
 dups <- read.table(file="data/gg_13_8/duplicate.analysis")
@@
 
-We next asked whether the lack of consistency in labeling would affect the taxonomic identification of the sequences.
-The `r sum(dups$V2)-nrow(dups)` duplicated V4 sequences represented `r nrow(dups)` unique sequences.
+We also observed that regardless of whether we used USEARCH or VSEARCH, the reference OTU labels that were assigned to each OTU differed between randomizations.
+When we used USEARCH to perform closed-reference clustering, an average of `r round(100*u_overlap["orig", "mean"], 1)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*u_overlap["orig", "min"], 1)` to `r round(100*u_overlap["orig", "max"], 1)`).
+If we instead used VSEARCH an average of `r round(100*v_overlap["orig", "mean"], 1)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*v_overlap["orig", "min"], 1)` to `r round(100*v_overlap["orig", "max"], 1)`).
+To better understand this result, we further analyzed the reference database that is used in QIIME.
+We hypothesized that within a given region there would be sequences that were more than 97% similar and possibly identical to each other.
+When a sequence was used to search the randomized databases, it would encounter a different reference sequence as the first match with each randomization.
+Among those reference sequences that fully overlap the V4 region, there were `r n_small_v19_dist` pairs of sequences that were more than 97% similar to each other over the full length of the 16S rRNA gene.
+When the extracted V4 sequences were dereplicated, we identified `r n_v4_unique` unique sequences.
+Among these dereplicated V4 sequences there were `r n_small_v4_dist` pairs of sequences that were more than 97% similar to each other.
+The presence of duplicate V4 reference sequences explains the lack of labeling stability when using either USEARCH or VSEARCH to carry out the closed-reference algorithm.
+We suspect that the reference database was designed to only include sequences that were at most 97% similar to each other way to overcome the limitations of the USEARCH search heuristic.
+
+Beyond comparing the abundance of specific OTUs across samples, the reference database is used in the open and closed-reference algorithms to generate OTU labels that are used in several downstream applications.
+to extract information from a reference phylogenetic tree to carrying out UniFrac-based analyses (REFS) and to identify reference genomes for performing analyses such as PICRUSt (REFS).
+Because these downstream applications depend on the correct and unique labeling of the OTUs, the lack of stability of the labeling is problematic.
+As one illustration of the effects that incorrect labels would have on an analysis, we asked whether the duplicate sequences had the same taxonomies.
 Among the `r sum(dups$V2==2)` reference sequences that had one duplicate, `r sum(dups$V2==2 & dups$V3==2)` had discordant taxonomies.
-Among those `r sum(dups$V2 > 2)` sequences with two or more duplicates, `r sum(dups$V2>2 & dups$V3>=2)` had discordant taxonomies.
+Furthermore, among those `r sum(dups$V2 > 2)` sequences with two or more duplicates, `r sum(dups$V2>2 & dups$V3>=2)` had discordant taxonomies.
 Two sequences mapped to `r max_taxa[1,2]-1` and `r max_taxa[2,2]-1` duplicate sequences and both contained `r max_taxa[1,3]` different taxonomies.
 There was also a sequence had `r max(dups[,2])-1` duplicates and contained `r dups[which.max(dups[,2]),3]` different taxonomies.
+Together, these results demonstrate some of the considerable problems with the reference-based clustering of sequences.
 
-***Evalution of an open-source alternative to USEARCH.***
+## Conclusions
+* Stability is meaningless in context of accuracy and precision of clustering assignments
+* *De novo* clustering algorithms outperform reference-based approaches.
+* The best algorithm is data-specific and researchers should evaluate the clustering obtained by multiple methods
+
+* Don't use USEARCH to cluster sequences
+* VSEARCH is an excellent alternative
+
+* Reference-based clustering has many problems
+
+* Many difficulties in clustering have to do with underlying data quality.
 

### 2015-10-01 Schloss_Cluster_PeerJ_2015.Rmd
Working on R&D

@@
 An alternative to USEARCH is VSEARCH, which is being developed in parallel to USEARCH as a open-source alternative.
-One subtle difference between the two programs is that USEARCH employs a heuristic to determine when to stop searching a database.
+One subtle difference between the two programs is that USEARCH employs a heuristic to determine when to stop searching a database because it think it has found the best hit or does not believe that the database contains a hit.
 This means that the best possible match is not necessarily returned once a highly similar hit is obtained.
@@
 
+[   Something, something, something.    ]
+
 This also explains the variation observed in the AGC and DGC algorithms, which rely upon the USEARCH algorithm.
-[   Something, something, something.    ]
 
@@
 
-***Problems with reference-based clustering.*** The He study indicated that the closed-reference algorithm generated perfectly stable OTUs.
+***Problems with reference-based clustering in general and as implemented in QIIME.*** The He study indicated that the closed-reference algorithm generated perfectly stable OTUs.
 This was unsurprising since, by definition, the algorithm represents a one-to-one mapping of reads to a reference and treats the input sequences independently.
@@
 v19_summary <- read.table(file="data/gg_13_8/gg_13_8_97.v19.summary", header=T)
-n_v4_overlap
+
 n_v4_unique <- length(scan("data/gg_13_8/gg_13_8_97.v4_ref.names", what="", sep="\n", quiet=T))
 
+u_overlap <- read.table(file="data/rand_ref/overlap.uclosed.summary")
 v_overlap <- read.table(file="data/rand_ref/overlap.vclosed.summary")
@@
 For the hierarchical clustering algorithms and to generate a distance matrix, the pairwise distances between sequences were calculated in mothur using the pairwise.seqs command with the default Needleman-Wunsch alignment algorithm and parameters.
-Execution of the clustering algorithms was performed as described in the original He study using mothur (v.1.37) and QIIME (v.1.XX) as appropriate.
+Execution of the clustering algorithms was performed as described in the original He study using mothur (v.1.37) and QIIME (v.1.9.1) as appropriate.

### 2015-10-01 Schloss_Cluster_PeerJ_2015.Rmd
Outlining sections and turn off echoing

@@
 opts_chunk$set("tidy" = TRUE)
-opts_chunk$set("echo" = TRUE)
+opts_chunk$set("echo" = FALSE)
 opts_chunk$set("eval" = TRUE)
@@
 
+* Compare MCC and its variation when using different algorithms
+* Address stability
+* Quality of algorithms are data-dependent
+
 ***Evalution of an open-source alternative to USEARCH.*** The AGC and DGC algorithms appear to perform as well and possibly than the hierarchical clustering algorithms for some datasets.
@@
 [   Something, something, something.    ]
+
+* MCC for VAGC/VDGC > AGC/DGC
+* MCC for VAGC and VDGC are basically the same
+* Stability of VAGC/VDGC goes up
 
@@
 
-***Roesch Canadian soil dataset***
+***454 FLX-generated Roesch Canadian soil dataset***
 After obtaining the 16S rRNA gene fragments from GenBank (accessions EF308591-EF361836), we followed the methods outlined by the He study by removing any sequence that contained an ambiguous base, was identified as being a chimera, and fell outside a defined sequence length.
@@
 Execution of the clustering algorithms was performed as described in the original He study using mothur (v.1.37) and QIIME (v.1.9.1) as appropriate.
+
+***MiSeq-generated Murine gut microbiota dataset***
+
+***Analysis of reference database***
+* used gg_13_8 with the 97_otus.fasta as the reference
+
+***Calculation of Matthew's Correlation Coefficient***
+
+-> Available within mothur as the sens.spec command
+
+***Software availability***
+QIIME v.1.9.1
+mothur v.1.37.0
+R v.3.2
+Reproducible workflow including this manuscript as a literate programming document is available at https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015

### 2015-10-02 code/build_figure1.R
Fix plots to play nicely with miseq dataset

@@ -12,7 +12,7 @@ build_ref_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
 
 	par(mar=c(5.5, 5, 0.5, 0.5))
 
-	plot(NA, ylim=c(0.5,1), xlim=c(0.1,1), axes=F, ylab="", xlab="")
+	plot(NA, ylim=c(0,1), xlim=c(0,1), axes=F, ylab="", xlab="")
 	clrs <- rainbow(n_methods)
 	abline(v=threshold, col="gray", lwd=2)
 
@@ -24,7 +24,7 @@ build_ref_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
 	axis(2, las=1)
 	box()
 
-	text(x=0.12, y=0.98, label="A", cex=1.5, font=2)
+	text(x=0.02, y=0.98, label="A", cex=1.5, font=2)
 
 	title(ylab="MCC value relative to\nfull dataset")
 

### 2015-10-02 code/build_figure2.R

@@ -1,4 +1,4 @@
-build_figure2 <- function(dataset, output_file_name){
+build_figure2 <- function(dataset, output_file_name, position='bottomright'){
 	rare <- read.table(file=paste0("data/process/", dataset, ".rarefaction.summary"), header=T, stringsAsFactors=T)
 
 	offset <- 0.1
@@ -6,6 +6,9 @@ build_figure2 <- function(dataset, output_file_name){
 	n_fractions <- length(levels(factor(rare$fraction)))
 
 	methods <- c("fn", "nn", "an", "dgc", "agc", "closed", "open")
+
+	rare_methods <- rare[rare$method %in% methods,]
+
 	n_methods <- length(methods)
 
 	clrs <- rainbow(n_fractions)
@@ -14,11 +17,11 @@ build_figure2 <- function(dataset, output_file_name){
 
 	pdf(file=output_file_name, width=4, height=4)
 	par(mar=c(5,5,0.5,0.5))
-	plot(NA, xlim=c(0.8,length(methods)+0.2), ylim=c(0,max(rare$rare_uci)), axes=F, xlab="", ylab="")
+	plot(NA, xlim=c(0.8,length(methods)+0.2), ylim=c(0,max(rare_methods$rare_uci)), axes=F, xlab="", ylab="")
 	par(cex=0.8)
 	for(m in 1:n_methods){
 
-		subset <- rare[rare$method==methods[m],]
+		subset <- rare_methods[rare_methods$method==methods[m],]
 
 	#	the 95% confidence intervals were smaller or similar in size to the cex=1 plotting symbol
 	#	arrows(x0=m-offset, y0=subset[,"sobs"], y1=subset[,"sobs_lci"], col=clrs, length=0.05, angle=90)
@@ -38,7 +41,8 @@ build_figure2 <- function(dataset, output_file_name){
 
 	title(ylab="Number of OTUs", line=3.5)
 
-	pos <- legend(x=6.2, y=1400, legend=c("20%", "40%", "60%", "80%", "100%"),pch=19, col="white", cex=0.7)
+#	pos <- legend(x=6.2, y=1400, legend=c("20%", "40%", "60%", "80%", "100%"),pch=19, col="white", cex=0.7)
+	pos <- legend(position, legend=c("20%", "40%", "60%", "80%", "100%"),pch=19, col="white", cex=0.7, bty="n")
[…]

### 2015-10-02 Schloss_Cluster_PeerJ_2015.Rmd
Editing
Add commas between numbers
Writing and editing

@@
 options(scipen=999)
+
+inline_hook <- function(x){
+  if(is.numeric(x)){
+    paste(format(x,digits=2))
+  }
+}
+knitr::knit_hooks$set(big.sep=',')
 ```
+\newpage
+
+## Abstract
+
+\newpage
 
@@
 
-It is worth noting that the entire design of the He study is artificial.
-First, their analysis was based on a single sample.
-Researchers generally have dozens or hundreds of samples that are pooled and clustered together to enable comparison across samples.
-Second, all of the sequence data from these datasets is pooled for a single analysis.
-No one would ever perform an analysis based on a subset of their data.
-Because of these points, the value of identifying stable OTUs is unclear.
-
 ***Methods vary in their accuracy.*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs.
@@
 Using the two datasets, we determined whether the AGC and DGC algorithms as implemented by the two methods yielded sequencing assignments of similar quality.
+
+* Able to parallelize
 
@@
 ## Conclusions
-* Stability is meaningless in context of accuracy and precision of clustering assignments
-* *De novo* clustering algorithms outperform reference-based approaches.
-* The best algorithm is data-specific and researchers should evaluate the clustering obtained by multiple methods
+It is worth noting that the entire design of the He study was artificial.
+First, their analysis was based on a single soil sample.
+Researchers generally have dozens or hundreds of samples that are pooled and clustered together to enable comparison across samples.
+Second, all of the sequence data from these datasets is pooled for a single analysis.
+It is unclear why anyone would ever perform an analysis based on a subset of their data (REFS).
+Because of these points, the value of identifying stable OTUs is unclear.
+Greater emphasis should be placed on obtaining an optimal balance between splitting similar sequences into separate OTUs and merging disparate sequences into the same OTU.
+Through the use of the pairwise distances between sequences, we were able to use the MCC to demonstrate that, in general, the AL, AGC, and DGC methods perform better than the others.
+Although there is concern that running the algorithms multiple times yields different clusterings, we have shown that there is little variation in their MCC values.
+This suggests that the different clusterings by the same algorithm are equally good.
+Finally, it is impossible to obtain a clustering with no false positives or false negatives and the optimal algorithm may vary by dataset.
+With this in mind, researchers are encouraged to run the AL and VSEARCH DGC algorithms and calculate and report their MCC values.
+MCC values can be calculated in the mothur software package using the `sens.spec` command.
 
-* Don't use USEARCH to cluster sequences
-* VSEARCH is an excellent alternative
+Our analysis of those algorithms that implemented USEARCH as a method for clustering sequences revealed that its heuristic limited its sensitivity.
+When we replaced USEARCH with VSEARCH, the clustering quality improved.
+Although there may be parameters in USEARCH that can be tuned to improve the heuristic, these parameters are likely dataset dependent.
+Based on the data presented in this study, its availability as an open source, and free program, VSEARCH should replace USEARCH in these clustering algorithms.
+Furthermore, although not tested in our study, VSEARCH can be parallelized leading to potentially significant improvements in speed.
+Although USEARCh and VSEARCH do utilize aligned sequences, it is important to note that a sequence curation pipeline including denoising, alignment, trimming to a consistent region of the 16S rRNA gene, and chimera checking are critical to making proper inferences (REF).
 
 * Reference-based clustering has many problems
+* Open-reference clustering issues
+* Implemented in numerous pipelines for large projects
 
@@
 Reproducible workflow including this manuscript as a literate programming document is available at https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015
+
+\newpage
+
+## References

### 2015-10-02 Schloss_Cluster_PeerJ_2015.Rmd
Work on conclusion section

@@
 title: "**Accuracy and precision of 16S rRNA gene sequence assignments to OTUs is more important than the stability of OTUs**"
-author: Patrick D. Schloss
+author: Patrick D. Schloss* and Sarah L. Westcott
 date: '`r format(Sys.Date(), format="%B %d, %Y")`'
@@
   if(is.numeric(x)){
-    paste(format(x,digits=2))
+    paste(format(x,big.mark=','))
   }
 }
-knitr::knit_hooks$set(big.sep=',')
+knitr::knit_hooks$set(inline=inline_hook)
 ```
+
+* To whom correspondence should be addressed.
+pschloss@umich.edu
+
+Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
+
 \newpage
@@
 
-* Reference-based clustering has many problems
-* Open-reference clustering issues
-* Implemented in numerous pipelines for large projects
+For the first time, we assessed the ability of reference-based clustering algorithms to capture the actual similarity of the sequences in a dataset.
+Several studies have lauded both the open and closed-reference approaches for generating reproducible clusterings, yet we showed that both reference-based approaches did a poor job of representing the similarity between the sequences compared to the *de novo* approaches.
+So, although the clusterings are reproducible and stable across a range of library sizes, the clusterings are a poor representation of the data.
+We also observed that the clusterings were not actually reproducible when the order of the reference sequences was randomize.
+When USEARCH was used, the actual number of sequences that mapped to the reference changed depending on the order of the reference.
+Perhaps most alarming is that the default order of the database provided the worst sensitivity of any of the randomizations we attempted.
+Even when we used VSEARCH to perform closed-reference clustering and were able to obtain a consistent clusterings, we observed that the labels on the OTUs differed between randomizations.
+Because the OTU labels are frequently used to identify representative sequences for those OTUs, variation in labels, often representing different taxonomic groups, will have a detrimental effect on the interpretation of downstream analyses.
 
-* Many difficulties in clustering have to do with underlying data quality.
+Because the open-reference algorithm is a hybrid of the closed-reference and DGC algorithms, it is also negatively affected by the problems with using USEARCH for both algorithms.
+An added problem with the open-reference algorithm is that the two phases of the algorithm employ different thresholds.
+In the closed-reference step, sequences must be within a threshold of a reference to be in the same OTU.
+This means that two sequences that are 97% similar to a reference and are joined into the same OTU, may only be 94% similar to each other.
+In the DGC step, the goal is to approximate the AL algorithm which requires that, on average, the sequences within an OTU are at least 97% similar to each other.
+The end result of the open-reference approach is that sequences that are similar to previously observed sequences are clustered with one threshold while those that are not similar to previously observed sequences are clustered with a different threshold.
+
+As the throughput of sequencing technologies have improved, algorithm development to keep pace.
+*De novo* clustering algorithms are considerably slower and more computationally intensive than reference-based algorithms and the greedy *de novo* algorithms are faster than the hierarchical.
+In our experience (REFS), the most significant detriment to execution speed of the *de novo* algorithms has been in adequate removal of sequencing error.
+As the rate of sequencing error increases so do the number of unique sequences that must be clustered.
+The speed of the *de novo* algorithms scales approximately quadratically, so that doubling the number of sequences results in a four-fold increase in the time required to execute the algorithm.
+The rapid expansion in sequencing throughput has been likened to the Red Queen in Lewis Carroll's, *Through the Looking-Glass* who must run in place to keep up to her changing surroundings (REF).
+Microbial ecologists must continue to refine clustering algorithms to better handle the size of the datasets, but they must also take steps to improve the quality of the underlying data.
+Ultimately, objective standards must be applied to assess the quality of the data and the quality of OTU clustering.
 

### 2015-10-02 code/build_figure4.R
Build figure 4

@@ -1,0 +1,41 @@
+he_file <- "data/process/he.mcc.summary"
+miseq_file <- "data/process/miseq.mcc.summary"
+
+he_mcc <- read.table(file=he_file, header=T)
+miseq_mcc <- read.table(file=miseq_file, header=T)
+
+methods <- c("an", "agc", "vagc", "dgc", "vdgc")
+he_mcc <- he_mcc[he_mcc$method %in% methods,]
+miseq_mcc <- miseq_mcc[miseq_mcc$method %in% methods,]
+
+pdf("results/figures/figure_4.pdf", width=4.5, height=3)
+par(mar=c(4, 5, 0.5, 0.5))
+
+plot(NA, xlim=c(0.5,11.5), ylim=c(0,1), axes=F, ylab="MCC value", xlab="")
+
+for(i in 1:length(methods)){
+	row <- he_mcc[he_mcc$method==methods[i] & he_mcc$fraction==1,]
+	arrows(x0=i, y0=row$mean, y1=row$uci+1e-4, angle=90, length=0.1)
+	arrows(x0=i, y0=row$mean, y1=row$lci-1e-4, angle=90, length=0.1)
+	points(i, row$mean, pch=19)
+}
+
+for(i in 1:length(methods)){
+	row <- miseq_mcc[he_mcc$method==methods[i] & miseq_mcc$fraction==1,]
+	arrows(x0=i+6, y0=row$mean, y1=row$uci+(1e-3), angle=90, length=0.1)
+	arrows(x0=i+6, y0=row$mean, y1=row$lci-(1e-3), angle=90, length=0.1)
+	points(i+6, row$mean, pch=19)
+}
+
+axis(1, at=c(1:5, 7:11), label=c("AL", "U", "V", "U", "V", "AL", "U", "V", "U", "V"), cex.axis=0.9)
+axis(2, las=2)
+box()
+mtext(side=1, line=2, at=2.5, "AGC")
+mtext(side=1, line=2, at=4.5, "DGC")
+
+mtext(side=1, line=2, at=8.5, "AGC")
+mtext(side=1, line=2, at=10.5, "DGC")
+abline(v=6, col="gray")
+text(x=3, y=0.9, label="Canadian soil")
+text(x=9, y=0.9, label="Murine gut")
+dev.off()

### 2015-10-02 code/build_figure4.R
Add stability data to Figure 4

@@ -1,41 +1,56 @@
-he_file <- "data/process/he.mcc.summary"
-miseq_file <- "data/process/miseq.mcc.summary"
-
-he_mcc <- read.table(file=he_file, header=T)
-miseq_mcc <- read.table(file=miseq_file, header=T)
-
 methods <- c("an", "agc", "vagc", "dgc", "vdgc")
-he_mcc <- he_mcc[he_mcc$method %in% methods,]
-miseq_mcc <- miseq_mcc[miseq_mcc$method %in% methods,]
 
 pdf("results/figures/figure_4.pdf", width=4.5, height=3)
-par(mar=c(4, 5, 0.5, 0.5))
+par(mar=c(3, 4, 0.5, 0.5))
 
 plot(NA, xlim=c(0.5,11.5), ylim=c(0,1), axes=F, ylab="MCC value", xlab="")
 
+he_mcc <- read.table(file="data/process/he.mcc.summary", header=T)
+he_ref <- read.table(file="data/process/he.mcc_ref.summary", header=T)
+he_mcc <- he_mcc[he_mcc$method %in% methods,]
+he_ref <- he_ref[he_mcc$method %in% methods,]
+
 for(i in 1:length(methods)){
 	row <- he_mcc[he_mcc$method==methods[i] & he_mcc$fraction==1,]
-	arrows(x0=i, y0=row$mean, y1=row$uci+1e-4, angle=90, length=0.1)
-	arrows(x0=i, y0=row$mean, y1=row$lci-1e-4, angle=90, length=0.1)
+	arrows(x0=i, y0=row$mean, y1=row$uci+1e-3, angle=90, length=0.1)
+	arrows(x0=i, y0=row$mean, y1=row$lci-1e-3, angle=90, length=0.1)
 	points(i, row$mean, pch=19)
+
+	row <- he_ref[he_ref$method==methods[i] & he_ref$fraction==0.6,]
+	arrows(x0=i, y0=row$mean, y1=row$uci+1e-3, angle=90, length=0.1)
+	arrows(x0=i, y0=row$mean, y1=row$lci-1e-3, angle=90, length=0.1)
+	points(i, row$mean, pch=21, bg="white")
 }
+
+miseq_mcc <- read.table(file="data/process/miseq.mcc.summary", header=T)
+miseq_ref <- read.table(file="data/process/miseq.mcc_ref.summary", header=T)
[…]

### 2015-10-02 Schloss_Cluster_PeerJ_2015.Rmd
Add commentary on Figure 4

@@
 ---
+
+## To Do:
+* miseq analysis
+* methods section
+* references
+* figure captions
 
@@
 ```{r}
-z <- read.table(file="data/process/he.mcc_ref.summary", header=T)
+he_stability <- read.table(file="data/process/he.mcc_ref.summary", header=T)
 ```
@@
 In the He study, when comparing the OTUs formed using the AL algorithm with 60% of the data, they observed a mean MCC value of approximately 0.63 (95% confidence interval between approximately 0.15 and 0.75).
-In contrast, we observed a mean value of `r round(z[z$method=="an" & z$fraction==0.6, "mean"], 2)` (95% confidence interval between `r round(z[z$method=="an" & z$fraction==0.6, "lci"], 2)` and `r round(z[z$method=="an" & z$fraction==0.6, "uci"], 2)`).
+In contrast, we observed a mean value of `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "mean"], 2)` (95% confidence interval between `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "lci"], 2)` and `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "uci"], 2)`).
 This result suggests that the AL algorithm was far more stabile than indicated in the He study.
@@
 Using the two datasets, we determined whether the AGC and DGC algorithms as implemented by the two methods yielded sequencing assignments of similar quality.
-
-* Able to parallelize
-
-[   Something, something, something.    ]
-
-* MCC for VAGC/VDGC > AGC/DGC
-* MCC for VAGC and VDGC are basically the same
-* Stability of VAGC/VDGC goes up
-
-This also explains the variation observed in the AGC and DGC algorithms, which rely upon the USEARCH algorithm.
+In general the overall trends that we observed with the USEARCH-version of AGC and DGC were also observed with the VSEARCH-verion of the algorithms.
+When we compared the two implementations of the AGC and DGC algorithms, the OTUs generated by the VSEARCH-verion of the algorithms were as stable or more stable than the USEARCH-version when using 60% of the datasets.
+In addition, the MCC values for the entire datasets, calculated relative to the distance matrix, were virtually indistinguishable (Figure 4).
+These results are a strong indication that VSEARCH is a suitable and possibly better replacement for USEARCH for executing the AGC and DGC algorithms.
 
@@
 ## References
+
+## Figures

### 2015-10-06 code/build_figure1.R
Fix x-axis range

@@ -12,7 +12,7 @@ build_ref_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
 
 	par(mar=c(5.5, 5, 0.5, 0.5))
 
-	plot(NA, ylim=c(0,1), xlim=c(0,1), axes=F, ylab="", xlab="")
+	plot(NA, ylim=c(0,1), xlim=c(0.1,1), axes=F, ylab="", xlab="")
 	clrs <- rainbow(n_methods)
 	abline(v=threshold, col="gray", lwd=2)
 

### 2015-10-06 code/build_figure1.R
Fix placement of 'A' in left panel

@@ -24,7 +24,7 @@ build_ref_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
 	axis(2, las=1)
 	box()
 
-	text(x=0.02, y=0.98, label="A", cex=1.5, font=2)
+	text(x=0.12, y=0.98, label="A", cex=1.5, font=2)
 
 	title(ylab="MCC value relative to\nfull dataset")
 

### 2015-10-06 code/build_figure1.R
Add x-axis label to panel A

@@ -12,7 +12,7 @@ build_ref_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
 
 	par(mar=c(5.5, 5, 0.5, 0.5))
 
-	plot(NA, ylim=c(0,1), xlim=c(0.1,1), axes=F, ylab="", xlab="")
+	plot(NA, ylim=c(0,1), xlim=c(0.1,1), axes=F, ylab="Fraction of dataset used", xlab="")
 	clrs <- rainbow(n_methods)
 	abline(v=threshold, col="gray", lwd=2)
 

### 2015-10-06 code/build_figure1.R
Add x-axis label to panel A

@@ -12,7 +12,7 @@ build_ref_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
 
 	par(mar=c(5.5, 5, 0.5, 0.5))
 
-	plot(NA, ylim=c(0,1), xlim=c(0.1,1), axes=F, ylab="Fraction of dataset used", xlab="")
+	plot(NA, ylim=c(0,1), xlim=c(0.1,1), axes=F, xlab="Fraction of dataset used", ylab="")
 	clrs <- rainbow(n_methods)
 	abline(v=threshold, col="gray", lwd=2)
 

### 2015-10-06 code/build_figure3.R
Fix x and y-axis labels

@@ -26,8 +26,8 @@ build_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
 
 	text(x=0.12, y=0.98, label="A", cex=1.5, font=2)
 
-	title(ylab="MCC value relative to\nfull dataset")
-
+	title(ylab="MCC value relative to\ndistance matrix")
+	title(xlab="Fraction of dataset used")
 
 
 

### 2015-10-06 code/build_figure1.R
Fix colors

@@ -13,7 +13,8 @@ build_ref_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
 	par(mar=c(5.5, 5, 0.5, 0.5))
 
 	plot(NA, ylim=c(0,1), xlim=c(0.1,1), axes=F, xlab="Fraction of dataset used", ylab="")
-	clrs <- rainbow(n_methods)
+	#clrs <- rainbow(n_methods)
+	clrs <- c("red", "orange", "yellow", rainbow(n_methods)[4], rainbow(n_methods)[5], rainbow(n_methods)[6])
 	abline(v=threshold, col="gray", lwd=2)
 
 	for(m in 1:n_methods){

### 2015-10-06 code/build_figure1.R
Fix colors

@@ -14,7 +14,7 @@ build_ref_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
 
 	plot(NA, ylim=c(0,1), xlim=c(0.1,1), axes=F, xlab="Fraction of dataset used", ylab="")
 	#clrs <- rainbow(n_methods)
-	clrs <- c("red", "orange", "yellow", rainbow(n_methods)[4], rainbow(n_methods)[5], rainbow(n_methods)[6])
+	clrs <- c("red", "orange", "yellow", rainbow(n_methods)[4], rainbow(n_methods)[5], rainbow(n_methods)[6], rainbow(n_methods)[7])
 	abline(v=threshold, col="gray", lwd=2)
 
 	for(m in 1:n_methods){

### 2015-10-06 code/build_figure1.R
Fix colors

@@ -1,3 +1,4 @@
+
 build_ref_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
 
 	data <- read.table(file=paste0("data/process/", dataset, ".mcc_ref.summary"), header=T, stringsAsFactors=F)
@@ -11,10 +12,10 @@ build_ref_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
 	layout(matrix(c(1,2), nrow=1), widths=c(4, 3.5))
 
 	par(mar=c(5.5, 5, 0.5, 0.5))
+	clrs <- c("red", "orange", "yellow", rainbow(n_methods)[4:n_methods])
 
 	plot(NA, ylim=c(0,1), xlim=c(0.1,1), axes=F, xlab="Fraction of dataset used", ylab="")
-	#clrs <- rainbow(n_methods)
-	clrs <- c("red", "orange", "yellow", rainbow(n_methods)[4], rainbow(n_methods)[5], rainbow(n_methods)[6], rainbow(n_methods)[7])
+
 	abline(v=threshold, col="gray", lwd=2)
 
 	for(m in 1:n_methods){

### 2015-10-06 code/build_figure3.R

@@ -1,3 +1,4 @@
+
 build_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
 
 	data <- read.table(file=paste0("data/process/", dataset, ".mcc.summary"), header=T, stringsAsFactors=F)
@@ -11,9 +12,10 @@ build_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
 	layout(matrix(c(1,2), nrow=1), widths=c(4, 3.5))
 
 	par(mar=c(5.5, 5, 0.5, 0.5))
+	clrs <- c("red", "orange", "yellow", rainbow(n_methods)[4:n_methods])
 
 	plot(NA, ylim=c(0,1), xlim=c(0.1,1), axes=F, ylab="", xlab="")
-	clrs <- rainbow(n_methods)
+
 	abline(v=threshold, col="gray", lwd=2)
 
 	for(m in 1:n_methods){

### 2015-10-06 code/build_figure1.R
Combine figure 1 and 3

@@ -1,61 +1,0 @@
-
-build_ref_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
-
-	data <- read.table(file=paste0("data/process/", dataset, ".mcc_ref.summary"), header=T, stringsAsFactors=F)
-
-	methods <- c("fn", "nn", "an", "dgc", "agc", "closed", "open")
-	n_methods <- length(methods)
-
-	orig_methods <- data[data$method %in% methods,]
-
-	pdf(file=output_file_name, width=7.5, height=4)
-	layout(matrix(c(1,2), nrow=1), widths=c(4, 3.5))
-
-	par(mar=c(5.5, 5, 0.5, 0.5))
-	clrs <- c("red", "orange", "yellow", rainbow(n_methods)[4:n_methods])
-
-	plot(NA, ylim=c(0,1), xlim=c(0.1,1), axes=F, xlab="Fraction of dataset used", ylab="")
-
-	abline(v=threshold, col="gray", lwd=2)
-
-	for(m in 1:n_methods){
-		subset <- orig_methods[orig_methods$method==methods[m],]
-		points(subset$mean~subset$fraction, type="l", col=clrs[m], lwd=2)
-	}
-	axis(1)
-	axis(2, las=1)
-	box()
-
-	text(x=0.12, y=0.98, label="A", cex=1.5, font=2)
-
-	title(ylab="MCC value relative to\nfull dataset")
-
-
-
-
-
-	par(mar=c(5.5, 0.5, 0.5, 0.5))
-
-	sixty <- orig_methods[orig_methods$fraction==threshold,]
-	sixty$method <- factor(sixty$method, levels=methods)
-	sixty[sixty$method=="closed", c("lci", "uci")] <- c(0.999, 1.001)
-	plot(NA, ylim=c(0.5,1), xlim=c(0.75,7.25), axes=F, ylab="", xlab="")
-
-	for(m in 1:n_methods){
-		points(x=m, y=sixty[sixty$method==methods[m],"mean"], pch=19, col=clrs[m])
-
-		arrows(x0=m, y0=sixty[sixty$method==methods[m],"mean"], x1=m, y1=sixty[sixty$method==methods[m],"lci"], lwd=2, angle=90, length=0.1, col=clrs[m])
[…]

### 2015-10-06 code/build_figure3.R

@@ -1,59 +1,0 @@
-
-build_mcc_plot <- function(dataset, output_file_name, threshold=0.6){
-
-	data <- read.table(file=paste0("data/process/", dataset, ".mcc.summary"), header=T, stringsAsFactors=F)
-
-	methods <- c("fn", "nn", "an", "dgc", "agc", "closed", "open")
-	n_methods <- length(methods)
-
-	orig_methods <- data[data$method %in% methods,]
-
-	pdf(file=output_file_name, width=7.5, height=4)
-	layout(matrix(c(1,2), nrow=1), widths=c(4, 3.5))
-
-	par(mar=c(5.5, 5, 0.5, 0.5))
-	clrs <- c("red", "orange", "yellow", rainbow(n_methods)[4:n_methods])
-
-	plot(NA, ylim=c(0,1), xlim=c(0.1,1), axes=F, ylab="", xlab="")
-
-	abline(v=threshold, col="gray", lwd=2)
-
-	for(m in 1:n_methods){
-		subset <- orig_methods[orig_methods$method==methods[m],]
-		points(subset$mean~subset$fraction, type="l", col=clrs[m], lwd=2)
-	}
-	axis(1)
-	axis(2, las=1)
-	box()
-
-	text(x=0.12, y=0.98, label="A", cex=1.5, font=2)
-
-	title(ylab="MCC value relative to\ndistance matrix")
-	title(xlab="Fraction of dataset used")
-
-
-
-
-	par(mar=c(5.5, 0.5, 0.5, 0.5))
-
-	sixty <- orig_methods[orig_methods$fraction==threshold,]
-	sixty$method <- factor(sixty$method, levels=methods)
-
-	plot(NA, ylim=c(0.0,1), xlim=c(0.75,7.25), axes=F, ylab="", xlab="")
-
-	for(m in 1:n_methods){
-		points(x=m, y=sixty[sixty$method==methods[m],"mean"], pch=19, col=clrs[m])
-
-		arrows(x0=m, y0=sixty[sixty$method==methods[m],"mean"], x1=m, y1=sixty[sixty$method==methods[m],"lci"], lwd=2, angle=90, length=0.1, col=clrs[m])
-		arrows(x0=m, y0=sixty[sixty$method==methods[m],"mean"], x1=m, y1=sixty[sixty$method==methods[m],"uci"], lwd=2, angle=90, length=0.1, col=clrs[m])
[…]

### 2015-10-06 code/build_mcc_plot.R

@@ -1,0 +1,126 @@
+build_mcc_plots <- function(dataset, output_file_name){
+
+	methods <- c("fn", "nn", "an", "dgc", "agc", "closed", "open")
+	n_methods <- length(methods)
+	clrs <- c("brown", "red", "orange", rainbow(n_methods)[4:n_methods])
+
+
+
+
+	data <- read.table(file=paste0("data/process/", dataset, ".mcc_ref.summary"), header=T, stringsAsFactors=F)
+
+	orig_methods <- data[data$method %in% methods,]
+
+	pdf(file=output_file_name, width=6.5, height=6.5)
+
+
+
+	layout(matrix(c(1, 2, 3, 4, 5, 6, 0, 7, 0), nrow=3, byrow=T), widths=c(1.0,3.5, 3.5), heights=c(3.5,3.5,1.5))
+
+	threshold <- 0.6
+
+	par(mar=c(0,0,0,0))
+	plot.new()
+	text(x=0.3, y=0.5, label="MCC value relative to\nfull dataset", cex=1.5, srt=90, xpd=TRUE)
+
+	### A ###
+	par(mar=c(0.5, 0.5, 0.5, 0.5))
+	plot(NA, ylim=c(0,1), xlim=c(0.1,1), axes=F, xlab="", ylab="")
+
+	abline(v=threshold, col="gray", lwd=2)
+
+	for(m in 1:n_methods){
+		subset <- orig_methods[orig_methods$method==methods[m],]
+		points(subset$mean~subset$fraction, type="l", col=clrs[m], lwd=3)
+	}
+	axis(1, at=seq(0.2,1.0,0.2), labels=rep("", 5))
+	axis(2, las=1, cex.axis=1.5)
+	box()
+
+	text(x=0.12, y=0.98, label="A", cex=2.5, font=2)
+
+	#title(xlab="Fraction of dataset used")
+
+	### B ###
+	par(mar=c(0.5, 0.5, 0.5, 0.5))
+
+	sixty <- orig_methods[orig_methods$fraction==threshold,]
+	sixty$method <- factor(sixty$method, levels=methods)
+	sixty[sixty$method=="closed", c("lci", "uci")] <- c(0.999, 1.001)
+	plot(NA, ylim=c(0.0,1), xlim=c(0.75,7.25), axes=F, ylab="", xlab="")
+
+	for(m in 1:n_methods){
[…]

### 2015-10-06 Schloss_Cluster_PeerJ_2015.Rmd
Add results for MiSeq data

@@
 ## To Do:
-* miseq analysis
+* Figure 5
 * methods section
@@
 We made three important observations.
-First, each of the *de novo* methods varied in how sensitive their MCC values were to additional sequences (Figure 3A).
+First, each of the *de novo* methods varied in how sensitive their MCC values were to additional sequences (Figure 1C).
 The SL and CL algorithms were the most sensitive; however, the MCC values using 80 and 100% of the data did not vary meaningfully when using *de novo* algorithms.
-Second, the AL algorithm out-performed the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL.
+Second, the AL algorithm out-performed the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL (Figure 1D).
 Third, with the possible exception of the CL algorithm, the MCC values for each of the only demonstrated a small amount of variation between runs of the algorithm with a different ordering of the input sequences.
@@
 n_groups <- ncol(read.table(file="data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.denovo.uchime.pick.pick.count_table")) - 2 #subtract 2 to remove the sequence name and total number of sequences
+
+miseq_stability <- read.table(file="data/process/miseq.mcc_ref.summary", header=T)
+miseq_mcc <- read.table(file="data/process/miseq.mcc.summary", header=T)
 ```
@@
 Although no dataset is perfect for exhaustively testing these clustering algorithms, this dataset was useful for demonstrating several points.
-
-[   Something, something, something.    ]
-
-* Compare MCC and its variation when using different algorithms
-* Address stability
-* Quality of algorithms are data-dependent
+First, when using 60% of the data the stability relationships amongst the different methods were similar to what we observed using the He dataset (Figure 3AB).
+With the exception for the clusters generated using CL, the methods all performed very well with stabilities greater than `r round(min(miseq_stability[miseq_stability$fraction=="0.6" & miseq_stability$method != "fn","mean"]), 2)`.
+Second, the MCC values calculated relative to the distances between sequences were generally higher than was observed for the He dataset for all of the methods except for CL and SL.
+Surprisingly, the MCC values for the DGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "dgc","mean"],2)`)  and AGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "agc","mean"],2)`) algorithms were comparable to the AL algorithm (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "an","mean"],2)`; Figure 3CD).
+This result suggests that the optimal algorithm may be database-dependent.
+Finally, as was observed with the He dataset, there was little variation in MCC values observed among the 30 randomizations.
+Therefore, although the algorithms have a stochastic component, the OTU assignments do not vary meaningfully.
+The results from both the Canadian soil and murine microbiota datasets demonstrate that the *de novo* algorithms can generate very stable OTU assignments, as defined by the He study, and that the assignments are highly reproducible.
+Most importantly, these analyses demonstrate that the OTU assignments using the AL, AGC, and DGC *de novo* algorithms are consistently more robust than either of the reference-based methods.
 
@@
 QIIME v.1.9.1
-mothur v.1.37.0
+mothur v.1.37.0 - includes sens.spec command and vsearch enabled clustering
 R v.3.2
+vsearch
 Reproducible workflow including this manuscript as a literate programming document is available at https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015

### 2015-10-08 code/summarize_rand_ref.R
Output the number of references that map in a counts file

@@ -22,6 +22,11 @@ get_hits_summary <- function(rand_hits, orig_hits, method){
 	output_data <- c(summary_stats(rand_n_hits), original=orig_n_hits)
 	output_filename <- paste0("data/rand_ref/hits.", method, "closed.summary")
 	write.table(file=output_filename, x=output_data, row.names=TRUE, col.names=FALSE, quote=FALSE, sep="\t")
+
+       output_filename <- paste0("data/rand_ref/hits.", method, "closed.counts")
+       write.table(file=output_filename, c(rand_n_hits, orig_n_hits), row.names=c(1:30, "original"), col.names=FALSE, sep="\t", quote=FALSE)
+
+
 }
 
 

### 2015-10-08 Schloss_Cluster_PeerJ_2015.Rmd
Add formatting option to inline hook
Flesh out Methods section

@@
 ## To Do:
-* Figure 5
-* methods section
 * references
 * figure captions
+* revise
 
@@
   if(is.numeric(x)){
-    paste(format(x,big.mark=','))
+    paste(format(x,big.mark=',', digits=2))
   }
@@
 Beyond comparing the abundance of specific OTUs across samples, the reference database is used in the open and closed-reference algorithms to generate OTU labels that are used in several downstream applications.
-to extract information from a reference phylogenetic tree to carrying out UniFrac-based analyses (REFS) and to identify reference genomes for performing analyses such as PICRUSt (REFS).
+It is commonly used to extract information from a reference phylogenetic tree to carrying out UniFrac-based analyses (REFS) and to identify reference genomes for performing analyses such as PICRUSt (REFS).
 Because these downstream applications depend on the correct and unique labeling of the OTUs, the lack of stability of the labeling is problematic.
@@
 With this in mind, researchers are encouraged to run the AL and VSEARCH DGC algorithms and calculate and report their MCC values.
-MCC values can be calculated in the mothur software package using the `sens.spec` command.
 
@@
 The order of the sequences in the full dataset was randomly permuted without replacement to generate an additional 30 datasets.
-For the hierarchical clustering algorithms and to generate a distance matrix, the pairwise distances between sequences were calculated in mothur using the pairwise.seqs command with the default Needleman-Wunsch alignment algorithm and parameters.
-Execution of the clustering algorithms was performed as described in the original He study using mothur (v.1.37) and QIIME (v.1.9.1) as appropriate.
+To perform the hierarchical clustering algorithms and to generate a distance matrix we followed their approach of the He study by calculating distances based on pairwise global alignments using the `pairwise.dist` command in mothur using the default Needleman-Wunsch alignment algorithm and parameters.
+It should be noted that this method has been strongly discouraged (REFS).
+Execution of the hierarchical clustering algorithms was performed as described in the original He study using mothur (v.1.37) and using the QIIME (v.1.9.1) parameter profiles provided in the supplementary material from the He study for the greedy and reference-based clustering algorithms.
 
 ***MiSeq-generated Murine gut microbiota dataset***
+The murine 16S rRNA gene sequence data generated from the V4 region using an Illumina MiSeq was obtained from http:/www.mothur.org/MiSeqDevelopmentData/StabilityNoMetaG.tar and was processed as outlined in the original study (REF).
+Briefly, 250 nt long read pairs were assembled into contigs by aligning the reads and correcting discordant base calls by requiring one of the base calls to have a Phred quality score at least 6 points higher than the other.
+Sequences where it was not possible to resolve the disagreement were culled from the dataset.
+The sequences were then aligned to a SILVA reference alignment (REF) and any reads that aligned outside of the V4 region were removed from the dataset.
+Sequences were pre-clustered by combining the abundances of sequences that were within 2 nt of a more abundant sequence.
+Each of the samples was then screened for chimeric sequences using the default parameters in UCHIME (REF).
+The resulting sequences were processed in the same manner as the Canadian soil dataset with the exception that the distance matrices were calculated based on the SILVA-based alignment.
 
 ***Analysis of reference database***
-* used gg_13_8 with the 97_otus.fasta as the reference
+We utilized the 97% OTUs greengenes reference sequence and taxonomy data (v.13.8) that accompanies the QIIME installation.
+Because the greengenes reference alignment does a poor job of representing the secondary structure of the 16S rRNA gene (REF), we realigned the fasta sequences to a SILVA reference alignment to identify the V4 region of the sequences.
 
-***Calculation of Matthew's Correlation Coefficient***
+***Calculation of Matthew's Correlation Coefficient (MCC)***
+The MCC was calculated by two approaches in this study.
+In both methods we used only the dereplicated sequence lists.
+First, we calculated the MCC to determine the stability of OTU assignments following the approach of the He study.
+We assumed that the clusters obtained from the 30 randomized full datasets were correct.
+We counted the number of sequence pairs that were in the same OTU for the subsetted dataset and the full dataset (i.e.
+true positives; TP), that were in different OTUs for the subsetted dataset and the full dataset (i.e.
+true negatives; TN), that were in the same OTU for the subsetted dataset and different OTUs in the full dataset (i.e.
+false positives; FP), and that were in different OTUs for the subsetted dataset and the same OTU in the full dataset (i.e.
+false negatives; FN).
+For each set of 30 random subsamplings of the dataset, we counted these parameters against the 30 randomizations of the full dataset.
+This gave 900 comparisons for each fraction of sequences being used in the analysis.
+The Matthew's correlation coefficient was then calculated as:
 
--> Available within mothur as the sens.spec command
+$$
+MCC = \frac{TP \times TN - FP \times FN}{\sqrt{(TP +FP)(TP+FN)(TN+FP)(TN+FN)} }
+$$
+
+Second, we calculated MCC to determine the quality of the clusterings as previously described (REF).
+Briefly, we compared the OTU assignments for pairs of sequences to the distance matrix that was calculated between all pairs of aligned sequences.
+For each dataset that was clustered, those sequences that were in the same OTU and had a similarity greater than 97% were TPs, those that were in different OTUs and had a similarity less than 97% were TNs, those that were in the same OTU and had a similarity less than 97% were FPs, and those that were in different OTUs and had a similarity greater than 97% were FNs.
+The MCC was counted for each dataset using the formula above as implemented in the `sens.spec` command in mothur.
 
 ***Software availability***
-QIIME v.1.9.1
-mothur v.1.37.0 - includes sens.spec command and vsearch enabled clustering
-R v.3.2
-vsearch
-Reproducible workflow including this manuscript as a literate programming document is available at https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015
+A reproducible workflow including all scripts and this manuscript as a literate programming document are available at https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015.
+The workflow utilizes QIIME (v.1.9.1; REFS), mothur (v.1.37.0; REFS), USEARCH (v.1.5.0; REFS), VSEARCH (v.1.5.0; https://github.com/torognes/vsearch), and R (v.3.2; REFS).
 

### 2015-10-08 code/build_figure4.R
Change distane matrix to distances

@@ -39,7 +39,7 @@ for(i in 1:length(methods)){
 	points(i+6, row$mean, pch=21, bg="white")
 }
 
-legend(x=7.5, y=0.4, legend=c("Relative to full dataset", "Relative to distance"), pch=c(21,19), pt.cex=0.8, pt.bg=c("white", "black"), cex=0.6, bty="n")
+legend(x=7.5, y=0.4, legend=c("Relative to full dataset", "Relative to distances"), pch=c(21,19), pt.cex=0.8, pt.bg=c("white", "black"), cex=0.6, bty="n")
 axis(1, at=c(1:5, 7:11), label=c("AL", "U", "V", "U", "V", "AL", "U", "V", "U", "V"), cex.axis=0.9)
 mtext(side=1, line=2, at=2.5, "AGC")
 mtext(side=1, line=2, at=4.5, "DGC")

### 2015-10-08 code/build_mcc_plot.R

@@ -72,7 +72,7 @@ build_mcc_plots <- function(dataset, output_file_name){
 
 	par(mar=c(0,0,0,0))
 	plot.new()
-	text(x=0.3, y=0.5, label="MCC value relative to\ndistance matrix", cex=1.5, srt=90, xpd=TRUE)
+	text(x=0.3, y=0.5, label="MCC value relative to\ndistances", cex=1.5, srt=90, xpd=TRUE)
 
 	### C ###
 	par(mar=c(0.5, 0.5, 0.5, 0.5))

### 2015-10-08 Schloss_Cluster_PeerJ_2015.Rmd
Add Figure legends and some editing

@@
 * references
-* figure captions
 * revise
+* subsample vs. sub-sample
 
@@
 Such comparisons have enabled researchers to characterize the microbiota associated with the human body (REF), soil (REF), aquatic ecosystems (REF), and numerous other environments.
-Within the field of microbial ecology a convention has emerged where sequences are clustered into OTUs using a threshold of 3% dissimilarity or 97% similarity.
+Within the field of microbial ecology a convention has emerged where sequences are clustered into OTUs using a threshold of 97% similarity or a distance of 3%.
 One advantage of the OTU-based approach is that the definition of the bins is operational and can be changed to suit the needs of the particular project.
-However, with the dissemination of clustering algorithms within packages such as mothur and QIIME and stand-alone tools (REFS), it is important to understand how different clustering methods implement this conventional OTU threshold.
-Furthermore, it is necessary to understand how algorithm choice affects the precision and accuracy of assigning sequences to OTUs.
+However, with the dissemination of clustering methods within packages such as mothur and QIIME and stand-alone tools (REFS), it is important to understand how different clustering methods implement this conventional OTU threshold.
+Furthermore, it is necessary to understand how method choice affects the precision and accuracy of assigning sequences to OTUs.
 Broadly speaking, three approaches have been developed to assign sequences to OTUs.
@@
 The second approach has been referred to as *de novo* clustering.
-In this method, the similarity between sequences is used to cluster sequence rather than the similarity to a reference database.
-In contrast to the efficiency of closed-reference clustering, the speed of hierarchical *de novo* clustering algorithms scale quadratically with the number of unique sequences.
+In this method, the distance between sequences is used to cluster sequence rather than the distance to a reference database.
+In contrast to the efficiency of closed-reference clustering, the speed of hierarchical *de novo* clustering methods scale quadratically with the number of unique sequences.
 The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences.
@@
 As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods.
-Previous comparisons of hierarchical and heuristic algorithms have shown that the average neighbor algorithm, a hierarchical algorithm outperforms the other *de novo* clustering algorithm.
+Previous comparisons of hierarchical and heuristic methods have shown that the average neighbor method, a hierarchical method outperforms the other *de novo* clustering method.
 An early analysis using 16S rRNA gene sequence data generated using the 454 sequencing platform found that if these methods were applied to the same dataset using a subset of the sequences then a different number of OTUs were observed compared to clustering the full dataset and rarefying to the same number of sequences.
-Because the initial subsampling of the dataset was not repeated in the earlier analysis, it is unclear whether the result was product of the algorithm or sampling.
+Because the initial subsampling of the dataset was not repeated in the earlier analysis, it is unclear whether the result was product of the method or sampling.
 One explanation could be that the clusters are getting better with additional sampling.
@@
 An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels (REF).
-For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average neighbor algorithm.
-Those sequences that did not classify to a known family would also be clustered using the average neighbor algorithm.
+For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average neighbor method.
+Those sequences that did not classify to a known family would also be clustered using the average neighbor method.
 An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group (e.g.
@@
 Some have focused on the time and memory required to process a dataset (REFS).
-These are valid parameters to assess when judging a clustering algorithm, but have little to say about the quality of the clustering.
-Others have attempted to judge the quality of an algorithm by its ability to generate data that parallels classification data.
+These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the clustering.
+Others have attempted to judge the quality of an method by its ability to generate data that parallels classification data.
 This approach is problematic because bacterial taxonomy often reflects the biases amongst bacterial systematicists and the rates of evolution across lineages are not the same.
-We recently proposed an approach for evaluating OTU assignments using the similarity between sequences (REF).
+We recently proposed an approach for evaluating OTU assignments using the distance between sequences (REF).
 Those sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives.
@@
 Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and ratios using the Matthew's correlation coefficient (MCC; REF).
-That analysis focused on assessing *de novo* clustering algorithms and found that the average neighbor algorithm outperformed the other hierarchical and heuristic algorithms.
+That analysis focused on assessing *de novo* clustering methods and found that the average neighbor method outperformed the other hierarchical and heuristic methods.
 
 A recent analysis by He and colleagues (REF) attempted to characterize the three general clustering approaches by focusing on what they called stability.
-They defined stability as the ability of an algorithm to provide the same clustering on a subset of the data as was found in the full dataset.
-Related to this, the authors expressed concern that because some algorithms use a random number generator to break ties there may be further instability between executions of algorithm.
+They defined stability as the ability of an method to provide the same clustering on a subset of the data as was found in the full dataset.
+Related to this, the authors expressed concern that because some methods use a random number generator to break ties there may be further instability between executions of method.
 Their concept of stability does not account for the accuracy of the clustering and instead focuses on the precision of the clustering.
 A method may be very precise, but low in accuracy.
-In the current analysis, we attempted to assess the accuracy and precision of the various clustering algorithms.
-Building on our previous analysis of clustering algorithms, our hypothesis was that the algorithms praised by the He study for their stability actually suffered a lack of accuracy.
+In the current analysis, we attempted to assess the accuracy and precision of the various clustering methods.
+Building on our previous analysis of clustering methods, our hypothesis was that the methods praised by the He study for their stability actually suffered a lack of accuracy.
 In addition, we assess these parameters in light of sequence quality using the original 454 dataset and a larger and more modern dataset generated using the MiSeq platform.
@@
 The He study found that when they used the open and closed-reference methods clusters formed using the subset data most closely resembled those of the full dataset.
-Among the *de novo* clustering methods they observed that the abundance-based greedy clustering (AGC) algorithm followed by single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) algorithms.
+Among the *de novo* clustering methods they observed that the abundance-based greedy clustering (AGC) method followed by single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) methods.
 They observed a broad range of MCC values among their AL replicates, which was not explained by the authors.
-We first sought to assess calculated the MCC for for each of the clustering algorithms using 20, 40, 60, and 80% relative to the clusters formed by the algorithms using the full dataset (Figure 1A).
-Because a random number generator is used in some of algorithms to break ties where pairs of sequences have the same distance between them, similar to the He study, we replicated each algorithm and subsample 30 times.
-Across these sequencing depths, we observed that the stability of the SL and CL algorithms were highly sensitive to sampling effort relative to the AL, AGC, and DGC algorithms (Figure 1A).
+We first sought to assess calculated the MCC for for each of the clustering methods using 20, 40, 60, and 80% relative to the clusters formed by the methods using the full dataset (Figure 1A).
+Because a random number generator is used in some of methods to break ties where pairs of sequences have the same distance between them, similar to the He study, we replicated each method and subsample 30 times.
+Across these sequencing depths, we observed that the stability of the SL and CL methods were highly sensitive to sampling effort relative to the AL, AGC, and DGC methods (Figure 1A).
 Our results (Figure 1B) largely confirmed those of Figure 4C of He study with one notable exception.
-In the He study, when comparing the OTUs formed using the AL algorithm with 60% of the data, they observed a mean MCC value of approximately 0.63 (95% confidence interval between approximately 0.15 and 0.75).
+In the He study, when comparing the OTUs formed using the AL method with 60% of the data, they observed a mean MCC value of approximately 0.63 (95% confidence interval between approximately 0.15 and 0.75).
 In contrast, we observed a mean value of `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "mean"], 2)` (95% confidence interval between `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "lci"], 2)` and `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "uci"], 2)`).
-This result suggests that the AL algorithm was far more stabile than indicated in the He study.
+This result suggests that the AL method was far more stabile than indicated in the He study.
 Regardless, it supports the assertion that the clustering observed for the subset of the data does not represent the same clustering that was found with the full dataset; however, the significance of these differences is unclear.
@@
 Second, rarefaction curves calculated using clusters obtained using a portion of the dataset did not overlap with rarefaction curves generated using clusters generated from the full dataset.
-This result was originally observed in the Roesch study using the complete linkage algorithm and was reproduced in the He study where this result was more pronounced problem when using the CL, SL, and DGC algorithms relative to the other algorithms.
-The He and Roesch studies both found that the CL algorithm produced fewer OTUs in the subset than in the rarefied data.
-Expanding the analysis to other algorithms, the He study found that the SL algorithm produced more OTUs, the AGC produced fewer, and the other algorithms produced similar numbers of OTUs than expected when comparing the subsetted data to the rarefied data.
+This result was originally observed in the Roesch study using the complete linkage method and was reproduced in the He study where this result was more pronounced problem when using the CL, SL, and DGC methods relative to the other methods.
+The He and Roesch studies both found that the CL method produced fewer OTUs in the subset than in the rarefied data.
+Expanding the analysis to other methods, the He study found that the SL method produced more OTUs, the AGC produced fewer, and the other methods produced similar numbers of OTUs than expected when comparing the subsetted data to the rarefied data.
 Our results support those of these previous studies (Figure 2).
 It was clear that inter-method differences were generally more pronounced than the differences observed between rarefying from the full dataset and from clustering the subsetted data.
-The number of OTUs observed was largest using the CL algorithm, followed by the open-reference algorithm.
-The AL, AGC, and DGC algorithms all provided comparable numbers of OTUs.
-Finally, the closed-referene and SL algorithms generated the fewest number of OTUs.
+The number of OTUs observed was largest using the CL method, followed by the open-reference method.
+The AL, AGC, and DGC methods all provided comparable numbers of OTUs.
+Finally, the closed-referene and SL methods generated the fewest number of OTUs.
 
@@
 The He study observed adonis R statistics of 0.02 (closed-reference), 0.03 (open-reference), 0.07 (CL, AGC, DGC), and 0.16 (SL and AL).
-Regardless of the statistical or biological significance of these results, the analysis does not make sense since the *de novo* and open-reference approaches do not consistently label the OTUs that sequences belong to when the clustering algorithms are run multiple times with different random number seeds.
+Regardless of the statistical or biological significance of these results, the analysis does not make sense since the *de novo* and open-reference approaches do not consistently label the OTUs that sequences belong to when the clustering methods are run multiple times with different random number seeds.
 To overcome this, the authors selected representative sequences from each OTU and used those representative sequences to link OTU assignments between the different sized sequence sets.
@@
 First, each of the *de novo* methods varied in how sensitive their MCC values were to additional sequences (Figure 1C).
-The SL and CL algorithms were the most sensitive; however, the MCC values using 80 and 100% of the data did not vary meaningfully when using *de novo* algorithms.
-Second, the AL algorithm out-performed the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL (Figure 1D).
-Third, with the possible exception of the CL algorithm, the MCC values for each of the only demonstrated a small amount of variation between runs of the algorithm with a different ordering of the input sequences.
-This indicates that although there may be variation between executions of the same algorithm, they produce OTU assignments that are equally good.
+The SL and CL methods were the most sensitive; however, the MCC values using 80 and 100% of the data did not vary meaningfully when using *de novo* methods.
+Second, the AL method out-performed the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL (Figure 1D).
+Third, with the possible exception of the CL method, the MCC values for each of the only demonstrated a small amount of variation between runs of the method with a different ordering of the input sequences.
+This indicates that although there may be variation between executions of the same method, they produce OTU assignments that are equally good.
 Revisiting the concept of stability, we question the value of obtaining stable OTUs when the the full dataset is not optimally assigned to OTUs.
-Our analysis indicates that the best algorithm for assigning the Canadian soils sequences to OTUs using a 97% threshold is the AL algorithm.
+Our analysis indicates that the best method for assigning the Canadian soils sequences to OTUs using a 97% threshold is the AL method.
 
@@
 
-***Deep sampling of 16S rRNA genes.*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering algorithms.
+***Deep sampling of 16S rRNA genes.*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering methods.
 First, the Canadian soil dataset that the He study used to analyze the stability of OTUs is one of the earliest 16S rRNA gene sequence datasets published using the 454 FLX platform.
@@
 In addition, because the original phred quality scores and flowgram data are not available, it was not possible to adequately remove sequencing errors (REFS).
-The large number of sequences that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering algorithms.
+The large number of sequences that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering methods.
 Second, the dataset used in the He study covered the V9 region of the 16S rRNA gene.
-For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference algorithms.
+For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference methods.
 Of the `r nrow(gg_summary)` sequences in the default QIIME database, only `r n_v4_overlap` fully cover the V9 region.
@@
 Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene.
-With these points in mind, we compared the clustering assignment for each of these algorithms using a time series experiment that was obtained using mouse stool (REFS).
+With these points in mind, we compared the clustering assignment for each of these methods using a time series experiment that was obtained using mouse stool (REFS).
 The MiSeq platform was used to generate `r n_miseq_seqs` sequences from the V4 region of the 16S rRNA gene of `r n_groups` samples.
 Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02% (REFS).
-Although no dataset is perfect for exhaustively testing these clustering algorithms, this dataset was useful for demonstrating several points.
+Although no dataset is perfect for exhaustively testing these clustering methods, this dataset was useful for demonstrating several points.
 First, when using 60% of the data the stability relationships amongst the different methods were similar to what we observed using the He dataset (Figure 3AB).
@@
 Second, the MCC values calculated relative to the distances between sequences were generally higher than was observed for the He dataset for all of the methods except for CL and SL.
-Surprisingly, the MCC values for the DGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "dgc","mean"],2)`)  and AGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "agc","mean"],2)`) algorithms were comparable to the AL algorithm (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "an","mean"],2)`; Figure 3CD).
-This result suggests that the optimal algorithm may be database-dependent.
+Surprisingly, the MCC values for the DGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "dgc","mean"],2)`)  and AGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "agc","mean"],2)`) methods were comparable to the AL method (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "an","mean"],2)`; Figure 3CD).
+This result suggests that the optimal method may be database-dependent.
 Finally, as was observed with the He dataset, there was little variation in MCC values observed among the 30 randomizations.
-Therefore, although the algorithms have a stochastic component, the OTU assignments do not vary meaningfully.
-The results from both the Canadian soil and murine microbiota datasets demonstrate that the *de novo* algorithms can generate very stable OTU assignments, as defined by the He study, and that the assignments are highly reproducible.
-Most importantly, these analyses demonstrate that the OTU assignments using the AL, AGC, and DGC *de novo* algorithms are consistently more robust than either of the reference-based methods.
+Therefore, although the methods have a stochastic component, the OTU assignments do not vary meaningfully.
+The results from both the Canadian soil and murine microbiota datasets demonstrate that the *de novo* methods can generate very stable OTU assignments, as defined by the He study, and that the assignments are highly reproducible.
+Most importantly, these analyses demonstrate that the OTU assignments using the AL, AGC, and DGC *de novo* methods are consistently more robust than either of the reference-based methods.
 
-***Evalution of an open-source alternative to USEARCH.*** The AGC and DGC algorithms appear to perform as well and possibly than the hierarchical clustering algorithms for some datasets.
-These algorithms utilize the USEARCH program (REF), which is not available as open source code and is only available for free to academic users as a 32-bit program.
+***Evalution of an open-source alternative to USEARCH.*** The AGC and DGC methods appear to perform as well and possibly than the hierarchical clustering methods for some datasets.
+These methods utilize the USEARCH program (REF), which is not available as open source code and is only available for free to academic users as a 32-bit program.
 Access for non-academic users and those needing the 64-bit version is available commercially from the developer.
@@
 In contrast, VSEARCH does not utilize the heuristic, which the VSEARCH developers claim enhances the sensitivity relative to USEARCH.
-Using the two datasets, we determined whether the AGC and DGC algorithms as implemented by the two methods yielded sequencing assignments of similar quality.
-In general the overall trends that we observed with the USEARCH-version of AGC and DGC were also observed with the VSEARCH-verion of the algorithms.
-When we compared the two implementations of the AGC and DGC algorithms, the OTUs generated by the VSEARCH-verion of the algorithms were as stable or more stable than the USEARCH-version when using 60% of the datasets.
+Using the two datasets, we determined whether the AGC and DGC methods as implemented by the two methods yielded sequencing assignments of similar quality.
+In general the overall trends that we observed with the USEARCH-version of AGC and DGC were also observed with the VSEARCH-verion of the methods.
+When we compared the two implementations of the AGC and DGC methods, the OTUs generated by the VSEARCH-verion of the methods were as stable or more stable than the USEARCH-version when using 60% of the datasets.
 In addition, the MCC values for the entire datasets, calculated relative to the distance matrix, were virtually indistinguishable (Figure 4).
-These results are a strong indication that VSEARCH is a suitable and possibly better replacement for USEARCH for executing the AGC and DGC algorithms.
+These results are a strong indication that VSEARCH is a suitable and possibly better replacement for USEARCH for executing the AGC and DGC methods.
 
@@
 
-***Problems with reference-based clustering in general and as implemented in QIIME.*** The He study indicated that the closed-reference algorithm generated perfectly stable OTUs.
-This was unsurprising since, by definition, the algorithm represents a one-to-one mapping of reads to a reference and treats the input sequences independently.
+***Problems with reference-based clustering in general and as implemented in QIIME.*** The He study indicated that the closed-reference method generated perfectly stable OTUs.
+This was unsurprising since, by definition, the method represents a one-to-one mapping of reads to a reference and treats the input sequences independently.
 An important test that was not performed in the He study was to determine whether the clustering was sensitive to the order of the sequences in the database.
 The default QIIME database, which was used in the He study and by others, contains full-length sequences that are at most 97% similar to each other.
-We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference algorithm with the full MiSeq dataset, which contained `r n_unique_miseq_seqs` unique sequences.
+We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference method with the full MiSeq dataset, which contained `r n_unique_miseq_seqs` unique sequences.
 Surprisingly, we observed that the number of OTUs generated was not the same in each of the randomizations.
@@
 This surprising result was likely due to the performance of the USEARCH heuristic.
-As with the AGC and DGC algorithms, if the order of the reference sequences differed between runs, then the heuristic criteria would be triggered at different points.
-To test this further, we substituted VSEARCH for USEARCH in the closed-reference algorithm.
+As with the AGC and DGC methods, if the order of the reference sequences differed between runs, then the heuristic criteria would be triggered at different points.
+To test this further, we substituted VSEARCH for USEARCH in the closed-reference method.
 When we used VSEARCH the original ordering of the reference sequences and all randomizations were able to map `r rand_ref_v["original", 1]` sequences to reference OTUs.
@@
 Among these dereplicated V4 sequences there were `r n_small_v4_dist` pairs of sequences that were more than 97% similar to each other.
-The presence of duplicate V4 reference sequences explains the lack of labeling stability when using either USEARCH or VSEARCH to carry out the closed-reference algorithm.
+The presence of duplicate V4 reference sequences explains the lack of labeling stability when using either USEARCH or VSEARCH to carry out the closed-reference method.
 We suspect that the reference database was designed to only include sequences that were at most 97% similar to each other way to overcome the limitations of the USEARCH search heuristic.
 
-Beyond comparing the abundance of specific OTUs across samples, the reference database is used in the open and closed-reference algorithms to generate OTU labels that are used in several downstream applications.
+Beyond comparing the abundance of specific OTUs across samples, the reference database is used in the open and closed-reference methods to generate OTU labels that are used in several downstream applications.
 It is commonly used to extract information from a reference phylogenetic tree to carrying out UniFrac-based analyses (REFS) and to identify reference genomes for performing analyses such as PICRUSt (REFS).
@@
 Through the use of the pairwise distances between sequences, we were able to use the MCC to demonstrate that, in general, the AL, AGC, and DGC methods perform better than the others.
-Although there is concern that running the algorithms multiple times yields different clusterings, we have shown that there is little variation in their MCC values.
-This suggests that the different clusterings by the same algorithm are equally good.
-Finally, it is impossible to obtain a clustering with no false positives or false negatives and the optimal algorithm may vary by dataset.
-With this in mind, researchers are encouraged to run the AL and VSEARCH DGC algorithms and calculate and report their MCC values.
+Although there is concern that running the methods multiple times yields different clusterings, we have shown that there is little variation in their MCC values.
+This suggests that the different clusterings by the same method are equally good.
+Finally, it is impossible to obtain a clustering with no false positives or false negatives and the optimal method may vary by dataset.
+With this in mind, researchers are encouraged to run the AL and VSEARCH DGC methods and calculate and report their MCC values.
 
-Our analysis of those algorithms that implemented USEARCH as a method for clustering sequences revealed that its heuristic limited its sensitivity.
+Our analysis of those methods that implemented USEARCH as a method for clustering sequences revealed that its heuristic limited its sensitivity.
 When we replaced USEARCH with VSEARCH, the clustering quality improved.
 Although there may be parameters in USEARCH that can be tuned to improve the heuristic, these parameters are likely dataset dependent.
-Based on the data presented in this study, its availability as an open source, and free program, VSEARCH should replace USEARCH in these clustering algorithms.
+Based on the data presented in this study, its availability as an open source, and free program, VSEARCH should replace USEARCH in these clustering methods.
 Furthermore, although not tested in our study, VSEARCH can be parallelized leading to potentially significant improvements in speed.
@@
 
-For the first time, we assessed the ability of reference-based clustering algorithms to capture the actual similarity of the sequences in a dataset.
-Several studies have lauded both the open and closed-reference approaches for generating reproducible clusterings, yet we showed that both reference-based approaches did a poor job of representing the similarity between the sequences compared to the *de novo* approaches.
+For the first time, we assessed the ability of reference-based clustering methods to capture the actual distance between the sequences in a dataset.
+Several studies have lauded both the open and closed-reference approaches for generating reproducible clusterings, yet we showed that both reference-based approaches did a poor job of representing the distance between the sequences compared to the *de novo* approaches.
 So, although the clusterings are reproducible and stable across a range of library sizes, the clusterings are a poor representation of the data.
@@
 
-Because the open-reference algorithm is a hybrid of the closed-reference and DGC algorithms, it is also negatively affected by the problems with using USEARCH for both algorithms.
-An added problem with the open-reference algorithm is that the two phases of the algorithm employ different thresholds.
+Because the open-reference method is a hybrid of the closed-reference and DGC methods, it is also negatively affected by the problems with using USEARCH for both methods.
+An added problem with the open-reference method is that the two phases of the method employ different thresholds.
 In the closed-reference step, sequences must be within a threshold of a reference to be in the same OTU.
 This means that two sequences that are 97% similar to a reference and are joined into the same OTU, may only be 94% similar to each other.
-In the DGC step, the goal is to approximate the AL algorithm which requires that, on average, the sequences within an OTU are at least 97% similar to each other.
+In the DGC step, the goal is to approximate the AL method which requires that, on average, the sequences within an OTU are at least 97% similar to each other.
 The end result of the open-reference approach is that sequences that are similar to previously observed sequences are clustered with one threshold while those that are not similar to previously observed sequences are clustered with a different threshold.
 
-As the throughput of sequencing technologies have improved, algorithm development to keep pace.
-*De novo* clustering algorithms are considerably slower and more computationally intensive than reference-based algorithms and the greedy *de novo* algorithms are faster than the hierarchical.
-In our experience (REFS), the most significant detriment to execution speed of the *de novo* algorithms has been in adequate removal of sequencing error.
+As the throughput of sequencing technologies have improved, method development to keep pace.
+*De novo* clustering methods are considerably slower and more computationally intensive than reference-based methods and the greedy *de novo* methods are faster than the hierarchical.
+In our experience (REFS), the most significant detriment to execution speed of the *de novo* methods has been in adequate removal of sequencing error.
 As the rate of sequencing error increases so do the number of unique sequences that must be clustered.
-The speed of the *de novo* algorithms scales approximately quadratically, so that doubling the number of sequences results in a four-fold increase in the time required to execute the algorithm.
+The speed of the *de novo* methods scales approximately quadratically, so that doubling the number of sequences results in a four-fold increase in the time required to execute the method.
 The rapid expansion in sequencing throughput has been likened to the Red Queen in Lewis Carroll's, *Through the Looking-Glass* who must run in place to keep up to her changing surroundings (REF).
-Microbial ecologists must continue to refine clustering algorithms to better handle the size of the datasets, but they must also take steps to improve the quality of the underlying data.
+Microbial ecologists must continue to refine clustering methods to better handle the size of the datasets, but they must also take steps to improve the quality of the underlying data.
 Ultimately, objective standards must be applied to assess the quality of the data and the quality of OTU clustering.
@@
 The order of the sequences in the full dataset was randomly permuted without replacement to generate an additional 30 datasets.
-To perform the hierarchical clustering algorithms and to generate a distance matrix we followed their approach of the He study by calculating distances based on pairwise global alignments using the `pairwise.dist` command in mothur using the default Needleman-Wunsch alignment algorithm and parameters.
+To perform the hierarchical clustering methods and to generate a distance matrix we followed their approach of the He study by calculating distances based on pairwise global alignments using the `pairwise.dist` command in mothur using the default Needleman-Wunsch alignment method and parameters.
 It should be noted that this method has been strongly discouraged (REFS).
-Execution of the hierarchical clustering algorithms was performed as described in the original He study using mothur (v.1.37) and using the QIIME (v.1.9.1) parameter profiles provided in the supplementary material from the He study for the greedy and reference-based clustering algorithms.
+Execution of the hierarchical clustering methods was performed as described in the original He study using mothur (v.1.37) and using the QIIME (v.1.9.1) parameter profiles provided in the supplementary material from the He study for the greedy and reference-based clustering methods.
 
@@
 Briefly, we compared the OTU assignments for pairs of sequences to the distance matrix that was calculated between all pairs of aligned sequences.
-For each dataset that was clustered, those sequences that were in the same OTU and had a similarity greater than 97% were TPs, those that were in different OTUs and had a similarity less than 97% were TNs, those that were in the same OTU and had a similarity less than 97% were FPs, and those that were in different OTUs and had a similarity greater than 97% were FNs.
+For each dataset that was clustered, those sequences that were in the same OTU and had a distance less than 3% were TPs, those that were in different OTUs and had a distance greater than 3% were TNs, those that were in the same OTU and had a distance greater than 3% were FPs, and those that were in different OTUs and had a distance less than 3% were FNs.
 The MCC was counted for each dataset using the formula above as implemented in the `sens.spec` command in mothur.
@@
 ## Figures
+
+**Figure 1.
+Comparison of the stability (A, B) and quality (C, D) of *de novo* and reference-based clustering methods using the Canadian soil dataset.** The average stability of the OTUs were determined by calculating the MCC with respect to the OTU assignments for the full dataset using varying sized subsamples (A).
+Thirty randomizations were performed for each fraction of the dataset and the average and 95% confidence interval are presented when using 60% of the data.
+The quality of the OTUs were determined by calculating the MCC with respect to the distances between the sequences using varying sized subsamples (C).
+Thirty randomizations were performed for each fraction of the dataset and the average and 95% confidence interval are presented when using the full dataset (D).
+The vertical gray line indicates in A and C indicates the fraction of the dataset represented in B and D, respectively.
+
+**Figure 2.
+The clustering methods varied in their ability to generate the same number of OTUs using a subset of the data as were observed when the full dataset was rarefied.** The subsetted data are depicted by closed circles and the data from the rarefied full dataset is depicted by the open circles.
+
+**Figure 3.
+Comparison of the stability (A, B) and quality (C, D) of *de novo* and reference-based clustering methods using the murine dataset.** The average stability of the OTUs were determined by calculating the MCC with respect to the OTU assignments for the full dataset using varying sized subsamples (A).
+Thirty randomizations were performed for each fraction of the dataset and the average and 95% confidence interval are presented when using 60% of the data.
+The quality of the OTUs were determined by calculating the MCC with respect to the distances between the sequences using varying sized subsamples (C).
+Thirty randomizations were performed for each fraction of the dataset and the average and 95% confidence interval are presented when using the full dataset (D).
+The vertical gray line indicates in A and C indicates the fraction of the dataset represented in B and D, respectively.
+
+**Figure 4. The VSEARCH OTUs generated by the AGC and DGC methods were comparable to those generated using USEARCH.**

### 2015-10-08 Schloss_Cluster_PeerJ_2015.Rmd
Change title

@@
 ---
-title: "**Accuracy and precision of 16S rRNA gene sequence assignments to OTUs is more important than the stability of OTUs**"
+title: "***De novo* clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units**"
 author: Patrick D. Schloss* and Sarah L. Westcott
@@
 * references
+* abstract
 * revise
-* subsample vs. sub-sample
 

### 2015-10-08 Schloss_Cluster_PeerJ_2015.Rmd
Write abstract

@@
 * references
-* abstract
 * revise
@@
 \newpage
+## Abstract
 
-## Abstract
+**Background.** 16S rRNA gene sequences are routinely assigned to operational taxonomic units (OTUs) that are then used to analyze complex microbial communities.
+A number of methods have been employed to carry out the assignment of 16S rRNA gene sequences to OTUs leading to confusion over which method is the best.
+A recent study has suggested that a clustering method should be selected based on its ability to generate stable OTU assignments as additional sequences are added to the dataset rather than the ability of the method to properly represent the distances between the sequences.
+It was thus necessary to replicate the previous analysis and assess the quality of the OTU assignments.
+
+**Methods.** Our analysis implemented five *de novo* clustering algorithms including the single linkage, complete linkage, average linkage, abundance-based greedy clustering, distance-based greedy clustering and two reference-based methods including the open and closed-reference methods.
+By varying the number of sequences sampled from each of two previously published datasets we used the Matthew’s Correlation Coefficient (MCC) to assess the stability of OTU assignments and their quality.
+
+**Results.** Our analysis demonstrated that the stability of OTU assignments does not predict to the ability of the methods to reflect the distance between the sequences.
+We found that the average linkage and the distance and abundance-based greedy clustering methods out performed the open and closed-reference methods.
+Depending on the dataset being analyzed, the average linkage or greedy clustering algorithms provided the best dataset.
+Further exploration of the reference-based methods indicated that when using USEARCH to identify the closest reference the results were sensitive to the order of the reference sequences.
+When USEARCH was replaced by VSEARCH the quality and stability of the OTU assignments by the reference-based methods were improved.
+Finally, we demonstrated that for the greedy algorithms VSEARCH produced assignments that were comparable to those produced by USEARCH making VSEARCH a viable free and open source alternative to USEARCH.
+
+**Discussion.** This study demonstrates that the quality of clustering assignments needs to be assessed for multiple methods per dataset to identify the optimal clustering method for the dataset.
+The analysis also calls into question the quality and stability of OTU assignments generated by the open and closed-reference methods as implemented in current versions of QIIME.
 

### 2015-10-12 code/process_mice.sh
Keep filter file

@@ -17,6 +17,8 @@
 # Produces...
 # * *.precluster.pick.pick.fasta
 # * *.precluster.uchime.pick.pick.count_table
+# * *.precluster.pick.pds.wang.pick.taxonomy
+# * *.filter
 #
 ################################################################################
 
@@ -77,9 +79,9 @@ rm $MISEQ_PATH/*.trim.contigs.good.unique.good.filter.unique.precluster.denovo.u
 rm $MISEQ_PATH/*.trim.contigs.good.unique.good.filter.unique.precluster.pick.fasta
 rm $MISEQ_PATH/*.trim.contigs.good.unique.good.filter.unique.precluster.pick.pds.wang.tax.summary
 rm $MISEQ_PATH/*.trim.contigs.good.unique.good.filter.unique.precluster.pick.pds.wang.taxonomy
-rm $MISEQ_PATH/*.filter
 
 #keeping...
 # $MISEQ_PATH/*.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.fasta
 # $MISEQ_PATH/*.trim.contigs.good.unique.good.filter.unique.precluster.uchime.pick.pick.count_table
 # $MISEQ_PATH/*.trim.contigs.good.unique.good.filter.unique.precluster.pick.pds.wang.pick.taxonomy
+# $MISEQ_PATH/*.filter

### 2015-10-12 code/process_mice.sh
Changed mind about using filter

@@ -18,7 +18,6 @@
 # * *.precluster.pick.pick.fasta
 # * *.precluster.uchime.pick.pick.count_table
 # * *.precluster.pick.pds.wang.pick.taxonomy
-# * *.filter
 #
 ################################################################################
 
@@ -79,9 +78,9 @@ rm $MISEQ_PATH/*.trim.contigs.good.unique.good.filter.unique.precluster.denovo.u
 rm $MISEQ_PATH/*.trim.contigs.good.unique.good.filter.unique.precluster.pick.fasta
 rm $MISEQ_PATH/*.trim.contigs.good.unique.good.filter.unique.precluster.pick.pds.wang.tax.summary
 rm $MISEQ_PATH/*.trim.contigs.good.unique.good.filter.unique.precluster.pick.pds.wang.taxonomy
+rm $MISEQ_PATH/*.filter
 
 #keeping...
 # $MISEQ_PATH/*.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.fasta
 # $MISEQ_PATH/*.trim.contigs.good.unique.good.filter.unique.precluster.uchime.pick.pick.count_table
 # $MISEQ_PATH/*.trim.contigs.good.unique.good.filter.unique.precluster.pick.pds.wang.pick.taxonomy
-# $MISEQ_PATH/*.filter

### 2015-10-12 Schloss_Cluster_PeerJ_2015.Rmd
Incease scipen option
Entered some references

@@
 * revise
+* add sensitivity/specificity of closed-reference searching
 
@@
 
-options(scipen=999)
+options(scipen=99999)
 
@@
 To simplify the complexity of these large datasets, it is helpful to cluster sequences into meaningful bins.
-These bins, known as operational taxonomic units (OTUs), are commonly used to compare the biodiversity contained within and between different samples.
-Such comparisons have enabled researchers to characterize the microbiota associated with the human body (REF), soil (REF), aquatic ecosystems (REF), and numerous other environments.
+These bins, commonly known as operational taxonomic units, are  used to compare the biodiversity contained within and between different samples [@schloss2011assessing].
+Such comparisons have enabled researchers to characterize the microbiota associated with the human body [@human2012structure], soil [@shade2013streptomycin], aquatic ecosystems [@gilbert2012defining], and numerous other environments.
 Within the field of microbial ecology a convention has emerged where sequences are clustered into OTUs using a threshold of 97% similarity or a distance of 3%.
 One advantage of the OTU-based approach is that the definition of the bins is operational and can be changed to suit the needs of the particular project.
-However, with the dissemination of clustering methods within packages such as mothur and QIIME and stand-alone tools (REFS), it is important to understand how different clustering methods implement this conventional OTU threshold.
+However, with the dissemination of clustering methods within software such as mothur [@schloss2009introducing], QIIME [@caporaso2010qiime], and other tools [@edgar2010search, @sun2009esprit, @10.7717/peerj.593], it is important to understand how different clustering methods implement this conventional OTU threshold.
 Furthermore, it is necessary to understand how method choice affects the precision and accuracy of assigning sequences to OTUs.
@@
 
-The first approach has been referred to as phylotyping or closed reference clustering.
+The first approach has been referred to as phylotyping [@schloss2011assessing] or closed reference clustering [@navas2013advancing].
 This approach involves comparing sequences to a curated database and then clustering sequences together that are similar to the same reference sequence.
@@
 The strength of the reference based approach is that the methods are generally fast, scaling linearly with the number of sequences being clustered.
-A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level in the Linnean taxonomic hierarchy.
+A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@schloss2011assessing].
 Ultimately, this approach is an example of supervised classification.
 
-The second approach has been referred to as *de novo* clustering.
+The second approach has been referred to as distance-based [@schloss2011assessing] or *de novo* clustering [@navas2013advancing].
 In this method, the distance between sequences is used to cluster sequence rather than the distance to a reference database.
@@
 The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences.
-If error rates can be reduced through stringent quality control measures then these problems can be overcome (REF).
-As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods.
-Previous comparisons of hierarchical and heuristic methods have shown that the average neighbor method, a hierarchical method outperforms the other *de novo* clustering method.
+If error rates can be reduced through stringent quality control measures then these problems can be overcome [@kozich2013development].
+As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@edgar2010search, @sun2009esprit, @10.7717/peerj.593].
+Previous comparisons of hierarchical and heuristic methods have shown that the average neighbor method, a hierarchical method, outperformed the other *de novo* clustering methods.
 An early analysis using 16S rRNA gene sequence data generated using the 454 sequencing platform found that if these methods were applied to the same dataset using a subset of the sequences then a different number of OTUs were observed compared to clustering the full dataset and rarefying to the same number of sequences.
@@
 The strength of *de novo* clustering is its independence of references for carrying out the clustering step.
-After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU (ref).
+After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU [@schloss2011assessing].
 For this reason, *de novo* clustering has been preferred across the field.
-In contrast to closed-reference clustering, *de novo* clustering is an example of unsupervised clustering.
 
@@
 This approach has been called open-reference clustering.
-This method involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference.
+This method involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference [@navas2013advancing].
 This method should exploit the strengths of both closed-referene and *de novo* clustering; however, the different OTU definitions generated by both approaches poses a possible problem when the methods are combined.
-An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels (REF).
+An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels [@schloss2011assessing].
 For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average neighbor method.
@@
 
+\bibliography
+
 ## Figures

### 2015-10-12 Schloss_Cluster_PeerJ_2015.Rmd
Remove slow steps from Rmd file

@@
 ```{r, cache=TRUE}
-gg_summary <- read.table(file="data/gg_13_8/gg_13_8_97.v19.summary", header=T)
-n_v4_overlap <- sum(gg_summary$start <= 3967 & gg_summary$end >= 6116)
-n_v9_overlap <- sum(gg_summary$start <= 10179 & gg_summary$end >= 10683)
+n_overlap <- read.table(file="data/gg_13_8/gg_13_8_97.overlap.count", row.names=1)
+n_v4_overlap <- n_overlap["v4", 1]
+n_v9_overlap <- n_overlap["v9", 1]
 
@@
 ```{r, cache=TRUE}
-v4_dist <- read.table(file="data/gg_13_8/gg_13_8_97.v4_ref.unique.dist")
-n_small_v4_dist <- sum(v4_dist$V3<=0.03)
+small_dists <- read.table(file="data/gg_13_8/gg_13_8_97.small_dist.count", row.names=1)
+n_small_v4_dist <- small_dists["v4",1]
+n_small_v19_dist <- small_dists["v19",1]
 
-v19_dist <- read.table(file="data/gg_13_8/gg_13_8_97.v19_ref.unique.dist")
-n_small_v19_dist <- sum(v19_dist$V3<=0.03)
-
-v19_summary <- read.table(file="data/gg_13_8/gg_13_8_97.v19.summary", header=T)
-
-n_v4_unique <- length(scan("data/gg_13_8/gg_13_8_97.v4_ref.names", what="", sep="\n", quiet=T))
+n_v4_unique <- length(scan(file="data/gg_13_8/gg_13_8_97.v4_ref.names", what="", sep="\n", quiet=T))
 

### 2015-10-12 Schloss_Cluster_PeerJ_2015.Rmd
Remove slow to calculate values from Rmd file

@@
 
-n_miseq_seqs <- nrow(read.table(file="data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.redundant.summary"))
-
-n_unique_miseq_seqs <- nrow(read.table(file="data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.summary"))
-
-n_groups <- ncol(read.table(file="data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.denovo.uchime.pick.pick.count_table")) - 2 #subtract 2 to remove the sequence name and total number of sequences
+miseq_info <- read.table(file="data/miseq/miseq.seq.info", row.names=1, quote=F)
+n_groups <- miseq_info["n_groups",1]
+n_miseq_seqs <- miseq_info["n_miseq_seqs",1]
+n_unique_miseq_seqs <- miseq_info["n_unique_miseq_seqs",1]
 

### 2015-10-12 code/count_region_overlap.R
Add number of full length sequences

@@ -1,5 +1,6 @@
 gg_summary <- read.table(file="data/gg_13_8/gg_13_8_97.v19.summary", header=T)
 n_v4_overlap <- sum(gg_summary$start <= 3967 & gg_summary$end >= 6116)
 n_v9_overlap <- sum(gg_summary$start <= 10179 & gg_summary$end >= 10683)
+n_v19_total <- nrow(gg_summary)
 
-write.table(file="data/gg_13_8/gg_13_8_97.overlap.count", c(v4=n_v4_overlap, v9=n_v9_overlap), col.names=F, quote=F)
+write.table(file="data/gg_13_8/gg_13_8_97.overlap.count", c(v4=n_v4_overlap, v9=n_v9_overlap, v19=n_v19_total), col.names=F, quote=F)

### 2015-10-12 Schloss_Cluster_PeerJ_2015.Rmd
Fix scipen value
Fix problem with inline code

@@
 
-options(scipen=99999)
+options(scipen=9999999)
 
@@
 n_v9_overlap <- n_overlap["v9", 1]
+n_v19_total <- n_overlap["v19", 1]
 
-miseq_info <- read.table(file="data/miseq/miseq.seq.info", row.names=1, quote=F)
+miseq_info <- read.table(file="data/miseq/miseq.seq.info", row.names=1)
 n_groups <- miseq_info["n_groups",1]
@@
 For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference methods.
-Of the `r nrow(gg_summary)` sequences in the default QIIME database, only `r n_v4_overlap` fully cover the V9 region.
+Of the `r n_v19_total` sequences in the default QIIME database, only `r n_v4_overlap` fully cover the V9 region.
 In contrast, `r n_v4_overlap` of the sequences fully covered the V4 region.

### 2015-10-12 Schloss_Cluster_PeerJ_2015.Rmd
Fix scipen value
More tweaking of bibliographic formatting

@@
 
-options(scipen=9999999)
+options(scipen=999999999)
 
@@
 One advantage of the OTU-based approach is that the definition of the bins is operational and can be changed to suit the needs of the particular project.
-However, with the dissemination of clustering methods within software such as mothur [@schloss2009introducing], QIIME [@caporaso2010qiime], and other tools [@edgar2010search, @sun2009esprit, @10.7717/peerj.593], it is important to understand how different clustering methods implement this conventional OTU threshold.
+However, with the dissemination of clustering methods within software such as mothur [@schloss2009introducing], QIIME [@Caporaso2010], and other tools [@edgar2010search, @sun2009esprit, @10.7717/peerj.593], it is important to understand how different clustering methods implement this conventional OTU threshold.
 Furthermore, it is necessary to understand how method choice affects the precision and accuracy of assigning sequences to OTUs.
@@
 The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences.
-If error rates can be reduced through stringent quality control measures then these problems can be overcome [@kozich2013development].
+If error rates can be reduced through stringent quality control measures then these problems can be overcome [@Kozich2013].
 As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@edgar2010search, @sun2009esprit, @10.7717/peerj.593].

### 2015-10-13 Schloss_Cluster_PeerJ_2015.Rmd
Does printbibliography work?
Change scipen
Turn off scientific notation
Turn off caching and name chunks

@@
 
-```{r knitr.settings, eval=TRUE, echo=FALSE, cache=TRUE}
+```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
 opts_chunk$set("tidy" = TRUE)
@@
 
-options(scipen=999999999)
-
 inline_hook <- function(x){
   if(is.numeric(x)){
-    paste(format(x,big.mark=',', digits=2))
+    paste(format(x,big.mark=',', digits=2, scientific=FALSE))
   }
@@
 ## Results and Discussion
-```{r}
+```{r he_stability_analysis}
 he_stability <- read.table(file="data/process/he.mcc_ref.summary", header=T)
@@
 
-```{r, cache=TRUE}
+```{r region_overlap}
 n_overlap <- read.table(file="data/gg_13_8/gg_13_8_97.overlap.count", row.names=1)
@@
 n_v19_total <- n_overlap["v19", 1]
+```
 
+```{r miseq_analysis}
 miseq_info <- read.table(file="data/miseq/miseq.seq.info", row.names=1)
@@
 
-```{r, cache=TRUE}
+```{r database_analysis}
 small_dists <- read.table(file="data/gg_13_8/gg_13_8_97.small_dist.count", row.names=1)
@@
 
-## References
-
-\bibliography
-
 ## Figures
@@
 **Figure 4. The VSEARCH OTUs generated by the AGC and DGC methods were comparable to those generated using USEARCH.**
+
+\newpage
+
+## References

### 2015-10-13 Schloss_Cluster_PeerJ_2015.Rmd
Add rules for rendering paper

@@
 After obtaining the 16S rRNA gene fragments from GenBank (accessions EF308591-EF361836), we followed the methods outlined by the He study by removing any sequence that contained an ambiguous base, was identified as being a chimera, and fell outside a defined sequence length.
-Although they reported observing a total of 50,542 sequences that were represented by 13,293 unique sequences, we obtained a total of `r length(scan("data/he/canada_soil.good.unique.pick.redundant.fasta", what="", sep="\n", quiet=T))/2` sequences that were represented by `r length(scan("data/he/canada_soil.good.unique.pick.fasta", what="", sep="\n", quiet=T))/2` unique sequences.
+Although they reported observing a total of 50,542 sequences that were represented by 13,293 unique sequences, we obtained a total of `r length(scan(file="data/he/canada_soil.good.unique.pick.redundant.fasta", what="", sep="\n", quiet=T))/2` sequences that were represented by `r length(scan(file="data/he/canada_soil.good.unique.pick.fasta", what="", sep="\n", quiet=T))/2` unique sequences.
 Similar to the He study, we randomly sampled, without replacement, 20, 40, 60, and 80% of the sequences from the full data set.

### 2015-10-15 code/distance.cpp
Write new mapping script that accounts for equally different matches

@@ -1,0 +1,41 @@
+#include <Rcpp.h>
+using namespace Rcpp;
+
+// [[Rcpp::export]]
+
+List map_to_reference(std::string sequence, std::vector< std::string > references) {
+
+	int n_refs = references.size();
+	int seq_length = sequence.length();
+
+	std::vector<int> matches;
+	double min_dist = 1.0;
+
+	for(int r=0;r<n_refs;r++){
+		std::string ref = references[r];
+		double difference = 0;
+		double total = 0;
+
+		for(int l=0;l<seq_length;l++){
+			char base = sequence[l];
+
+			if(!(base == '.' || ref[l] == '.'  || (base == '-' && ref[l] == '-'))){
+				if(base != ref[l]){	difference++;	}
+				total++;
+			}
+		}
+
+		double distance = (double) difference / total;
+
+		if(distance < min_dist){
+			min_dist = distance;
+			matches.resize(1);
+			matches[0] = r+1;
+		} else if (distance == min_dist){
+			matches.push_back(r+1);
+		}
+
+	}
+
+	return List::create(min_dist, matches);
+}

### 2015-10-15 code/map_mouse_to_gg.sh

@@ -2,17 +2,21 @@
 #	data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.fasta
 #	data/gg_13_8/97_otus.align
 #	data/references/silva.bact_archaea.align
+#
+#	code/map_to_reference.R
+#	code/distance.cpp
+#
+# Output:
+#	data/rand_ref/miseq.ref.mapping
 
-mothur "#align.seqs(fasta=data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.fasta, reference=data/references/silva.bact_archaea.align, outputdir=data/rand_ref, processors=4);
-	filter.seqs(fasta=data/rand_ref/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.align-data/gg_13_8/97_otus.align, vertical=T);
-	classify.seqs(fasta=data/rand_ref/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.filter.fasta,reference=data/rand_ref/97_otus.filter.fasta, taxonomy=data/references/97_otus.taxonomy, method=knn, numwanted=1, search=distance)"
+mothur "#align.seqs(fasta=data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.fasta, reference=data/references/silva.bact_archaea.align, outputdir=data/rand_ref, processors=4); screen.seqs(fasta=data/gg_13_8/97_otus.align, start=13862, end=23443, outputdir=data/rand_ref, processors=8); filter.seqs(fasta=data/rand_ref/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.align-data/rand_ref/97_otus.good.align, vertical=T, trump=., processors=8);"
+
+R -e "source('code/map_to_reference.R')"
 
 rm data/rand_ref/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.align
 rm data/rand_ref/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.align.report
+rm data/rand_ref/97_otus.good.align
+rm data/rand_ref/97_otus.bad.accnos
 rm data/rand_ref/miseq97_otus.filter
 rm data/rand_ref/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.filter.fasta
[…]

### 2015-10-15 code/map_mouse_to_gg.sh
Change number of processors to 4

@@ -9,7 +9,7 @@
 # Output:
 #	data/rand_ref/miseq.ref.mapping
 
-mothur "#align.seqs(fasta=data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.fasta, reference=data/references/silva.bact_archaea.align, outputdir=data/rand_ref, processors=4); screen.seqs(fasta=data/gg_13_8/97_otus.align, start=13862, end=23443, outputdir=data/rand_ref, processors=8); filter.seqs(fasta=data/rand_ref/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.align-data/rand_ref/97_otus.good.align, vertical=T, trump=., processors=8);"
+mothur "#align.seqs(fasta=data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.fasta, reference=data/references/silva.bact_archaea.align, outputdir=data/rand_ref, processors=4); screen.seqs(fasta=data/gg_13_8/97_otus.align, start=13862, end=23443, outputdir=data/rand_ref, processors=4); filter.seqs(fasta=data/rand_ref/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.align-data/rand_ref/97_otus.good.align, vertical=T, trump=., processors=4);"
 
 R -e "source('code/map_to_reference.R')"
 

### 2015-10-15 Schloss_Cluster_PeerJ_2015.Rmd
Edit Abstract and Introduction

@@
 A number of methods have been employed to carry out the assignment of 16S rRNA gene sequences to OTUs leading to confusion over which method is the best.
-A recent study has suggested that a clustering method should be selected based on its ability to generate stable OTU assignments as additional sequences are added to the dataset rather than the ability of the method to properly represent the distances between the sequences.
-It was thus necessary to replicate the previous analysis and assess the quality of the OTU assignments.
+A recent study suggested that a clustering method should be selected based on its ability to generate stable OTU assignments that do not change as additional sequences are added to the dataset.
+In contrast, we contend that the ability of the method to properly represent the distances between the sequences is more important.
 
 **Methods.** Our analysis implemented five *de novo* clustering algorithms including the single linkage, complete linkage, average linkage, abundance-based greedy clustering, distance-based greedy clustering and two reference-based methods including the open and closed-reference methods.
-By varying the number of sequences sampled from each of two previously published datasets we used the Matthew’s Correlation Coefficient (MCC) to assess the stability of OTU assignments and their quality.
+By varying the number of sequences sampled from each of two previously published datasets we used the Matthew’s Correlation Coefficient (MCC) to assess the quality and stability of OTU assignments.
 
-**Results.** Our analysis demonstrated that the stability of OTU assignments does not predict to the ability of the methods to reflect the distance between the sequences.
-We found that the average linkage and the distance and abundance-based greedy clustering methods out performed the open and closed-reference methods.
-Depending on the dataset being analyzed, the average linkage or greedy clustering algorithms provided the best dataset.
-Further exploration of the reference-based methods indicated that when using USEARCH to identify the closest reference the results were sensitive to the order of the reference sequences.
+**Results.** The stability of OTU assignments did not reflect the quality of the assignments.
+Depending on the dataset being analyzed, the average linkage and the distance and abundance-based greedy clustering methods generated more robust OTUs than the open and closed-reference methods.
+Further interrogation of the reference-based methods indicated that when USEARCH is used to identify the closest reference, the OTU assignments were sensitive to the order of the reference sequences.
 When USEARCH was replaced by VSEARCH the quality and stability of the OTU assignments by the reference-based methods were improved.
@@
 To simplify the complexity of these large datasets, it is helpful to cluster sequences into meaningful bins.
-These bins, commonly known as operational taxonomic units, are  used to compare the biodiversity contained within and between different samples [@schloss2011assessing].
-Such comparisons have enabled researchers to characterize the microbiota associated with the human body [@human2012structure], soil [@shade2013streptomycin], aquatic ecosystems [@gilbert2012defining], and numerous other environments.
-Within the field of microbial ecology a convention has emerged where sequences are clustered into OTUs using a threshold of 97% similarity or a distance of 3%.
+These bins, commonly known as operational taxonomic units (OTUs), are  used to compare the biodiversity contained within and between different samples [@schloss2011assessing].
+Such comparisons have enabled researchers to characterize the microbiota associated with the human body [e.g.
+@human2012structure], soil [e.g.
+@shade2013streptomycin], aquatic ecosystems [e.g.
+@gilbert2012defining], and numerous other environments.
+Within the field of microbial ecology, a convention has emerged where sequences are clustered into OTUs using a threshold of 97% similarity or a distance of 3%.
 One advantage of the OTU-based approach is that the definition of the bins is operational and can be changed to suit the needs of the particular project.
@@
 The first approach has been referred to as phylotyping [@schloss2011assessing] or closed reference clustering [@navas2013advancing].
-This approach involves comparing sequences to a curated database and then clustering sequences together that are similar to the same reference sequence.
-Reference-based clustering methods suffer when the reference sequences do not reflect the composition of the community.
+This approach involves comparing sequences to a curated database and then clustering sequences into the same OTU that are similar to the same reference sequence.
+Reference-based clustering methods suffer when the reference does not adequately reflect the biodiversity of the community.
 If a large fraction of sequences are novel, then they cannot be assigned to an OTU.
@@
 Thus, a sequence representing a fragment of the gene may be more than 97% similar to multiple reference sequences.
-Because of this, defining OTUs in the closed-reference approach is complicated because two sequences might be 97% similar to the same reference sequence; however, they may only be 94% similar to each other.
+Therefore, defining OTUs in the closed-reference approach is complicated because although two sequences might be 97% similar to the same reference sequence, they may only be 94% similar to each other.
+A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@schloss2011assessing].
 The strength of the reference based approach is that the methods are generally fast, scaling linearly with the number of sequences being clustered.
-A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@schloss2011assessing].
-Ultimately, this approach is an example of supervised classification.
 
 The second approach has been referred to as distance-based [@schloss2011assessing] or *de novo* clustering [@navas2013advancing].
-In this method, the distance between sequences is used to cluster sequence rather than the distance to a reference database.
+In this method, the distance between sequences is used to cluster sequences into OTUs rather than the distance to a reference database.
 In contrast to the efficiency of closed-reference clustering, the speed of hierarchical *de novo* clustering methods scale quadratically with the number of unique sequences.
@@
 As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@edgar2010search, @sun2009esprit, @10.7717/peerj.593].
-Previous comparisons of hierarchical and heuristic methods have shown that the average neighbor method, a hierarchical method, outperformed the other *de novo* clustering methods.
-An early analysis using 16S rRNA gene sequence data generated using the 454 sequencing platform found that if these methods were applied to the same dataset using a subset of the sequences then a different number of OTUs were observed compared to clustering the full dataset and rarefying to the same number of sequences.
-Because the initial subsampling of the dataset was not repeated in the earlier analysis, it is unclear whether the result was product of the method or sampling.
-One explanation could be that the clusters are getting better with additional sampling.
-Another observed problem with *de novo* clustering is that because it is necessary to break ties when assigning sequences to OTUs, clustering the same data multiple times may result in different clustering assignments.
-Whether the differences in assignments is meaningful is unclear; however the variation in results could represent equally valid clustering of the data.
+One critique of *de novo* approaches is that OTU assignments are sensitive to the input order of the sequences [REFS].
+Whether the differences in assignments is meaningful is unclear; however, the variation in results could represent equally valid clustering of the data.
 The strength of *de novo* clustering is its independence of references for carrying out the clustering step.
-After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU [@schloss2011assessing].
+After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU [@Schloss2011].
 For this reason, *de novo* clustering has been preferred across the field.
 
-The third approach is a hybrid of the closed-reference and *de novo* approaches.
-This approach has been called open-reference clustering.
-This method involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference [@navas2013advancing].
-This method should exploit the strengths of both closed-referene and *de novo* clustering; however, the different OTU definitions generated by both approaches poses a possible problem when the methods are combined.
-An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels [@schloss2011assessing].
-For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average neighbor method.
-Those sequences that did not classify to a known family would also be clustered using the average neighbor method.
-An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group (e.g.
-each family) is seen as being independent and can be processed separately.
+The third approach, open-reference clustering, is a hybrid of the closed-reference and *de novo* approaches [@Navas2013].
+Open-reference clustering method involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference.
+In theory, this method should exploit the strengths of both closed-referene and *de novo* clustering; however, the different OTU definitions employed by both approaches poses a possible problem when the methods are combined.
+An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels using the average linkage method [@Schloss2011].
+For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average linkage method using a 3% distance threshold.
+Those sequences that did not classify to a known family would also be clustered using the average linkage method.
+An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group is seen as being independent and can be processed separately.
 Such an approach would overcome the difficulty of mixing OTU definitions between the closed-reference and *de novo* approaches.
@@
 These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the clustering.
-Others have attempted to judge the quality of an method by its ability to generate data that parallels classification data.
-This approach is problematic because bacterial taxonomy often reflects the biases amongst bacterial systematicists and the rates of evolution across lineages are not the same.
+Others have attempted to judge the quality of an method by its ability to generate data that parallels classification data (REFS).
+This approach is problematic because bacterial taxonomy often reflects historical biases amongst bacterial systematicists.
+Furthermore, it is well known that the rates of evolution across lineages are not the same (REFS).
 We recently proposed an approach for evaluating OTU assignments using the distance between sequences (REF).
@@
 Meanwhile, those sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives.
-Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and ratios using the Matthew's correlation coefficient (MCC; REF).
-That analysis focused on assessing *de novo* clustering methods and found that the average neighbor method outperformed the other hierarchical and heuristic methods.
+Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and negatives using the Matthew's correlation coefficient (MCC; REF).
+This is an objective approach to assessing the quality of the OTU assignments.
 
@@
 They defined stability as the ability of an method to provide the same clustering on a subset of the data as was found in the full dataset.
-Related to this, the authors expressed concern that because some methods use a random number generator to break ties there may be further instability between executions of method.
-Their concept of stability does not account for the accuracy of the clustering and instead focuses on the precision of the clustering.
+Their concept of stability did not account for the accuracy of the OTU assignments and instead focused on the precision of the assignments.
 A method may be very precise, but low in accuracy.
-In the current analysis, we attempted to assess the accuracy and precision of the various clustering methods.
+In the current analysis, we assessed the accuracy and precision of the various clustering methods.
 Building on our previous analysis of clustering methods, our hypothesis was that the methods praised by the He study for their stability actually suffered a lack of accuracy.

### 2015-10-15 Schloss_Cluster_PeerJ_2015.Rmd
Editing the Results section

@@
 
-***Summary and replication of He study.*** We sought to identify the more critical analyses performed in the He study.
-Similar to their study, we obtained the Canadian soil dataset from Roesch et al.
-(REF) and processed the sequences as described in their analysis.
-In our opinion there were three important tests.
+***Summary and replication of He study.*** We obtained the Canadian soil dataset from Roesch et al.
+(REF) and processed the sequences as described by He and colleagues.
+Using these data, we reconsidered three of the more critical analyses performed in the He study.
 
-First, we sought to quantify whether the clustering observed for a subset of the data represented the same clustering that was found with the full dataset.
-The He study found that when they used the open and closed-reference methods clusters formed using the subset data most closely resembled those of the full dataset.
-Among the *de novo* clustering methods they observed that the abundance-based greedy clustering (AGC) method followed by single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) methods.
-They observed a broad range of MCC values among their AL replicates, which was not explained by the authors.
-We first sought to assess calculated the MCC for for each of the clustering methods using 20, 40, 60, and 80% relative to the clusters formed by the methods using the full dataset (Figure 1A).
-Because a random number generator is used in some of methods to break ties where pairs of sequences have the same distance between them, similar to the He study, we replicated each method and subsample 30 times.
-Across these sequencing depths, we observed that the stability of the SL and CL methods were highly sensitive to sampling effort relative to the AL, AGC, and DGC methods (Figure 1A).
-Our results (Figure 1B) largely confirmed those of Figure 4C of He study with one notable exception.
-In the He study, when comparing the OTUs formed using the AL method with 60% of the data, they observed a mean MCC value of approximately 0.63 (95% confidence interval between approximately 0.15 and 0.75).
+First, we sought to quantify whether the OTU assignments observed for a subset of the data represented the same assignments that were found with the full dataset.
+The He study found that when they used the open and closed-reference methods the OTUs formed using the subsetted data most closely resembled those of the full dataset.
+Among the *de novo* methods they observed that the abundance-based greedy clustering (AGC) method generated the most stable OTUs followed by the single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) methods.
+We first sought to assess calculated the MCC for the OTU assignments generated by each of the clustering methods using 20, 40, 60, and 80% of the sequences relative to the OTU composition formed by the methods using the full dataset (Figure 1A).
+Similar to the He study, we replicated each method and subsample 30 times because a random number generator is used in some of methods to break ties where pairs of sequences have the same distance between them.
+Across these sequencing depths, we observed that the stability of the OTUs generated by the SL and CL methods were highly sensitive to sampling effort relative to the OTUs generated by the AL, AGC, and DGC methods (Figure 1A).
+Our results (Figure 1B) largely confirmed those of Figure 4C in the He study with one notable exception.
+The He study observed a broad range of MCC values among their AL replicates when analyzing OTUs generated using 60% of the data.
+This result appeared out of character and was not explained by the authors.
+They observed a mean MCC value of approximately 0.63 (95% confidence interval between approximately 0.15 and 0.75).
 In contrast, we observed a mean value of `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "mean"], 2)` (95% confidence interval between `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "lci"], 2)` and `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "uci"], 2)`).
-This result suggests that the AL method was far more stabile than indicated in the He study.
-Regardless, it supports the assertion that the clustering observed for the subset of the data does not represent the same clustering that was found with the full dataset; however, the significance of these differences is unclear.
+This result indicates that the AL assignments were far more stable than indicated in the He study.
+Regardless, although the assignments are quite stable, it does support the assertion that the OTU assignments observed for the subset of the data do not perfectly match the assignments that were found with the full dataset as they did with the reference-based methods; however, the significance of these differences is unclear.
 
-Second, rarefaction curves calculated using clusters obtained using a portion of the dataset did not overlap with rarefaction curves generated using clusters generated from the full dataset.
-This result was originally observed in the Roesch study using the complete linkage method and was reproduced in the He study where this result was more pronounced problem when using the CL, SL, and DGC methods relative to the other methods.
+Second, the He study and the original Roesch study showed that rarefaction curves calculated using CL-generated OTU assignments obtained using a portion of the dataset did not overlap with rarefaction curves generated using OTU assignments generated from the full dataset.
 The He and Roesch studies both found that the CL method produced fewer OTUs in the subset than in the rarefied data.
-Expanding the analysis to other methods, the He study found that the SL method produced more OTUs, the AGC produced fewer, and the other methods produced similar numbers of OTUs than expected when comparing the subsetted data to the rarefied data.
+In addition, the He study found that the SL method produced more OTUs, the AGC produced fewer, and the other methods produced similar numbers of OTUs than expected when comparing the subsetted data to the rarefied data.
 Our results support those of these previous studies (Figure 2).
@@
 The AL, AGC, and DGC methods all provided comparable numbers of OTUs.
-Finally, the closed-referene and SL methods generated the fewest number of OTUs.
+Finally, the closed-reference and SL methods generated the fewest number of OTUs.
 
-Third, the authors attempted to describe the effects of the clustering instability on comparisons of communities.
-They used Adonis to test whether the community structure represented in subsetted communities resembled that of the full dataset.
-Inspection of these results indicates that although they were able to detect significant p-values, they were were of marginal biological significance.
+Third, the authors attempted to describe the effects of the OTU assignment instability on comparisons of communities.
+They used Adonis to test whether the community structure represented in subsetted communities resembled that of the full dataset when only using the unstable OTUs (REFS).
+Although they were able to detect significant p-values, they appeared to be of marginal biological significance.
 Adonis R statistics close to zero indicate the community structures from the full and subsetted datasets overlapped while values of one indicate the communities are completely different.
 The He study observed adonis R statistics of 0.02 (closed-reference), 0.03 (open-reference), 0.07 (CL, AGC, DGC), and 0.16 (SL and AL).
-Regardless of the statistical or biological significance of these results, the analysis does not make sense since the *de novo* and open-reference approaches do not consistently label the OTUs that sequences belong to when the clustering methods are run multiple times with different random number seeds.
+Regardless of the statistical or biological significance of these results, the analysis does not make sense since, by definition, representing communities based on their unstable OTUs would yield differences.
+Furthermore, the *de novo* and open-reference approaches do not consistently label the OTUs that sequences belong to when the clustering methods are run multiple times with different random number seeds.
 To overcome this, the authors selected representative sequences from each OTU and used those representative sequences to link OTU assignments between the different sized sequence sets.
@@
 It is not surprising that the only analysis that did not provide a significant p-value was for the closed-reference analysis, which is the only analysis that provides consistent OTU labels.
+Finally, the authors built off of this analysis to count the number of OTUs that were differentially represented between the subsetted and full datasets by each method.
+This analysis assumes that the OTUs generated using the full dataset were correct, which was an unsubstantiated assumption since the authors did not assess the quality of the OTU assignments.
 Because this analysis was so poorly designed, we did not seek to reproduce it.
 
-***Methods vary in their accuracy.*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs.
-A method can generate highly stable OTUs, but the OTUs may be meaningless by poorly representing a specified cutoff in the assignment of sequences to those OTUs.
-To assess the accuracy and precision of the various methods, we made use of the pairwise distance between the unique sequences to count the number of true positives and negatives and the number of false positives and negatives for each method and sampling depth so that we could calculate the average MCC value as a measure of a method's accuracy and its variation as a measure of its precision.
+***OTU assignment methods vary in their accuracy.*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs.
+A method can generate highly stable OTUs, but the OTU assignments may be meaningless if they poorly represent the specified cutoff and the actual distance bewteen the sequences.
+To assess the quality of OTU assignments by the various methods, we made use of the pairwise distance between the unique sequences to count the number of true positives and negatives and the number of false positives and negatives for each method and sampling depth.
+This enabled us to calculate the average MCC value as a measure of a method's accuracy and its variation as a measure of its precision.
 We made three important observations.
 First, each of the *de novo* methods varied in how sensitive their MCC values were to additional sequences (Figure 1C).
-The SL and CL methods were the most sensitive; however, the MCC values using 80 and 100% of the data did not vary meaningfully when using *de novo* methods.
-Second, the AL method out-performed the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL (Figure 1D).
+The SL and CL methods were the most sensitive; however, the accuracy of the OTU assignments did not meaningfully differ when 80 or 100% of the data were assigned to OTUs using the *de novo* methods.
+Second, the AL method had higher MCC values than the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL (Figure 1D).
 Third, with the possible exception of the CL method, the MCC values for each of the only demonstrated a small amount of variation between runs of the method with a different ordering of the input sequences.

### 2015-10-15 Schloss_Cluster_PeerJ_2015.Rmd
Edit the Results and Conclusion section

@@
 ***Deep sampling of 16S rRNA genes.*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering methods.
-First, the Canadian soil dataset that the He study used to analyze the stability of OTUs is one of the earliest 16S rRNA gene sequence datasets published using the 454 FLX platform.
-Developments in sequencing technology now permits the sequencing of millions of sequences for a study.
+First, it was one of the earliest 16S rRNA gene sequence datasets published using the 454 FLX platform.
+Developments in sequencing technology now permit the sequencing of millions of sequences for a study.
 In addition, because the original phred quality scores and flowgram data are not available, it was not possible to adequately remove sequencing errors (REFS).
@@
 Inadequate coverage of the V9 region would adversely affect the ability of the reference-based methods to assign sequences to OTUs.
-Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene.
-With these points in mind, we compared the clustering assignment for each of these methods using a time series experiment that was obtained using mouse stool (REFS).
+Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene (REFS).
+With these points in mind, we compared the clustering assignment for each of these methods using a time series experiment that was obtained using mouse feces (REFS).
 The MiSeq platform was used to generate `r n_miseq_seqs` sequences from the V4 region of the 16S rRNA gene of `r n_groups` samples.
@@
 With the exception for the clusters generated using CL, the methods all performed very well with stabilities greater than `r round(min(miseq_stability[miseq_stability$fraction=="0.6" & miseq_stability$method != "fn","mean"]), 2)`.
-Second, the MCC values calculated relative to the distances between sequences were generally higher than was observed for the He dataset for all of the methods except for CL and SL.
+Second, the MCC values calculated relative to the distances between sequences were generally higher than was observed for the Canadian soil dataset for all of the methods except the CL and SL methods.
 Surprisingly, the MCC values for the DGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "dgc","mean"],2)`)  and AGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "agc","mean"],2)`) methods were comparable to the AL method (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "an","mean"],2)`; Figure 3CD).
-This result suggests that the optimal method may be database-dependent.
-Finally, as was observed with the He dataset, there was little variation in MCC values observed among the 30 randomizations.
-Therefore, although the methods have a stochastic component, the OTU assignments do not vary meaningfully.
-The results from both the Canadian soil and murine microbiota datasets demonstrate that the *de novo* methods can generate very stable OTU assignments, as defined by the He study, and that the assignments are highly reproducible.
+This result suggests that the optimal method is likely to be database-dependent.
+Finally, as was observed with the Canadian soil dataset, there was little variation in MCC values observed among the 30 randomizations.
+Therefore, although the methods have a stochastic component, the OTU assignments do not vary meaningfully between runs.
+The results from both the Canadian soil and murine microbiota datasets demonstrate that the *de novo* methods can generate stable OTU assignments and that the assignments are highly reproducible.
 Most importantly, these analyses demonstrate that the OTU assignments using the AL, AGC, and DGC *de novo* methods are consistently more robust than either of the reference-based methods.
 
-***Evalution of an open-source alternative to USEARCH.*** The AGC and DGC methods appear to perform as well and possibly than the hierarchical clustering methods for some datasets.
-These methods utilize the USEARCH program (REF), which is not available as open source code and is only available for free to academic users as a 32-bit program.
+***Evalution of an open-source alternative to USEARCH.*** For some datasets the AGC and DGC methods appear to perform as well or better than the hierarchical clustering methods.
+As originally described in the He study, the AGC and DGC methods utilized the USEARCH program (REF), which is not available as open source code and is only available for free to academic users as a 32-bit program.
 Access for non-academic users and those needing the 64-bit version is available commercially from the developer.
 An alternative to USEARCH is VSEARCH, which is being developed in parallel to USEARCH as a open-source alternative.
-One subtle difference between the two programs is that USEARCH employs a heuristic to determine when to stop searching a database because it think it has found the best hit or does not believe that the database contains a hit.
-This means that the best possible match is not necessarily returned once a highly similar hit is obtained.
-In contrast, VSEARCH does not utilize the heuristic, which the VSEARCH developers claim enhances the sensitivity relative to USEARCH.
-Using the two datasets, we determined whether the AGC and DGC methods as implemented by the two methods yielded sequencing assignments of similar quality.
-In general the overall trends that we observed with the USEARCH-version of AGC and DGC were also observed with the VSEARCH-verion of the methods.
+One subtle difference between the two programs is that USEARCH employs a heuristic to generate candidate alignments whereas VSEARCH generates the actual global alignments.
+The VSEARCH developers claim that this difference enhances the sensitivity of VSEARCH relative to USEARCH.
+Using the two datasets, we determined whether the AGC and DGC methods, as implemented by the two programs, yielded OTU assignments of similar quality.
+In general the overall trends that we observed with the USEARCH-version of AGC and DGC were also observed with the VSEARCH-verion of the methods (Figure 4).
 When we compared the two implementations of the AGC and DGC methods, the OTUs generated by the VSEARCH-verion of the methods were as stable or more stable than the USEARCH-version when using 60% of the datasets.
-In addition, the MCC values for the entire datasets, calculated relative to the distance matrix, were virtually indistinguishable (Figure 4).
+In addition, the MCC values for the entire datasets, calculated relative to the distance matrix, were virtually indistinguishable.
 These results are a strong indication that VSEARCH is a suitable and possibly better replacement for USEARCH for executing the AGC and DGC methods.
@@
 
-***Problems with reference-based clustering in general and as implemented in QIIME.*** The He study indicated that the closed-reference method generated perfectly stable OTUs.
+***Problems with reference-based clustering in general and as implemented in QIIME.*** The He study and our replication attempt validated that the closed-reference method generated perfectly stable OTUs.
 This was unsurprising since, by definition, the method represents a one-to-one mapping of reads to a reference and treats the input sequences independently.
 An important test that was not performed in the He study was to determine whether the clustering was sensitive to the order of the sequences in the database.
-The default QIIME database, which was used in the He study and by others, contains full-length sequences that are at most 97% similar to each other.
+The default database used in QIIME, which was also used in the He study and by others, contains full-length sequences that are at most 97% similar to each other.
 We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference method with the full MiSeq dataset, which contained `r n_unique_miseq_seqs` unique sequences.
@@
 This surprising result was likely due to the performance of the USEARCH heuristic.
-As with the AGC and DGC methods, if the order of the reference sequences differed between runs, then the heuristic criteria would be triggered at different points.
 To test this further, we substituted VSEARCH for USEARCH in the closed-reference method.
@@
 If we instead used VSEARCH an average of `r round(100*v_overlap["orig", "mean"], 1)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*v_overlap["orig", "min"], 1)` to `r round(100*v_overlap["orig", "max"], 1)`).
-To better understand this result, we further analyzed the reference database that is used in QIIME.
+To better understand this result, we further analyzed QIIME's reference database.
 We hypothesized that within a given region there would be sequences that were more than 97% similar and possibly identical to each other.
@@
 There was also a sequence had `r max(dups[,2])-1` duplicates and contained `r dups[which.max(dups[,2]),3]` different taxonomies.
+
+ADD SOMETHING ABOUT CONFLICT BETWEEN TAXONOMIES OF REFERENCES THAT WERE TIES TO QUERY SEQUENCE.
+
 Together, these results demonstrate some of the considerable problems with the reference-based clustering of sequences.
@@
 Second, all of the sequence data from these datasets is pooled for a single analysis.
-It is unclear why anyone would ever perform an analysis based on a subset of their data (REFS).
+It is unclear why anyone would ever perform an analysis based on a subset of their data.
 Because of these points, the value of identifying stable OTUs is unclear.
@@
 Finally, it is impossible to obtain a clustering with no false positives or false negatives and the optimal method may vary by dataset.
-With this in mind, researchers are encouraged to run the AL and VSEARCH DGC methods and calculate and report their MCC values.
+With this in mind, researchers are encouraged to run the AL and VSEARCH AGC or DGC methods and calculate and report their MCC values.
 
-Our analysis of those methods that implemented USEARCH as a method for clustering sequences revealed that its heuristic limited its sensitivity.
+Our analysis of those methods that implemented USEARCH as a method for clustering sequences revealed that its heuristic limited its specificity.
 When we replaced USEARCH with VSEARCH, the clustering quality improved.
@@
 Furthermore, although not tested in our study, VSEARCH can be parallelized leading to potentially significant improvements in speed.
-Although USEARCh and VSEARCH do utilize aligned sequences, it is important to note that a sequence curation pipeline including denoising, alignment, trimming to a consistent region of the 16S rRNA gene, and chimera checking are critical to making proper inferences (REF).
+Although USEARCH and VSEARCH do not utilize aligned sequences, it is important to note that a sequence curation pipeline including denoising, alignment, trimming to a consistent region of the 16S rRNA gene, and chimera checking are critical to making proper inferences (REF).
 
-For the first time, we assessed the ability of reference-based clustering methods to capture the actual distance between the sequences in a dataset.
-Several studies have lauded both the open and closed-reference approaches for generating reproducible clusterings, yet we showed that both reference-based approaches did a poor job of representing the distance between the sequences compared to the *de novo* approaches.
-So, although the clusterings are reproducible and stable across a range of library sizes, the clusterings are a poor representation of the data.
-We also observed that the clusterings were not actually reproducible when the order of the reference sequences was randomize.
-When USEARCH was used, the actual number of sequences that mapped to the reference changed depending on the order of the reference.
-Perhaps most alarming is that the default order of the database provided the worst sensitivity of any of the randomizations we attempted.
+We have assessed the ability of reference-based clustering methods to capture the actual distance between the sequences in a dataset in parallel with *de novo* methods.
+Several studies have lauded both the open and closed-reference approaches for generating reproducible clusterings (REFS), yet we have shown that both reference-based approaches did a poor job of representing the distance between the sequences compared to the *de novo* approaches.
+Although the OTU assignments are reproducible and stable across a range of library sizes, the reference-based OTU assignments are a poor representation of the data.
+We also observed that the assignments were not actually reproducible when the order of the reference sequences was randomize.
+When USEARCH was used, the actual number of sequences that mapped to the reference changed depended on the order of the reference.
+Perhaps most alarming was that the default order of the database provided the worst MCC of any of the randomizations we attempted.
 Even when we used VSEARCH to perform closed-reference clustering and were able to obtain a consistent clusterings, we observed that the labels on the OTUs differed between randomizations.
@@
 Because the open-reference method is a hybrid of the closed-reference and DGC methods, it is also negatively affected by the problems with using USEARCH for both methods.
-An added problem with the open-reference method is that the two phases of the method employ different thresholds.
+An added problem with the open-reference method is that the two phases of the method employ different thresholds to define its OTUs.
 In the closed-reference step, sequences must be within a threshold of a reference to be in the same OTU.
 This means that two sequences that are 97% similar to a reference and are joined into the same OTU, may only be 94% similar to each other.
-In the DGC step, the goal is to approximate the AL method which requires that, on average, the sequences within an OTU are at least 97% similar to each other.
+In the DGC step, the goal is to approximate the AL method which requires that, on average, the sequences within an OTU are, on average, 97% similar to each other.
 The end result of the open-reference approach is that sequences that are similar to previously observed sequences are clustered with one threshold while those that are not similar to previously observed sequences are clustered with a different threshold.
 
-As the throughput of sequencing technologies have improved, method development to keep pace.
-*De novo* clustering methods are considerably slower and more computationally intensive than reference-based methods and the greedy *de novo* methods are faster than the hierarchical.
-In our experience (REFS), the most significant detriment to execution speed of the *de novo* methods has been in adequate removal of sequencing error.
+As the throughput of sequencing technologies have improved, development of clustering algorithms must continue to keep pace.
+*De novo* clustering methods are considerably slower and more computationally intensive than reference-based methods and the greedy *de novo* methods are faster than the hierarchical methods.
+In our experience (REFS), the most significant detriment to execution speed of the *de novo* methods has been the inadequate removal of sequencing error and chimeras.
 As the rate of sequencing error increases so do the number of unique sequences that must be clustered.

### 2015-10-15 Schloss_Cluster_PeerJ_2015.Rmd
Edit Methods section

@@
 The order of the sequences in the full dataset was randomly permuted without replacement to generate an additional 30 datasets.
-To perform the hierarchical clustering methods and to generate a distance matrix we followed their approach of the He study by calculating distances based on pairwise global alignments using the `pairwise.dist` command in mothur using the default Needleman-Wunsch alignment method and parameters.
-It should be noted that this method has been strongly discouraged (REFS).
+To perform the hierarchical clustering methods and to generate a distance matrix we followed the approach of the He study by calculating distances based on pairwise global alignments using the `pairwise.dist` command in mothur using the default Needleman-Wunsch alignment method and parameters.
+It should be noted that this approach has been strongly discouraged (REFS).
 Execution of the hierarchical clustering methods was performed as described in the original He study using mothur (v.1.37) and using the QIIME (v.1.9.1) parameter profiles provided in the supplementary material from the He study for the greedy and reference-based clustering methods.
@@
 The murine 16S rRNA gene sequence data generated from the V4 region using an Illumina MiSeq was obtained from http:/www.mothur.org/MiSeqDevelopmentData/StabilityNoMetaG.tar and was processed as outlined in the original study (REF).
-Briefly, 250 nt long read pairs were assembled into contigs by aligning the reads and correcting discordant base calls by requiring one of the base calls to have a Phred quality score at least 6 points higher than the other.
+Briefly, 250 nt read pairs were assembled into contigs by aligning the reads and correcting discordant base calls by requiring one of the base calls to have a Phred quality score at least 6 points higher than the other.
 Sequences where it was not possible to resolve the disagreement were culled from the dataset.
 The sequences were then aligned to a SILVA reference alignment (REF) and any reads that aligned outside of the V4 region were removed from the dataset.
-Sequences were pre-clustered by combining the abundances of sequences that were within 2 nt of a more abundant sequence.
+Sequences were pre-clustered by combining the abundances of sequences that differed by 2 or fewer nucleotides of a more abundant sequence.
 Each of the samples was then screened for chimeric sequences using the default parameters in UCHIME (REF).
@@
 ***Calculation of Matthew's Correlation Coefficient (MCC)***
-The MCC was calculated by two approaches in this study.
-In both methods we used only the dereplicated sequence lists.
+The MCC was calculated by two approaches in this study using only the dereplicated sequence lists.
 First, we calculated the MCC to determine the stability of OTU assignments following the approach of the He study.
@@
 
-Second, we calculated MCC to determine the quality of the clusterings as previously described (REF).
+Second, we calculated the MCC to determine the quality of the clusterings as previously described (REF).
 Briefly, we compared the OTU assignments for pairs of sequences to the distance matrix that was calculated between all pairs of aligned sequences.

### 2015-10-15 Schloss_Cluster_PeerJ_2015.Rmd
Add code for R package version numbers

@@
 A reproducible workflow including all scripts and this manuscript as a literate programming document are available at https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015.
-The workflow utilizes QIIME (v.1.9.1; REFS), mothur (v.1.37.0; REFS), USEARCH (v.1.5.0; REFS), VSEARCH (v.1.5.0; https://github.com/torognes/vsearch), and R (v.3.2; REFS).
+The workflow utilized QIIME (v.1.9.1; REFS), mothur (v.1.37.0; REFS), USEARCH (v.1.5.0; REFS), VSEARCH (v.1.5.0; https://github.com/torognes/vsearch), and R (v.3.2; REFS).
+The Rcpp (REFS), rentrez (REFS), and jsonlite (REFS) packages were used within R.
 

### 2015-10-15 Schloss_Cluster_PeerJ_2015.Rmd
Add code for R package version numbers

@@
 The workflow utilized QIIME (v.1.9.1; REFS), mothur (v.1.37.0; REFS), USEARCH (v.1.5.0; REFS), VSEARCH (v.1.5.0; https://github.com/torognes/vsearch), and R (v.3.2; REFS).
-The Rcpp (REFS), rentrez (REFS), and jsonlite (REFS) packages were used within R.
+The knitr (`r packageDescription("knitr")$Version`, REFS), Rcpp (`r packageDescription("Rcpp")$Version`, REFS), rentrez (`r packageDescription("rentrez")$Version`, REFS), and jsonlite (`r packageDescription("jsonlite")$Version`, REFS) packages were used within R.
 

### 2015-10-16 code/closed_ref_analysis.R
Add sens/spec analysis of closed ref mapping

@@ -1,0 +1,104 @@
+split_line <- function(line){
+	sub_vector_names <- sort(unlist(strsplit(line, ',')))
+	sub_vector <- rep(sub_vector_names[1], length(sub_vector_names))
+	names(sub_vector) <- sub_vector_names
+	return(sub_vector)
+}
+
+split_duplicates <- function(duplicates){
+	unlist(unname(lapply(duplicates, split_line)))
+}
+
+
+unique_mapping <- function(mappings, redundant_ref_map){
+	unlist(lapply(mappings, function(x){paste(sort(unique(redundant_ref_map[unlist(strsplit(x, ','))])), collapse=',')}))
+}
+
+
+get_sens_spec_data <- function(search_file, true_mapping, duplicate_lookup){
+
+	small_dists <- true_mapping[true_mapping$distance <= 0.03,]
+	true_map <- small_dists$uniqued
+	names(true_map) <- gsub('-', '_', rownames(small_dists))
+
+	clusters <- read.table(file=search_file, stringsAsFactors=F)
+	clusters <- clusters[clusters$V1=='H',c(9,10)]
+	obs_map <- as.character(clusters$V10)
+	names(obs_map) <- gsub("-\\d*;size=.*;", "", clusters$V9)
+	names(obs_map) <- gsub('-', "_", names(obs_map))
+	obs_map_unique <- duplicate_lookup[obs_map]
+	names(obs_map_unique) <- names(obs_map)
+
+	tp <- 0
+	fp <- 0
+	tn <- 0
+	fn <- 0
+
+	total <- nrow(true_map)
+
+	seq_overlap <- intersect(names(true_map), names(obs_map_unique))
+	real_overlap <- true_map[seq_overlap]
+	obs_overlap <- obs_map_unique[seq_overlap]
+
+	tp <- sum(real_overlap == obs_overlap)
+
+	real_diff <- real_overlap[real_overlap != obs_overlap]
+	obs_diff <- obs_overlap[real_overlap != obs_overlap]
+	found <- sapply(1:length(real_diff), function(x){grepl(obs_diff[x], real_diff[x])})
+	tp <- tp+sum(found)
+	fp <- sum(!found)
+
+	fn <- fn + sum(!names(true_map) %in% seq_overlap)
+	fp <- fp + sum(!names(obs_map_unique) %in% seq_overlap)
+	tn <- total - length(union(names(obs_map_unique), names(true_map)))
+
[…]

### 2015-10-16 code/closed_ref_analysis.R
Fix how total number of sequences was determined

@@ -34,7 +34,7 @@ get_sens_spec_data <- function(search_file, true_mapping, duplicate_lookup){
 	tn <- 0
 	fn <- 0
 
-	total <- nrow(true_map)
+	total <- nrow(true_mapping)
 
 	seq_overlap <- intersect(names(true_map), names(obs_map_unique))
 	real_overlap <- true_map[seq_overlap]
@@ -69,7 +69,7 @@ run_sens_spec_analysis <- function(method){
 	true_mapping <- read.table(file="data/rand_ref/miseq.ref.mapping", row.names=1, header=T, stringsAsFactors=F)
 	true_mapping$uniqued <- unique_mapping(true_mapping$references, redundant_ref_map)
 
-	files <- list.files(path="./data/rand_ref", paste0("*.", method, "closed.", method, "c"), include.dirs=TRUE)
+	files <- list.files(path="./data/rand_ref", paste0("*.", method, "closed.", method, "c"), full.names=TRUE)
 	results <- t(sapply(files, get_sens_spec_data, true_mapping, redundant_ref_map))
 
 	write.table(results, paste0('data/rand_ref/closed_ref.', method, 'search.sensspec'), quote=F, sep='\t')

### 2015-10-16 code/run_rand_uref.sh
Make run_rand_?ref.sh scripts run the same way

@@ -2,9 +2,12 @@ REF=$1
 FASTA=data/rand_ref/miseq.fasta
 CLOSED_PATH=$(echo $REF | sed 's/fasta/uclosed/')
 rm -rf $CLOSED_PATH/
+mkdir $CLOSED_PATH
+mkdir $CLOSED_PATH/usearch
 
-pick_closed_reference_otus.py -i $FASTA -r $REF -o $CLOSED_PATH -p code/closedref.params.txt
-mv $CLOSED_PATH/usearch61_ref_picked_otus/ref_clustered.uc $CLOSED_PATH.uc
+usearch61 --sizeout --derep_fulllength $FASTA --minseqlength 30 --threads 1 --uc $CLOSED_PATH/usearch/abundance_sorted.uc --output $CLOSED_PATH/usearch/abundance_sorted.fna --strand both --log $CLOSED_PATH/usearch/abundance_sorted.log 
+usearch61 --maxaccepts 16 --id 0.97 --minseqlength 30 --threads 1 --wordlength 8 --uc $CLOSED_PATH/usearch/ref_clustered.uc --maxrejects 64 --strand both --db $REF --log $CLOSED_PATH/usearch/ref_clustered.log --usearch_global $CLOSED_PATH/usearch/abundance_sorted.fna
 
+mv $CLOSED_PATH/usearch/ref_clustered.uc $CLOSED_PATH.uc
 rm -rf $CLOSED_PATH/
 

### 2015-10-16 code/run_rand_vref.sh

@@ -6,7 +6,7 @@ mkdir $CLOSED_PATH
 mkdir $CLOSED_PATH/vsearch
 
 code/vsearch/vsearch --sizeout --derep_fulllength $FASTA --minseqlength 30 --threads 1 --uc $CLOSED_PATH/vsearch/abundance_sorted.uc --output $CLOSED_PATH/vsearch/abundance_sorted.fna --strand both --log $CLOSED_PATH/vsearch/abundance_sorted.log 
-code/vsearch/vsearch --maxaccepts 16 --id 0.97 --minseqlength 30 --threads 1 --wordlength 8 --uc $CLOSED_PATH/vsearch/ref_clustered.uc --maxrejects 64 --strand both --db $REF --log $CLOSED_PATH/vsearch/ref_clustered.log --usearch_global $CLOSED_PATH/vsearch/abundance_sorted.fna 
+code/vsearch/vsearch --maxaccepts 16 --id 0.97 --minseqlength 30 --threads 1 --wordlength 8 --uc $CLOSED_PATH/vsearch/ref_clustered.uc --maxrejects 64 --strand both --db $REF --log $CLOSED_PATH/vsearch/ref_clustered.log --usearch_global $CLOSED_PATH/vsearch/abundance_sorted.fna
 
 mv $CLOSED_PATH/vsearch/ref_clustered.uc $CLOSED_PATH.vc
 rm -rf $CLOSED_PATH/

### 2015-10-19 Schloss_Cluster_PeerJ_2015.Rmd
Expand closed ref analysis

@@
 rand_ref_v <- read.table(file="data/rand_ref/hits.vclosed.summary", row.names=1)
+
+uclosed <- read.table(file="data/rand_ref/closed_ref.usearch.sensspec", header=T, row.names=1)[2:31,]
+vclosed <- read.table(file="data/rand_ref/closed_ref.vsearch.sensspec", header=T, row.names=1)[2:31,]
+
+miseq_closed_ref <- read.table(file="data/rand_ref/closed_ref.redundancy.analysis", header=T, row.names=1)
 ```
@@
 The default database used in QIIME, which was also used in the He study and by others, contains full-length sequences that are at most 97% similar to each other.
-We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference method with the full MiSeq dataset, which contained `r n_unique_miseq_seqs` unique sequences.
+We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference method with the full murine dataset, which contained `r n_unique_miseq_seqs` unique sequences.
 Surprisingly, we observed that the number of OTUs generated was not the same in each of the randomizations.
@@
 When we used VSEARCH the original ordering of the reference sequences and all randomizations were able to map `r rand_ref_v["original", 1]` sequences to reference OTUs.
-That VSEARCH mapped more sequences than even the maximum number of sequences found across the USEARCH randomizations supports the VSEARCH developers' assertion that it has greater sensitivity than USEARCH.
+When we calculated the true distance between each of the murine sequences and the references, we were able to map `r nrow(miseq_closed_ref)` of the murine sequences to the reference sequences when using a 97% similarity threshold without the use of a heuristic.
+This result indicates that the closed reference approach, whether using USEARCH or VSEARCH, does not exhaustively or accurately map reads to the closest reference.
+To quantify this further, we calculated the MCC for the USEARCH and VSEARCH assignments relative to the assignments using the non-heuristic approach.
+Using USEARCH the average MCC was `r mean(uclosed$mcc)` (range: `r min(uclosed$mcc)` to `r max(uclosed$mcc)`) and using VSEARCH the average MCC was `r mean(vclosed$mcc)` (range: `r min(vclosed$mcc)` to `r max(vclosed$mcc)`).
+The two methods had similar sensitivities (USEARCH: `r mean(uclosed$sensitivity)` and VSEARCH: `r mean(vclosed$sensitivity)`), but the USEARCH specificity (`r mean(uclosed$specificity)`) was considerably higher than VSEARCH `r mean(uclosed$specificity)`.
+Overall, these results indicate that although heuristic approaches may be fast, relative to non-heuristic approaches, they do a poor job of mapping reads to the correct reference sequence.
 
@@
 dups <- read.table(file="data/gg_13_8/duplicate.analysis")
+n_true_mappings <- nrow(dups)
 max_taxa <- dups[dups$V3 == max(dups$V3),]
@@
 
+```{r }
+n_actual_map <- nrow(miseq_closed_ref)
+
+n_map_more_than_one_expanded <- sum(miseq_closed_ref$n_dups_expanded > 1)
+n_map_one_uniqued <- sum(miseq_closed_ref$n_dups_expanded>1 & miseq_closed_ref$n_dups_uniqued==1)
+n_map_more_than_one_uniqued <- sum(miseq_closed_ref$n_dups_expanded>1 & miseq_closed_ref$n_dups_uniqued>1)
+n_mismatch_taxonomy <- sum(table(miseq_closed_ref$n_taxa_expanded, miseq_closed_ref$n_dups_expanded)[-1,-1])
+```
+
 Beyond comparing the abundance of specific OTUs across samples, the reference database is used in the open and closed-reference methods to generate OTU labels that are used in several downstream applications.
@@
 Two sequences mapped to `r max_taxa[1,2]-1` and `r max_taxa[2,2]-1` duplicate sequences and both contained `r max_taxa[1,3]` different taxonomies.
-There was also a sequence had `r max(dups[,2])-1` duplicates and contained `r dups[which.max(dups[,2]),3]` different taxonomies.
-
-ADD SOMETHING ABOUT CONFLICT BETWEEN TAXONOMIES OF REFERENCES THAT WERE TIES TO QUERY SEQUENCE.
-
+Among the sequences within the database, there was also a sequence had `r max(dups[,2])-1` duplicates and contained `r dups[which.max(dups[,2]),3]` different taxonomies.
+When we analyzed the `r n_actual_map` sequences that mapped to the reference sequences using a non-heuristic approach, we observed that `r n_map_more_than_one_expanded` of the sequences mapped to more than one reference sequence.
+Of these sequences, `r n_map_one_uniqued` (`r 100*n_map_one_uniqued/n_map_more_than_one_expanded`%) mapped to references that were identical over the V4 region and `r n_map_more_than_one_uniqued` (`r 100*n_map_more_than_one_uniqued/n_map_more_than_one_expanded`%) mapped equally well to two or more references that were not identical over the V4 region.
+Among the combined `r n_map_more_than_one_expanded` sequences that mapped to multiple reference sequences, the taxonomy of the multiple reference sequences conflicted for `r n_mismatch_taxonomy` (`r 100*n_mismatch_taxonomy/n_map_more_than_one_expanded`%).
 Together, these results demonstrate some of the considerable problems with the reference-based clustering of sequences.

### 2015-10-19 code/closed_ref_analysis.R

@@ -1,5 +1,5 @@
 split_line <- function(line){
-	sub_vector_names <- sort(unlist(strsplit(line, ',')))
+	sub_vector_names <- unlist(strsplit(line, ','))
 	sub_vector <- rep(sub_vector_names[1], length(sub_vector_names))
 	names(sub_vector) <- sub_vector_names
 	return(sub_vector)
@@ -82,11 +82,15 @@ map_to_taxonomy <- function(line, taxonomy_data){
 	length(unique(taxonomy_data[otus]))
 }
 
+expand_names <- function(line, names_data){
+	otus <- unlist(strsplit(line, ','))
+	paste(unique(names_data[otus,1]), collapse=',')
+}
 
 run_redundancy_analysis <- function(){
 
-	names_file <- read.table(file="data/gg_13_8/gg_13_8_97.v4_ref.names", stringsAsFactors=F)
-	redundant_ref_map <- split_duplicates(names_file$V2)
+	names_file_map <- read.table(file="data/gg_13_8/gg_13_8_97.v4_ref.names", stringsAsFactors=F, row.names=1)
+	redundant_ref_map <- split_duplicates(names_file_map$V2)
 
 	true_mapping <- read.table(file="data/rand_ref/miseq.ref.mapping", row.names=1, header=T, stringsAsFactors=F)
 	true_mapping$uniqued <- unique_mapping(true_mapping$references, redundant_ref_map)
@@ -96,9 +100,14 @@ run_redundancy_analysis <- function(){
 	names(taxonomy_map) <- rownames(taxonomy_file)
 
 	close_data <- true_mapping[true_mapping$distance <= 0.03,]
-	n_taxa <- sapply(close_data$uniqued, map_to_taxonomy, taxonomy_data=taxonomy_map)
-	n_dups <- nchar(gsub("[^,]", "", close_data$uniqued)) + 1
+	n_taxa_uniqued <- sapply(close_data$uniqued, map_to_taxonomy, taxonomy_data=taxonomy_map)
+	n_dups_uniqued <- nchar(gsub("[^,]", "", close_data$uniqued)) + 1
 
-	write.table(table(n_taxa, n_dups), 'data/rand_ref/closed_ref.redundancy.analysis', quote=F, sep='\t')
+	close_data$expanded <- sapply(close_data$uniqued, expand_names, names_file_map)
[…]

### 2015-10-19 Schloss_Cluster_PeerJ_2015.Rmd
Insert citations

@@
 title: "***De novo* clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units**"
-author: Patrick D. Schloss* and Sarah L. Westcott
-date: '`r format(Sys.Date(), format="%B %d, %Y")`'
 bibliography: references.bib
@@
 ---
-
-## To Do:
-* references
-* revise
-* add sensitivity/specificity of closed-reference searching
 
@@
 
-* To whom correspondence should be addressed.
-pschloss@umich.edu
+\begin{center}
+\vspace{25mm}
+Patrick D. Schloss${^\dagger}$ and Sarah L. Westcott
+
+\vspace{30mm}
+
+$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu
 
 Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
+\end{center}
 
 \newpage
+\linenumbers
+
 ## Abstract
@@
 To simplify the complexity of these large datasets, it is helpful to cluster sequences into meaningful bins.
-These bins, commonly known as operational taxonomic units (OTUs), are  used to compare the biodiversity contained within and between different samples [@schloss2011assessing].
+These bins, commonly known as operational taxonomic units (OTUs), are  used to compare the biodiversity contained within and between different samples [@Schloss2011].
 Such comparisons have enabled researchers to characterize the microbiota associated with the human body [e.g.
-@human2012structure], soil [e.g.
-@shade2013streptomycin], aquatic ecosystems [e.g.
-@gilbert2012defining], and numerous other environments.
+@Huttenhower2012], soil [e.g.
+@Shade2013], aquatic ecosystems [e.g.
+@Gilbert2011], and numerous other environments.
 Within the field of microbial ecology, a convention has emerged where sequences are clustered into OTUs using a threshold of 97% similarity or a distance of 3%.
 One advantage of the OTU-based approach is that the definition of the bins is operational and can be changed to suit the needs of the particular project.
-However, with the dissemination of clustering methods within software such as mothur [@schloss2009introducing], QIIME [@Caporaso2010], and other tools [@edgar2010search, @sun2009esprit, @10.7717/peerj.593], it is important to understand how different clustering methods implement this conventional OTU threshold.
+However, with the dissemination of clustering methods within software such as mothur [@Schloss2009], QIIME [@Caporaso2010], and other tools [@Edgar2010; @Sun2009; @Mah2014; @Edgar2013; @Cai2011], it is important to understand how different clustering methods implement this conventional OTU threshold.
 Furthermore, it is necessary to understand how method choice affects the precision and accuracy of assigning sequences to OTUs.
@@
 
-The first approach has been referred to as phylotyping [@schloss2011assessing] or closed reference clustering [@navas2013advancing].
+The first approach has been referred to as phylotyping [@Schloss2011] or closed reference clustering [@NavasMolina2013].
 This approach involves comparing sequences to a curated database and then clustering sequences into the same OTU that are similar to the same reference sequence.
@@
 Therefore, defining OTUs in the closed-reference approach is complicated because although two sequences might be 97% similar to the same reference sequence, they may only be 94% similar to each other.
-A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@schloss2011assessing].
+A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@Schloss2011].
 The strength of the reference based approach is that the methods are generally fast, scaling linearly with the number of sequences being clustered.
 
-The second approach has been referred to as distance-based [@schloss2011assessing] or *de novo* clustering [@navas2013advancing].
+The second approach has been referred to as distance-based [@Schloss2011] or *de novo* clustering [@NavasMolina2013].
 In this method, the distance between sequences is used to cluster sequences into OTUs rather than the distance to a reference database.
@@
 If error rates can be reduced through stringent quality control measures then these problems can be overcome [@Kozich2013].
-As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@edgar2010search, @sun2009esprit, @10.7717/peerj.593].
-One critique of *de novo* approaches is that OTU assignments are sensitive to the input order of the sequences [REFS].
+As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@Edgar2010, @Sun2009, @Mah2014].
+One critique of *de novo* approaches is that OTU assignments are sensitive to the input order of the sequences [@He2015; @Hamady2009].
 Whether the differences in assignments is meaningful is unclear; however, the variation in results could represent equally valid clustering of the data.
@@
 
-The third approach, open-reference clustering, is a hybrid of the closed-reference and *de novo* approaches [@Navas2013].
+The third approach, open-reference clustering, is a hybrid of the closed-reference and *de novo* approaches [@NavasMolina2013].
 Open-reference clustering method involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference.
@@
 It has been difficult to objectively assess the quality of OTU assignments.
-Some have focused on the time and memory required to process a dataset (REFS).
+Some have focused on the time and memory required to process a dataset [@Sun2009; @Cai2011; @Rideout2014].
 These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the clustering.
-Others have attempted to judge the quality of an method by its ability to generate data that parallels classification data (REFS).
+Others have attempted to judge the quality of an method by its ability to generate data that parallels classification data [@Cai2011; @Sun2011; @Chen2013; @Edgar2013].
 This approach is problematic because bacterial taxonomy often reflects historical biases amongst bacterial systematicists.
-Furthermore, it is well known that the rates of evolution across lineages are not the same (REFS).
-We recently proposed an approach for evaluating OTU assignments using the distance between sequences (REF).
+Furthermore, it is well known that the rates of evolution across lineages are not the same [@Wang2007; @Schloss2010].
+Others have assessed the quality of clustering based on their ability to generate the same OTUs generated by other methods [@Schmidt2014; @Rideout2014].
+This is problematic because it does not solve the fundamental question of which method is most correct.
+We recently proposed an approach for evaluating OTU assignments using the distances between pairs of sequences [@Schloss2011].
 Those sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives.
 Meanwhile, those sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives.
-Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and negatives using the Matthew's correlation coefficient (MCC; REF).
+Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and negatives using the Matthew's correlation coefficient [MCC; @Matthews1975].
 This is an objective approach to assessing the quality of the OTU assignments.
 
-A recent analysis by He and colleagues (REF) attempted to characterize the three general clustering approaches by focusing on what they called stability.
+A recent analysis by He and colleagues @He2015 attempted to characterize the three general clustering approaches by focusing on what they called stability.
 They defined stability as the ability of an method to provide the same clustering on a subset of the data as was found in the full dataset.
@@
 ***Summary and replication of He study.*** We obtained the Canadian soil dataset from Roesch et al.
-(REF) and processed the sequences as described by He and colleagues.
+@Roesch2007 and processed the sequences as described by He and colleagues.
 Using these data, we reconsidered three of the more critical analyses performed in the He study.
@@
 Third, the authors attempted to describe the effects of the OTU assignment instability on comparisons of communities.
-They used Adonis to test whether the community structure represented in subsetted communities resembled that of the full dataset when only using the unstable OTUs (REFS).
+They used Adonis to test whether the community structure represented in subsetted communities resembled that of the full dataset when only using the unstable OTUs [@Anderson2001].
 Although they were able to detect significant p-values, they appeared to be of marginal biological significance.
@@
 ***OTU assignment methods vary in their accuracy.*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs.
-A method can generate highly stable OTUs, but the OTU assignments may be meaningless if they poorly represent the specified cutoff and the actual distance bewteen the sequences.
+A method can generate highly stable OTUs, but the OTU assignments may be meaningless if they poorly represent the specified cutoff and the actual distance between the sequences.
 To assess the quality of OTU assignments by the various methods, we made use of the pairwise distance between the unique sequences to count the number of true positives and negatives and the number of false positives and negatives for each method and sampling depth.
@@
 Developments in sequencing technology now permit the sequencing of millions of sequences for a study.
-In addition, because the original phred quality scores and flowgram data are not available, it was not possible to adequately remove sequencing errors (REFS).
+In addition, because the original Phred quality scores and flowgram data are not available, it was not possible for us to adequately remove sequencing errors [@Schloss2011Reducing].
 The large number of sequences that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering methods.
@@
 Inadequate coverage of the V9 region would adversely affect the ability of the reference-based methods to assign sequences to OTUs.
-Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene (REFS).
-With these points in mind, we compared the clustering assignment for each of these methods using a time series experiment that was obtained using mouse feces (REFS).
+Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene [@Schloss2010].
+With these points in mind, we compared the clustering assignment for each of these methods using a time series experiment that was obtained using mouse feces [@Schloss2012; @Kozich2013].
 The MiSeq platform was used to generate `r n_miseq_seqs` sequences from the V4 region of the 16S rRNA gene of `r n_groups` samples.
-Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02% (REFS).
+Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02% [@Kozich2013].
 Although no dataset is perfect for exhaustively testing these clustering methods, this dataset was useful for demonstrating several points.
@@
 ***Evalution of an open-source alternative to USEARCH.*** For some datasets the AGC and DGC methods appear to perform as well or better than the hierarchical clustering methods.
-As originally described in the He study, the AGC and DGC methods utilized the USEARCH program (REF), which is not available as open source code and is only available for free to academic users as a 32-bit program.
+As originally described in the He study, the AGC and DGC methods utilized the USEARCH program [@Edgar2010], which is not available as open source code and is only available for free to academic users as a 32-bit program.
 Access for non-academic users and those needing the 64-bit version is available commercially from the developer.
@@
 Beyond comparing the abundance of specific OTUs across samples, the reference database is used in the open and closed-reference methods to generate OTU labels that are used in several downstream applications.
-It is commonly used to extract information from a reference phylogenetic tree to carrying out UniFrac-based analyses (REFS) and to identify reference genomes for performing analyses such as PICRUSt (REFS).
+It is commonly used to extract information from a reference phylogenetic tree to carrying out UniFrac-based analyses [@Hamady2009] and to identify reference genomes for performing analyses such as PICRUSt [@Langille2013].
 Because these downstream applications depend on the correct and unique labeling of the OTUs, the lack of stability of the labeling is problematic.
@@
 Furthermore, although not tested in our study, VSEARCH can be parallelized leading to potentially significant improvements in speed.
-Although USEARCH and VSEARCH do not utilize aligned sequences, it is important to note that a sequence curation pipeline including denoising, alignment, trimming to a consistent region of the 16S rRNA gene, and chimera checking are critical to making proper inferences (REF).
+Although USEARCH and VSEARCH do not utilize aligned sequences, it is important to note that a sequence curation pipeline including denoising, alignment, trimming to a consistent region of the 16S rRNA gene, and chimera checking are critical to making proper inferences [@Schloss2011Reducing; @Kozich2013; @Schloss2012Secondary].
 
 We have assessed the ability of reference-based clustering methods to capture the actual distance between the sequences in a dataset in parallel with *de novo* methods.
-Several studies have lauded both the open and closed-reference approaches for generating reproducible clusterings (REFS), yet we have shown that both reference-based approaches did a poor job of representing the distance between the sequences compared to the *de novo* approaches.
+Several studies have lauded both the open and closed-reference approaches for generating reproducible clusterings [@He2015; @Rideout2014; @NavasMolina2013], yet we have shown that both reference-based approaches did a poor job of representing the distance between the sequences compared to the *de novo* approaches.
 Although the OTU assignments are reproducible and stable across a range of library sizes, the reference-based OTU assignments are a poor representation of the data.
@@
 *De novo* clustering methods are considerably slower and more computationally intensive than reference-based methods and the greedy *de novo* methods are faster than the hierarchical methods.
-In our experience (REFS), the most significant detriment to execution speed of the *de novo* methods has been the inadequate removal of sequencing error and chimeras.
+In our experience [@Kozich2013], the most significant detriment to execution speed of the *de novo* methods has been the inadequate removal of sequencing error and chimeras.
 As the rate of sequencing error increases so do the number of unique sequences that must be clustered.
 The speed of the *de novo* methods scales approximately quadratically, so that doubling the number of sequences results in a four-fold increase in the time required to execute the method.
-The rapid expansion in sequencing throughput has been likened to the Red Queen in Lewis Carroll's, *Through the Looking-Glass* who must run in place to keep up to her changing surroundings (REF).
+The rapid expansion in sequencing throughput has been likened to the Red Queen in Lewis Carroll's, *Through the Looking-Glass* who must run in place to keep up to her changing surroundings [@Schloss2009].
 Microbial ecologists must continue to refine clustering methods to better handle the size of the datasets, but they must also take steps to improve the quality of the underlying data.
@@
 To perform the hierarchical clustering methods and to generate a distance matrix we followed the approach of the He study by calculating distances based on pairwise global alignments using the `pairwise.dist` command in mothur using the default Needleman-Wunsch alignment method and parameters.
-It should be noted that this approach has been strongly discouraged (REFS).
+It should be noted that this approach has been strongly discouraged [@Schloss2012Secondary].
 Execution of the hierarchical clustering methods was performed as described in the original He study using mothur (v.1.37) and using the QIIME (v.1.9.1) parameter profiles provided in the supplementary material from the He study for the greedy and reference-based clustering methods.
@@
 ***MiSeq-generated Murine gut microbiota dataset***
-The murine 16S rRNA gene sequence data generated from the V4 region using an Illumina MiSeq was obtained from http:/www.mothur.org/MiSeqDevelopmentData/StabilityNoMetaG.tar and was processed as outlined in the original study (REF).
-Briefly, 250 nt read pairs were assembled into contigs by aligning the reads and correcting discordant base calls by requiring one of the base calls to have a Phred quality score at least 6 points higher than the other.
+The murine 16S rRNA gene sequence data generated from the V4 region using an Illumina MiSeq was obtained from http:/www.mothur.org/MiSeqDevelopmentData/StabilityNoMetaG.tar and was processed as outlined in the original study [@Kozich2013].
+Briefly, 250-nt read pairs were assembled into contigs by aligning the reads and correcting discordant base calls by requiring one of the base calls to have a Phred quality score at least 6 points higher than the other.
 Sequences where it was not possible to resolve the disagreement were culled from the dataset.
-The sequences were then aligned to a SILVA reference alignment (REF) and any reads that aligned outside of the V4 region were removed from the dataset.
+The sequences were then aligned to a SILVA reference alignment [@Pruesse2007] and any reads that aligned outside of the V4 region were removed from the dataset.
 Sequences were pre-clustered by combining the abundances of sequences that differed by 2 or fewer nucleotides of a more abundant sequence.
-Each of the samples was then screened for chimeric sequences using the default parameters in UCHIME (REF).
+Each of the samples was then screened for chimeric sequences using the default parameters in UCHIME [@Edgar2011].
 The resulting sequences were processed in the same manner as the Canadian soil dataset with the exception that the distance matrices were calculated based on the SILVA-based alignment.
@@
 We utilized the 97% OTUs greengenes reference sequence and taxonomy data (v.13.8) that accompanies the QIIME installation.
-Because the greengenes reference alignment does a poor job of representing the secondary structure of the 16S rRNA gene (REF), we realigned the fasta sequences to a SILVA reference alignment to identify the V4 region of the sequences.
+Because the greengenes reference alignment does a poor job of representing the secondary structure of the 16S rRNA gene [@Schloss2010], we realigned the fasta sequences to a SILVA reference alignment to identify the V4 region of the sequences.
 
@@
 
-Second, we calculated the MCC to determine the quality of the clusterings as previously described (REF).
+Second, we calculated the MCC to determine the quality of the clusterings as previously described [@Schloss2011].
 Briefly, we compared the OTU assignments for pairs of sequences to the distance matrix that was calculated between all pairs of aligned sequences.
@@
 A reproducible workflow including all scripts and this manuscript as a literate programming document are available at https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015.
-The workflow utilized QIIME (v.1.9.1; REFS), mothur (v.1.37.0; REFS), USEARCH (v.1.5.0; REFS), VSEARCH (v.1.5.0; https://github.com/torognes/vsearch), and R (v.3.2; REFS).
-The knitr (`r packageDescription("knitr")$Version`, REFS), Rcpp (`r packageDescription("Rcpp")$Version`, REFS), rentrez (`r packageDescription("rentrez")$Version`, REFS), and jsonlite (`r packageDescription("jsonlite")$Version`, REFS) packages were used within R.
+The workflow utilized QIIME [v.1.9.1; @Caporaso2010], mothur [v.1.37.0; @Schloss2009], USEARCH [v.6.1; @Edgar2010], VSEARCH [v.1.5.0; @Rognes2015], and R (v.3.2) [@language2015].
+The knitr (`r packageDescription("knitr")$Version`) [@xie2013dynamic], Rcpp (`r packageDescription("Rcpp")$Version`) [@eddelbuettel2013seamless], rentrez (`r packageDescription("rentrez")$Version`) [@winter2015rentrez], and jsonlite (`r packageDescription("jsonlite")$Version`) [@ooms2014jsonlite]) packages were used within R.
 

### 2015-10-19 Schloss_Cluster_PeerJ_2015.Rmd
Fix inline_hook function to correctly format text

@@
 output:
+  markdown
   pdf_document:
@@
     paste(format(x,big.mark=',', digits=2, scientific=FALSE))
+    } else {
+        paste(x)      
     }
@@
 
-***Analysis of reference database***
+***Analysis of reference database.***
 We utilized the 97% OTUs greengenes reference sequence and taxonomy data (v.13.8) that accompanies the QIIME installation.
@@
 
-***Calculation of Matthew's Correlation Coefficient (MCC)***
+***Calculation of Matthew's Correlation Coefficient (MCC).***
 The MCC was calculated by two approaches in this study using only the dereplicated sequence lists.
@@
 
-***Software availability***
+***Software availability.***
 A reproducible workflow including all scripts and this manuscript as a literate programming document are available at https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015.
-The workflow utilized QIIME [v.1.9.1; @Caporaso2010], mothur [v.1.37.0; @Schloss2009], USEARCH [v.6.1; @Edgar2010], VSEARCH [v.1.5.0; @Rognes2015], and R (v.3.2) [@language2015].
-The knitr (`r packageDescription("knitr")$Version`) [@xie2013dynamic], Rcpp (`r packageDescription("Rcpp")$Version`) [@eddelbuettel2013seamless], rentrez (`r packageDescription("rentrez")$Version`) [@winter2015rentrez], and jsonlite (`r packageDescription("jsonlite")$Version`) [@ooms2014jsonlite]) packages were used within R.
+The workflow utilized QIIME [v.1.9.1; @Caporaso2010], mothur [v.1.37.0; @Schloss2009], USEARCH [v.6.1; @Edgar2010], VSEARCH [v.1.5.0; @Rognes2015], and R [v.`r paste(version$major, version$minor, sep='.')`; @language2015].
+The knitr [v.`r packageDescription("knitr")$Version`; @xie2013dynamic], Rcpp [v.
+`r packageDescription("Rcpp")$Version`; @eddelbuettel2013seamless], rentrez [v.
+`r packageDescription("rentrez")$Version`; @winter2015rentrez], and jsonlite [v.
+`r packageDescription("jsonlite")$Version`; @ooms2014jsonlite] packages were used within R.
 

### 2015-10-19 Schloss_Cluster_PeerJ_2015.Rmd
Editing

@@
 output:
-  markdown
   pdf_document:
@@
 Depending on the dataset being analyzed, the average linkage and the distance and abundance-based greedy clustering methods generated more robust OTUs than the open and closed-reference methods.
-Further interrogation of the reference-based methods indicated that when USEARCH is used to identify the closest reference, the OTU assignments were sensitive to the order of the reference sequences.
-When USEARCH was replaced by VSEARCH the quality and stability of the OTU assignments by the reference-based methods were improved.
-Finally, we demonstrated that for the greedy algorithms VSEARCH produced assignments that were comparable to those produced by USEARCH making VSEARCH a viable free and open source alternative to USEARCH.
+We also demonstrated that for the greedy algorithms VSEARCH produced assignments that were comparable to those produced by USEARCH making VSEARCH a viable free and open source alternative to USEARCH.
+Further interrogation of the reference-based methods indicated that when USEARCH is used to identify the closest reference, the OTU assignments were sensitive to the order of the reference sequences because the reference sequences can be identical over the region being considered.
+More troubling was the observation that while both USEARCH and VSEARCH have a high level of sensitivity to detect reference sequences, the specificity of those matches was poor relative to the true best match.
 
 **Discussion.** This study demonstrates that the quality of clustering assignments needs to be assessed for multiple methods per dataset to identify the optimal clustering method for the dataset.
-The analysis also calls into question the quality and stability of OTU assignments generated by the open and closed-reference methods as implemented in current versions of QIIME.
+The analysis also calls into question the quality and stability of OTU assignments generated by the open and closed-reference methods as implemented in current version of QIIME.
 
@@
 If a large fraction of sequences are novel, then they cannot be assigned to an OTU.
-In addition, the reference sequences used in this application are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene.
+In addition, the reference sequences used in this application are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene [@Schloss2011].
 Thus, a sequence representing a fragment of the gene may be more than 97% similar to multiple reference sequences.
@@
 The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences.
-If error rates can be reduced through stringent quality control measures then these problems can be overcome [@Kozich2013].
+If error rates can be reduced through stringent quality control measures, then these problems can be overcome [@Kozich2013].
 As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@Edgar2010, @Sun2009, @Mah2014].
@@
 
-The third approach, open-reference clustering, is a hybrid of the closed-reference and *de novo* approaches [@NavasMolina2013].
+The third approach, open-reference clustering, is a hybrid of the closed-reference and *de novo* approaches [@NavasMolina2013; @Rideout2014].
 Open-reference clustering method involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference.

### 2015-10-20 Schloss_Cluster_PeerJ_2015.Rmd
Add rentrez doi number and spacing btwn parargraphs

@@
 
-***454 FLX-generated Roesch Canadian soil dataset***
+***454 FLX-generated Roesch Canadian soil dataset.***
 After obtaining the 16S rRNA gene fragments from GenBank (accessions EF308591-EF361836), we followed the methods outlined by the He study by removing any sequence that contained an ambiguous base, was identified as being a chimera, and fell outside a defined sequence length.
@@
 
-***MiSeq-generated Murine gut microbiota dataset***
+***MiSeq-generated Murine gut microbiota dataset.***
 The murine 16S rRNA gene sequence data generated from the V4 region using an Illumina MiSeq was obtained from http:/www.mothur.org/MiSeqDevelopmentData/StabilityNoMetaG.tar and was processed as outlined in the original study [@Kozich2013].

### 2015-10-28 Schloss_Cluster_PeerJ_2015.Rmd
Incorporate Marc, Kaitlin, and Geof's comments

@@
 inline_hook <- function(x){
+	print(x)
 	if(is.numeric(x)){
-    paste(format(x,big.mark=',', digits=2, scientific=FALSE))
+		if(abs(x - round(x)) < .Machine$double.eps^0.5){
+			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
+		} else {
+			paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
+		}
+
 	} else {
@@
 **Background.** 16S rRNA gene sequences are routinely assigned to operational taxonomic units (OTUs) that are then used to analyze complex microbial communities.
-A number of methods have been employed to carry out the assignment of 16S rRNA gene sequences to OTUs leading to confusion over which method is the best.
+A number of methods have been employed to carry out the assignment of 16S rRNA gene sequences to OTUs leading to confusion over which method is the most rigorous.
 A recent study suggested that a clustering method should be selected based on its ability to generate stable OTU assignments that do not change as additional sequences are added to the dataset.
@@
 
-**Discussion.** This study demonstrates that the quality of clustering assignments needs to be assessed for multiple methods per dataset to identify the optimal clustering method for the dataset.
-The analysis also calls into question the quality and stability of OTU assignments generated by the open and closed-reference methods as implemented in current version of QIIME.
+**Discussion.** Our analysis calls into question the quality and stability of OTU assignments generated by the open and closed-reference methods as implemented in current version of QIIME.
+This study demonstrates that *de novo* methods are the most rigorous and that the quality of clustering assignments needs to be assessed for multiple methods to identify the optimal clustering method for the dataset.
 
@@
 These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the clustering.
-Others have attempted to judge the quality of an method by its ability to generate data that parallels classification data [@Cai2011; @Sun2011; @Chen2013; @Edgar2013].
+Others have attempted to judge the quality of a method by its ability to generate data that parallels classification data [@Cai2011; @Sun2011; @Chen2013; @Edgar2013].
 This approach is problematic because bacterial taxonomy often reflects historical biases amongst bacterial systematicists.
@@
 
-A recent analysis by He and colleagues @He2015 attempted to characterize the three general clustering approaches by focusing on what they called stability.
-They defined stability as the ability of an method to provide the same clustering on a subset of the data as was found in the full dataset.
+A recent analysis by He and colleagues [-@He2015] attempted to characterize the three general clustering approaches by focusing on what they called stability.
+They defined stability as the ability of a method to provide the same clustering on a subset of the data as was found in the full dataset.
 Their concept of stability did not account for the accuracy of the OTU assignments and instead focused on the precision of the assignments.
@@
 ***Summary and replication of He study.*** We obtained the Canadian soil dataset from Roesch et al.
-@Roesch2007 and processed the sequences as described by He and colleagues.
+[-@Roesch2007] and processed the sequences as described by He and colleagues.
 Using these data, we reconsidered three of the more critical analyses performed in the He study.
@@
 Among the *de novo* methods they observed that the abundance-based greedy clustering (AGC) method generated the most stable OTUs followed by the single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) methods.
-We first sought to assess calculated the MCC for the OTU assignments generated by each of the clustering methods using 20, 40, 60, and 80% of the sequences relative to the OTU composition formed by the methods using the full dataset (Figure 1A).
+We first sought to assess the calculated the MCC for the OTU assignments generated by each of the clustering methods using 20, 40, 60, and 80% of the sequences relative to the OTU composition formed by the methods using the full dataset (Figure 1A).
 Similar to the He study, we replicated each method and subsample 30 times because a random number generator is used in some of methods to break ties where pairs of sequences have the same distance between them.
@@
 Second, the AL method had higher MCC values than the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL (Figure 1D).
-Third, with the possible exception of the CL method, the MCC values for each of the only demonstrated a small amount of variation between runs of the method with a different ordering of the input sequences.
+Third, with the possible exception of the CL method, the MCC values for each of the methods only demonstrated a small amount of variation between runs of the method with a different ordering of the input sequences.
 This indicates that although there may be variation between executions of the same method, they produce OTU assignments that are equally good.
-Revisiting the concept of stability, we question the value of obtaining stable OTUs when the the full dataset is not optimally assigned to OTUs.
-Our analysis indicates that the best method for assigning the Canadian soils sequences to OTUs using a 97% threshold is the AL method.
+Revisiting the concept of stability, we question the value of obtaining stable OTUs when the full dataset is not optimally assigned to OTUs.
+Our analysis indicates that the most rigorous method for assigning the Canadian soils sequences to OTUs using a 97% threshold is the AL method.
 
@@
 n_groups <- miseq_info["n_groups",1]
-n_miseq_seqs <- miseq_info["n_miseq_seqs",1]
-n_unique_miseq_seqs <- miseq_info["n_unique_miseq_seqs",1]
+n_miseq_seqs <- miseq_info["total_seqs",1]
+n_unique_miseq_seqs <- miseq_info["unique_seqs",1]
 
@@
 For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference methods.
-Of the `r n_v19_total` sequences in the default QIIME database, only `r n_v4_overlap` fully cover the V9 region.
+Of the `r n_v19_total` sequences in the default QIIME database, only `r n_v9_overlap` fully cover the V9 region.
 In contrast, `r n_v4_overlap` of the sequences fully covered the V4 region.
@@
 This result suggests that the optimal method is likely to be database-dependent.
+
 Finally, as was observed with the Canadian soil dataset, there was little variation in MCC values observed among the 30 randomizations.
@@
 Access for non-academic users and those needing the 64-bit version is available commercially from the developer.
-An alternative to USEARCH is VSEARCH, which is being developed in parallel to USEARCH as a open-source alternative.
+An alternative to USEARCH is VSEARCH, which is being developed in parallel to USEARCH as an open-source alternative.
 One subtle difference between the two programs is that USEARCH employs a heuristic to generate candidate alignments whereas VSEARCH generates the actual global alignments.
@@
 Surprisingly, we observed that the number of OTUs generated was not the same in each of the randomizations.
-On average there were `r round(rand_ref_u["mean", 1], 2)` sequences that mapped to a reference OTU per randomization (range from `r round(rand_ref_u["min", 1], 2)` to `r round(rand_ref_u["max", 1], 2)`).
+On average there were `r round(rand_ref_u["mean", 1])` sequences that mapped to a reference OTU per randomization (range from `r rand_ref_u["min", 1]` to `r rand_ref_u["max", 1]`).
 The original ordering of the reference resulted in `r rand_ref_u["original", 1]` sequences being mapped, less than the minimum observed number of mapped sequences when the references were randomized.
@@
 Using USEARCH the average MCC was `r mean(uclosed$mcc)` (range: `r min(uclosed$mcc)` to `r max(uclosed$mcc)`) and using VSEARCH the average MCC was `r mean(vclosed$mcc)` (range: `r min(vclosed$mcc)` to `r max(vclosed$mcc)`).
-The two methods had similar sensitivities (USEARCH: `r mean(uclosed$sensitivity)` and VSEARCH: `r mean(vclosed$sensitivity)`), but the USEARCH specificity (`r mean(uclosed$specificity)`) was considerably higher than VSEARCH `r mean(uclosed$specificity)`.
+The two methods had similar sensitivities (USEARCH: `r mean(uclosed$sensitivity)` and VSEARCH: `r mean(vclosed$sensitivity)`), but the USEARCH specificity (`r mean(uclosed$specificity)`) was considerably higher than VSEARCH (`r mean(vclosed$specificity)`).
 Overall, these results indicate that although heuristic approaches may be fast, relative to non-heuristic approaches, they do a poor job of mapping reads to the correct reference sequence.
@@
 We also observed that regardless of whether we used USEARCH or VSEARCH, the reference OTU labels that were assigned to each OTU differed between randomizations.
-When we used USEARCH to perform closed-reference clustering, an average of `r round(100*u_overlap["orig", "mean"], 1)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*u_overlap["orig", "min"], 1)` to `r round(100*u_overlap["orig", "max"], 1)`).
-If we instead used VSEARCH an average of `r round(100*v_overlap["orig", "mean"], 1)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*v_overlap["orig", "min"], 1)` to `r round(100*v_overlap["orig", "max"], 1)`).
+When we used USEARCH to perform closed-reference clustering, an average of `r round(100*u_overlap["orig", "mean"], 2)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*u_overlap["orig", "min"], 2)` to `r round(100*u_overlap["orig", "max"], 2)`).
+If we instead used VSEARCH an average of `r round(100*v_overlap["orig", "mean"], 2)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*v_overlap["orig", "min"], 2)` to `r round(100*v_overlap["orig", "max"], 2)`).
 To better understand this result, we further analyzed QIIME's reference database.
@@
 The presence of duplicate V4 reference sequences explains the lack of labeling stability when using either USEARCH or VSEARCH to carry out the closed-reference method.
-We suspect that the reference database was designed to only include sequences that were at most 97% similar to each other way to overcome the limitations of the USEARCH search heuristic.
+We suspect that the reference database was designed to only include sequences that were at most 97% similar to each other as a way to overcome the limitations of the USEARCH search heuristic.
 
@@
 Two sequences mapped to `r max_taxa[1,2]-1` and `r max_taxa[2,2]-1` duplicate sequences and both contained `r max_taxa[1,3]` different taxonomies.
-Among the sequences within the database, there was also a sequence had `r max(dups[,2])-1` duplicates and contained `r dups[which.max(dups[,2]),3]` different taxonomies.
+Among the sequences within the database, there was also a sequence that had `r max(dups[,2])-1` duplicates and contained `r dups[which.max(dups[,2]),3]` different taxonomies.
 When we analyzed the `r n_actual_map` sequences that mapped to the reference sequences using a non-heuristic approach, we observed that `r n_map_more_than_one_expanded` of the sequences mapped to more than one reference sequence.
@@
 Although the OTU assignments are reproducible and stable across a range of library sizes, the reference-based OTU assignments are a poor representation of the data.
-We also observed that the assignments were not actually reproducible when the order of the reference sequences was randomize.
+We also observed that the assignments were not actually reproducible when the order of the reference sequences was randomized.
 When USEARCH was used, the actual number of sequences that mapped to the reference changed depended on the order of the reference.
@@
 
+\includepdf{results/figures/figure_1.pdf}
+\includepdf{results/figures/figure_2.pdf}
+\includepdf{results/figures/figure_3.pdf}
+\includepdf{results/figures/figure_4.pdf}
+
+\newpage
+
 ## References

### 2015-10-28 code/build_mcc_plot.R
Add swarm data to plots

@@ -1,8 +1,8 @@
 build_mcc_plots <- function(dataset, output_file_name){
 
-	methods <- c("fn", "nn", "an", "dgc", "agc", "closed", "open")
+	methods <- c("fn", "nn", "an", "dgc", "agc", "closed", "open", "swarm")
 	n_methods <- length(methods)
-	clrs <- c("brown", "red", "orange", rainbow(n_methods)[4:n_methods])
+	clrs <- c("brown", "red", "orange", rainbow(n_methods-1)[4:7], "black")
 
 
 
@@ -47,7 +47,7 @@ build_mcc_plots <- function(dataset, output_file_name){
 	sixty <- orig_methods[orig_methods$fraction==threshold,]
 	sixty$method <- factor(sixty$method, levels=methods)
 	sixty[sixty$method=="closed", c("lci", "uci")] <- c(0.999, 1.001)
-	plot(NA, ylim=c(0.0,1), xlim=c(0.75,7.25), axes=F, ylab="", xlab="")
+	plot(NA, ylim=c(0.0,1), xlim=c(0.75,n_methods+0.25), axes=F, ylab="", xlab="")
 
 	for(m in 1:n_methods){
 		points(x=m, y=sixty[sixty$method==methods[m],"mean"], pch=19, col=clrs[m], cex=2.0)
@@ -67,6 +67,10 @@ build_mcc_plots <- function(dataset, output_file_name){
 	data <- read.table(file=paste0("data/process/", dataset, ".mcc.summary"), header=T, stringsAsFactors=F)
 	orig_methods <- data[data$method %in% methods,]
 
+	swarm_data <- read.table(file=paste0("data/", dataset, "/", dataset, ".swarm.opt.sensspec"), header=T, stringsAsFactors=F)
+	opt_swarm_data <- do.call(rbind, lapply(split(swarm_data, swarm_data$label), function(chunk) chunk[which.max(chunk$mcc),c("cutoff", "mcc")]))
+	fraction <- gsub(".*_(\\d.\\d)_.*", "\\1", rownames(opt_swarm_data))
+	swarm_mcc_by_fraction <- aggregate(opt_swarm_data, by=list(fraction), function(x)c(mean(x), range(x)))
 
 	threshold <- 1.0
 
@@ -102,7 +106,7 @@ build_mcc_plots <- function(dataset, output_file_name){
[…]

### 2015-10-28 Schloss_Cluster_PeerJ_2015.Rmd
Add Swarm paragraph

@@
 As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@Edgar2010, @Sun2009, @Mah2014].
-One critique of *de novo* approaches is that OTU assignments are sensitive to the input order of the sequences [@He2015; @Hamady2009].
+One critique of *de novo* approaches is that OTU assignments are sensitive to the input order of the sequences [@He2015, @Mah2014].
 Whether the differences in assignments is meaningful is unclear; however, the variation in results could represent equally valid clustering of the data.
@@
 It has been difficult to objectively assess the quality of OTU assignments.
-Some have focused on the time and memory required to process a dataset [@Sun2009; @Cai2011; @Rideout2014].
+Some have focused on the time and memory required to process a dataset [@Sun2009; @Cai2011; @Rideout2014, @Mah2014].
 These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the clustering.
@@
 Although no dataset is perfect for exhaustively testing these clustering methods, this dataset was useful for demonstrating several points.
-First, when using 60% of the data the stability relationships amongst the different methods were similar to what we observed using the He dataset (Figure 3AB).
+First, when using 60% of the data the stability relationships amongst the different methods were similar to what we observed using the Canadian soil dataset (Figure 3AB).
 With the exception for the clusters generated using CL, the methods all performed very well with stabilities greater than `r round(min(miseq_stability[miseq_stability$fraction=="0.6" & miseq_stability$method != "fn","mean"]), 2)`.
@@
 These results are a strong indication that VSEARCH is a suitable and possibly better replacement for USEARCH for executing the AGC and DGC methods.
+
+***Evalution of Swarm as an alternative de novo clustering algorithm.*** The Swarm algorithm is a recently proposed *de novo* method for assigning sequences to OTUs that identifies clusters of sequences based on the number of differences between each other without applying a distance threshold [@Mah2014].
+Swarm was originally validated by comparing the results against the expected clusters formed based on the taxonomic composition of a mock community.
+Similar to the authors of the He study, the Swarm developers suggest that method are needed that are insensitive to input order.
+Use of Swarm on the Canadian soil and murine datasets demonstrated that similar to the other *de novo* methods, Swarm's OTU assignments changed as sequences were added (Figures 1A and 3A).
+When we compared the OTU assignments for both datasets when using all of the sequence data, the variation in MCC values across the 30 randomizations were not meaningfully different (Figures 1D and 3D).
+Most importantly, when we selected the distance threshold that optimized the MCC value, the quality of the OTU assignments was close to that of the AL assignments when using the Canadian soil dataset and considerably worse than that of the murine dataset (Figures 1D and 3D).
+Interestingly, the distance thresholds that resulted in the largest MCC values were 0.03 and 0.02 for the Canadian soil and murine datasets, respectively.
+This suggests that OTU definitions are not consistent across datasets when using the Swarm algorithm.
+Finally, the Swarm developers indicated that hierarchical *de novo* algorithms were too impractical to use on large MiSeq-generated datasets.
+Our ability to apply AL to the large mouse dataset and even larger datasets suggests [e.g.
+@Schubert2015, @Zackular2015] that it is not necessary to sacrifice OTU assignment quality for speed.
 
@@
 The MCC was counted for each dataset using the formula above as implemented in the `sens.spec` command in mothur.
+To judge the quality of the Swarm-generated OTU assignments we calculated the MCC value using thresholds incremented by 0.01 between 0.00 and 0.05 and selected the threshold that provided the optimal MCC value.
 
@@
 A reproducible workflow including all scripts and this manuscript as a literate programming document are available at https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015.
-The workflow utilized QIIME [v.1.9.1; @Caporaso2010], mothur [v.1.37.0; @Schloss2009], USEARCH [v.6.1; @Edgar2010], VSEARCH [v.1.5.0; @Rognes2015], and R [v.`r paste(version$major, version$minor, sep='.')`; @language2015].
+The workflow utilized QIIME [v.1.9.1; @Caporaso2010], mothur [v.1.37.0; @Schloss2009], USEARCH [v.6.1; @Edgar2010], VSEARCH [v.1.5.0; @Rognes2015], Swarm [v.2.1.1; @Mah2014], and R [v.`r paste(version$major, version$minor, sep='.')`; @language2015].
+The SL, AL, and CL methods are called nearest neighbor (NN), average neighbor (AN), and furthest neighbor (FN) in mothur; we have used the terminology from the He study to minimize confusion.
 The knitr [v.`r packageDescription("knitr")$Version`; @xie2013dynamic], Rcpp [v.
@@
 The vertical gray line indicates in A and C indicates the fraction of the dataset represented in B and D, respectively.
+The optimum threshold for the Swarm-generated assignments was 0.03.
 
@@
 The vertical gray line indicates in A and C indicates the fraction of the dataset represented in B and D, respectively.
+The optimum threshold for the Swarm-generated assignments was 0.02.
 

### 2015-10-28 Schloss_Cluster_PeerJ_2015.Rmd
Edit manuscript

@@
 
-**Methods.** Our analysis implemented five *de novo* clustering algorithms including the single linkage, complete linkage, average linkage, abundance-based greedy clustering, distance-based greedy clustering and two reference-based methods including the open and closed-reference methods.
-By varying the number of sequences sampled from each of two previously published datasets we used the Matthew’s Correlation Coefficient (MCC) to assess the quality and stability of OTU assignments.
+**Methods.** Our analysis implemented five *de novo* clustering algorithms including the single linkage, complete linkage, average linkage, abundance-based greedy clustering, distance-based greedy clustering, and the open and closed-reference methods.
+Using two previously published datasets we used the Matthew’s Correlation Coefficient (MCC) to assess the stability and quality of OTU assignments.
 
@@
 **Discussion.** Our analysis calls into question the quality and stability of OTU assignments generated by the open and closed-reference methods as implemented in current version of QIIME.
-This study demonstrates that *de novo* methods are the most rigorous and that the quality of clustering assignments needs to be assessed for multiple methods to identify the optimal clustering method for the dataset.
+This study demonstrates that *de novo* methods are the most rigorous and that the quality of clustering assignments needs to be assessed for multiple methods to identify the optimal clustering method for a particular dataset.
 
@@
 One advantage of the OTU-based approach is that the definition of the bins is operational and can be changed to suit the needs of the particular project.
-However, with the dissemination of clustering methods within software such as mothur [@Schloss2009], QIIME [@Caporaso2010], and other tools [@Edgar2010; @Sun2009; @Mah2014; @Edgar2013; @Cai2011], it is important to understand how different clustering methods implement this conventional OTU threshold.
-Furthermore, it is necessary to understand how method choice affects the precision and accuracy of assigning sequences to OTUs.
+However, with the dissemination of clustering methods within software such as mothur [@Schloss2009], QIIME [@Caporaso2010], and other tools [@Edgar2010, @Sun2009, @Mah2014, @Edgar2013, @Cai2011], it is important to understand how different clustering methods implement this conventional OTU threshold.
+Furthermore, it is necessary to understand how the selected method affects the precision and accuracy of assigning sequences to OTUs.
 Broadly speaking, three approaches have been developed to assign sequences to OTUs.
@@
 If a large fraction of sequences are novel, then they cannot be assigned to an OTU.
-In addition, the reference sequences used in this application are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene [@Schloss2011].
+In addition, the reference sequences are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene [@Schloss2011].
 Thus, a sequence representing a fragment of the gene may be more than 97% similar to multiple reference sequences.
-Therefore, defining OTUs in the closed-reference approach is complicated because although two sequences might be 97% similar to the same reference sequence, they may only be 94% similar to each other.
+Therefore, defining OTUs in the closed-reference approach is problematic because two sequences might be 97% similar to the same reference sequence, but they may only be 94% similar to each other.
+Alternatively, a seuquence may be equally similar to two or more reference sequences.
 A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@Schloss2011].
@@
 The second approach has been referred to as distance-based [@Schloss2011] or *de novo* clustering [@NavasMolina2013].
-In this method, the distance between sequences is used to cluster sequences into OTUs rather than the distance to a reference database.
+In this approach, the distance between sequences is used to cluster sequences into OTUs rather than the distance to a reference database.
 In contrast to the efficiency of closed-reference clustering, the speed of hierarchical *de novo* clustering methods scale quadratically with the number of unique sequences.
@@
 The third approach, open-reference clustering, is a hybrid of the closed-reference and *de novo* approaches [@NavasMolina2013; @Rideout2014].
-Open-reference clustering method involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference.
-In theory, this method should exploit the strengths of both closed-referene and *de novo* clustering; however, the different OTU definitions employed by both approaches poses a possible problem when the methods are combined.
+Open-reference clustering involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference.
+In theory, this method should exploit the strengths of both closed-reference and *de novo* clustering; however, the different OTU definitions employed by closed-reference and *de novo* clustering pose a possible problem when the methods are combined.
 An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels using the average linkage method [@Schloss2011].
@@
 Some have focused on the time and memory required to process a dataset [@Sun2009; @Cai2011; @Rideout2014, @Mah2014].
-These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the clustering.
+These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the OTU assignments.
 Others have attempted to judge the quality of a method by its ability to generate data that parallels classification data [@Cai2011; @Sun2011; @Chen2013; @Edgar2013].
@@
 We first sought to assess the calculated the MCC for the OTU assignments generated by each of the clustering methods using 20, 40, 60, and 80% of the sequences relative to the OTU composition formed by the methods using the full dataset (Figure 1A).
-Similar to the He study, we replicated each method and subsample 30 times because a random number generator is used in some of methods to break ties where pairs of sequences have the same distance between them.
+Similar to the He study, we replicated each method and subsampled to the desired fraction of the dataset 30 times.
+Multiple subsamplings was necessary because a random number generator is used in some of the methods to break ties where pairs of sequences have the same distance between them.
 Across these sequencing depths, we observed that the stability of the OTUs generated by the SL and CL methods were highly sensitive to sampling effort relative to the OTUs generated by the AL, AGC, and DGC methods (Figure 1A).
@@
 Revisiting the concept of stability, we question the value of obtaining stable OTUs when the full dataset is not optimally assigned to OTUs.
-Our analysis indicates that the most rigorous method for assigning the Canadian soils sequences to OTUs using a 97% threshold is the AL method.
+Our analysis indicates that the most rigorous method for assigning the Canadian soils sequences to OTUs using a 97% threshold was the AL method.
 
@@
 Although no dataset is perfect for exhaustively testing these clustering methods, this dataset was useful for demonstrating several points.
-First, when using 60% of the data the stability relationships amongst the different methods were similar to what we observed using the Canadian soil dataset (Figure 3AB).
+First, when using 60% of the data, the stability relationships amongst the different methods were similar to what we observed using the Canadian soil dataset (Figure 3AB).
 With the exception for the clusters generated using CL, the methods all performed very well with stabilities greater than `r round(min(miseq_stability[miseq_stability$fraction=="0.6" & miseq_stability$method != "fn","mean"]), 2)`.
@@
 
+***Evalution of Swarm as an alternative de novo clustering algorithm.*** The Swarm algorithm is a recently proposed *de novo* method for assigning sequences to OTUs that identifies clusters of sequences based on the number of differences between each other without applying a distance threshold [@Mah2014].
+Swarm was originally validated by comparing the results against the expected clusters formed based on the taxonomic composition of a mock community.
+Similar to the authors of the He study, the Swarm developers suggest that method are needed that are insensitive to input order.
+Use of Swarm on the Canadian soil and murine datasets demonstrated that similar to the other *de novo* methods, Swarm's OTU assignments changed as sequences were added (Figures 1A and 3A).
+When we compared the OTU assignments for both datasets when using all of the sequence data, the variation in MCC values across the 30 randomizations were not meaningfully different (Figures 1D and 3D).
+Most importantly, when we selected the distance threshold that optimized the MCC value, the quality of the OTU assignments was close to that of the AL assignments when using the Canadian soil dataset and considerably worse than that of the murine dataset (Figures 1D and 3D).
+Interestingly, the distance thresholds that resulted in the largest MCC values were 3 and 2% for the Canadian soil and murine datasets, respectively.
+This suggests that OTU definitions are not consistent across datasets when using the Swarm algorithm.
+Finally, the Swarm developers indicated that hierarchical *de novo* algorithms were too impractical to use on large MiSeq-generated datasets.
+Our ability to apply AL to the large mouse dataset and even larger datasets suggests [e.g.
+@Schubert2015, @Zackular2015] that it is not necessary to sacrifice OTU assignment quality for speed.
+
 ***Evalution of an open-source alternative to USEARCH.*** For some datasets the AGC and DGC methods appear to perform as well or better than the hierarchical clustering methods.
-As originally described in the He study, the AGC and DGC methods utilized the USEARCH program [@Edgar2010], which is not available as open source code and is only available for free to academic users as a 32-bit program.
+As originally described in the He study, the AGC and DGC methods utilized the USEARCH program and the DGC method is used for clustering in UPARSE [@Edgar2010, @Edgar2013].
+The source code for USEARCH is not publicly available and only the 32-bit executables are available for free to academic users.
 Access for non-academic users and those needing the 64-bit version is available commercially from the developer.
@@
 In addition, the MCC values for the entire datasets, calculated relative to the distance matrix, were virtually indistinguishable.
-These results are a strong indication that VSEARCH is a suitable and possibly better replacement for USEARCH for executing the AGC and DGC methods.
-
-***Evalution of Swarm as an alternative de novo clustering algorithm.*** The Swarm algorithm is a recently proposed *de novo* method for assigning sequences to OTUs that identifies clusters of sequences based on the number of differences between each other without applying a distance threshold [@Mah2014].
-Swarm was originally validated by comparing the results against the expected clusters formed based on the taxonomic composition of a mock community.
-Similar to the authors of the He study, the Swarm developers suggest that method are needed that are insensitive to input order.
-Use of Swarm on the Canadian soil and murine datasets demonstrated that similar to the other *de novo* methods, Swarm's OTU assignments changed as sequences were added (Figures 1A and 3A).
-When we compared the OTU assignments for both datasets when using all of the sequence data, the variation in MCC values across the 30 randomizations were not meaningfully different (Figures 1D and 3D).
-Most importantly, when we selected the distance threshold that optimized the MCC value, the quality of the OTU assignments was close to that of the AL assignments when using the Canadian soil dataset and considerably worse than that of the murine dataset (Figures 1D and 3D).
-Interestingly, the distance thresholds that resulted in the largest MCC values were 0.03 and 0.02 for the Canadian soil and murine datasets, respectively.
-This suggests that OTU definitions are not consistent across datasets when using the Swarm algorithm.
-Finally, the Swarm developers indicated that hierarchical *de novo* algorithms were too impractical to use on large MiSeq-generated datasets.
-Our ability to apply AL to the large mouse dataset and even larger datasets suggests [e.g.
-@Schubert2015, @Zackular2015] that it is not necessary to sacrifice OTU assignment quality for speed.
+These results are a strong indication that VSEARCH is a suitable and possibly better option for executing the AGC and DGC methods.
 
@@
 ***Problems with reference-based clustering in general and as implemented in QIIME.*** The He study and our replication attempt validated that the closed-reference method generated perfectly stable OTUs.
-This was unsurprising since, by definition, the method represents a one-to-one mapping of reads to a reference and treats the input sequences independently.
+This was unsurprising since, by definition, the method is designed to return one-to-one mapping of reads to a reference.
+Furthermore, because it treats the input sequences independently the input order or use of a random number generator is not an issue.
 An important test that was not performed in the He study was to determine whether the clustering was sensitive to the order of the sequences in the database.
-The default database used in QIIME, which was also used in the He study and by others, contains full-length sequences that are at most 97% similar to each other.
+The default database used in QIIME, which was also used in the He study, contains full-length sequences that are at most 97% similar to each other.
 We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference method with the full murine dataset, which contained `r n_unique_miseq_seqs` unique sequences.
@@
 The two methods had similar sensitivities (USEARCH: `r mean(uclosed$sensitivity)` and VSEARCH: `r mean(vclosed$sensitivity)`), but the USEARCH specificity (`r mean(uclosed$specificity)`) was considerably higher than VSEARCH (`r mean(vclosed$specificity)`).
-Overall, these results indicate that although heuristic approaches may be fast, relative to non-heuristic approaches, they do a poor job of mapping reads to the correct reference sequence.
+Overall, these results indicate that although heuristic approaches may be fast, they do a poor job of mapping reads to the correct reference sequence relative to non-heuristic approaches.
 
@@
 Greater emphasis should be placed on obtaining an optimal balance between splitting similar sequences into separate OTUs and merging disparate sequences into the same OTU.
-Through the use of the pairwise distances between sequences, we were able to use the MCC to demonstrate that, in general, the AL, AGC, and DGC methods perform better than the others.
+Through the use of the pairwise distances between sequences, we were able to use the MCC to demonstrate that, in general, the AL method is consistently robust, but that Swarm, AGC, and DGC sometimes perform as well as AL.
+At least for the murine dataset, Swarm also could be among the least robust methods.
 Although there is concern that running the methods multiple times yields different clusterings, we have shown that there is little variation in their MCC values.
@@
 Finally, it is impossible to obtain a clustering with no false positives or false negatives and the optimal method may vary by dataset.
-With this in mind, researchers are encouraged to run the AL and VSEARCH AGC or DGC methods and calculate and report their MCC values.
+With this in mind, researchers are encouraged to calculate and report their MCC values for the AL method and at least one other method.
 
 Our analysis of those methods that implemented USEARCH as a method for clustering sequences revealed that its heuristic limited its specificity.
-When we replaced USEARCH with VSEARCH, the clustering quality improved.
+When we replaced USEARCH with VSEARCH, the clustering quality was as good or better.
 Although there may be parameters in USEARCH that can be tuned to improve the heuristic, these parameters are likely dataset dependent.
-Based on the data presented in this study, its availability as an open source, and free program, VSEARCH should replace USEARCH in these clustering methods.
+Based on the data presented in this study, its availability as an open source, and free program, VSEARCH should replace USEARCH in the *de novo* clustering methods; however, USEARCH performed better than VSEARCH for closed-reference clustering.
 Furthermore, although not tested in our study, VSEARCH can be parallelized leading to potentially significant improvements in speed.
@@
 
-Because the open-reference method is a hybrid of the closed-reference and DGC methods, it is also negatively affected by the problems with using USEARCH for both methods.
+Because the open-reference method is a hybrid of the closed-reference and DGC methods, it is also negatively affected by the various problems using USEARCH.
 An added problem with the open-reference method is that the two phases of the method employ different thresholds to define its OTUs.
@@
 We utilized the 97% OTUs greengenes reference sequence and taxonomy data (v.13.8) that accompanies the QIIME installation.
-Because the greengenes reference alignment does a poor job of representing the secondary structure of the 16S rRNA gene [@Schloss2010], we realigned the fasta sequences to a SILVA reference alignment to identify the V4 region of the sequences.
+Because the greengenes reference alignment does a poor job of representing the secondary structure of the 16S rRNA gene [@Schloss2010], we realigned the FASTA sequences to a SILVA reference alignment to identify the V4 region of the sequences.
 
@@
 The MCC was counted for each dataset using the formula above as implemented in the `sens.spec` command in mothur.
-To judge the quality of the Swarm-generated OTU assignments we calculated the MCC value using thresholds incremented by 0.01 between 0.00 and 0.05 and selected the threshold that provided the optimal MCC value.
+To judge the quality of the Swarm-generated OTU assignments we calculated the MCC value using thresholds incremented by 1% between 0 and 5% and selected the threshold that provided the optimal MCC value.
 
@@
 The vertical gray line indicates in A and C indicates the fraction of the dataset represented in B and D, respectively.
-The optimum threshold for the Swarm-generated assignments was 0.03.
+The optimum threshold for the Swarm-generated assignments was 3%.
 
@@
 The vertical gray line indicates in A and C indicates the fraction of the dataset represented in B and D, respectively.
-The optimum threshold for the Swarm-generated assignments was 0.02.
+The optimum threshold for the Swarm-generated assignments was 2%.
 

### 2015-10-29 Schloss_Cluster_PeerJ_2015.Rmd
verion -> version typo

@@
 Using the two datasets, we determined whether the AGC and DGC methods, as implemented by the two programs, yielded OTU assignments of similar quality.
-In general the overall trends that we observed with the USEARCH-version of AGC and DGC were also observed with the VSEARCH-verion of the methods (Figure 4).
-When we compared the two implementations of the AGC and DGC methods, the OTUs generated by the VSEARCH-verion of the methods were as stable or more stable than the USEARCH-version when using 60% of the datasets.
+In general the overall trends that we observed with the USEARCH-version of AGC and DGC were also observed with the VSEARCH-version of the methods (Figure 4).
+When we compared the two implementations of the AGC and DGC methods, the OTUs generated by the VSEARCH-version of the methods were as stable or more stable than the USEARCH-version when using 60% of the datasets.
 In addition, the MCC values for the entire datasets, calculated relative to the distance matrix, were virtually indistinguishable.

### 2015-11-12 rebuttal.md
Incorporate and respond to critique of R1

> Reviewer 1 (Anonymous)

> Basic reporting

> Code to reproduce the results has been provided in a github repository.

> See general comments.

> Experimental design

> See general comments.

> Validity of the findings

> See general comments.

> Comments for the author

> This paper evaluates the "stability" and "quality" of several different OTU clustering methods. "Stability" is defined here as the consistency between OTU assignments made on subsets of the data and the full dataset. "Quality" is defined here as assigning sequencing reads that are <3% different to the same OTU, and sequencing reads >3% different to different OTUs. The authors broad conclusions are that while reference based clustering is more "stable", that hierarchical de novo methods, particularly average linkage, produce OTUs of higher "quality". The authors also make an important observation that the OTU assignments of commonly used reference methods depend on the ordering of the reference database.

> This paper is largely a response to He et al. "Stability of operational taxonomic units: an important but neglected property for analyzing microbial diversity" which argued for reference based methods based on the "stability" of their OTU assignments.

> -------

> This paper is well-written, makes several useful observations, and is clearly suitable for publication. It can be improved by certain clarifications. I recommend its acceptance with minor revisions, as detailed below.

> More important points are preceded by an asterisk.

> -------

> *Terminology: Over the course of the paper, (at least) five different terms are used to refer to the goodness of a clustering: stability, rigor, robustness, quality and accuracy. Yet only two distinct concepts are explored (stability and quality/accuracy). I strongly recommend that the words "rigorous" and "robust" be struck from the manuscript entirely. They are never defined, are unclear, and the first impression they give does not correspond well to the actual results they are being used to describe.

We thank this Reviewer (and Reviewer 2) for expressing their concern that we are not sufficiently clear in our terminology. Part of the problem comes down to the multiple definitions of accuracy, precision, etc. in the classification and statistics literature. The multiple terms is symptomatic of our obviously failed attempt to be clear. Throughout the manuscript we will stick with stability and quality as the reviewer recommends.

> Abstract...

> L5: replace "rigorous" with something else, eg. "optimal", or even "best". Rigor is not being evaluated here.
> L27: replace "rigorous". What is described in this paper is that the de novo methods produce OTU clusterings that coincide better with the pairwise distances between the sequencing reads.
> L209: Remove "rigorous". You've used "accuracy" in this section (which appears identical to "quality" elsewhere, consider consolidating), so replace rigorous with accurate.

As suggested, we have replaced rigor/rigorous with optimal and accuracy with quality throughout the manuscript.

> L7-L8: Recommend introducing "quality" here alongside its initial operational definition.

Thank you for this recommendation and it has been incorporated.

> L16: replace "robust". Robust is a particularly poor choice of word here, as it brings to mind stability, which is not what is being referred to.
> L245: Remove "robust". The de novo methods produce clusterings that are more consistent with the pairwise separations of the individual sequencing reads. That does not correspond to any normal usage of the word robust.
> L362: replace "robust"
> L364: replace "robust"

As suggested, we have replaced robust throughout the manuscript

> *L20-L21: USEARCH -> USEARCH and VSEARCH. According to your results, both methods produced OTU assignments that depended on the ordering of the reference database (~55-60% label consistency between reorderings of the database).

We have corrected the text to indicate that both programs are deficient in this regard.

> Introduction...

> L46-47: Move the last sentence of this paragraph to the first sentence of the next paragraph.

This has been edited as suggested.

> L57-59: This sentence does not follow from the previous (don't need "Therefore,") and is not necessarily a shortcoming, but is rather an observation. All methods other than complete-linkage can join sequences together that are are less than 97% similar.

This has been edited as suggested.

> L60: seuquence -> sequence. And drop "subtle".

This has been edited as suggested.

> *L64: The other major advantage of reference based methods is that the OTU assignments can be compared across experiments and even studies. This should be mentioned here.

This has been edited as suggested.

> L68: speed -> computational cost (or something similar)

This has been edited as suggested.

> L76: drop "however"

This has been edited as suggested.

> L80: Move last sentence ahead of previous sentence, as this follows from the reference independence property, not the ability to subsequently classify.

This has been edited as suggested.

> L85: Clarify that this is a potential issue with "current" or "common" open-reference methods. It is not an inherent issue with any open-reference approach, as one can use coherent OTU assignment strategies for both parts (eg. join reference sequence or open-de-novo cluster if within 3%).

This has been edited as suggested.

> *L95-105: Evaluation of accuracy on mock communities needs to be included in this discussion of previous methods to evaluate OTU methods.

A sentence describing the use of mock communities has been added as suggested. Previously, these references were wrapped within the reference mapping approach.

> L109: Those sequences that... -> Those pairs of sequence that... The unit of this analysis is pairs of sequences, not individual sequences.

This has been edited as suggested here and in the Methods section.

> L117: attempted to characterize -> characterized

This has been edited as suggested.

> L137-139: Either explain how the stability MCC is calculated, or reference the Methods at this point.

This has been edited by referring the reader to the Methods section.

> L141: Multiple subsamplings was -> were

This has been edited as suggested.

> L148-152: This is almost an extreme level of disagreement on this point between essentially identical AL analyses in the He study and this study. I find the He results rather unbelievable, but can the authors shed any further light on what might have happened? THe He study doesn't provide a full workflow, but did provide the command they executed: cluster(column=unique.dist,name=names,cutoff=0.05,precision=1000,method=average)

My suspicion is that they did not compare apples-to-apples. I suspect they may have compared OTUs generated at one threshold for the subsampled data to OTUs generated at a different threshold for the full dataset. But it could really be anything. I have sought to clarify several questions with He and it has been a rather difficult process and decided to not bother with this point.

> *L182-190: Comparing OTUs on the basis of their representative sequence is not "specious" and the He analysis was not "so poorly designed". The most abundant sequence in an OTU is a biologically meaningful object, and we would like to think that OTUs are biologically meaningful and not just abstract objects. This section needs to be removed, or rewritten with less polemic language and with an explanation of how comparing representative sequences can break down (for example for very low abundance OTUs).

We appreciate the concern regarding our polemical language and have edited the text to be less confrontational.

> L243: the assignments are highly reproducible -> the overall quality of the assignments are consistent.
> The reproducibility of the assignments themselves was not evaluated here.

This has been edited as suggested.

> L246: Evalution -> Evaluation

This has been edited as suggested.

> L246-249: I don't really understand this. Swarm implements single-linkage clustering, in a way that is very clever computationally, but that is still single-linkage clustering.

We have clarified the description of the method to indicate that it is a SL method that then breaks up the long chains the method generally produces with a user-defined parameter.

> L251: method -> methods

This has been edited as suggested.

> L326-328: It would be useful to the reader to include the number of V4 covering reference sequences here.
> L323-328: Numbers of sequences and numbers of pairs of sequences are mixed together here in a confusing fashion. It would be better to do one or the other (eg. 7,785 _pairs of sequences_ that were more than 97% similar... -> XXX sequences within 97% of another reference sequence...). This allows the reader to easily evaluate the fraction of the reference database this corresponds to.

The first point has been edited as suggested and we believe that this helps to clarify the second point.

> L341-345: Is this referring to sequences duplicated over V4? Or over the whole 16S gene?

This has been edited to indicate that these numbers were over the V4 region.

> L354: "It is worth noting that the entire design of the He study was artificial." - This is unnecessarily polemic. There are ways in which the He study was not ideal, but essentially every OTU evaluation article uses an "artificial" design. The points being made here stand on their own, and are only obscured by this kind of language.

We thank the reviewer for encouraging us to take a less polemic tone. We have restructured this paragraph to make our point more clearly without being argumentative.

> L390-395: It appears to me that the dependence on the order of the reference database has the potential to introduce _systematic_ rather than just random errors, which is somewhat troubling.

We agree and have inserted a sentence here to underscore the point.

> Reviewer 2 (Thomas Schmidt)

> Basic reporting

> In their manuscript, 'De novo clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units', Schloss & Westcott report a series of quality benchmarks on several established and recent 16S sequence clustering approaches and methods. The authors critically reproduced and extended a recent study by He et al. (Microbiome, 2015); they provide arguments and data that OTU "stability" is not necessarily indicative of OTU "quality"; they provide OTU quality benchmarks for very recent algorithms/software (SWARM and VSEARCH); and they report a novel perspective and tests on reference-based OTU demarcation. I believe that this work has merit, that the science is generally sound (but see comments below) and that the manuscript is timely and important for the growing community of microbial ecologists.

> My main concern is that the study design is somewhat disparate, as the individual reported findings are not inherently inter-connected (see also comments on "Validity of the Findings"). Also, while the data and results provided generally support the authors' conclusions, I am not convinced that benchmarks based on the Matthew's Correlation Coefficient alone are the most informative approach for all the tests conducted (again, see comments below). Finally, there are several issues regarding the general reporting in the manuscript that could make it hard to follow for non-expert readers.

> Following the journal's recommendation, I have tried to organize all comments into the appropriate sections ("Basic Reporting", "Experimental Design" and "Validity of the Findings"). Within these sections, I tried to organize all major comments in decreasing order of importance and minor comments chronologically, following the manuscript. However, several points touch more than one of the specified comment sections, and I have tried to disentangle these as much as possible in order to achieve a readable review.

> Comments on "Basic Reporting".

> - The manuscript is overall very readable, clearly written, the global structure is clear and all individual sections are well structured. That being said, the structure of the study is also somewhat disparate: the individual parts stand somewhat independently from each other and do not form a fully coherent body of research. In the current form, the authors do not make it sufficiently clear how the first part of the manuscript (a critical reproduction of the study by He et al) relates to the second (quality benchmarks of the very recent methods SWARM and VSEARCH) and, in turn, to the third (general problems with reference-based OTU demarcation). While all of these findings are valid and interesting in themselves, the manuscript seems to be telling several unrelated stories at once. One general topic is a comparison of de novo vs reference-based OTUs, another is the question of OTU "stability" vs "accuracy" and a third thread refers to the clustering quality of novel methods, in particular a comparison of USEARCH to its open-source clone VSEARCH. As I said, the manuscript is very readable as is, but these findings are not sufficiently synthesized in the Conclusions, in my opinion.

> - All Figures, in their current form, are really non-obvious to interpret. Frankly speaking, when I first read through the manuscript, I understood all results from the text but really struggled at interpreting the figures. I believe they would profit from a careful workover.
> => Figures 1&3 lack color legends (it took me some time to realize that colors in the left panel corresponded to the categorical axis in right panel). Also, being partially color-blind, I had a hard time distinguishing the thick lines in Fig 1&3, left panels. Ordinate axis labeling in Fig 1&3 was not intuitive on a first read-through; the meaning of "MCC relative to full dataset" and "to distances" was not clear, and only a close look at the corresponding Methods section (but not the Figure caption) helped resolve this.
> => In my opinion, Fig 2 presents the data in a non-intuitive way. I am not sure which point the authors are trying to emphasize here: the illustration draws the focus to the generally varying number in absolute OTU counts, but from the arguments in the text I believe that relative differences in counts between re-clustered sub-setted datasets and rarefied full datasets are more important. The latter, however, are not clear to interpret from the figure as is. Also, it is not clear to me why clustering methods are the categorical axis while down-sampling steps are scattered vertically. Moreover, a comparison of SWARM is missing from this figure for some reason (?). I would suggest to split this figure into two parts: part A providing more classical rarefaction curves per method (with sub-setted clustering points as overlay scatter) and part B providing the (log) relative OTU count between rarefaction and sub-setting across down-sampling steps.
> => Figure 4 lacks a caption that is explanatory beyond the figure title.

> - The Introduction is generally very clear, concise and readable. However, there is a large body of previous work on testing methods for OTU clustering, and the authors focus on only a very limited subset of studies. While I think that a comprehensive review of all studies on OTU benchmarking ever published is certainly outside the scope of the present work, I believe that the manuscript would much profit from a more in-depth discussion of at least some. Other works possibly worth mentioning include, but are not limited to:
> => Huse et al., Env Microbiol, 2010, DOI: 10.1111/j.1462-2920.2010.02193.x
> => White et al., BMC Bioinformatics, 2010, doi:10.1186/1471-2105-11-152
> => Barriuso et al., BMC Bioinformatics, 2011, doi:10.1186/1471-2105-12-473, who indeed provided an early investigation into "output variability" of OTU clustering (Table 7)
> => Bonder et al., Bioinformatics, 2012, doi: 10.1093/bioinformatics/bts552
> => Koeppel & Wu, NAR, 2013, doi: 10.1093/nar/gkt241
> => Schmidt et al., PLOS Comp Biol, 2014, DOI: 10.1371/journal.pcbi.1003594
> => May et al., Bioinformatics, 2014, doi: 10.1093/bioinformatics/btu085
> However, these are merely suggestions and the authors should not feel obliged to include any or all of these as new references in the Introduction – in particular, since I also pointed out one work of which I am co-author. Nevertheless, I do believe that the manuscript would generally profit from a discussion of (some) further previous work.

> - The Introduction provides a detailed, accessible and balanced explanation of the three general approaches to OTU demarcation discussed in the paper: de novo, closed reference and open reference. Although the authors point to two comprehensive references which give further information on these points (Schloss & Westcott, 2011; Navas-Molina et al., 2013), I would suggest that the authors provide at least some visual/structural guidance to the reader (an overview figure or table). However, the manuscript is certainly understandable as is, so please consider this comment as a mere suggestion.

> - Related to this, the terms "abundance-based" and "distance-based greedy clustering" (AGC and DGC) are not properly introduced. Although the authors point out that they stick with the terminology of the He et al paper, these terms need to be explained somewhere in the manuscript, preferably in the Introduction.

> - Similarly, the last part in the Results and Discussion, "Problems with reference-based clustering [...]", could profit from an illustration. This part is very clearly written, and the presented results are striking (see also comments under "Validity of the Findings"). Nevertheless, a visual presentation of the data could additionally help to guide the reader through the conducted tests and illustrate just how striking the reported results are. Even a simple table summarizing the results would be helpful.

> - The authors very clearly point out weaknesses in the criticized study by He et al; they support all their arguments with data and/or sound reasoning and do not, in my opinion, state any unfair or unsupported claims against the criticized work. Nevertheless, I personally felt that the wording is sometimes a bit harsh: e.g., p.9, l.190 ("this analysis was so poorly designed"); p.15, l.354 ("the entire design of the He study was artificial").

> Minor comments:

> - Please re-check for typos and grammar across the text (e.g., p.5, l.60 "seuquence"; p.5, l.69 "scale[s]"; p.6, l.88 "assigned"; p.7, l.137 "assess the calculated the MCC"; etc.)

> - Several examples given throughout the text appear rather arbitrary. For example, p.5, l.59 ("may only be 94% similar", same in p.17, ll.402-3) or p.6, l.89 ("all sequences assigned to the Porphyromonadaceae would then be assigned to OTUs [...]") are out of context, the sudden mention of a concrete clustering threshold or an arbitrary bacterial family could confuse the reader more than illustrate the point being made.

> - p.4, l.56. I believe that the provided reference for the statement that "the commonly used variable regions [...] do not evolve at the same rate" is not fully appropriate. Citing Schloss, Plos Comp Biol, 2010 here (as is done elsewhere in the text) seems more meaningful, as this study is arguably the most comprehensive to date regarding this point. Moreover, it is probably also worth mentioning the related contribution by Kim et al. (Journal Micobiol Methods, 2011, http://www.ncbi.nlm.nih.gov/pubmed/21047533).

> - When enumerating the pros and cons of closed ref clustering (p.5, ll. 48f), several points in favor of this approach could be added: it is trivially parallelizable, has a low memory footprint and arguably gets better as reference databases improve (which is the case as more and more sequence data accumulates).

> - p.6, l.93-4. This statement is phrased too strongly. The approach advocated by the authors (taxonomic classification, followed by de novo clustering within taxonomic bins) has been shown to be very useful in practice, but presenting it as an ideal compromise between closed reference and de novo approaches is an overstatement and potentially misleading. The authors, themselves, point out just a few lines later that taxonomic classification is often far from perfect (it could be added that it depends strongly on reference database quality and classifier performance); however, if the classification step is not perfect, the entire clustering outcome will be (strongly) affected. Others have suggested database-agnostic alternatives in a similar spirit, such as SL pre-clustering (Huse et al., Env Microbiol, 2010) or the kmer-pre-clustering implemented in ESPRIT-Tree (Cai & Sun, NAR, 2011) which would be worth mentioning here.

> - p.6, ll.109ff. This paragraph explaining the MCC (in detail) feels a little out of place here; I suggest to relocate it to the (early) Results section.

> Experimental design

> - Methods are reported clearly and concisely, links to available data are provided, all tested tools are specified down to the version number. More Methods sections should look like this! In particular, I would like to commend the authors for uploading all code as reproducible workflow to Github.

> - Matthew's Correlation Coefficient (MCC) is used throughout the study as the sole indicator of OTU "quality". The MCC is an objective summary statistic for classification problems and has been successfully used for OTU benchmarking before (originally introduced by the authors in their 2011 paper and since then picked up by several other teams). Nevertheless, I would like to point out three general objections to using the MCC as an OTU quality criterion and then argue that the MCC, when used as a sole index, is not the most informative choice for (all) the questions that the authors pursue in this study.

> => In general, the MCC is a summary statistic for binary classification problems. However, (OTU) clustering is arguably not a problem that boils down to binary true/false assignments only. Of course, violations of a "hard" sequence similarity clustering threshold provide an objective criterion, but different algorithms implement inherently different concepts of a "threshold", and are consequentially prone to different types of errors. For example, by definition, complete linkage or USEARCH would never show "false positive" clustering (but many false negatives can be expected), while single linkage would never make "false negative" mistakes (while a wealth of false positives is expected). In other words, the MCC was not designed for this kind of application and OTU "quality" is not (only) a binary classification problem; nevertheless, MCC values are arguably informative, if interpreted carefully.
> => My second "general" issue with the MCC concerns its formulation as a summary statistic: frankly speaking, I find it rather non-intuitive to interpret. For example, in the present study, it would be very informative to see what actually goes "right" or "wrong" for the different methods, driving MCC values (e.g., by showing distributions of TPR, FPR, etc. per clustering method).
> => Finally, I suspect that the MCC is sensitive to cluster count and size distribution, although I am not aware that this has ever been tested. In general, more "errors" can be expected the further a clustering algorithm proceeds: sets with fewer and larger OTUs would be more prone to erroneous assignments than sets with many small OTUs. This is problematic, because the "hard" clustering threshold is implemented very differently by different algorithms, and it is well established that at a given threshold of, say, 97% the different methods provide very different cluster count and size distributions. In the present study, such effects would be especially striking for the SWARM algorithm.

> In spite of these comments, the use of the MCC in the second use case of the present study ("MCC value relative to distances") is arguably established and meaningful, if interpreted cautiously. However, in my opinion, the MCC is an insufficient metric in the first use case ("MCC value relative to full dataset"). The authors generated "reference" OTU sets based on the full datasets and subsequently compared OTU sets based on randomly down-sampled datasets to these using the MCC. In this setup, TPs are sequence pairs which end up in the same OTU under full and down-sampled clustering, etc. Arguably, this corresponds even less to the original use case of the MCC, and putative differential cluster size and count effects would be even more striking (based on the authors' findings in Figure 2). What appears particularly confusing is that the interpretation of the MCC changes drastically between both use cases: the "MCC on distances" is interpretable in line with previous work, but the "MCC relative to full dataset clustering" means something completely different.

> Several metrics have been developed which appear more appropriate for the latter use case. The down-sampling experiment, in my opinion, would be more straightforward to interpret if clustering comparison metrics, e.g. based on mutual information were to be used. I freely admit to a "conflict of interest" here: in a previous work I was involved in (DOI: 10.1111/1462-2920.12610), we used Adjusted Mutual Information (Vinh et al., 2010, http://www.jmlr.org/papers/v11/vinh10a.html), Normalized Mutual Information (Fred 7 Jain, 2003, http://ieeexplore.ieee.org/stamp/stamp.jsp?arnumber=1211462) and the Adjusted Rand Index (Hubert & Arabie, 1985, http://link.springer.com/article/10.1007%2FBF01908075) in a very similar setup (testing clustering robustness to selective down-sampling). While I generally consider it very impolite to push one's own work as a reviewer, I do believe that in this case, there is a clear scientific motivation for doing so. For the tests conducted in Fig 1AB & 3AB, at least, I believe that the above measures would provide a highly informative complement, if not replacement, for the MCC. They are straightforward indices of OTU "stability", in the sense defined in the present manuscript. Indeed, based on our previous observations (Fig 5 in Schmidt et al., Environ Microbiol, 2015), I would expect that the trends observed in the present study would be even more pronounced. Taken together, MCC and AMI/NMI/ARI could then indeed be interpreted as OTU "accuracy" and "stability". Likewise, the comparison between USEARCH and VSEARCH is a model use case for which the above-mentioned methods were originally designed.

> Validity of the findings

> - As detailed above, I believe that all conclusions are generally justified by the analyses conducted, albeit reservations regarding the use of the MCC as sole index in the stated cases. However, as likewise detailed above, I also believe that several findings stand independently of each other, and their connections are not immediately obvious.

> - One main rationale of the present study is that OTU assignment "accuracy" trumps "stability". The authors provide clear reasoning to support this argument, and their data is generally in line with this notion. Nevertheless, one could argue that indeed, both "stability" and "accuracy" are important features of OTUs, and that their relative significance depends on the application. It is neither desirable to work with "stable but wrong" nor "accurate but noisy" OTUs. I agree with the authors that accuracy is the more desirable property in general, but putting their study in context with other recent findings could suggest that in fact, accuracy and stability tend to be correlated, which would lend even more gravity to their argument.

> - The points raised in critical response to the He et al. study are all well illustrated, argued and supported by data. Moreover, by repeating several of the tests on a more recent (and more reasonably chosen), more deeply sequenced dataset, the authors provide a clear added value.

> - The inclusion of tests on the very recent SWARM and VSEARCH algorithms makes this manuscript very timely and useful for researchers in the field. To my knowledge, this is the first available independent test on VSEARCH performance.

> - The findings on "problems with reference-based clustering in general and as implemented in QIIME" are striking and, indeed, slightly disturbing. To be honest, I would not have expected such strong effects of sequence order for mappings against the QIIME reference database. I believe that these results will contribute to more debates in the field, and ultimately lead to further long-needed benchmarks on standard operating procedures for widely used software.

> - p.12, ll.260-1. This statement is too harsh based on the given data. The SWARM algorithm is designed to not work with "hard" thresholds in the classical sense, so it is not surprising that in the given testing framework, different "thresholds" would appear optimal for differentially deeply sequenced datasets.

> - p.15, ll.368-9. This solution appears slightly impractical. While it would clearly be desirable if researchers provided clustering quality metrics along with their results (and ideally, for more than one clustering method), calculating the MCC would always require to calculate and store full sequence distance matrices. Also, it is unclear what an MCC comparison across different studies would actually mean (considering additional confounding factors such as sequencing depth, clustering threshold, alignment, etc.). Nevertheless, I fully agree with the authors that researchers should ideally perform (and report) different clusterings on their data.

> Comments for the author

> I would like to again commend the authors on their manuscript and again apologize for pushing a study their way in which I was involved. I was very hesitant to include the points on the use of the MCC vs AMI/NMI/ARI in this review, but I honestly believe that they make sense scientifically.

### 2015-11-12 Schloss_Cluster_PeerJ_2015.Rmd
Fix authorship order
Incorporate and respond to critique of R1

@@
 \vspace{25mm}
-Patrick D. Schloss${^\dagger}$ and Sarah L. Westcott
+Sarah L. Westcott and Patrick D. Schloss${^\dagger}$
 
@@
 **Background.** 16S rRNA gene sequences are routinely assigned to operational taxonomic units (OTUs) that are then used to analyze complex microbial communities.
-A number of methods have been employed to carry out the assignment of 16S rRNA gene sequences to OTUs leading to confusion over which method is the most rigorous.
+A number of methods have been employed to carry out the assignment of 16S rRNA gene sequences to OTUs leading to confusion over which method is optimal.
 A recent study suggested that a clustering method should be selected based on its ability to generate stable OTU assignments that do not change as additional sequences are added to the dataset.
-In contrast, we contend that the ability of the method to properly represent the distances between the sequences is more important.
+In contrast, we contend that the quality of the OTU assignments, the ability of the method to properly represent the distances between the sequences, is more important.
 
@@
 **Results.** The stability of OTU assignments did not reflect the quality of the assignments.
-Depending on the dataset being analyzed, the average linkage and the distance and abundance-based greedy clustering methods generated more robust OTUs than the open and closed-reference methods.
+Depending on the dataset being analyzed, the average linkage and the distance and abundance-based greedy clustering methods generated OTUs that were more likely to represent the actual distances between sequences than the open and closed-reference methods.
 We also demonstrated that for the greedy algorithms VSEARCH produced assignments that were comparable to those produced by USEARCH making VSEARCH a viable free and open source alternative to USEARCH.
-Further interrogation of the reference-based methods indicated that when USEARCH is used to identify the closest reference, the OTU assignments were sensitive to the order of the reference sequences because the reference sequences can be identical over the region being considered.
+Further interrogation of the reference-based methods indicated that when USEARCH or VSEARCH were used to identify the closest reference, the OTU assignments were sensitive to the order of the reference sequences because the reference sequences can be identical over the region being considered.
 More troubling was the observation that while both USEARCH and VSEARCH have a high level of sensitivity to detect reference sequences, the specificity of those matches was poor relative to the true best match.
@@
 **Discussion.** Our analysis calls into question the quality and stability of OTU assignments generated by the open and closed-reference methods as implemented in current version of QIIME.
-This study demonstrates that *de novo* methods are the most rigorous and that the quality of clustering assignments needs to be assessed for multiple methods to identify the optimal clustering method for a particular dataset.
+This study demonstrates that *de novo* methods are the optimal method of assigning sequences into OTUs and that the quality of these assignments needs to be assessed for multiple methods to identify the optimal clustering method for a particular dataset.
 
@@
 One advantage of the OTU-based approach is that the definition of the bins is operational and can be changed to suit the needs of the particular project.
-However, with the dissemination of clustering methods within software such as mothur [@Schloss2009], QIIME [@Caporaso2010], and other tools [@Edgar2010, @Sun2009, @Mah2014, @Edgar2013, @Cai2011], it is important to understand how different clustering methods implement this conventional OTU threshold.
+However, with the dissemination of clustering methods within software such as mothur [@Schloss2009], QIIME [@Caporaso2010], and other tools [@Edgar2010; @Sun2009; @Mah2014; @Edgar2013; @Cai2011], it is important to understand how different clustering methods implement this conventional OTU threshold.
 Furthermore, it is necessary to understand how the selected method affects the precision and accuracy of assigning sequences to OTUs.
+
 Broadly speaking, three approaches have been developed to assign sequences to OTUs.
-
 The first approach has been referred to as phylotyping [@Schloss2011] or closed reference clustering [@NavasMolina2013].
@@
 Thus, a sequence representing a fragment of the gene may be more than 97% similar to multiple reference sequences.
-Therefore, defining OTUs in the closed-reference approach is problematic because two sequences might be 97% similar to the same reference sequence, but they may only be 94% similar to each other.
-Alternatively, a seuquence may be equally similar to two or more reference sequences.
-A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@Schloss2011].
-The strength of the reference based approach is that the methods are generally fast, scaling linearly with the number of sequences being clustered.
+Defining OTUs in the closed-reference approach is problematic because two sequences might be 97% similar to the same reference sequence, but they may only be 94% similar to each other.
+Alternatively, a sequence may be equally similar to two or more reference sequences.
+A alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@Schloss2011].
+The strength of the reference based approach is that the methods are generally fast, scaling linearly with the number of sequences being clustered and that OTU assignments can be compared across studies.
 
@@
 In this approach, the distance between sequences is used to cluster sequences into OTUs rather than the distance to a reference database.
-In contrast to the efficiency of closed-reference clustering, the speed of hierarchical *de novo* clustering methods scale quadratically with the number of unique sequences.
+In contrast to the efficiency of closed-reference clustering, the computational cost of hierarchical *de novo* clustering methods scale quadratically with the number of unique sequences.
 The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences.
 If error rates can be reduced through stringent quality control measures, then these problems can be overcome [@Kozich2013].
-As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@Edgar2010, @Sun2009, @Mah2014].
-One critique of *de novo* approaches is that OTU assignments are sensitive to the input order of the sequences [@He2015, @Mah2014].
-Whether the differences in assignments is meaningful is unclear; however, the variation in results could represent equally valid clustering of the data.
+As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@Edgar2010; @Sun2009; @Mah2014].
+One critique of *de novo* approaches is that OTU assignments are sensitive to the input order of the sequences [@He2015; @Mah2014].
+Whether the differences in assignments is meaningful is unclear and the variation in results could represent equally valid clustering of the data.
 The strength of *de novo* clustering is its independence of references for carrying out the clustering step.
+For this reason, *de novo* clustering has been preferred across the field.
 After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU [@Schloss2011].
-For this reason, *de novo* clustering has been preferred across the field.
 
@@
 Open-reference clustering involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference.
-In theory, this method should exploit the strengths of both closed-reference and *de novo* clustering; however, the different OTU definitions employed by closed-reference and *de novo* clustering pose a possible problem when the methods are combined.
+In theory, this method should exploit the strengths of both closed-reference and *de novo* clustering; however, the different OTU definitions employed by commonly used closed-reference and *de novo* clustering implementations pose a possible problem when the methods are combined.
 An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels using the average linkage method [@Schloss2011].
@@
 It has been difficult to objectively assess the quality of OTU assignments.
-Some have focused on the time and memory required to process a dataset [@Sun2009; @Cai2011; @Rideout2014, @Mah2014].
+Some have focused on the time and memory required to process a dataset [@Sun2009; @Cai2011; @Rideout2014; @Mah2014].
 These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the OTU assignments.
-Others have attempted to judge the quality of a method by its ability to generate data that parallels classification data [@Cai2011; @Sun2011; @Chen2013; @Edgar2013].
+Others have attempted to judge the quality of a method by its ability to generate data that parallels classification data [@Cai2011; @Sun2011; @Chen2013].
 This approach is problematic because bacterial taxonomy often reflects historical biases amongst bacterial systematicists.
 Furthermore, it is well known that the rates of evolution across lineages are not the same [@Wang2007; @Schloss2010].
+A related approach has used clustering of mock community data to evaluate methods [@Edgar2013; @Mah2014].
+Yet these approaches ignore the effects of sequencing errors that tend to accumulate with sequencing depth and represent highly idealized communities [@Kozich2013; @Schloss2011Reducing].
 Others have assessed the quality of clustering based on their ability to generate the same OTUs generated by other methods [@Schmidt2014; @Rideout2014].
@@
 We recently proposed an approach for evaluating OTU assignments using the distances between pairs of sequences [@Schloss2011].
-Those sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives.
-Meanwhile, those sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives.
+Those pairs of sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives.
+Meanwhile, those pairs of sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives.
 Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and negatives using the Matthew's correlation coefficient [MCC; @Matthews1975].
@@
 
-A recent analysis by He and colleagues [-@He2015] attempted to characterize the three general clustering approaches by focusing on what they called stability.
+A recent analysis by He and colleagues [-@He2015] characterized the three general clustering approaches by focusing on what they called stability.
 They defined stability as the ability of a method to provide the same clustering on a subset of the data as was found in the full dataset.
@@
 Among the *de novo* methods they observed that the abundance-based greedy clustering (AGC) method generated the most stable OTUs followed by the single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) methods.
-We first sought to assess the calculated the MCC for the OTU assignments generated by each of the clustering methods using 20, 40, 60, and 80% of the sequences relative to the OTU composition formed by the methods using the full dataset (Figure 1A).
+We first calculated the MCC for the OTU assignments generated by each of the clustering methods using 20, 40, 60, and 80% of the sequences relative to the OTU composition formed by the methods using the full dataset (see Methods for description; Figure 1A).
 Similar to the He study, we replicated each method and subsampled to the desired fraction of the dataset 30 times.
-Multiple subsamplings was necessary because a random number generator is used in some of the methods to break ties where pairs of sequences have the same distance between them.
+Multiple subsamplings were necessary because a random number generator is used in some of the methods to break ties where pairs of sequences have the same distance between them.
 Across these sequencing depths, we observed that the stability of the OTUs generated by the SL and CL methods were highly sensitive to sampling effort relative to the OTUs generated by the AL, AGC, and DGC methods (Figure 1A).
@@
 The He study observed adonis R statistics of 0.02 (closed-reference), 0.03 (open-reference), 0.07 (CL, AGC, DGC), and 0.16 (SL and AL).
-Regardless of the statistical or biological significance of these results, the analysis does not make sense since, by definition, representing communities based on their unstable OTUs would yield differences.
+Regardless of the statistical or biological significance of these results, the analysis was tautological since, by definition, representing communities based on their unstable OTUs would yield differences.
 Furthermore, the *de novo* and open-reference approaches do not consistently label the OTUs that sequences belong to when the clustering methods are run multiple times with different random number seeds.
 To overcome this, the authors selected representative sequences from each OTU and used those representative sequences to link OTU assignments between the different sized sequence sets.
-The justification for this analysis is specious as the OTU assignments are based on the data available in the dataset when the sequences are clustered and comparing assignments in this manner are irreconcilable.
 It is not surprising that the only analysis that did not provide a significant p-value was for the closed-reference analysis, which is the only analysis that provides consistent OTU labels.
 Finally, the authors built off of this analysis to count the number of OTUs that were differentially represented between the subsetted and full datasets by each method.
-This analysis assumes that the OTUs generated using the full dataset were correct, which was an unsubstantiated assumption since the authors did not assess the quality of the OTU assignments.
-Because this analysis was so poorly designed, we did not seek to reproduce it.
+This entire analysis assumed that the OTUs generated using the full dataset were correct, which was an unsubstantiated assumption since the authors did not assess the quality of the OTU assignments.
 
-***OTU assignment methods vary in their accuracy.*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs.
+***OTU assignment methods vary in their quality.*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs.
 A method can generate highly stable OTUs, but the OTU assignments may be meaningless if they poorly represent the specified cutoff and the actual distance between the sequences.
 To assess the quality of OTU assignments by the various methods, we made use of the pairwise distance between the unique sequences to count the number of true positives and negatives and the number of false positives and negatives for each method and sampling depth.
-This enabled us to calculate the average MCC value as a measure of a method's accuracy and its variation as a measure of its precision.
+This enabled us to calculate the average MCC value as a measure of a method's quality and its variation as a measure of its consistency.
 We made three important observations.
 First, each of the *de novo* methods varied in how sensitive their MCC values were to additional sequences (Figure 1C).
-The SL and CL methods were the most sensitive; however, the accuracy of the OTU assignments did not meaningfully differ when 80 or 100% of the data were assigned to OTUs using the *de novo* methods.
+The SL and CL methods were the most sensitive; however, the quality of the OTU assignments did not meaningfully differ when 80 or 100% of the data were assigned to OTUs using the *de novo* methods.
 Second, the AL method had higher MCC values than the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL (Figure 1D).
@@
 Revisiting the concept of stability, we question the value of obtaining stable OTUs when the full dataset is not optimally assigned to OTUs.
-Our analysis indicates that the most rigorous method for assigning the Canadian soils sequences to OTUs using a 97% threshold was the AL method.
+Our analysis indicates that the most optimal method for assigning the Canadian soils sequences to OTUs using a 97% threshold was the AL method.
 
@@
 Therefore, although the methods have a stochastic component, the OTU assignments do not vary meaningfully between runs.
-The results from both the Canadian soil and murine microbiota datasets demonstrate that the *de novo* methods can generate stable OTU assignments and that the assignments are highly reproducible.
-Most importantly, these analyses demonstrate that the OTU assignments using the AL, AGC, and DGC *de novo* methods are consistently more robust than either of the reference-based methods.
+The results from both the Canadian soil and murine microbiota datasets demonstrate that the *de novo* methods can generate stable OTU assignments and that the the overall quality of the assignments were consistent.
+Most importantly, these analyses demonstrate that the OTU assignments using the AL, AGC, and DGC *de novo* methods were consistently better than either of the reference-based methods.
 
-***Evalution of Swarm as an alternative de novo clustering algorithm.*** The Swarm algorithm is a recently proposed *de novo* method for assigning sequences to OTUs that identifies clusters of sequences based on the number of differences between each other without applying a distance threshold [@Mah2014].
+***Evaluation of Swarm as an alternative de novo clustering algorithm.*** The Swarm algorithm is a recently proposed *de novo* method for assigning sequences to OTUs that uses user-defined parameters to break up chains generated by SL clustering [@Mah2014].
 Swarm was originally validated by comparing the results against the expected clusters formed based on the taxonomic composition of a mock community.
-Similar to the authors of the He study, the Swarm developers suggest that method are needed that are insensitive to input order.
+Similar to the authors of the He study, the Swarm developers suggest that methods are needed that are insensitive to input order.
 Use of Swarm on the Canadian soil and murine datasets demonstrated that similar to the other *de novo* methods, Swarm's OTU assignments changed as sequences were added (Figures 1A and 3A).
@@
 Our ability to apply AL to the large mouse dataset and even larger datasets suggests [e.g.
-@Schubert2015, @Zackular2015] that it is not necessary to sacrifice OTU assignment quality for speed.
+@Schubert2015; @Zackular2015] that it is not necessary to sacrifice OTU assignment quality for speed.
 
-***Evalution of an open-source alternative to USEARCH.*** For some datasets the AGC and DGC methods appear to perform as well or better than the hierarchical clustering methods.
-As originally described in the He study, the AGC and DGC methods utilized the USEARCH program and the DGC method is used for clustering in UPARSE [@Edgar2010, @Edgar2013].
+***Evaluation of an open-source alternative to USEARCH.*** For some datasets the AGC and DGC methods appear to perform as well or better than the hierarchical clustering methods.
+As originally described in the He study, the AGC and DGC methods utilized the USEARCH program and the DGC method is used for clustering in UPARSE [@Edgar2010; @Edgar2013].
 The source code for USEARCH is not publicly available and only the 32-bit executables are available for free to academic users.
@@
 When a sequence was used to search the randomized databases, it would encounter a different reference sequence as the first match with each randomization.
-Among those reference sequences that fully overlap the V4 region, there were `r n_small_v19_dist` pairs of sequences that were more than 97% similar to each other over the full length of the 16S rRNA gene.
+Among the `r n_v4_overlap` reference sequences that fully overlap the V4 region, there were `r n_small_v19_dist` pairs of sequences that were more than 97% similar to each other over the full length of the 16S rRNA gene.
 When the extracted V4 sequences were dereplicated, we identified `r n_v4_unique` unique sequences.
@@
 As one illustration of the effects that incorrect labels would have on an analysis, we asked whether the duplicate sequences had the same taxonomies.
-Among the `r sum(dups$V2==2)` reference sequences that had one duplicate, `r sum(dups$V2==2 & dups$V3==2)` had discordant taxonomies.
-Furthermore, among those `r sum(dups$V2 > 2)` sequences with two or more duplicates, `r sum(dups$V2>2 & dups$V3>=2)` had discordant taxonomies.
-Two sequences mapped to `r max_taxa[1,2]-1` and `r max_taxa[2,2]-1` duplicate sequences and both contained `r max_taxa[1,3]` different taxonomies.
-Among the sequences within the database, there was also a sequence that had `r max(dups[,2])-1` duplicates and contained `r dups[which.max(dups[,2]),3]` different taxonomies.
-When we analyzed the `r n_actual_map` sequences that mapped to the reference sequences using a non-heuristic approach, we observed that `r n_map_more_than_one_expanded` of the sequences mapped to more than one reference sequence.
+Among the `r sum(dups$V2==2)` V4 reference sequences that had one duplicate, `r sum(dups$V2==2 & dups$V3==2)` had discordant taxonomies.
+Furthermore, among those `r sum(dups$V2 > 2)` V4 reference sequences with two or more duplicates, `r sum(dups$V2>2 & dups$V3>=2)` had discordant taxonomies.
+Two V4 reference sequences mapped to `r max_taxa[1,2]-1` and `r max_taxa[2,2]-1` duplicate sequences and both contained `r max_taxa[1,3]` different taxonomies.
+Among the V4 sequences within the database, there was also a sequence that had `r max(dups[,2])-1` duplicates and represented `r dups[which.max(dups[,2]),3]` different taxonomies.
+When we analyzed the `r n_actual_map` sequences that mapped to the V4 reference sequences using a non-heuristic approach, we observed that `r n_map_more_than_one_expanded` of the sequences mapped to more than one reference sequence.
 Of these sequences, `r n_map_one_uniqued` (`r 100*n_map_one_uniqued/n_map_more_than_one_expanded`%) mapped to references that were identical over the V4 region and `r n_map_more_than_one_uniqued` (`r 100*n_map_more_than_one_uniqued/n_map_more_than_one_expanded`%) mapped equally well to two or more references that were not identical over the V4 region.
@@
 ## Conclusions
-It is worth noting that the entire design of the He study was artificial.
+It is worth noting that the analysis from the Roesch study that motivated the He study is not typical of microbial ecology studies.
 First, their analysis was based on a single soil sample.
@@
 Second, all of the sequence data from these datasets is pooled for a single analysis.
-It is unclear why anyone would ever perform an analysis based on a subset of their data.
-Because of these points, the value of identifying stable OTUs is unclear.
+Rarely would a researcher rarefy their data prior to clustering since it can be more efficiently done after all of the data are assigned to OTUs.
+Third, the CL method used in the original Roesch study has since been shown to not generate optimal OTUs [@Schloss2011].
+As for the approach used in the He study, the value of identifying stable OTUs is unclear.
+Although there is concern that running the methods multiple times yields different clusterings, we have shown that there is little variation in their quality.
+This suggests that the different clusterings by the same method are equally good.
 Greater emphasis should be placed on obtaining an optimal balance between splitting similar sequences into separate OTUs and merging disparate sequences into the same OTU.
-Through the use of the pairwise distances between sequences, we were able to use the MCC to demonstrate that, in general, the AL method is consistently robust, but that Swarm, AGC, and DGC sometimes perform as well as AL.
-At least for the murine dataset, Swarm also could be among the least robust methods.
-Although there is concern that running the methods multiple times yields different clusterings, we have shown that there is little variation in their MCC values.
-This suggests that the different clusterings by the same method are equally good.
-Finally, it is impossible to obtain a clustering with no false positives or false negatives and the optimal method may vary by dataset.
+
+The approach of the current study quantified the effects of merging and splitting by using an objective metric.
+Through the use of the pairwise distances between sequences, we were able to use the MCC to demonstrate that, in general, the AL method was consistently the optimal method for each dataset, but that Swarm, AGC, and DGC sometimes perform as well as AL.
+At least for the murine dataset, Swarm also could be among the methods that performed poorly.
+It is impossible to obtain a clustering with no false positives or false negatives and the optimal method may vary by dataset.
 With this in mind, researchers are encouraged to calculate and report their MCC values for the AL method and at least one other method.
@@
 Perhaps most alarming was that the default order of the database provided the worst MCC of any of the randomizations we attempted.
+This has the potential to introduce systematic a bias rather than a random error.
 Even when we used VSEARCH to perform closed-reference clustering and were able to obtain a consistent clusterings, we observed that the labels on the OTUs differed between randomizations.
@@
 Briefly, we compared the OTU assignments for pairs of sequences to the distance matrix that was calculated between all pairs of aligned sequences.
-For each dataset that was clustered, those sequences that were in the same OTU and had a distance less than 3% were TPs, those that were in different OTUs and had a distance greater than 3% were TNs, those that were in the same OTU and had a distance greater than 3% were FPs, and those that were in different OTUs and had a distance less than 3% were FNs.
+For each dataset that was clustered, those pairs of sequences that were in the same OTU and had a distance less than 3% were TPs, those that were in different OTUs and had a distance greater than 3% were TNs, those that were in the same OTU and had a distance greater than 3% were FPs, and those that were in different OTUs and had a distance less than 3% were FNs.
 The MCC was counted for each dataset using the formula above as implemented in the `sens.spec` command in mothur.

### 2015-11-13 rebuttal.md
Incorporate and respond to critique of R2

@@
 
-This has been edited as suggested.
+We would prefer to leave this sentence where it is because it introduces the following three paragraphs.
 
@@
 
+The reviewer raises a great point.
+We have added a paragraph at the end of the section where we describe our reproduction of the He study.
+The paragraph lays out five additional questions that we raised because of this reproduction analysis.
+These questions now serve as the sub-section headings for the next five sub-sections of the Results & Discussion section.
+We are optimistic that this paragraph helps to tie together the various parts of the analysis.
+
 > - All Figures, in their current form, are really non-obvious to interpret.
@@
 I believe they would profit from a careful workover.
+
+We appreciate the reviewer's feedback and have altered the presentation of the figures and their captions. Hopefully, their presentation is more clear in the current draft.
+
 > => Figures 1&3 lack color legends (it took me some time to realize that colors in the left panel corresponded to the categorical axis in right panel).
@@
 Ordinate axis labeling in Fig 1&3 was not intuitive on a first read-through; the meaning of "MCC relative to full dataset" and "to distances" was not clear, and only a close look at the corresponding Methods section (but not the Figure caption) helped resolve this.
+
+We have added a plotting symbol to the lines in panels A and C corresponding to where the gray line crosses the colored lines.
+These plotting symbols have been used in panels B and D.
+Because the lines in A and C overlap, there is a risk that including plotting symbols for all of the methods would be impossible to read.
+In addition, we have re-labeled the y-axes as Stability and Quality and then defined those terms in the figure caption.
+
 > => In my opinion, Fig 2 presents the data in a non-intuitive way.
@@
 I would suggest to split this figure into two parts: part A providing more classical rarefaction curves per method (with sub-setted clustering points as overlay scatter) and part B providing the (log) relative OTU count between rarefaction and sub-setting across down-sampling steps.
+
+We have added the Swarm data to this plot.
+Although we see the reviewers point of presenting the absolute difference between the sub-sampled and rarefied datasets, we also wanted to present the difference in the inferred richness using the different methods.
+As the reviewer indicated, we could present traditional rarefaction plots, but this would result in a figure with 8 panels.
+We feel that this presentation does the best job of showing that there really is little difference between the subsampled and rarefied data for most methods and that there are large differences in the number of OTUs generated by the various methods.
+
 > => Figure 4 lacks a caption that is explanatory beyond the figure title.
+
+Similar to the previous comment, we have changed the Relative to full dataset / Relative to distances descriptions to Stability and Quality and defined these terms in the figure caption.
+We also expended the description of the plot.
 
@@
 
+We appreciate the reviewer's suggestions for additional papers to cite in our Introduction.
+Many of these have been added to the text as appropriate.
+As the reviewer can appreciate there have been many papers published in this area of varying quality.
+We did consider developing an overview figure or table to describe the various methods, but realized that the text was quickly becoming a review article.
+In addition, if this paper is read in conjunction with the original He study, much of the material is already covered in their study and we did not feel it was necessary to recover the same material.
+
+>>>>>>>>>>>>>>>>>>>>>
+hmmmmm
+>>>>>>>>>>>>>>>>>>>>>
+
 > - Related to this, the terms "abundance-based" and "distance-based greedy clustering" (AGC and DGC) are not properly introduced.
 Although the authors point out that they stick with the terminology of the He et al paper, these terms need to be explained somewhere in the manuscript, preferably in the Introduction.
+
+We have added text in the paragraph describing *de novo* approaches defining AGC and DGC.
 
@@
 
+>>>>>>>>>>>>>>>>>>>>>
+hmmmmm
+>>>>>>>>>>>>>>>>>>>>>
+
 > - The authors very clearly point out weaknesses in the criticized study by He et al; they support all their arguments with data and/or sound reasoning and do not, in my opinion, state any unfair or unsupported claims against the criticized work.
 Nevertheless, I personally felt that the wording is sometimes a bit harsh: e.g., p.9, l.190 ("this analysis was so poorly designed"); p.15, l.354 ("the entire design of the He study was artificial").
+
+Thank you for this comment, we have edited these sections to be less confrontational.
 
@@
 
+Thank you for pointing out these errors. We have corrected these and have gone back through the text to identify any additional problems.
+
 > - Several examples given throughout the text appear rather arbitrary.
 For example, p.5, l.59 ("may only be 94% similar", same in p.17, ll.402-3) or p.6, l.89 ("all sequences assigned to the Porphyromonadaceae would then be assigned to OTUs [...]") are out of context, the sudden mention of a concrete clustering threshold or an arbitrary bacterial family could confuse the reader more than illustrate the point being made.
+
+On the first point, the 94% similar follows from the previous sentence indicating that two sequences could be 97% similar to a reference.
+The 94% would be a worst case scenario where those two sequences are similar to the reference, but only 94% similar to each other (i.e.
+100 - 3 - 3 = 94).
+In the second case, this is described as an example of a hybrid approach.
+We have edited the text around these comments to hopefully make these points more clear.
 
@@
 
+This has been edited as suggested.
+
 > - When enumerating the pros and cons of closed ref clustering (p.5, ll.
 48f), several points in favor of this approach could be added: it is trivially parallelizable, has a low memory footprint and arguably gets better as reference databases improve (which is the case as more and more sequence data accumulates).
+
+This has been edited as suggested.
 
@@
 
+We have added a clause to the end of this paragraph indicating that our previously proposed hybrid approach still suffers from database quality and classification errors.
+
 > - p.6, ll.109ff. This paragraph explaining the MCC (in detail) feels a little out of place here; I suggest to relocate it to the (early) Results section.
+
+Thank you for the suggestion; we have moved this section to the paragraph in the Results & Discussion section describing the quality of the clustering methods.
 
@@
 In particular, I would like to commend the authors for uploading all code as reproducible workflow to Github.
+
+Thanks! We try to lead by example :).
 
@@
 
+I appreciate the reviewer's comments and can judge from their comments elsewhere i this review that they were a bit on the fence about whether to recommend we use the ARI/NMI/AMI in place of the MCC for the measurement of stability in the first part of the manuscript where we try to replicate the He paper’s work.
+Our initial inclination is to stick with the MCC approach, which was the method used by the original He study.
+If we were to do one of the other methods instead, then I feel like that would confuse the issue and would be a different type of replication.
+I feel that indicating that they perhaps picked the wrong metric would add unnecessarily to the figures and text when the point of the first section of the Results section is that their idea of “stability” is subservient to the quality of OTU assignments.
+Compounding this complexity would be the decision to pick one or all of the three choices to replace or supplement MCC for measuring stability.
+Furthermore, we feel that the consistency of that quality over multiple random seeds is much a better metric of “stability” than to compare the subsampled OTU assignments to the full dataset assignments.
+That being said, I feel like we’ve done a good job of addressing the quality/stability issue and the reviewers both seemed to agree.
+If the reviewer and editor feel that this would be an important issue to incorporate into the manuscript, then we’ll do it.
+At this point, it just seems like piling on when we’ve already made the point that their concept of “stability” is problematic.
+
+To the point of interpreting the MCC value, we have added a sentence to the end of the penultimate paragraph in the Introduction that provides an interpretation.
+
 > Validity of the findings
@@
 However, as likewise detailed above, I also believe that several findings stand independently of each other, and their connections are not immediately obvious.
+
+We hope that the inclusion of the paragraph to the Results & Discussion section outlining the questions that were raised by our replication of the He study provides a sufficient roadmap to help tie the paper together better.
 
@@
 I agree with the authors that accuracy is the more desirable property in general, but putting their study in context with other recent findings could suggest that in fact, accuracy and stability tend to be correlated, which would lend even more gravity to their argument.
+
+We agree that some metric of stability is important.
+We feel that by running the methods 30 times on the same dataset and essentially getting the same quality value (i.e.
+MCC) relative to the distances between sequences, is a superior metric of stability than comparing the subsampled OTU assignments to those from the full dataset.
 
@@
 
+Thank you for highlighting these three points.
+
 > - p.12, ll.260-1.
@@
 The SWARM algorithm is designed to not work with "hard" thresholds in the classical sense, so it is not surprising that in the given testing framework, different "thresholds" would appear optimal for differentially deeply sequenced datasets.
+
+We attempted to soften this language. Regardless, in our opinion, the inability to define a "hard" thredhold makes it difficult to compare results across studies.
 
@@
 
+We attempted to soften this language.
+
 > Comments for the author
@@
 I was very hesitant to include the points on the use of the MCC vs AMI/NMI/ARI in this review, but I honestly believe that they make sense scientifically.
+
+Thank you for your openness and thoroughness. Both reviews have gone a long way to improving the quality of the manuscript.

### 2015-11-13 Schloss_Cluster_PeerJ_2015.Rmd
Incorporate and respond to critique of R2

@@
 Furthermore, it is necessary to understand how the selected method affects the precision and accuracy of assigning sequences to OTUs.
+Broadly speaking, three approaches have been developed to assign sequences to OTUs:
 
-Broadly speaking, three approaches have been developed to assign sequences to OTUs.
 The first approach has been referred to as phylotyping [@Schloss2011] or closed reference clustering [@NavasMolina2013].
@@
 If a large fraction of sequences are novel, then they cannot be assigned to an OTU.
-In addition, the reference sequences are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene [@Schloss2011].
+In addition, the reference sequences are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene [@Schloss2010; @Kim2011].
 Thus, a sequence representing a fragment of the gene may be more than 97% similar to multiple reference sequences.
@@
 A alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@Schloss2011].
-The strength of the reference based approach is that the methods are generally fast, scaling linearly with the number of sequences being clustered and that OTU assignments can be compared across studies.
+The strengths of the reference-based methods include their speed, potential for trivial parallelization, ability to compare OTU assignments across studies, and the hope that as databases improve, the OTU assignments will also improve.
 
@@
 In this approach, the distance between sequences is used to cluster sequences into OTUs rather than the distance to a reference database.
-In contrast to the efficiency of closed-reference clustering, the computational cost of hierarchical *de novo* clustering methods scale quadratically with the number of unique sequences.
+In contrast to the efficiency of closed-reference clustering, the computational cost of hierarchical *de novo* clustering methods scales quadratically with the number of unique sequences.
 The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences.
@@
 As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@Edgar2010; @Sun2009; @Mah2014].
+Two related heuristics implemented in USEARCH were recently described: distance-based greedy clustering (DGC) and abundance-based greedy clustering (AGC) [@Edgar2010; @He2015].
+These greedy methods cluster sequences within a defined similarity threshold of an index sequence or creates a new index sequence.
+If a sequence is more similar than the defined threshold, it is assigned to the closest centroid based (i.e.
+DGC) or the most abundant centroid (i.e.
+AGC).
 One critique of *de novo* approaches is that OTU assignments are sensitive to the input order of the sequences [@He2015; @Mah2014].
@@
 In theory, this method should exploit the strengths of both closed-reference and *de novo* clustering; however, the different OTU definitions employed by commonly used closed-reference and *de novo* clustering implementations pose a possible problem when the methods are combined.
-An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels using the average linkage method [@Schloss2011].
+An alternative to this approach has been to classify sequences to a bacterial family or genus and then assign those sequences to OTUs within those taxonomic groups using the average linkage method [@Schloss2011].
 For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average linkage method using a 3% distance threshold.
@@
 An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group is seen as being independent and can be processed separately.
-Such an approach would overcome the difficulty of mixing OTU definitions between the closed-reference and *de novo* approaches.
+Such an approach would overcome the difficulty of mixing OTU definitions between the closed-reference and *de novo* approaches; however, it would still suffer from the problems associated with database quality and classification error.
 
@@
 We recently proposed an approach for evaluating OTU assignments using the distances between pairs of sequences [@Schloss2011].
-Those pairs of sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives.
-Meanwhile, those pairs of sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives.
-Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and negatives using the Matthew's correlation coefficient [MCC; @Matthews1975].
-This is an objective approach to assessing the quality of the OTU assignments.
+We were able to synthesize the relationship between OTU assignments and the distances between sequences using the Matthew's correlation coefficient [MCC; @Matthews1975].
+MCC is an objective approach to assessing the quality of the OTU assignments that can be interpreted as representing the correlation between the observed and expected classifications and can vary between -1.0 and 1.0.
 
@@
 
-***OTU assignment methods vary in their quality.*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs.
+This re-analysis of the He study raised five complementary questions.
+First, how do the various methods vary in the quality of their OTU assignments?
+Second, how generalizable are these results to modern datasets generated using a large number of sequences that were deeply sequenced?
+Third, how does the stability and quality of OTU assignments generated by new methods compare to those analyzed in the He study?
+Fourth, are there open-source alternatives to USEARCH that perform just as well?
+Finally, although the stability of reference-based methods did not appear to be impacted by the input order of the sequences to be assigned to OTUs, is the stability of reference-based methods impacted by the order of the reference sequences?
+In the remainder of the Results and Discussion section we address each of these questions.
+
+***How do the various methods vary in the quality of their OTU assignments?*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs.
 A method can generate highly stable OTUs, but the OTU assignments may be meaningless if they poorly represent the specified cutoff and the actual distance between the sequences.
 To assess the quality of OTU assignments by the various methods, we made use of the pairwise distance between the unique sequences to count the number of true positives and negatives and the number of false positives and negatives for each method and sampling depth.
-This enabled us to calculate the average MCC value as a measure of a method's quality and its variation as a measure of its consistency.
+Those pairs of sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives.
+Meanwhile, those pairs of sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives.
+Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and negatives using the MCC.
+We used the average MCC value as a measure of a method's quality and its variation as a measure of its consistency.
 We made three important observations.
@@
 
-***Deep sampling of 16S rRNA genes.*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering methods.
+***How generalizable are these results to modern datasets generated using a large number of sequences that were deeply sequenced?*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering methods.
 First, it was one of the earliest 16S rRNA gene sequence datasets published using the 454 FLX platform.
@@
 
-***Evaluation of Swarm as an alternative de novo clustering algorithm.*** The Swarm algorithm is a recently proposed *de novo* method for assigning sequences to OTUs that uses user-defined parameters to break up chains generated by SL clustering [@Mah2014].
+***How does the stability and quality of OTU assignments generated by new methods compare to those analyzed in the He study?*** The Swarm algorithm is a recently proposed *de novo* method for assigning sequences to OTUs that uses user-defined parameters to break up chains generated by SL clustering [@Mah2014].
 Swarm was originally validated by comparing the results against the expected clusters formed based on the taxonomic composition of a mock community.
@@
 Interestingly, the distance thresholds that resulted in the largest MCC values were 3 and 2% for the Canadian soil and murine datasets, respectively.
-This suggests that OTU definitions are not consistent across datasets when using the Swarm algorithm.
+This suggests that distance-based OTU definitions are not consistent across datasets when using the Swarm algorithm, although they do appear to be within the neighborhood of 3%.
 Finally, the Swarm developers indicated that hierarchical *de novo* algorithms were too impractical to use on large MiSeq-generated datasets.
@@
 
-***Evaluation of an open-source alternative to USEARCH.*** For some datasets the AGC and DGC methods appear to perform as well or better than the hierarchical clustering methods.
+***Are there open-source alternatives to USEARCH that perform just as well?*** For some datasets the AGC and DGC methods appear to perform as well or better than the hierarchical clustering methods.
 As originally described in the He study, the AGC and DGC methods utilized the USEARCH program and the DGC method is used for clustering in UPARSE [@Edgar2010; @Edgar2013].
@@
 
-***Problems with reference-based clustering in general and as implemented in QIIME.*** The He study and our replication attempt validated that the closed-reference method generated perfectly stable OTUs.
+***Is the stability of reference-based methods impacted by the order of the reference sequences?*** The He study and our replication attempt validated that the closed-reference method generated perfectly stable OTUs.
 This was unsurprising since, by definition, the method is designed to return one-to-one mapping of reads to a reference.
@@
 It is impossible to obtain a clustering with no false positives or false negatives and the optimal method may vary by dataset.
-With this in mind, researchers are encouraged to calculate and report their MCC values for the AL method and at least one other method.
+With this in mind, researchers are encouraged to calculate and report their MCC values and to use these values to justify using methods other than the AL.
 
@@
 In the closed-reference step, sequences must be within a threshold of a reference to be in the same OTU.
-This means that two sequences that are 97% similar to a reference and are joined into the same OTU, may only be 94% similar to each other.
+This means that in the worst case scenario two sequences that are 97% similar to a reference and are joined into the same OTU, may only be 94% similar to each other.
 In the DGC step, the goal is to approximate the AL method which requires that, on average, the sequences within an OTU are, on average, 97% similar to each other.
@@
 **Figure 1.
-Comparison of the stability (A, B) and quality (C, D) of *de novo* and reference-based clustering methods using the Canadian soil dataset.** The average stability of the OTUs were determined by calculating the MCC with respect to the OTU assignments for the full dataset using varying sized subsamples (A).
+Comparison of the stability (A, B) and quality (C, D) of *de novo* and reference-based clustering methods using the Canadian soil dataset.** The average stability of the OTUs was determined by calculating the MCC with respect to the OTU assignments for the full dataset using varying sized subsamples.
+The quality of the OTUs was determined by calculating the MCC with respect to the distances between the sequences using varying sized subsamples.
 Thirty randomizations were performed for each fraction of the dataset and the average and 95% confidence interval are presented when using 60% of the data.
-The quality of the OTUs were determined by calculating the MCC with respect to the distances between the sequences using varying sized subsamples (C).
-Thirty randomizations were performed for each fraction of the dataset and the average and 95% confidence interval are presented when using the full dataset (D).
-The vertical gray line indicates in A and C indicates the fraction of the dataset represented in B and D, respectively.
+The vertical gray lines in A and C indicates the fraction of the dataset represented in B and D, respectively.
+The color and shape of the plotting symbol is the same between the different panels and is described along the x-axis of panel D.
 The optimum threshold for the Swarm-generated assignments was 3%.
@@
 **Figure 3.
-Comparison of the stability (A, B) and quality (C, D) of *de novo* and reference-based clustering methods using the murine dataset.** The average stability of the OTUs were determined by calculating the MCC with respect to the OTU assignments for the full dataset using varying sized subsamples (A).
+Comparison of the stability (A, B) and quality (C, D) of *de novo* and reference-based clustering methods using the murine dataset.** The average stability of the OTUs was determined by calculating the MCC with respect to the OTU assignments for the full dataset using varying sized subsamples.
+The quality of the OTUs was determined by calculating the MCC with respect to the distances between the sequences using varying sized subsamples.
 Thirty randomizations were performed for each fraction of the dataset and the average and 95% confidence interval are presented when using 60% of the data.
-The quality of the OTUs were determined by calculating the MCC with respect to the distances between the sequences using varying sized subsamples (C).
-Thirty randomizations were performed for each fraction of the dataset and the average and 95% confidence interval are presented when using the full dataset (D).
-The vertical gray line indicates in A and C indicates the fraction of the dataset represented in B and D, respectively.
+The vertical gray lines in A and C indicates the fraction of the dataset represented in B and D, respectively.
+The color and shape of the plotting symbol is the same between the different panels and is described along the x-axis of panel D.
 The optimum threshold for the Swarm-generated assignments was 2%.
 
-**Figure 4. The VSEARCH OTUs generated by the AGC and DGC methods were comparable to those generated using USEARCH.**
+**Figure 4.
+The stability and quality of USEARCH and VSEARCH OTUs generated by the AGC and DGC methods were similar.**  The stability of the OTUs was determined by calculating the MCC for OTUs calculated using 60% of the data relative to the OTU assignments for the full dataset.
+The quality of the OTUs was determined by calculating the MCC of the OTUs calculated using the full dataset with respect to the distances between the sequences.
+The error bars represent the 95% confidence interval across the 30 randomizations.
 

### 2015-11-13 code/build_mcc_plot.R
Altered plots to resond to reviewers

@@ -4,9 +4,8 @@ build_mcc_plots <- function(dataset, output_file_name){
 	n_methods <- length(methods)
 	clrs <- c("brown", "red", "orange", rainbow(n_methods-1)[4:7], "black")
 
-
-
-
+	my_pch <- c(15, 21, 16, 22, 17, 23, 18, 24)
+	my_cex <-  c(2,2,2,2,2,2,3,2)
 	data <- read.table(file=paste0("data/process/", dataset, ".mcc_ref.summary"), header=T, stringsAsFactors=F)
 
 	orig_methods <- data[data$method %in% methods,]
@@ -21,7 +20,7 @@ build_mcc_plots <- function(dataset, output_file_name){
 
 	par(mar=c(0,0,0,0))
 	plot.new()
-	text(x=0.3, y=0.5, label="MCC value relative to\nfull dataset", cex=1.5, srt=90, xpd=TRUE)
+	text(x=0.3, y=0.5, label="Stability", cex=1.5, srt=90, xpd=TRUE)
 
 	### A ###
 	par(mar=c(0.5, 0.5, 0.5, 0.5))
@@ -31,7 +30,8 @@ build_mcc_plots <- function(dataset, output_file_name){
 
 	for(m in 1:n_methods){
 		subset <- orig_methods[orig_methods$method==methods[m],]
-		points(subset$mean~subset$fraction, type="l", col=clrs[m], lwd=3)
+		points(subset$mean~subset$fraction, type="l", col=clrs[m], lwd=2)
+		points(0.2, subset$mean[1], pch=my_pch[m], col=clrs[m], cex=my_cex[m], bg="white")
 	}
 	axis(1, at=seq(0.2,1.0,0.2), labels=rep("", 5))
 	axis(2, las=1, cex.axis=1.5)
@@ -50,10 +50,11 @@ build_mcc_plots <- function(dataset, output_file_name){
 	plot(NA, ylim=c(0.0,1), xlim=c(0.75,n_methods+0.25), axes=F, ylab="", xlab="")
 
 	for(m in 1:n_methods){
-		points(x=m, y=sixty[sixty$method==methods[m],"mean"], pch=19, col=clrs[m], cex=2.0)
 
[…]

### 2015-11-13 code/build_figure2.R
Added Swarm data per reviewer's request

@@ -1,11 +1,12 @@
 build_figure2 <- function(dataset, output_file_name, position='bottomright'){
+
 	rare <- read.table(file=paste0("data/process/", dataset, ".rarefaction.summary"), header=T, stringsAsFactors=T)
 
 	offset <- 0.1
 
 	n_fractions <- length(levels(factor(rare$fraction)))
 
-	methods <- c("fn", "nn", "an", "dgc", "agc", "closed", "open")
+	methods <- c("fn", "nn", "an", "dgc", "agc", "closed", "open", "swarm")
 
 	rare_methods <- rare[rare$method %in% methods,]
 
@@ -34,7 +35,7 @@ build_figure2 <- function(dataset, output_file_name, position='bottomright'){
 		points(rep(m+offset, n_fractions), subset[,"rarefied"], col=clrs, pch=21, cex=1.0)
 	}
 
-	method_labels <- c("CL", "SL", "AL", "DGC", "AGC", "Closed-ref", "Open-ref")
+	method_labels <- c("CL", "SL", "AL", "DGC", "AGC", "Closed-ref", "Open-ref", "Swarm")
 	axis(1, at=1:n_methods, label=method_labels, las=2)
 	axis(2, las=2, at=seq(0,6000,1000), label=c("0", "1,000", "2,000", "3,000", "4,000", "5,000", "6,000"))
 	box()
@@ -47,6 +48,6 @@ build_figure2 <- function(dataset, output_file_name, position='bottomright'){
 	    y=rep(pos$text$y, times=2),
 	    pch=rep(c(19,21), times=2), col=clrs, cex=0.7)
 
-	dev.off()
 	par <- orig_par
+	dev.off()
 }

### 2015-11-13 code/build_figure4.R
Change relative to to stability/quality closing issue #5

@@ -39,7 +39,7 @@ for(i in 1:length(methods)){
 	points(i+6, row$mean, pch=21, bg="white")
 }
 
-legend(x=7.5, y=0.4, legend=c("Relative to full dataset", "Relative to distances"), pch=c(21,19), pt.cex=0.8, pt.bg=c("white", "black"), cex=0.6, bty="n")
+legend(x=7.5, y=0.4, legend=c("Stability", "Quality"), pch=c(21,19), pt.cex=1.0, pt.bg=c("white", "black"), cex=0.8, bty="n")
 axis(1, at=c(1:5, 7:11), label=c("AL", "U", "V", "U", "V", "AL", "U", "V", "U", "V"), cex.axis=0.9)
 mtext(side=1, line=2, at=2.5, "AGC")
 mtext(side=1, line=2, at=4.5, "DGC")

### 2015-11-13 Schloss_Cluster_PeerJ_2015.Rmd
Add Figure 5

@@
 The default database used in QIIME, which was also used in the He study, contains full-length sequences that are at most 97% similar to each other.
-We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference method with the full murine dataset, which contained `r n_unique_miseq_seqs` unique sequences.
+We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference method with the full murine dataset, which contained `r n_unique_miseq_seqs` unique sequences (Figure 5).
 Surprisingly, we observed that the number of OTUs generated was not the same in each of the randomizations.
@@
 
+**Figure 5.
+The number of closed-reference OTUs observed in the murine dataset when using USEARCH, VSEARCH, and without a heuristic.** In addition to the default ordering of the references provided with the QIIME package, the reference sequences were randomized 30 times; the order of the murine dataset was not randomized.
+Regardless of whether the default or randomized ordering was used, the number of OTUs generated using VSEARCH did not differ.
+The non-heuristic approach calculated the exact distance between the murine sequences and the the reference sequences and assigned the sequences to the reference with the smallest distance.
+
 \newpage
@@
 \includepdf{results/figures/figure_4.pdf}
+\includepdf{results/figures/figure_5.pdf}
 

### 2015-11-13 code/build_figure5.R

@@ -1,0 +1,36 @@
+usearch_counts <- read.table(file="data/rand_ref/hits.uclosed.counts")
+usearch_original <- usearch_counts[nrow(usearch_counts), 2]
+usearch_counts <-  usearch_counts[-nrow(usearch_counts), 2]
+
+vsearch_counts <- read.table(file="data/rand_ref/hits.vclosed.counts")
+vsearch_original <- vsearch_counts[nrow(vsearch_counts), 2]
+vsearch_counts <-  vsearch_counts[-nrow(vsearch_counts), 2]
+stopifnot(sd(vsearch_counts) == 0, mean(vsearch_counts) == vsearch_original)
+
+
+expected_counts <- nrow(read.table(file="data/rand_ref/closed_ref.redundancy.analysis", header=T))
+
+
+pdf(file="results/figures/figure_5.pdf", width=4, height=4)
+par(mar=c(4,4,0.5,0.5), cex=0.8)
+
+hist(usearch_counts, xlim=c(27600, 28350), ylim=c(0,10), breaks=4, main="", xlab="", axes=F, col="gray")
+arrows(x0=usearch_original, y0=2, y1=0, length=0.10)
+text(x=usearch_original, y=3.0, label="Number of hits\nusing USEARCH\nwith default\nordering", cex=0.8)
+
+arrows(x0=vsearch_original, y0=1, y1=0, length=0.10)
+text(x=vsearch_original, y=1.5, label="Number of hits\nusing VSEARCH", cex=0.8)
+
+arrows(x0=expected_counts, y0=0.75, y1=0, length=0.10)
+text(x=expected_counts, y=1.5, label="Expected\nnumber\nof hits", cex=0.8)
+
+
+text(x=median(usearch_counts), y=8.9, label="Number of hits\nusing USEARCH\nwith randomized\nordering", cex=0.8)
+
+
+polygon(x=c(0, 0, 28350, 28350),y=c(0, 10, 10, 0))
+title(xlab="Number of Sequences that\nMapped to the Reference", line=3)
+axis(1, at=seq(27600,28300,100), label=format(seq(27600,28300,100), big.mark=','), pos=0)
+axis(2, las=2)
+
+dev.off()

### 2015-11-13 rebuttal.md
Rewrite paragraphon methods of assessing quality

@@
 
-> *Terminology: Over the course of the paper, (at least) five different terms are used to refer to the goodness of a clustering: stability, rigor, robustness, quality and accuracy.
+> \*Terminology: Over the course of the paper, (at least) five different terms are used to refer to the goodness of a clustering: stability, rigor, robustness, quality and accuracy.
 Yet only two distinct concepts are explored (stability and quality/accuracy).
@@
 
-> *L20-L21: USEARCH -> USEARCH and VSEARCH.
+> \*L20-L21: USEARCH -> USEARCH and VSEARCH.
 According to your results, both methods produced OTU assignments that depended on the ordering of the reference database (~55-60% label consistency between reorderings of the database).
@@
 
-> *L64: The other major advantage of reference based methods is that the OTU assignments can be compared across experiments and even studies. This should be mentioned here.
+> \*L64: The other major advantage of reference based methods is that the OTU assignments can be compared across experiments and even studies. This should be mentioned here.
 
@@
 
-> *L95-105: Evaluation of accuracy on mock communities needs to be included in this discussion of previous methods to evaluate OTU methods.
+> \*L95-105: Evaluation of accuracy on mock communities needs to be included in this discussion of previous methods to evaluate OTU methods.
 
@@
 
-> *L182-190: Comparing OTUs on the basis of their representative sequence is not "specious" and the He analysis was not "so poorly designed".
+> \*L182-190: Comparing OTUs on the basis of their representative sequence is not "specious" and the He analysis was not "so poorly designed".
 The most abundant sequence in an OTU is a biologically meaningful object, and we would like to think that OTUs are biologically meaningful and not just abstract objects.
@@
 
-We have added a plotting symbol to the lines in panels A and C corresponding to where the gray line crosses the colored lines.
+We have added a plotting symbol to the lines in panels A and C.
 These plotting symbols have been used in panels B and D.
-Because the lines in A and C overlap, there is a risk that including plotting symbols for all of the methods would be impossible to read.
+Because the lines in A and C overlap, there is a risk that including plotting symbols for all of the fractions would be impossible to read.
+To make lines easier to differentiate in A & C we have made them thinner.
 In addition, we have re-labeled the y-axes as Stability and Quality and then defined those terms in the figure caption.
@@
 
->>>>>>>>>>>>>>>>>>>>>
-hmmmmm
->>>>>>>>>>>>>>>>>>>>>
-
 > - Related to this, the terms "abundance-based" and "distance-based greedy clustering" (AGC and DGC) are not properly introduced.
@@
 
->>>>>>>>>>>>>>>>>>>>>
-hmmmmm
->>>>>>>>>>>>>>>>>>>>>
+We thank the reviewer for this recommendation.
+We have added a new figure showing the distribution of OTU numbers when the databases are randomized for both USEARCH and VSEARCH.
+We also show the expected number of OTUs based on calculating the actual distances between the murine dataset sequences and the references.
 
@@
 At this point, it just seems like piling on when we’ve already made the point that their concept of “stability” is problematic.
+Finally, we have added a sentence to the section of the Results section where we attempt to replicate the He study indicating that information-based methods have previously been used and are likely better, but that we chose to stick with the original approach.
 

### 2015-11-13 Schloss_Cluster_PeerJ_2015.Rmd
Rewrite paragraphon methods of assessing quality

@@
 To simplify the complexity of these large datasets, it is helpful to cluster sequences into meaningful bins.
-These bins, commonly known as operational taxonomic units (OTUs), are  used to compare the biodiversity contained within and between different samples [@Schloss2011].
+These bins, commonly known as operational taxonomic units (OTUs), are  used to compare the biodiversity contained within and between different samples [@Schloss2011Assessing].
 Such comparisons have enabled researchers to characterize the microbiota associated with the human body [e.g.
@@
 
-The first approach has been referred to as phylotyping [@Schloss2011] or closed reference clustering [@NavasMolina2013].
+The first approach has been referred to as phylotyping [@Schloss2011Assessing] or closed reference clustering [@NavasMolina2013].
 This approach involves comparing sequences to a curated database and then clustering sequences into the same OTU that are similar to the same reference sequence.
@@
 Alternatively, a sequence may be equally similar to two or more reference sequences.
-A alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@Schloss2011].
+A alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@Schloss2011Assessing].
 The strengths of the reference-based methods include their speed, potential for trivial parallelization, ability to compare OTU assignments across studies, and the hope that as databases improve, the OTU assignments will also improve.
 
-The second approach has been referred to as distance-based [@Schloss2011] or *de novo* clustering [@NavasMolina2013].
+The second approach has been referred to as distance-based [@Schloss2011Assessing] or *de novo* clustering [@NavasMolina2013].
 In this approach, the distance between sequences is used to cluster sequences into OTUs rather than the distance to a reference database.
@@
 For this reason, *de novo* clustering has been preferred across the field.
-After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU [@Schloss2011].
+After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU [@Schloss2011Assessing].
 
@@
 In theory, this method should exploit the strengths of both closed-reference and *de novo* clustering; however, the different OTU definitions employed by commonly used closed-reference and *de novo* clustering implementations pose a possible problem when the methods are combined.
-An alternative to this approach has been to classify sequences to a bacterial family or genus and then assign those sequences to OTUs within those taxonomic groups using the average linkage method [@Schloss2011].
+An alternative to this approach has been to classify sequences to a bacterial family or genus and then assign those sequences to OTUs within those taxonomic groups using the average linkage method [@Schloss2011Assessing].
 For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average linkage method using a 3% distance threshold.
@@
 These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the OTU assignments.
-Others have attempted to judge the quality of a method by its ability to generate data that parallels classification data [@Cai2011; @Sun2011; @Chen2013].
+Others have attempted to judge the quality of a method by its ability to generate data that parallels classification data [@Cai2011; @Sun2011; @White2010].
 This approach is problematic because bacterial taxonomy often reflects historical biases amongst bacterial systematicists.
 Furthermore, it is well known that the rates of evolution across lineages are not the same [@Wang2007; @Schloss2010].
-A related approach has used clustering of mock community data to evaluate methods [@Edgar2013; @Mah2014].
-Yet these approaches ignore the effects of sequencing errors that tend to accumulate with sequencing depth and represent highly idealized communities [@Kozich2013; @Schloss2011Reducing].
-Others have assessed the quality of clustering based on their ability to generate the same OTUs generated by other methods [@Schmidt2014; @Rideout2014].
+A related approach has used clustering of mock community data to evaluate methods [@Edgar2013; @Mah2014; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014].
+Yet these approaches ignore the effects of sequencing errors that tend to accumulate with sequencing depth and represent highly idealized communities that lack the phylogenetic diversity of real microbial communities [@Kozich2013; @Schloss2011Reducing].
+Others have assessed the quality of clustering based on their ability to generate the same OTUs generated by other methods [@Schmidt2014Limits; @Rideout2014].
 This is problematic because it does not solve the fundamental question of which method is most correct.
-We recently proposed an approach for evaluating OTU assignments using the distances between pairs of sequences [@Schloss2011].
+The concept of ecological consistency as a metric of quality asserts that sequences that cluster into the same OTU should share similar ecological affiliations [@Koeppel2013; @Schmidt2014Ecological; @Preheim2013].
+Although this is an intriguing approach and is a metric that can be quantified, it is unclear how the metric would be validated.
+We recently proposed an approach for evaluating OTU assignments using the distances between pairs of sequences [@Schloss2011Assessing].
 We were able to synthesize the relationship between OTU assignments and the distances between sequences using the Matthew's correlation coefficient [MCC; @Matthews1975].
-MCC is an objective approach to assessing the quality of the OTU assignments that can be interpreted as representing the correlation between the observed and expected classifications and can vary between -1.0 and 1.0.
+MCC is can be interpreted as representing the correlation between the observed and expected classifications and can vary between -1.0 and 1.0.
+The strength of the MCC, as implemented by Schloss et al.
+[-@Schloss2011Assessing], is that it is an objective approach to assessing the quality of the OTU assignments that can be calculated for any set of OTU assignments where there is a distance matrix and a specific threshold without relying on an external reference.
 
@@
 First, we sought to quantify whether the OTU assignments observed for a subset of the data represented the same assignments that were found with the full dataset.
+The He study used the MCC to quantify the degree to which pairs of sequences were in the same OTUs in subsampled and full datasets.
+A more robust alternative, which used metrics that quantify the mutual information held between two sets of clusterings, has been applied to assess inter-method variation [@Schmidt2014Limits].
+To maintain consistency with the original He study, we also calculated the MCC value as they described.
 The He study found that when they used the open and closed-reference methods the OTUs formed using the subsetted data most closely resembled those of the full dataset.
@@
 Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene [@Schloss2010].
-With these points in mind, we compared the clustering assignment for each of these methods using a time series experiment that was obtained using mouse feces [@Schloss2012; @Kozich2013].
+With these points in mind, we compared the clustering assignment for each of these methods using a time series experiment that was obtained using mouse feces [@Schloss2012Stabilization; @Kozich2013].
 The MiSeq platform was used to generate `r n_miseq_seqs` sequences from the V4 region of the 16S rRNA gene of `r n_groups` samples.
@@
 Rarely would a researcher rarefy their data prior to clustering since it can be more efficiently done after all of the data are assigned to OTUs.
-Third, the CL method used in the original Roesch study has since been shown to not generate optimal OTUs [@Schloss2011].
+Third, the CL method used in the original Roesch study has since been shown to not generate optimal OTUs [@Schloss2011Assessing].
 As for the approach used in the He study, the value of identifying stable OTUs is unclear.
@@
 With this in mind, researchers are encouraged to calculate and report their MCC values and to use these values to justify using methods other than the AL.
+As an alternative to the He study's method of measuring stability, we propose using the variation in the quality of the clustering of the full dataset.
+Given the tight 95% confidence intervals shown in Figures 1D and 3D, with the exception of CL, it is clear that this variation is quite small.
+This indicates that although the order of the sequences being clustered can affect the actual cluster assignments, the quality of those different clusterings is not meaningfully different.
 
@@
 
-Second, we calculated the MCC to determine the quality of the clusterings as previously described [@Schloss2011].
+Second, we calculated the MCC to determine the quality of the clusterings as previously described [@Schloss2011Assessing].
 Briefly, we compared the OTU assignments for pairs of sequences to the distance matrix that was calculated between all pairs of aligned sequences.

### 2015-11-13 ResponseToReviewers.md
Rename and have output be PDF

--- 2015-11-13 rebuttal.md
@@
+---
+title: "***Response to reviewers**"
+output:
+  pdf_document:
+    includes:
+      in_header: header.tex
+csl: peerj.csl
+fontsize: 11pt
+geometry: margin=1.0in
+---
+
 > Reviewer 1 (Anonymous)
@@
 
-We attempted to soften this language. Regardless, in our opinion, the inability to define a "hard" thredhold makes it difficult to compare results across studies.
+We attempted to soften this language. Regardless, in our opinion, the inability to define a "hard" threshold makes it difficult to compare results across studies.
 

### 2015-11-13 Schloss_Cluster_PeerJ_2015.Rmd
Editing

@@
 
-**Methods.** Our analysis implemented five *de novo* clustering algorithms including the single linkage, complete linkage, average linkage, abundance-based greedy clustering, distance-based greedy clustering, and the open and closed-reference methods.
+**Methods.** Our analysis implemented six *de novo* clustering algorithms including the single linkage, complete linkage, average linkage, abundance-based greedy clustering, distance-based greedy clustering, and Swarm and the open and closed-reference methods.
 Using two previously published datasets we used the Matthew’s Correlation Coefficient (MCC) to assess the stability and quality of OTU assignments.
@@
 An alternative to this approach has been to classify sequences to a bacterial family or genus and then assign those sequences to OTUs within those taxonomic groups using the average linkage method [@Schloss2011Assessing].
-For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average linkage method using a 3% distance threshold.
+For example, all sequences classified as belonging to the *Porphyromonadaceae* would then be assigned to OTUs using the average linkage method using a 3% distance threshold.
 Those sequences that did not classify to a known family would also be clustered using the average linkage method.
@@
 Others have assessed the quality of clustering based on their ability to generate the same OTUs generated by other methods [@Schmidt2014Limits; @Rideout2014].
-This is problematic because it does not solve the fundamental question of which method is most correct.
+This is problematic because it does not solve the fundamental question of which method is optimal.
 The concept of ecological consistency as a metric of quality asserts that sequences that cluster into the same OTU should share similar ecological affiliations [@Koeppel2013; @Schmidt2014Ecological; @Preheim2013].
-Although this is an intriguing approach and is a metric that can be quantified, it is unclear how the metric would be validated.
+Although this is an intriguing approach and is a quantitative metric, it is unclear how the metric would be objectively validated.
 We recently proposed an approach for evaluating OTU assignments using the distances between pairs of sequences [@Schloss2011Assessing].
@@
 They defined stability as the ability of a method to provide the same clustering on a subset of the data as was found in the full dataset.
-Their concept of stability did not account for the accuracy of the OTU assignments and instead focused on the precision of the assignments.
-A method may be very precise, but low in accuracy.
-In the current analysis, we assessed the accuracy and precision of the various clustering methods.
-Building on our previous analysis of clustering methods, our hypothesis was that the methods praised by the He study for their stability actually suffered a lack of accuracy.
+Their concept of stability did not account for the quality of the OTU assignments and instead focused on the precision of the assignments.
+A method may be very stable, but of poor quality.
+In the current analysis, we assessed the quality and stability of the various clustering methods.
+Building on our previous analysis of clustering methods, our hypothesis was that the methods praised by the He study for their stability actually suffered a lack of quality.
 In addition, we assess these parameters in light of sequence quality using the original 454 dataset and a larger and more modern dataset generated using the MiSeq platform.
@@
 The He study used the MCC to quantify the degree to which pairs of sequences were in the same OTUs in subsampled and full datasets.
-A more robust alternative, which used metrics that quantify the mutual information held between two sets of clusterings, has been applied to assess inter-method variation [@Schmidt2014Limits].
+A more robust approach would utilize metrics that quantify the mutual information held between two sets of clusterings and has been applied to assess inter-method variation in OTU composition [@Schmidt2014Limits].
 To maintain consistency with the original He study, we also calculated the MCC value as they described.
 The He study found that when they used the open and closed-reference methods the OTUs formed using the subsetted data most closely resembled those of the full dataset.
-Among the *de novo* methods they observed that the abundance-based greedy clustering (AGC) method generated the most stable OTUs followed by the single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) methods.
+Among the *de novo* methods they observed that the AGC method generated the most stable OTUs followed by the single linkage (SL), DGC, complete linkage (CL), and average linkage (AL) methods.
 We first calculated the MCC for the OTU assignments generated by each of the clustering methods using 20, 40, 60, and 80% of the sequences relative to the OTU composition formed by the methods using the full dataset (see Methods for description; Figure 1A).
@@
 
-Second, the He study and the original Roesch study showed that rarefaction curves calculated using CL-generated OTU assignments obtained using a portion of the dataset did not overlap with rarefaction curves generated using OTU assignments generated from the full dataset.
+Second, the He study and the original Roesch study showed that rarefaction curves calculated using CL-generated OTU assignments obtained using a subsample of the sequencing data did not overlap with rarefaction curves generated using OTU assignments generated from the full dataset.
 The He and Roesch studies both found that the CL method produced fewer OTUs in the subset than in the rarefied data.
@@
 To overcome this, the authors selected representative sequences from each OTU and used those representative sequences to link OTU assignments between the different sized sequence sets.
-It is not surprising that the only analysis that did not provide a significant p-value was for the closed-reference analysis, which is the only analysis that provides consistent OTU labels.
+It was not surprising that the only analysis that did not provide a significant p-value was for the closed-reference analysis, which is the only analysis that provides consistent OTU labels.
 Finally, the authors built off of this analysis to count the number of OTUs that were differentially represented between the subsetted and full datasets by each method.
@@
 Third, with the possible exception of the CL method, the MCC values for each of the methods only demonstrated a small amount of variation between runs of the method with a different ordering of the input sequences.
-This indicates that although there may be variation between executions of the same method, they produce OTU assignments that are equally good.
+This indicates that although there may be variation between executions of the same method, they produced OTU assignments that were of equal quality.
 Revisiting the concept of stability, we question the value of obtaining stable OTUs when the full dataset is not optimally assigned to OTUs.
@@
 In addition, because the original Phred quality scores and flowgram data are not available, it was not possible for us to adequately remove sequencing errors [@Schloss2011Reducing].
-The large number of sequences that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering methods.
+The large number of sequencing errors that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering methods.
 Second, the dataset used in the He study covered the V9 region of the 16S rRNA gene.
@@
 Finally, as was observed with the Canadian soil dataset, there was little variation in MCC values observed among the 30 randomizations.
-Therefore, although the methods have a stochastic component, the OTU assignments do not vary meaningfully between runs.
+Therefore, although the methods have a stochastic component, the OTU assignments did not vary meaningfully between runs.
 The results from both the Canadian soil and murine microbiota datasets demonstrate that the *de novo* methods can generate stable OTU assignments and that the the overall quality of the assignments were consistent.
@@
 Finally, the Swarm developers indicated that hierarchical *de novo* algorithms were too impractical to use on large MiSeq-generated datasets.
-Our ability to apply AL to the large mouse dataset and even larger datasets suggests [e.g.
-@Schubert2015; @Zackular2015] that it is not necessary to sacrifice OTU assignment quality for speed.
+Our ability to apply AL to the large mouse dataset and even larger datasets suggests that it is not necessary to sacrifice OTU assignment quality for speed [e.g.
+@Schubert2015; @Zackular2015].
 
@@
 We also observed that regardless of whether we used USEARCH or VSEARCH, the reference OTU labels that were assigned to each OTU differed between randomizations.
-When we used USEARCH to perform closed-reference clustering, an average of `r round(100*u_overlap["orig", "mean"], 2)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*u_overlap["orig", "min"], 2)` to `r round(100*u_overlap["orig", "max"], 2)`).
-If we instead used VSEARCH an average of `r round(100*v_overlap["orig", "mean"], 2)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*v_overlap["orig", "min"], 2)` to `r round(100*v_overlap["orig", "max"], 2)`).
+When we used USEARCH to perform closed-reference clustering, an average of `r round(100*u_overlap["orig", "mean"], 2)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*u_overlap["orig", "min"], 2)` to `r round(100*u_overlap["orig", "max"], 2)`%).
+If we instead used VSEARCH an average of `r round(100*v_overlap["orig", "mean"], 2)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*v_overlap["orig", "min"], 2)` to `r round(100*v_overlap["orig", "max"], 2)`%).
 To better understand this result, we further analyzed QIIME's reference database.
@@
 Among these dereplicated V4 sequences there were `r n_small_v4_dist` pairs of sequences that were more than 97% similar to each other.
-The presence of duplicate V4 reference sequences explains the lack of labeling stability when using either USEARCH or VSEARCH to carry out the closed-reference method.
+The presence of duplicate and highly similar V4 reference sequences explains the lack of labeling stability when using either USEARCH or VSEARCH to carry out the closed-reference method.
 We suspect that the reference database was designed to only include sequences that were at most 97% similar to each other as a way to overcome the limitations of the USEARCH search heuristic.
@@
 
-Beyond comparing the abundance of specific OTUs across samples, the reference database is used in the open and closed-reference methods to generate OTU labels that are used in several downstream applications.
+Beyond comparing the abundance of specific OTUs across samples, the reference database is used in the open and closed-reference methods to generate OTU labels that can be used in several downstream applications.
 It is commonly used to extract information from a reference phylogenetic tree to carrying out UniFrac-based analyses [@Hamady2009] and to identify reference genomes for performing analyses such as PICRUSt [@Langille2013].
-Because these downstream applications depend on the correct and unique labeling of the OTUs, the lack of stability of the labeling is problematic.
+Because these downstream applications depend on the correct and unique labeling of the OTUs, the lack of label stability is problematic.
 As one illustration of the effects that incorrect labels would have on an analysis, we asked whether the duplicate sequences had the same taxonomies.

### 2015-11-13 ResponseToReviewers.md
Change formatting to differentiate review and response

@@
 ---
-title: "***Response to reviewers**"
+title: "**Response to reviewers**"
 output:
@@
 
-> Reviewer 1 (Anonymous)
+**Reviewer 1 (Anonymous)**
 
-> Basic reporting
+**Basic reporting**
 
-> Code to reproduce the results has been provided in a github repository.
+**Code to reproduce the results has been provided in a github repository.**
 
-> See general comments.
+**See general comments.**
 
-> Experimental design
+**Experimental design**
 
-> See general comments.
+**See general comments.**
 
-> Validity of the findings
+**Validity of the findings**
 
-> See general comments.
+**See general comments.**
 
-> Comments for the author
+**Comments for the author**
 
-> This paper evaluates the "stability" and "quality" of several different OTU clustering methods.
+**This paper evaluates the "stability" and "quality" of several different OTU clustering methods.
 "Stability" is defined here as the consistency between OTU assignments made on subsets of the data and the full dataset.
@@
 The authors broad conclusions are that while reference based clustering is more "stable", that hierarchical de novo methods, particularly average linkage, produce OTUs of higher "quality".
-The authors also make an important observation that the OTU assignments of commonly used reference methods depend on the ordering of the reference database.
+The authors also make an important observation that the OTU assignments of commonly used reference methods depend on the ordering of the reference database.**
 
-> This paper is largely a response to He et al.
-"Stability of operational taxonomic units: an important but neglected property for analyzing microbial diversity" which argued for reference based methods based on the "stability" of their OTU assignments.
+**This paper is largely a response to He et al.
+"Stability of operational taxonomic units: an important but neglected property for analyzing microbial diversity" which argued for reference based methods based on the "stability" of their OTU assignments.**
 
-> -------
+**-------**
 
-> This paper is well-written, makes several useful observations, and is clearly suitable for publication.
+**This paper is well-written, makes several useful observations, and is clearly suitable for publication.
 It can be improved by certain clarifications.
-I recommend its acceptance with minor revisions, as detailed below.
+I recommend its acceptance with minor revisions, as detailed below.**
 
-> More important points are preceded by an asterisk.
+**More important points are preceded by an asterisk.**
 
-> -------
+**-------**
 
-> \*Terminology: Over the course of the paper, (at least) five different terms are used to refer to the goodness of a clustering: stability, rigor, robustness, quality and accuracy.
+**\*Terminology: Over the course of the paper, (at least) five different terms are used to refer to the goodness of a clustering: stability, rigor, robustness, quality and accuracy.
 Yet only two distinct concepts are explored (stability and quality/accuracy).
 I strongly recommend that the words "rigorous" and "robust" be struck from the manuscript entirely.
-They are never defined, are unclear, and the first impression they give does not correspond well to the actual results they are being used to describe.
+They are never defined, are unclear, and the first impression they give does not correspond well to the actual results they are being used to describe.**
 
@@
 
-> Abstract...
+**Abstract...**
 
-> L5: replace "rigorous" with something else, eg. "optimal", or even "best". Rigor is not being evaluated here.
-> L27: replace "rigorous". What is described in this paper is that the de novo methods produce OTU clusterings that coincide better with the pairwise distances between the sequencing reads.
-> L209: Remove "rigorous". You've used "accuracy" in this section (which appears identical to "quality" elsewhere, consider consolidating), so replace rigorous with accurate.
+**L5: replace "rigorous" with something else, eg. "optimal", or even "best". Rigor is not being evaluated here.**
+**L27: replace "rigorous". What is described in this paper is that the de novo methods produce OTU clusterings that coincide better with the pairwise distances between the sequencing reads.**
+**L209: Remove "rigorous". You've used "accuracy" in this section (which appears identical to "quality" elsewhere, consider consolidating), so replace rigorous with accurate.**
 
@@
 
-> L7-L8: Recommend introducing "quality" here alongside its initial operational definition.
+**L7-L8: Recommend introducing "quality" here alongside its initial operational definition.**
 
@@
 
-> L16: replace "robust". Robust is a particularly poor choice of word here, as it brings to mind stability, which is not what is being referred to.
-> L245: Remove "robust".
+**L16: replace "robust". Robust is a particularly poor choice of word here, as it brings to mind stability, which is not what is being referred to.**
+**L245: Remove "robust".
 The de novo methods produce clusterings that are more consistent with the pairwise separations of the individual sequencing reads.
-That does not correspond to any normal usage of the word robust.
-> L362: replace "robust"
-> L364: replace "robust"
+That does not correspond to any normal usage of the word robust.**
+**L362: replace "robust"**
+**L364: replace "robust"**
 
@@
 
-> \*L20-L21: USEARCH -> USEARCH and VSEARCH.
-According to your results, both methods produced OTU assignments that depended on the ordering of the reference database (~55-60% label consistency between reorderings of the database).
+**\*L20-L21: USEARCH -> USEARCH and VSEARCH.
+According to your results, both methods produced OTU assignments that depended on the ordering of the reference database (~55-60% label consistency between reorderings of the database).**
 
@@
 
-> Introduction...
+**Introduction...**
 
-> L46-47: Move the last sentence of this paragraph to the first sentence of the next paragraph.
+**L46-47: Move the last sentence of this paragraph to the first sentence of the next paragraph.**
 
@@
 
-> L57-59: This sentence does not follow from the previous (don't need "Therefore,") and is not necessarily a shortcoming, but is rather an observation.
-All methods other than complete-linkage can join sequences together that are are less than 97% similar.
+**L57-59: This sentence does not follow from the previous (don't need "Therefore,") and is not necessarily a shortcoming, but is rather an observation.
+All methods other than complete-linkage can join sequences together that are are less than 97% similar.**
 
@@
 
-> L60: seuquence -> sequence. And drop "subtle".
+**L60: seuquence -> sequence. And drop "subtle".**
 
@@
 
-> \*L64: The other major advantage of reference based methods is that the OTU assignments can be compared across experiments and even studies. This should be mentioned here.
+**\*L64: The other major advantage of reference based methods is that the OTU assignments can be compared across experiments and even studies. This should be mentioned here.**
 
@@
 
-> L68: speed -> computational cost (or something similar)
+**L68: speed -> computational cost (or something similar)**
 
@@
 
-> L76: drop "however"
+**L76: drop "however"**
 
@@
 
-> L80: Move last sentence ahead of previous sentence, as this follows from the reference independence property, not the ability to subsequently classify.
+**L80: Move last sentence ahead of previous sentence, as this follows from the reference independence property, not the ability to subsequently classify.**
 
@@
 
-> L85: Clarify that this is a potential issue with "current" or "common" open-reference methods.
+**L85: Clarify that this is a potential issue with "current" or "common" open-reference methods.
 It is not an inherent issue with any open-reference approach, as one can use coherent OTU assignment strategies for both parts (eg.
-join reference sequence or open-de-novo cluster if within 3%).
+join reference sequence or open-de-novo cluster if within 3%).**
 
@@
 
-> \*L95-105: Evaluation of accuracy on mock communities needs to be included in this discussion of previous methods to evaluate OTU methods.
+**\*L95-105: Evaluation of accuracy on mock communities needs to be included in this discussion of previous methods to evaluate OTU methods.**
 
@@
 
-> L109: Those sequences that... -> Those pairs of sequence that... The unit of this analysis is pairs of sequences, not individual sequences.
+**L109: Those sequences that... -> Those pairs of sequence that... The unit of this analysis is pairs of sequences, not individual sequences.**
 
@@
 
-> L117: attempted to characterize -> characterized
+**L117: attempted to characterize -> characterized**
 
@@
 
-> L137-139: Either explain how the stability MCC is calculated, or reference the Methods at this point.
+**L137-139: Either explain how the stability MCC is calculated, or reference the Methods at this point.**
 
@@
 
-> L141: Multiple subsamplings was -> were
+**L141: Multiple subsamplings was -> were**
 
@@
 
-> L148-152: This is almost an extreme level of disagreement on this point between essentially identical AL analyses in the He study and this study.
+**L148-152: This is almost an extreme level of disagreement on this point between essentially identical AL analyses in the He study and this study.
 I find the He results rather unbelievable, but can the authors shed any further light on what might have happened?
-THe He study doesn't provide a full workflow, but did provide the command they executed: cluster(column=unique.dist,name=names,cutoff=0.05,precision=1000,method=average)
+THe He study doesn't provide a full workflow, but did provide the command they executed: cluster(column=unique.dist,name=names,cutoff=0.05,precision=1000,method=average)**
 
@@
 
-> \*L182-190: Comparing OTUs on the basis of their representative sequence is not "specious" and the He analysis was not "so poorly designed".
+**\*L182-190: Comparing OTUs on the basis of their representative sequence is not "specious" and the He analysis was not "so poorly designed".
 The most abundant sequence in an OTU is a biologically meaningful object, and we would like to think that OTUs are biologically meaningful and not just abstract objects.
-This section needs to be removed, or rewritten with less polemic language and with an explanation of how comparing representative sequences can break down (for example for very low abundance OTUs).
+This section needs to be removed, or rewritten with less polemic language and with an explanation of how comparing representative sequences can break down (for example for very low abundance OTUs).**
 
@@
 
-> L243: the assignments are highly reproducible -> the overall quality of the assignments are consistent.
-> The reproducibility of the assignments themselves was not evaluated here.
+**L243: the assignments are highly reproducible -> the overall quality of the assignments are consistent.**
+**The reproducibility of the assignments themselves was not evaluated here.**
 
@@
 
-> L246: Evalution -> Evaluation
+**L246: Evalution -> Evaluation**
 
@@
 
-> L246-249: I don't really understand this. Swarm implements single-linkage clustering, in a way that is very clever computationally, but that is still single-linkage clustering.
+**L246-249: I don't really understand this. Swarm implements single-linkage clustering, in a way that is very clever computationally, but that is still single-linkage clustering.**
 
@@
 
-> L251: method -> methods
+**L251: method -> methods**
 
@@
 
-> L326-328: It would be useful to the reader to include the number of V4 covering reference sequences here.
-> L323-328: Numbers of sequences and numbers of pairs of sequences are mixed together here in a confusing fashion.
+**L326-328: It would be useful to the reader to include the number of V4 covering reference sequences here.**
+**L323-328: Numbers of sequences and numbers of pairs of sequences are mixed together here in a confusing fashion.
 It would be better to do one or the other (eg.
@@
 -> XXX sequences within 97% of another reference sequence...).
-This allows the reader to easily evaluate the fraction of the reference database this corresponds to.
+This allows the reader to easily evaluate the fraction of the reference database this corresponds to.**
 
@@
 
-> L341-345: Is this referring to sequences duplicated over V4? Or over the whole 16S gene?
+**L341-345: Is this referring to sequences duplicated over V4? Or over the whole 16S gene?**
 
@@
 
-> L354: "It is worth noting that the entire design of the He study was artificial." - This is unnecessarily polemic.
+**L354: "It is worth noting that the entire design of the He study was artificial." - This is unnecessarily polemic.
 There are ways in which the He study was not ideal, but essentially every OTU evaluation article uses an "artificial" design.
-The points being made here stand on their own, and are only obscured by this kind of language.
+The points being made here stand on their own, and are only obscured by this kind of language.**
 
@@
 
-> L390-395: It appears to me that the dependence on the order of the reference database has the potential to introduce _systematic_ rather than just random errors, which is somewhat troubling.
+**L390-395: It appears to me that the dependence on the order of the reference database has the potential to introduce _systematic_ rather than just random errors, which is somewhat troubling.**
 
@@
 
-> Reviewer 2 (Thomas Schmidt)
+**Reviewer 2 (Thomas Schmidt)**
 
-> Basic reporting
+**Basic reporting**
 
-> In their manuscript, 'De novo clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units', Schloss & Westcott report a series of quality benchmarks on several established and recent 16S sequence clustering approaches and methods.
+**In their manuscript, 'De novo clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units', Schloss & Westcott report a series of quality benchmarks on several established and recent 16S sequence clustering approaches and methods.
 The authors critically reproduced and extended a recent study by He et al.
 (Microbiome, 2015); they provide arguments and data that OTU "stability" is not necessarily indicative of OTU "quality"; they provide OTU quality benchmarks for very recent algorithms/software (SWARM and VSEARCH); and they report a novel perspective and tests on reference-based OTU demarcation.
-I believe that this work has merit, that the science is generally sound (but see comments below) and that the manuscript is timely and important for the growing community of microbial ecologists.
+I believe that this work has merit, that the science is generally sound (but see comments below) and that the manuscript is timely and important for the growing community of microbial ecologists.**
 
-> My main concern is that the study design is somewhat disparate, as the individual reported findings are not inherently inter-connected (see also comments on "Validity of the Findings").
+**My main concern is that the study design is somewhat disparate, as the individual reported findings are not inherently inter-connected (see also comments on "Validity of the Findings").
 Also, while the data and results provided generally support the authors' conclusions, I am not convinced that benchmarks based on the Matthew's Correlation Coefficient alone are the most informative approach for all the tests conducted (again, see comments below).
-Finally, there are several issues regarding the general reporting in the manuscript that could make it hard to follow for non-expert readers.
+Finally, there are several issues regarding the general reporting in the manuscript that could make it hard to follow for non-expert readers.**
 
-> Following the journal's recommendation, I have tried to organize all comments into the appropriate sections ("Basic Reporting", "Experimental Design" and "Validity of the Findings").
+**Following the journal's recommendation, I have tried to organize all comments into the appropriate sections ("Basic Reporting", "Experimental Design" and "Validity of the Findings").
 Within these sections, I tried to organize all major comments in decreasing order of importance and minor comments chronologically, following the manuscript.
-However, several points touch more than one of the specified comment sections, and I have tried to disentangle these as much as possible in order to achieve a readable review.
+However, several points touch more than one of the specified comment sections, and I have tried to disentangle these as much as possible in order to achieve a readable review.**
 
-> Comments on "Basic Reporting".
+**Comments on "Basic Reporting".**
 
-> - The manuscript is overall very readable, clearly written, the global structure is clear and all individual sections are well structured.
+**- The manuscript is overall very readable, clearly written, the global structure is clear and all individual sections are well structured.
 That being said, the structure of the study is also somewhat disparate: the individual parts stand somewhat independently from each other and do not form a fully coherent body of research.
@@
 One general topic is a comparison of de novo vs reference-based OTUs, another is the question of OTU "stability" vs "accuracy" and a third thread refers to the clustering quality of novel methods, in particular a comparison of USEARCH to its open-source clone VSEARCH.
-As I said, the manuscript is very readable as is, but these findings are not sufficiently synthesized in the Conclusions, in my opinion.
+As I said, the manuscript is very readable as is, but these findings are not sufficiently synthesized in the Conclusions, in my opinion.**
 
@@
 
-> - All Figures, in their current form, are really non-obvious to interpret.
+**- All Figures, in their current form, are really non-obvious to interpret.
 Frankly speaking, when I first read through the manuscript, I understood all results from the text but really struggled at interpreting the figures.
-I believe they would profit from a careful workover.
+I believe they would profit from a careful workover.**
 
@@
 
-> => Figures 1&3 lack color legends (it took me some time to realize that colors in the left panel corresponded to the categorical axis in right panel).
+**=> Figures 1&3 lack color legends (it took me some time to realize that colors in the left panel corresponded to the categorical axis in right panel).
 Also, being partially color-blind, I had a hard time distinguishing the thick lines in Fig 1&3, left panels.
-Ordinate axis labeling in Fig 1&3 was not intuitive on a first read-through; the meaning of "MCC relative to full dataset" and "to distances" was not clear, and only a close look at the corresponding Methods section (but not the Figure caption) helped resolve this.
+Ordinate axis labeling in Fig 1&3 was not intuitive on a first read-through; the meaning of "MCC relative to full dataset" and "to distances" was not clear, and only a close look at the corresponding Methods section (but not the Figure caption) helped resolve this.**
 
@@
 
-> => In my opinion, Fig 2 presents the data in a non-intuitive way.
+**=> In my opinion, Fig 2 presents the data in a non-intuitive way.
 I am not sure which point the authors are trying to emphasize here: the illustration draws the focus to the generally varying number in absolute OTU counts, but from the arguments in the text I believe that relative differences in counts between re-clustered sub-setted datasets and rarefied full datasets are more important.
@@
 Moreover, a comparison of SWARM is missing from this figure for some reason (?).
-I would suggest to split this figure into two parts: part A providing more classical rarefaction curves per method (with sub-setted clustering points as overlay scatter) and part B providing the (log) relative OTU count between rarefaction and sub-setting across down-sampling steps.
+I would suggest to split this figure into two parts: part A providing more classical rarefaction curves per method (with sub-setted clustering points as overlay scatter) and part B providing the (log) relative OTU count between rarefaction and sub-setting across down-sampling steps.**
 
@@
 
-> => Figure 4 lacks a caption that is explanatory beyond the figure title.
+**=> Figure 4 lacks a caption that is explanatory beyond the figure title.**
 
@@
 
-> - The Introduction is generally very clear, concise and readable.
+**- The Introduction is generally very clear, concise and readable.
 However, there is a large body of previous work on testing methods for OTU clustering, and the authors focus on only a very limited subset of studies.
 While I think that a comprehensive review of all studies on OTU benchmarking ever published is certainly outside the scope of the present work, I believe that the manuscript would much profit from a more in-depth discussion of at least some.
-Other works possibly worth mentioning include, but are not limited to:
-> => Huse et al., Env Microbiol, 2010, DOI: 10.1111/j.1462-2920.2010.02193.x
-> => White et al., BMC Bioinformatics, 2010, doi:10.1186/1471-2105-11-152
-> => Barriuso et al., BMC Bioinformatics, 2011, doi:10.1186/1471-2105-12-473, who indeed provided an early investigation into "output variability" of OTU clustering (Table 7)
-> => Bonder et al., Bioinformatics, 2012, doi: 10.1093/bioinformatics/bts552
-> => Koeppel & Wu, NAR, 2013, doi: 10.1093/nar/gkt241
-> => Schmidt et al., PLOS Comp Biol, 2014, DOI: 10.1371/journal.pcbi.1003594
-> => May et al., Bioinformatics, 2014, doi: 10.1093/bioinformatics/btu085
-> However, these are merely suggestions and the authors should not feel obliged to include any or all of these as new references in the Introduction – in particular, since I also pointed out one work of which I am co-author.
-Nevertheless, I do believe that the manuscript would generally profit from a discussion of (some) further previous work.
+Other works possibly worth mentioning include, but are not limited to:**
+**=> Huse et al., Env Microbiol, 2010, DOI: 10.1111/j.1462-2920.2010.02193.x**
+**=> White et al., BMC Bioinformatics, 2010, doi:10.1186/1471-2105-11-152**
+**=> Barriuso et al., BMC Bioinformatics, 2011, doi:10.1186/1471-2105-12-473, who indeed provided an early investigation into "output variability" of OTU clustering (Table 7)**
+**=> Bonder et al., Bioinformatics, 2012, doi: 10.1093/bioinformatics/bts552**
+**=> Koeppel & Wu, NAR, 2013, doi: 10.1093/nar/gkt241**
+**=> Schmidt et al., PLOS Comp Biol, 2014, DOI: 10.1371/journal.pcbi.1003594**
+**=> May et al., Bioinformatics, 2014, doi: 10.1093/bioinformatics/btu085**
+**However, these are merely suggestions and the authors should not feel obliged to include any or all of these as new references in the Introduction – in particular, since I also pointed out one work of which I am co-author.
+Nevertheless, I do believe that the manuscript would generally profit from a discussion of (some) further previous work.**
 
-> - The Introduction provides a detailed, accessible and balanced explanation of the three general approaches to OTU demarcation discussed in the paper: de novo, closed reference and open reference.
+**- The Introduction provides a detailed, accessible and balanced explanation of the three general approaches to OTU demarcation discussed in the paper: de novo, closed reference and open reference.
 Although the authors point to two comprehensive references which give further information on these points (Schloss & Westcott, 2011; Navas-Molina et al., 2013), I would suggest that the authors provide at least some visual/structural guidance to the reader (an overview figure or table).
-However, the manuscript is certainly understandable as is, so please consider this comment as a mere suggestion.
+However, the manuscript is certainly understandable as is, so please consider this comment as a mere suggestion.**
 
@@
 
-> - Related to this, the terms "abundance-based" and "distance-based greedy clustering" (AGC and DGC) are not properly introduced.
-Although the authors point out that they stick with the terminology of the He et al paper, these terms need to be explained somewhere in the manuscript, preferably in the Introduction.
+**- Related to this, the terms "abundance-based" and "distance-based greedy clustering" (AGC and DGC) are not properly introduced.
+Although the authors point out that they stick with the terminology of the He et al paper, these terms need to be explained somewhere in the manuscript, preferably in the Introduction.**
 
@@
 
-> - Similarly, the last part in the Results and Discussion, "Problems with reference-based clustering [...]", could profit from an illustration.
+**- Similarly, the last part in the Results and Discussion, "Problems with reference-based clustering [...]", could profit from an illustration.
 This part is very clearly written, and the presented results are striking (see also comments under "Validity of the Findings").
 Nevertheless, a visual presentation of the data could additionally help to guide the reader through the conducted tests and illustrate just how striking the reported results are.
-Even a simple table summarizing the results would be helpful.
+Even a simple table summarizing the results would be helpful.**
 
@@
 
-> - The authors very clearly point out weaknesses in the criticized study by He et al; they support all their arguments with data and/or sound reasoning and do not, in my opinion, state any unfair or unsupported claims against the criticized work.
-Nevertheless, I personally felt that the wording is sometimes a bit harsh: e.g., p.9, l.190 ("this analysis was so poorly designed"); p.15, l.354 ("the entire design of the He study was artificial").
+**- The authors very clearly point out weaknesses in the criticized study by He et al; they support all their arguments with data and/or sound reasoning and do not, in my opinion, state any unfair or unsupported claims against the criticized work.
+Nevertheless, I personally felt that the wording is sometimes a bit harsh: e.g., p.9, l.190 ("this analysis was so poorly designed"); p.15, l.354 ("the entire design of the He study was artificial").**
 
@@
 
-> Minor comments:
+**Minor comments:**
 
-> - Please re-check for typos and grammar across the text (e.g., p.5, l.60 "seuquence"; p.5, l.69 "scale[s]"; p.6, l.88 "assigned"; p.7, l.137 "assess the calculated the MCC"; etc.)
+**- Please re-check for typos and grammar across the text (e.g., p.5, l.60 "seuquence"; p.5, l.69 "scale[s]"; p.6, l.88 "assigned"; p.7, l.137 "assess the calculated the MCC"; etc.)**
 
@@
 
-> - Several examples given throughout the text appear rather arbitrary.
-For example, p.5, l.59 ("may only be 94% similar", same in p.17, ll.402-3) or p.6, l.89 ("all sequences assigned to the Porphyromonadaceae would then be assigned to OTUs [...]") are out of context, the sudden mention of a concrete clustering threshold or an arbitrary bacterial family could confuse the reader more than illustrate the point being made.
+**- Several examples given throughout the text appear rather arbitrary.
+For example, p.5, l.59 ("may only be 94% similar", same in p.17, ll.402-3) or p.6, l.89 ("all sequences assigned to the Porphyromonadaceae would then be assigned to OTUs [...]") are out of context, the sudden mention of a concrete clustering threshold or an arbitrary bacterial family could confuse the reader more than illustrate the point being made.**
 
@@
 
-> - p.4, l.56.
+**- p.4, l.56.
 I believe that the provided reference for the statement that "the commonly used variable regions [...] do not evolve at the same rate" is not fully appropriate.
@@
 Moreover, it is probably also worth mentioning the related contribution by Kim et al.
-(Journal Micobiol Methods, 2011, http://www.ncbi.nlm.nih.gov/pubmed/21047533).
+(Journal Micobiol Methods, 2011, http://www.ncbi.nlm.nih.gov/pubmed/21047533).**
 
@@
 
-> - When enumerating the pros and cons of closed ref clustering (p.5, ll.
-48f), several points in favor of this approach could be added: it is trivially parallelizable, has a low memory footprint and arguably gets better as reference databases improve (which is the case as more and more sequence data accumulates).
+**- When enumerating the pros and cons of closed ref clustering (p.5, ll.
+48f), several points in favor of this approach could be added: it is trivially parallelizable, has a low memory footprint and arguably gets better as reference databases improve (which is the case as more and more sequence data accumulates).**
 
@@
 
-> - p.6, l.93-4.
+**- p.6, l.93-4.
 This statement is phrased too strongly.
@@
 The authors, themselves, point out just a few lines later that taxonomic classification is often far from perfect (it could be added that it depends strongly on reference database quality and classifier performance); however, if the classification step is not perfect, the entire clustering outcome will be (strongly) affected.
-Others have suggested database-agnostic alternatives in a similar spirit, such as SL pre-clustering (Huse et al., Env Microbiol, 2010) or the kmer-pre-clustering implemented in ESPRIT-Tree (Cai & Sun, NAR, 2011) which would be worth mentioning here.
+Others have suggested database-agnostic alternatives in a similar spirit, such as SL pre-clustering (Huse et al., Env Microbiol, 2010) or the kmer-pre-clustering implemented in ESPRIT-Tree (Cai & Sun, NAR, 2011) which would be worth mentioning here.**
 
@@
 
-> - p.6, ll.109ff. This paragraph explaining the MCC (in detail) feels a little out of place here; I suggest to relocate it to the (early) Results section.
+**- p.6, ll.109ff. This paragraph explaining the MCC (in detail) feels a little out of place here; I suggest to relocate it to the (early) Results section.**
 
@@
 
-> Experimental design
+**Experimental design**
 
-> - Methods are reported clearly and concisely, links to available data are provided, all tested tools are specified down to the version number.
+**- Methods are reported clearly and concisely, links to available data are provided, all tested tools are specified down to the version number.
 More Methods sections should look like this!
-In particular, I would like to commend the authors for uploading all code as reproducible workflow to Github.
+In particular, I would like to commend the authors for uploading all code as reproducible workflow to Github.**
 
@@
 
-> - Matthew's Correlation Coefficient (MCC) is used throughout the study as the sole indicator of OTU "quality".
+**- Matthew's Correlation Coefficient (MCC) is used throughout the study as the sole indicator of OTU "quality".
 The MCC is an objective summary statistic for classification problems and has been successfully used for OTU benchmarking before (originally introduced by the authors in their 2011 paper and since then picked up by several other teams).
-Nevertheless, I would like to point out three general objections to using the MCC as an OTU quality criterion and then argue that the MCC, when used as a sole index, is not the most informative choice for (all) the questions that the authors pursue in this study.
+Nevertheless, I would like to point out three general objections to using the MCC as an OTU quality criterion and then argue that the MCC, when used as a sole index, is not the most informative choice for (all) the questions that the authors pursue in this study.**
 
-> => In general, the MCC is a summary statistic for binary classification problems.
+**=> In general, the MCC is a summary statistic for binary classification problems.
 However, (OTU) clustering is arguably not a problem that boils down to binary true/false assignments only.
@@
 For example, by definition, complete linkage or USEARCH would never show "false positive" clustering (but many false negatives can be expected), while single linkage would never make "false negative" mistakes (while a wealth of false positives is expected).
-In other words, the MCC was not designed for this kind of application and OTU "quality" is not (only) a binary classification problem; nevertheless, MCC values are arguably informative, if interpreted carefully.
-> => My second "general" issue with the MCC concerns its formulation as a summary statistic: frankly speaking, I find it rather non-intuitive to interpret.
+In other words, the MCC was not designed for this kind of application and OTU "quality" is not (only) a binary classification problem; nevertheless, MCC values are arguably informative, if interpreted carefully.**
+**=> My second "general" issue with the MCC concerns its formulation as a summary statistic: frankly speaking, I find it rather non-intuitive to interpret.
 For example, in the present study, it would be very informative to see what actually goes "right" or "wrong" for the different methods, driving MCC values (e.g., by showing distributions of TPR, FPR, etc.
-per clustering method).
-> => Finally, I suspect that the MCC is sensitive to cluster count and size distribution, although I am not aware that this has ever been tested.
+per clustering method).**
+**=> Finally, I suspect that the MCC is sensitive to cluster count and size distribution, although I am not aware that this has ever been tested.
 In general, more "errors" can be expected the further a clustering algorithm proceeds: sets with fewer and larger OTUs would be more prone to erroneous assignments than sets with many small OTUs.
 This is problematic, because the "hard" clustering threshold is implemented very differently by different algorithms, and it is well established that at a given threshold of, say, 97% the different methods provide very different cluster count and size distributions.
-In the present study, such effects would be especially striking for the SWARM algorithm.
+In the present study, such effects would be especially striking for the SWARM algorithm.**
 
-> In spite of these comments, the use of the MCC in the second use case of the present study ("MCC value relative to distances") is arguably established and meaningful, if interpreted cautiously.
+**In spite of these comments, the use of the MCC in the second use case of the present study ("MCC value relative to distances") is arguably established and meaningful, if interpreted cautiously.
 However, in my opinion, the MCC is an insufficient metric in the first use case ("MCC value relative to full dataset").
@@
 Arguably, this corresponds even less to the original use case of the MCC, and putative differential cluster size and count effects would be even more striking (based on the authors' findings in Figure 2).
-What appears particularly confusing is that the interpretation of the MCC changes drastically between both use cases: the "MCC on distances" is interpretable in line with previous work, but the "MCC relative to full dataset clustering" means something completely different.
+What appears particularly confusing is that the interpretation of the MCC changes drastically between both use cases: the "MCC on distances" is interpretable in line with previous work, but the "MCC relative to full dataset clustering" means something completely different.**
 
-> Several metrics have been developed which appear more appropriate for the latter use case.
+**Several metrics have been developed which appear more appropriate for the latter use case.
 The down-sampling experiment, in my opinion, would be more straightforward to interpret if clustering comparison metrics, e.g.
@@
 Taken together, MCC and AMI/NMI/ARI could then indeed be interpreted as OTU "accuracy" and "stability".
-Likewise, the comparison between USEARCH and VSEARCH is a model use case for which the above-mentioned methods were originally designed.
+Likewise, the comparison between USEARCH and VSEARCH is a model use case for which the above-mentioned methods were originally designed.**
 
@@
 
-> Validity of the findings
+**Validity of the findings**
 
-> - As detailed above, I believe that all conclusions are generally justified by the analyses conducted, albeit reservations regarding the use of the MCC as sole index in the stated cases.
-However, as likewise detailed above, I also believe that several findings stand independently of each other, and their connections are not immediately obvious.
+**- As detailed above, I believe that all conclusions are generally justified by the analyses conducted, albeit reservations regarding the use of the MCC as sole index in the stated cases.
+However, as likewise detailed above, I also believe that several findings stand independently of each other, and their connections are not immediately obvious.**
 
@@
 
-> - One main rationale of the present study is that OTU assignment "accuracy" trumps "stability".
+**- One main rationale of the present study is that OTU assignment "accuracy" trumps "stability".
 The authors provide clear reasoning to support this argument, and their data is generally in line with this notion.
@@
 It is neither desirable to work with "stable but wrong" nor "accurate but noisy" OTUs.
-I agree with the authors that accuracy is the more desirable property in general, but putting their study in context with other recent findings could suggest that in fact, accuracy and stability tend to be correlated, which would lend even more gravity to their argument.
+I agree with the authors that accuracy is the more desirable property in general, but putting their study in context with other recent findings could suggest that in fact, accuracy and stability tend to be correlated, which would lend even more gravity to their argument.**
 
@@
 
-> - The points raised in critical response to the He et al.
+**- The points raised in critical response to the He et al.
 study are all well illustrated, argued and supported by data.
-Moreover, by repeating several of the tests on a more recent (and more reasonably chosen), more deeply sequenced dataset, the authors provide a clear added value.
+Moreover, by repeating several of the tests on a more recent (and more reasonably chosen), more deeply sequenced dataset, the authors provide a clear added value.**
 
-> - The inclusion of tests on the very recent SWARM and VSEARCH algorithms makes this manuscript very timely and useful for researchers in the field.
-To my knowledge, this is the first available independent test on VSEARCH performance.
+**- The inclusion of tests on the very recent SWARM and VSEARCH algorithms makes this manuscript very timely and useful for researchers in the field.
+To my knowledge, this is the first available independent test on VSEARCH performance.**
 
-> - The findings on "problems with reference-based clustering in general and as implemented in QIIME" are striking and, indeed, slightly disturbing.
+**- The findings on "problems with reference-based clustering in general and as implemented in QIIME" are striking and, indeed, slightly disturbing.
 To be honest, I would not have expected such strong effects of sequence order for mappings against the QIIME reference database.
-I believe that these results will contribute to more debates in the field, and ultimately lead to further long-needed benchmarks on standard operating procedures for widely used software.
+I believe that these results will contribute to more debates in the field, and ultimately lead to further long-needed benchmarks on standard operating procedures for widely used software.**
 
@@
 
-> - p.12, ll.260-1.
+**- p.12, ll.260-1.
 This statement is too harsh based on the given data.
-The SWARM algorithm is designed to not work with "hard" thresholds in the classical sense, so it is not surprising that in the given testing framework, different "thresholds" would appear optimal for differentially deeply sequenced datasets.
+The SWARM algorithm is designed to not work with "hard" thresholds in the classical sense, so it is not surprising that in the given testing framework, different "thresholds" would appear optimal for differentially deeply sequenced datasets.**
 
@@
 
-> - p.15, ll.368-9.
+**- p.15, ll.368-9.
 This solution appears slightly impractical.
@@
 Also, it is unclear what an MCC comparison across different studies would actually mean (considering additional confounding factors such as sequencing depth, clustering threshold, alignment, etc.).
-Nevertheless, I fully agree with the authors that researchers should ideally perform (and report) different clusterings on their data.
+Nevertheless, I fully agree with the authors that researchers should ideally perform (and report) different clusterings on their data.**
 
@@
 
-> Comments for the author
+**Comments for the author**
 
-> I would like to again commend the authors on their manuscript and again apologize for pushing a study their way in which I was involved.
-I was very hesitant to include the points on the use of the MCC vs AMI/NMI/ARI in this review, but I honestly believe that they make sense scientifically.
+**I would like to again commend the authors on their manuscript and again apologize for pushing a study their way in which I was involved.
+I was very hesitant to include the points on the use of the MCC vs AMI/NMI/ARI in this review, but I honestly believe that they make sense scientifically.**
 

### 2015-11-13 ResponseToReviewers.md
Add cover letter

@@
 ---
-title: "**Response to reviewers**"
+title: ""
 output:
@@
 ---
+
+Dear Dr. Eren,
+
+Thank you for providing my co-author and I with a very fast review of our first submission.
+We were delighted by the generally positive and thorough comments provided by the two reviewers.
+Attached you will find a complete set of responses to each of their concerns.
+As you suggested, we streamlined our jargon and tempered our language.
+To the points raised by reviewer 2, we elected to use the MCC as was done in the original He paper for consistency.
+We have added text to the manuscript indicating that this probably wasn't the best choice of methods.
+Hopefully, the point is clear that this way of measuring stability is not ideal.
+
+Please do not hesitate to contact us if you have additional comments or concerns.
+
+Sincerely,
+
+Sarah Westcott & Pat Schloss
+
+\newpage
 

### 2015-11-13 Schloss_Cluster_PeerJ_2015.tex
Add latex file

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={De novo clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{\textbf{\emph{De novo} clustering methods out-perform reference-based
methods for assigning 16S rRNA gene sequences to operational taxonomic
units}}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}
\usepackage[all]{nowidow}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}

% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\begin{document}
\maketitle

\begin{center}
\vspace{25mm}
Sarah L. Westcott and Patrick D. Schloss${^\dagger}$

\vspace{30mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
\end{center}

\newpage

\linenumbers

\subsection{Abstract}\label{abstract}

\textbf{Background.} 16S rRNA gene sequences are routinely assigned to
operational taxonomic units (OTUs) that are then used to analyze complex
microbial communities. A number of methods have been employed to carry
out the assignment of 16S rRNA gene sequences to OTUs leading to
confusion over which method is optimal. A recent study suggested that a
clustering method should be selected based on its ability to generate
stable OTU assignments that do not change as additional sequences are
added to the dataset. In contrast, we contend that the quality of the
OTU assignments, the ability of the method to properly represent the
distances between the sequences, is more important.

\textbf{Methods.} Our analysis implemented six \emph{de novo} clustering
algorithms including the single linkage, complete linkage, average
linkage, abundance-based greedy clustering, distance-based greedy
clustering, and Swarm and the open and closed-reference methods. Using
two previously published datasets we used the Matthew's Correlation
Coefficient (MCC) to assess the stability and quality of OTU
assignments.

\textbf{Results.} The stability of OTU assignments did not reflect the
quality of the assignments. Depending on the dataset being analyzed, the
average linkage and the distance and abundance-based greedy clustering
methods generated OTUs that were more likely to represent the actual
distances between sequences than the open and closed-reference methods.
We also demonstrated that for the greedy algorithms VSEARCH produced
assignments that were comparable to those produced by USEARCH making
VSEARCH a viable free and open source alternative to USEARCH. Further
interrogation of the reference-based methods indicated that when USEARCH
or VSEARCH were used to identify the closest reference, the OTU
assignments were sensitive to the order of the reference sequences
because the reference sequences can be identical over the region being
considered. More troubling was the observation that while both USEARCH
and VSEARCH have a high level of sensitivity to detect reference
sequences, the specificity of those matches was poor relative to the
true best match.

\textbf{Discussion.} Our analysis calls into question the quality and
stability of OTU assignments generated by the open and closed-reference
methods as implemented in current version of QIIME. This study
demonstrates that \emph{de novo} methods are the optimal method of
assigning sequences into OTUs and that the quality of these assignments
needs to be assessed for multiple methods to identify the optimal
clustering method for a particular dataset.

\newpage

\subsection{Introduction}\label{introduction}

The ability to affordably generate millions of 16S rRNA gene sequences
has allowed microbial ecologists to thoroughly characterize the
microbial community composition of hundreds of samples. To simplify the
complexity of these large datasets, it is helpful to cluster sequences
into meaningful bins. These bins, commonly known as operational
taxonomic units (OTUs), are used to compare the biodiversity contained
within and between different samples (Schloss \& Westcott, 2011). Such
comparisons have enabled researchers to characterize the microbiota
associated with the human body (e.g. Huttenhower et al., 2012), soil
(e.g. Shade et al., 2013), aquatic ecosystems (e.g. Gilbert et al.,
2011), and numerous other environments. Within the field of microbial
ecology, a convention has emerged where sequences are clustered into
OTUs using a threshold of 97\% similarity or a distance of 3\%. One
advantage of the OTU-based approach is that the definition of the bins
is operational and can be changed to suit the needs of the particular
project. However, with the dissemination of clustering methods within
software such as mothur (Schloss et al., 2009), QIIME (Caporaso et al.,
2010), and other tools (Sun et al., 2009; Edgar, 2010, 2013; Cai \& Sun,
2011; Mahé et al., 2014), it is important to understand how different
clustering methods implement this conventional OTU threshold.
Furthermore, it is necessary to understand how the selected method
affects the precision and accuracy of assigning sequences to OTUs.
Broadly speaking, three approaches have been developed to assign
sequences to OTUs:

The first approach has been referred to as phylotyping (Schloss \&
Westcott, 2011) or closed reference clustering (Navas-Molina et al.,
2013). This approach involves comparing sequences to a curated database
and then clustering sequences into the same OTU that are similar to the
same reference sequence. Reference-based clustering methods suffer when
the reference does not adequately reflect the biodiversity of the
community. If a large fraction of sequences are novel, then they cannot
be assigned to an OTU. In addition, the reference sequences are selected
because they are less than 97\% similar to each other over the full
length of the gene; however, it is known that the commonly used variable
regions within the 16S rRNA gene do not evolve at the same rate as the
full-length gene (Schloss, 2010; Kim, Morrison \& Yu, 2011). Thus, a
sequence representing a fragment of the gene may be more than 97\%
similar to multiple reference sequences. Defining OTUs in the
closed-reference approach is problematic because two sequences might be
97\% similar to the same reference sequence, but they may only be 94\%
similar to each other. Alternatively, a sequence may be equally similar
to two or more reference sequences. A alternative to this approach is to
use a classifier to assign a taxonomy to each sequence so that sequences
can be clustered at a desired level within the Linnean taxonomic
hierarchy (Schloss \& Westcott, 2011). The strengths of the
reference-based methods include their speed, potential for trivial
parallelization, ability to compare OTU assignments across studies, and
the hope that as databases improve, the OTU assignments will also
improve.

The second approach has been referred to as distance-based (Schloss \&
Westcott, 2011) or \emph{de novo} clustering (Navas-Molina et al.,
2013). In this approach, the distance between sequences is used to
cluster sequences into OTUs rather than the distance to a reference
database. In contrast to the efficiency of closed-reference clustering,
the computational cost of hierarchical \emph{de novo} clustering methods
scales quadratically with the number of unique sequences. The expansion
in sequencing throughput combined with sequencing errors inflates the
number of unique sequences resulting in the need for large amounts of
memory and time to cluster the sequences. If error rates can be reduced
through stringent quality control measures, then these problems can be
overcome (Kozich et al., 2013). As an alternative, heuristics have been
developed to approximate the clustering of hierarchical methods (Sun et
al., 2009; Edgar, 2010; Mahé et al., 2014). Two related heuristics
implemented in USEARCH were recently described: distance-based greedy
clustering (DGC) and abundance-based greedy clustering (AGC) (Edgar,
2010; He et al., 2015). These greedy methods cluster sequences within a
defined similarity threshold of an index sequence or creates a new index
sequence. If a sequence is more similar than the defined threshold, it
is assigned to the closest centroid based (i.e.~DGC) or the most
abundant centroid (i.e.~AGC). One critique of \emph{de novo} approaches
is that OTU assignments are sensitive to the input order of the
sequences (Mahé et al., 2014; He et al., 2015). Whether the differences
in assignments is meaningful is unclear and the variation in results
could represent equally valid clustering of the data. The strength of
\emph{de novo} clustering is its independence of references for carrying
out the clustering step. For this reason, \emph{de novo} clustering has
been preferred across the field. After clustering, the classification of
each sequence can be used to obtain a consensus classification for the
OTU (Schloss \& Westcott, 2011).

The third approach, open-reference clustering, is a hybrid of the
closed-reference and \emph{de novo} approaches (Navas-Molina et al.,
2013; Rideout et al., 2014). Open-reference clustering involves
performing closed-reference clustering followed by \emph{de novo}
clustering on those sequences that are not sufficiently similar to the
reference. In theory, this method should exploit the strengths of both
closed-reference and \emph{de novo} clustering; however, the different
OTU definitions employed by commonly used closed-reference and \emph{de
novo} clustering implementations pose a possible problem when the
methods are combined. An alternative to this approach has been to
classify sequences to a bacterial family or genus and then assign those
sequences to OTUs within those taxonomic groups using the average
linkage method (Schloss \& Westcott, 2011). For example, all sequences
classified as belonging to the \emph{Porphyromonadaceae} would then be
assigned to OTUs using the average linkage method using a 3\% distance
threshold. Those sequences that did not classify to a known family would
also be clustered using the average linkage method. An advantage of this
approach is that it lends itself nicely to parallelization since each
taxonomic group is seen as being independent and can be processed
separately. Such an approach would overcome the difficulty of mixing OTU
definitions between the closed-reference and \emph{de novo} approaches;
however, it would still suffer from the problems associated with
database quality and classification error.

The growth in options for assigning sequences using each of these three
broad approaches has been considerable. It has been difficult to
objectively assess the quality of OTU assignments. Some have focused on
the time and memory required to process a dataset (Sun et al., 2009; Cai
\& Sun, 2011; Mahé et al., 2014; Rideout et al., 2014). These are valid
parameters to assess when judging a clustering method, but have little
to say about the quality of the OTU assignments. Others have attempted
to judge the quality of a method by its ability to generate data that
parallels classification data (White et al., 2010; Sun et al., 2011; Cai
\& Sun, 2011). This approach is problematic because bacterial taxonomy
often reflects historical biases amongst bacterial systematicists.
Furthermore, it is well known that the rates of evolution across
lineages are not the same (Wang et al., 2007; Schloss, 2010). A related
approach has used clustering of mock community data to evaluate methods
(Huse et al., 2010; Barriuso, Valverde \& Mellado, 2011; Bonder et al.,
2012; Chen et al., 2013; Edgar, 2013; Mahé et al., 2014; May et al.,
2014). Yet these approaches ignore the effects of sequencing errors that
tend to accumulate with sequencing depth and represent highly idealized
communities that lack the phylogenetic diversity of real microbial
communities (Schloss, Gevers \& Westcott, 2011; Kozich et al., 2013).
Others have assessed the quality of clustering based on their ability to
generate the same OTUs generated by other methods (Rideout et al., 2014;
Schmidt, Rodrigues \& Mering, 2014b). This is problematic because it
does not solve the fundamental question of which method is optimal. The
concept of ecological consistency as a metric of quality asserts that
sequences that cluster into the same OTU should share similar ecological
affiliations (Koeppel \& Wu, 2013; Preheim et al., 2013; Schmidt,
Rodrigues \& Mering, 2014a). Although this is an intriguing approach and
is a quantitative metric, it is unclear how the metric would be
objectively validated. We recently proposed an approach for evaluating
OTU assignments using the distances between pairs of sequences (Schloss
\& Westcott, 2011). We were able to synthesize the relationship between
OTU assignments and the distances between sequences using the Matthew's
correlation coefficient (MCC; Matthews, 1975). MCC is can be interpreted
as representing the correlation between the observed and expected
classifications and can vary between -1.0 and 1.0. The strength of the
MCC, as implemented by Schloss et al. (2011), is that it is an objective
approach to assessing the quality of the OTU assignments that can be
calculated for any set of OTU assignments where there is a distance
matrix and a specific threshold without relying on an external
reference.

A recent analysis by He and colleagues (2015) characterized the three
general clustering approaches by focusing on what they called stability.
They defined stability as the ability of a method to provide the same
clustering on a subset of the data as was found in the full dataset.
Their concept of stability did not account for the quality of the OTU
assignments and instead focused on the precision of the assignments. A
method may be very stable, but of poor quality. In the current analysis,
we assessed the quality and stability of the various clustering methods.
Building on our previous analysis of clustering methods, our hypothesis
was that the methods praised by the He study for their stability
actually suffered a lack of quality. In addition, we assess these
parameters in light of sequence quality using the original 454 dataset
and a larger and more modern dataset generated using the MiSeq platform.

\subsection{Results and Discussion}\label{results-and-discussion}

\textbf{\emph{Summary and replication of He study.}} We obtained the
Canadian soil dataset from Roesch et al. (2007) and processed the
sequences as described by He and colleagues. Using these data, we
reconsidered three of the more critical analyses performed in the He
study.

First, we sought to quantify whether the OTU assignments observed for a
subset of the data represented the same assignments that were found with
the full dataset. The He study used the MCC to quantify the degree to
which pairs of sequences were in the same OTUs in subsampled and full
datasets. A more robust approach would utilize metrics that quantify the
mutual information held between two sets of clusterings and has been
applied to assess inter-method variation in OTU composition (Schmidt,
Rodrigues \& Mering, 2014b). To maintain consistency with the original
He study, we also calculated the MCC value as they described. The He
study found that when they used the open and closed-reference methods
the OTUs formed using the subsetted data most closely resembled those of
the full dataset. Among the \emph{de novo} methods they observed that
the AGC method generated the most stable OTUs followed by the single
linkage (SL), DGC, complete linkage (CL), and average linkage (AL)
methods. We first calculated the MCC for the OTU assignments generated
by each of the clustering methods using 20, 40, 60, and 80\% of the
sequences relative to the OTU composition formed by the methods using
the full dataset (see Methods for description; Figure 1A). Similar to
the He study, we replicated each method and subsampled to the desired
fraction of the dataset 30 times. Multiple subsamplings were necessary
because a random number generator is used in some of the methods to
break ties where pairs of sequences have the same distance between them.
Across these sequencing depths, we observed that the stability of the
OTUs generated by the SL and CL methods were highly sensitive to
sampling effort relative to the OTUs generated by the AL, AGC, and DGC
methods (Figure 1A). Our results (Figure 1B) largely confirmed those of
Figure 4C in the He study with one notable exception. The He study
observed a broad range of MCC values among their AL replicates when
analyzing OTUs generated using 60\% of the data. This result appeared
out of character and was not explained by the authors. They observed a
mean MCC value of approximately 0.63 (95\% confidence interval between
approximately 0.15 and 0.75). In contrast, we observed a mean value of
0.93 (95\% confidence interval between 0.91 and 0.95). This result
indicates that the AL assignments were far more stable than indicated in
the He study. Regardless, although the assignments are quite stable, it
does support the assertion that the OTU assignments observed for the
subset of the data do not perfectly match the assignments that were
found with the full dataset as they did with the reference-based
methods; however, the significance of these differences is unclear.

Second, the He study and the original Roesch study showed that
rarefaction curves calculated using CL-generated OTU assignments
obtained using a subsample of the sequencing data did not overlap with
rarefaction curves generated using OTU assignments generated from the
full dataset. The He and Roesch studies both found that the CL method
produced fewer OTUs in the subset than in the rarefied data. In
addition, the He study found that the SL method produced more OTUs, the
AGC produced fewer, and the other methods produced similar numbers of
OTUs than expected when comparing the subsetted data to the rarefied
data. Our results support those of these previous studies (Figure 2). It
was clear that inter-method differences were generally more pronounced
than the differences observed between rarefying from the full dataset
and from clustering the subsetted data. The number of OTUs observed was
largest using the CL method, followed by the open-reference method. The
AL, AGC, and DGC methods all provided comparable numbers of OTUs.
Finally, the closed-reference and SL methods generated the fewest number
of OTUs.

Third, the authors attempted to describe the effects of the OTU
assignment instability on comparisons of communities. They used Adonis
to test whether the community structure represented in subsetted
communities resembled that of the full dataset when only using the
unstable OTUs (Anderson, 2001). Although they were able to detect
significant p-values, they appeared to be of marginal biological
significance. Adonis R statistics close to zero indicate the community
structures from the full and subsetted datasets overlapped while values
of one indicate the communities are completely different. The He study
observed adonis R statistics of 0.02 (closed-reference), 0.03
(open-reference), 0.07 (CL, AGC, DGC), and 0.16 (SL and AL). Regardless
of the statistical or biological significance of these results, the
analysis was tautological since, by definition, representing communities
based on their unstable OTUs would yield differences. Furthermore, the
\emph{de novo} and open-reference approaches do not consistently label
the OTUs that sequences belong to when the clustering methods are run
multiple times with different random number seeds. To overcome this, the
authors selected representative sequences from each OTU and used those
representative sequences to link OTU assignments between the different
sized sequence sets. It was not surprising that the only analysis that
did not provide a significant p-value was for the closed-reference
analysis, which is the only analysis that provides consistent OTU
labels. Finally, the authors built off of this analysis to count the
number of OTUs that were differentially represented between the
subsetted and full datasets by each method. This entire analysis assumed
that the OTUs generated using the full dataset were correct, which was
an unsubstantiated assumption since the authors did not assess the
quality of the OTU assignments.

This re-analysis of the He study raised five complementary questions.
First, how do the various methods vary in the quality of their OTU
assignments? Second, how generalizable are these results to modern
datasets generated using a large number of sequences that were deeply
sequenced? Third, how does the stability and quality of OTU assignments
generated by new methods compare to those analyzed in the He study?
Fourth, are there open-source alternatives to USEARCH that perform just
as well? Finally, although the stability of reference-based methods did
not appear to be impacted by the input order of the sequences to be
assigned to OTUs, is the stability of reference-based methods impacted
by the order of the reference sequences? In the remainder of the Results
and Discussion section we address each of these questions.

\textbf{\emph{How do the various methods vary in the quality of their
OTU assignments?}} More important than the stability of OTUs is whether
sequences are assigned to the correct OTUs. A method can generate highly
stable OTUs, but the OTU assignments may be meaningless if they poorly
represent the specified cutoff and the actual distance between the
sequences. To assess the quality of OTU assignments by the various
methods, we made use of the pairwise distance between the unique
sequences to count the number of true positives and negatives and the
number of false positives and negatives for each method and sampling
depth. Those pairs of sequences that were similar to each other and
found in the same OTU were called true positives while those that were
similar and found in different OTUs were called false negatives.
Meanwhile, those pairs of sequences that were different from each other
and found in the same OTU were called false positives and those that
were dissimilar and found in different OTUs were called true negatives.
Counting the frequency of these different classes allowed us to judge
how each method balanced the ratio of true positives and negatives to
false positives and negatives using the MCC. We used the average MCC
value as a measure of a method's quality and its variation as a measure
of its consistency. We made three important observations. First, each of
the \emph{de novo} methods varied in how sensitive their MCC values were
to additional sequences (Figure 1C). The SL and CL methods were the most
sensitive; however, the quality of the OTU assignments did not
meaningfully differ when 80 or 100\% of the data were assigned to OTUs
using the \emph{de novo} methods. Second, the AL method had higher MCC
values than the other methods followed by DGC, AGC, CL, open-reference,
and closed-reference, and SL (Figure 1D). Third, with the possible
exception of the CL method, the MCC values for each of the methods only
demonstrated a small amount of variation between runs of the method with
a different ordering of the input sequences. This indicates that
although there may be variation between executions of the same method,
they produced OTU assignments that were of equal quality. Revisiting the
concept of stability, we question the value of obtaining stable OTUs
when the full dataset is not optimally assigned to OTUs. Our analysis
indicates that the most optimal method for assigning the Canadian soils
sequences to OTUs using a 97\% threshold was the AL method.

\textbf{\emph{How generalizable are these results to modern datasets
generated using a large number of sequences that were deeply
sequenced?}} Three factors make the Canadian soil dataset less than
desirable to evaluate clustering methods. First, it was one of the
earliest 16S rRNA gene sequence datasets published using the 454 FLX
platform. Developments in sequencing technology now permit the
sequencing of millions of sequences for a study. In addition, because
the original Phred quality scores and flowgram data are not available,
it was not possible for us to adequately remove sequencing errors
(Schloss, Gevers \& Westcott, 2011). The large number of sequencing
errors that one would expect to remain in the dataset are likely to
negatively affect the performance of all of the clustering methods.
Second, the dataset used in the He study covered the V9 region of the
16S rRNA gene. For a variety of reasons, this region is not well
represented in databases, including the reference database used by the
closed and open-reference methods. Of the 99,322 sequences in the
default QIIME database, only 48,824 fully cover the V9 region. In
contrast, 99,310 of the sequences fully covered the V4 region.
Inadequate coverage of the V9 region would adversely affect the ability
of the reference-based methods to assign sequences to OTUs. Third, our
previous analysis has shown that the V9 region evolves at a rate much
slower than the rest of the gene (Schloss, 2010). With these points in
mind, we compared the clustering assignment for each of these methods
using a time series experiment that was obtained using mouse feces
(Schloss et al., 2012; Kozich et al., 2013). The MiSeq platform was used
to generate 2,825,000 sequences from the V4 region of the 16S rRNA gene
of 360 samples. Parallel sequencing of a mock community indicated that
the sequencing error rate was approximately 0.02\% (Kozich et al.,
2013). Although no dataset is perfect for exhaustively testing these
clustering methods, this dataset was useful for demonstrating several
points. First, when using 60\% of the data, the stability relationships
amongst the different methods were similar to what we observed using the
Canadian soil dataset (Figure 3AB). With the exception for the clusters
generated using CL, the methods all performed very well with stabilities
greater than 0.91. Second, the MCC values calculated relative to the
distances between sequences were generally higher than was observed for
the Canadian soil dataset for all of the methods except the CL and SL
methods. Surprisingly, the MCC values for the DGC (0.77) and AGC (0.76)
methods were comparable to the AL method (0.76; Figure 3CD). This result
suggests that the optimal method is likely to be database-dependent.

Finally, as was observed with the Canadian soil dataset, there was
little variation in MCC values observed among the 30 randomizations.
Therefore, although the methods have a stochastic component, the OTU
assignments did not vary meaningfully between runs. The results from
both the Canadian soil and murine microbiota datasets demonstrate that
the \emph{de novo} methods can generate stable OTU assignments and that
the the overall quality of the assignments were consistent. Most
importantly, these analyses demonstrate that the OTU assignments using
the AL, AGC, and DGC \emph{de novo} methods were consistently better
than either of the reference-based methods.

\textbf{\emph{How does the stability and quality of OTU assignments
generated by new methods compare to those analyzed in the He study?}}
The Swarm algorithm is a recently proposed \emph{de novo} method for
assigning sequences to OTUs that uses user-defined parameters to break
up chains generated by SL clustering (Mahé et al., 2014). Swarm was
originally validated by comparing the results against the expected
clusters formed based on the taxonomic composition of a mock community.
Similar to the authors of the He study, the Swarm developers suggest
that methods are needed that are insensitive to input order. Use of
Swarm on the Canadian soil and murine datasets demonstrated that similar
to the other \emph{de novo} methods, Swarm's OTU assignments changed as
sequences were added (Figures 1A and 3A). When we compared the OTU
assignments for both datasets when using all of the sequence data, the
variation in MCC values across the 30 randomizations were not
meaningfully different (Figures 1D and 3D). Most importantly, when we
selected the distance threshold that optimized the MCC value, the
quality of the OTU assignments was close to that of the AL assignments
when using the Canadian soil dataset and considerably worse than that of
the murine dataset (Figures 1D and 3D). Interestingly, the distance
thresholds that resulted in the largest MCC values were 3 and 2\% for
the Canadian soil and murine datasets, respectively. This suggests that
distance-based OTU definitions are not consistent across datasets when
using the Swarm algorithm, although they do appear to be within the
neighborhood of 3\%. Finally, the Swarm developers indicated that
hierarchical \emph{de novo} algorithms were too impractical to use on
large MiSeq-generated datasets. Our ability to apply AL to the large
mouse dataset and even larger datasets suggests that it is not necessary
to sacrifice OTU assignment quality for speed (e.g. Schubert, Sinani \&
Schloss, 2015; Zackular et al., 2015).

\textbf{\emph{Are there open-source alternatives to USEARCH that perform
just as well?}} For some datasets the AGC and DGC methods appear to
perform as well or better than the hierarchical clustering methods. As
originally described in the He study, the AGC and DGC methods utilized
the USEARCH program and the DGC method is used for clustering in UPARSE
(Edgar, 2010, 2013). The source code for USEARCH is not publicly
available and only the 32-bit executables are available for free to
academic users. Access for non-academic users and those needing the
64-bit version is available commercially from the developer. An
alternative to USEARCH is VSEARCH, which is being developed in parallel
to USEARCH as an open-source alternative. One subtle difference between
the two programs is that USEARCH employs a heuristic to generate
candidate alignments whereas VSEARCH generates the actual global
alignments. The VSEARCH developers claim that this difference enhances
the sensitivity of VSEARCH relative to USEARCH. Using the two datasets,
we determined whether the AGC and DGC methods, as implemented by the two
programs, yielded OTU assignments of similar quality. In general the
overall trends that we observed with the USEARCH-version of AGC and DGC
were also observed with the VSEARCH-version of the methods (Figure 4).
When we compared the two implementations of the AGC and DGC methods, the
OTUs generated by the VSEARCH-version of the methods were as stable or
more stable than the USEARCH-version when using 60\% of the datasets. In
addition, the MCC values for the entire datasets, calculated relative to
the distance matrix, were virtually indistinguishable. These results are
a strong indication that VSEARCH is a suitable and possibly better
option for executing the AGC and DGC methods.

\textbf{\emph{Is the stability of reference-based methods impacted by
the order of the reference sequences?}} The He study and our replication
attempt validated that the closed-reference method generated perfectly
stable OTUs. This was unsurprising since, by definition, the method is
designed to return one-to-one mapping of reads to a reference.
Furthermore, because it treats the input sequences independently the
input order or use of a random number generator is not an issue. An
important test that was not performed in the He study was to determine
whether the clustering was sensitive to the order of the sequences in
the database. The default database used in QIIME, which was also used in
the He study, contains full-length sequences that are at most 97\%
similar to each other. We randomized the order of the reference
sequences 30 times and used them to carry out the closed-reference
method with the full murine dataset, which contained 32,106 unique
sequences (Figure 5). Surprisingly, we observed that the number of OTUs
generated was not the same in each of the randomizations. On average
there were 28,059 sequences that mapped to a reference OTU per
randomization (range from 28,007 to 28,111). The original ordering of
the reference resulted in 27,876 sequences being mapped, less than the
minimum observed number of mapped sequences when the references were
randomized. This surprising result was likely due to the performance of
the USEARCH heuristic. To test this further, we substituted VSEARCH for
USEARCH in the closed-reference method. When we used VSEARCH the
original ordering of the reference sequences and all randomizations were
able to map 27,737 sequences to reference OTUs. When we calculated the
true distance between each of the murine sequences and the references,
we were able to map 28,238 of the murine sequences to the reference
sequences when using a 97\% similarity threshold without the use of a
heuristic. This result indicates that the closed reference approach,
whether using USEARCH or VSEARCH, does not exhaustively or accurately
map reads to the closest reference. To quantify this further, we
calculated the MCC for the USEARCH and VSEARCH assignments relative to
the assignments using the non-heuristic approach. Using USEARCH the
average MCC was 0.78 (range: 0.75 to 0.80) and using VSEARCH the average
MCC was 0.65 (range: 0.64 to 0.66). The two methods had similar
sensitivities (USEARCH: 0.98 and VSEARCH: 0.97), but the USEARCH
specificity (0.73) was considerably higher than VSEARCH (0.60). Overall,
these results indicate that although heuristic approaches may be fast,
they do a poor job of mapping reads to the correct reference sequence
relative to non-heuristic approaches.

We also observed that regardless of whether we used USEARCH or VSEARCH,
the reference OTU labels that were assigned to each OTU differed between
randomizations. When we used USEARCH to perform closed-reference
clustering, an average of 57.38\% of the labels were shared between
pairs of the 30 randomizations (range=56.14 to 59.55\%). If we instead
used VSEARCH an average of 56.23\% of the labels were shared between
pairs of the 30 randomizations (range=53.48 to 59.12\%). To better
understand this result, we further analyzed QIIME's reference database.
We hypothesized that within a given region there would be sequences that
were more than 97\% similar and possibly identical to each other. When a
sequence was used to search the randomized databases, it would encounter
a different reference sequence as the first match with each
randomization. Among the 99,310 reference sequences that fully overlap
the V4 region, there were 7,785 pairs of sequences that were more than
97\% similar to each other over the full length of the 16S rRNA gene.
When the extracted V4 sequences were dereplicated, we identified 88,347
unique sequences. Among these dereplicated V4 sequences there were
311,430 pairs of sequences that were more than 97\% similar to each
other. The presence of duplicate and highly similar V4 reference
sequences explains the lack of labeling stability when using either
USEARCH or VSEARCH to carry out the closed-reference method. We suspect
that the reference database was designed to only include sequences that
were at most 97\% similar to each other as a way to overcome the
limitations of the USEARCH search heuristic.

Beyond comparing the abundance of specific OTUs across samples, the
reference database is used in the open and closed-reference methods to
generate OTU labels that can be used in several downstream applications.
It is commonly used to extract information from a reference phylogenetic
tree to carrying out UniFrac-based analyses (Hamady, Lozupone \& Knight,
2009) and to identify reference genomes for performing analyses such as
PICRUSt (Langille et al., 2013). Because these downstream applications
depend on the correct and unique labeling of the OTUs, the lack of label
stability is problematic. As one illustration of the effects that
incorrect labels would have on an analysis, we asked whether the
duplicate sequences had the same taxonomies. Among the 3,132 V4
reference sequences that had one duplicate, 443 had discordant
taxonomies. Furthermore, among those 1,699 V4 reference sequences with
two or more duplicates, 698 had discordant taxonomies. Two V4 reference
sequences mapped to 30 and 10 duplicate sequences and both contained 7
different taxonomies. Among the V4 sequences within the database, there
was also a sequence that had 131 duplicates and represented 5 different
taxonomies. When we analyzed the 28,238 sequences that mapped to the V4
reference sequences using a non-heuristic approach, we observed that
18,315 of the sequences mapped to more than one reference sequence. Of
these sequences, 13,378 (73.04\%) mapped to references that were
identical over the V4 region and 4,937 (26.96\%) mapped equally well to
two or more references that were not identical over the V4 region. Among
the combined 18,315 sequences that mapped to multiple reference
sequences, the taxonomy of the multiple reference sequences conflicted
for 3,637 (19.86\%). Together, these results demonstrate some of the
considerable problems with the reference-based clustering of sequences.

\subsection{Conclusions}\label{conclusions}

It is worth noting that the analysis from the Roesch study that
motivated the He study is not typical of microbial ecology studies.
First, their analysis was based on a single soil sample. Researchers
generally have dozens or hundreds of samples that are pooled and
clustered together to enable comparison across samples. Second, all of
the sequence data from these datasets is pooled for a single analysis.
Rarely would a researcher rarefy their data prior to clustering since it
can be more efficiently done after all of the data are assigned to OTUs.
Third, the CL method used in the original Roesch study has since been
shown to not generate optimal OTUs (Schloss \& Westcott, 2011). As for
the approach used in the He study, the value of identifying stable OTUs
is unclear. Although there is concern that running the methods multiple
times yields different clusterings, we have shown that there is little
variation in their quality. This suggests that the different clusterings
by the same method are equally good. Greater emphasis should be placed
on obtaining an optimal balance between splitting similar sequences into
separate OTUs and merging disparate sequences into the same OTU.

The approach of the current study quantified the effects of merging and
splitting by using an objective metric. Through the use of the pairwise
distances between sequences, we were able to use the MCC to demonstrate
that, in general, the AL method was consistently the optimal method for
each dataset, but that Swarm, AGC, and DGC sometimes perform as well as
AL. At least for the murine dataset, Swarm also could be among the
methods that performed poorly. It is impossible to obtain a clustering
with no false positives or false negatives and the optimal method may
vary by dataset. With this in mind, researchers are encouraged to
calculate and report their MCC values and to use these values to justify
using methods other than the AL. As an alternative to the He study's
method of measuring stability, we propose using the variation in the
quality of the clustering of the full dataset. Given the tight 95\%
confidence intervals shown in Figures 1D and 3D, with the exception of
CL, it is clear that this variation is quite small. This indicates that
although the order of the sequences being clustered can affect the
actual cluster assignments, the quality of those different clusterings
is not meaningfully different.

Our analysis of those methods that implemented USEARCH as a method for
clustering sequences revealed that its heuristic limited its
specificity. When we replaced USEARCH with VSEARCH, the clustering
quality was as good or better. Although there may be parameters in
USEARCH that can be tuned to improve the heuristic, these parameters are
likely dataset dependent. Based on the data presented in this study, its
availability as an open source, and free program, VSEARCH should replace
USEARCH in the \emph{de novo} clustering methods; however, USEARCH
performed better than VSEARCH for closed-reference clustering.
Furthermore, although not tested in our study, VSEARCH can be
parallelized leading to potentially significant improvements in speed.
Although USEARCH and VSEARCH do not utilize aligned sequences, it is
important to note that a sequence curation pipeline including denoising,
alignment, trimming to a consistent region of the 16S rRNA gene, and
chimera checking are critical to making proper inferences (Schloss,
Gevers \& Westcott, 2011; Schloss, 2012; Kozich et al., 2013).

We have assessed the ability of reference-based clustering methods to
capture the actual distance between the sequences in a dataset in
parallel with \emph{de novo} methods. Several studies have lauded both
the open and closed-reference approaches for generating reproducible
clusterings (Navas-Molina et al., 2013; Rideout et al., 2014; He et al.,
2015), yet we have shown that both reference-based approaches did a poor
job of representing the distance between the sequences compared to the
\emph{de novo} approaches. Although the OTU assignments are reproducible
and stable across a range of library sizes, the reference-based OTU
assignments are a poor representation of the data. We also observed that
the assignments were not actually reproducible when the order of the
reference sequences was randomized. When USEARCH was used, the actual
number of sequences that mapped to the reference changed depended on the
order of the reference. Perhaps most alarming was that the default order
of the database provided the worst MCC of any of the randomizations we
attempted. This has the potential to introduce systematic a bias rather
than a random error. Even when we used VSEARCH to perform
closed-reference clustering and were able to obtain a consistent
clusterings, we observed that the labels on the OTUs differed between
randomizations. Because the OTU labels are frequently used to identify
representative sequences for those OTUs, variation in labels, often
representing different taxonomic groups, will have a detrimental effect
on the interpretation of downstream analyses.

Because the open-reference method is a hybrid of the closed-reference
and DGC methods, it is also negatively affected by the various problems
using USEARCH. An added problem with the open-reference method is that
the two phases of the method employ different thresholds to define its
OTUs. In the closed-reference step, sequences must be within a threshold
of a reference to be in the same OTU. This means that in the worst case
scenario two sequences that are 97\% similar to a reference and are
joined into the same OTU, may only be 94\% similar to each other. In the
DGC step, the goal is to approximate the AL method which requires that,
on average, the sequences within an OTU are, on average, 97\% similar to
each other. The end result of the open-reference approach is that
sequences that are similar to previously observed sequences are
clustered with one threshold while those that are not similar to
previously observed sequences are clustered with a different threshold.

As the throughput of sequencing technologies have improved, development
of clustering algorithms must continue to keep pace. \emph{De novo}
clustering methods are considerably slower and more computationally
intensive than reference-based methods and the greedy \emph{de novo}
methods are faster than the hierarchical methods. In our experience
(Kozich et al., 2013), the most significant detriment to execution speed
of the \emph{de novo} methods has been the inadequate removal of
sequencing error and chimeras. As the rate of sequencing error increases
so do the number of unique sequences that must be clustered. The speed
of the \emph{de novo} methods scales approximately quadratically, so
that doubling the number of sequences results in a four-fold increase in
the time required to execute the method. The rapid expansion in
sequencing throughput has been likened to the Red Queen in Lewis
Carroll's, \emph{Through the Looking-Glass} who must run in place to
keep up to her changing surroundings (Schloss et al., 2009). Microbial
ecologists must continue to refine clustering methods to better handle
the size of the datasets, but they must also take steps to improve the
quality of the underlying data. Ultimately, objective standards must be
applied to assess the quality of the data and the quality of OTU
clustering.

\subsection{Methods}\label{methods}

\textbf{\emph{454 FLX-generated Roesch Canadian soil dataset.}} After
obtaining the 16S rRNA gene fragments from GenBank (accessions
EF308591-EF361836), we followed the methods outlined by the He study by
removing any sequence that contained an ambiguous base, was identified
as being a chimera, and fell outside a defined sequence length. Although
they reported observing a total of 50,542 sequences that were
represented by 13,293 unique sequences, we obtained a total of 50,946
sequences that were represented by 13,393 unique sequences. Similar to
the He study, we randomly sampled, without replacement, 20, 40, 60, and
80\% of the sequences from the full data set. The random sampling was
repeated 30 times. The order of the sequences in the full dataset was
randomly permuted without replacement to generate an additional 30
datasets. To perform the hierarchical clustering methods and to generate
a distance matrix we followed the approach of the He study by
calculating distances based on pairwise global alignments using the
\texttt{pairwise.dist} command in mothur using the default
Needleman-Wunsch alignment method and parameters. It should be noted
that this approach has been strongly discouraged (Schloss, 2012).
Execution of the hierarchical clustering methods was performed as
described in the original He study using mothur (v.1.37) and using the
QIIME (v.1.9.1) parameter profiles provided in the supplementary
material from the He study for the greedy and reference-based clustering
methods.

\textbf{\emph{MiSeq-generated Murine gut microbiota dataset.}} The
murine 16S rRNA gene sequence data generated from the V4 region using an
Illumina MiSeq was obtained from
\url{http:/www.mothur.org/MiSeqDevelopmentData/StabilityNoMetaG.tar} and
was processed as outlined in the original study (Kozich et al., 2013).
Briefly, 250-nt read pairs were assembled into contigs by aligning the
reads and correcting discordant base calls by requiring one of the base
calls to have a Phred quality score at least 6 points higher than the
other. Sequences where it was not possible to resolve the disagreement
were culled from the dataset. The sequences were then aligned to a SILVA
reference alignment (Pruesse et al., 2007) and any reads that aligned
outside of the V4 region were removed from the dataset. Sequences were
pre-clustered by combining the abundances of sequences that differed by
2 or fewer nucleotides of a more abundant sequence. Each of the samples
was then screened for chimeric sequences using the default parameters in
UCHIME (Edgar et al., 2011). The resulting sequences were processed in
the same manner as the Canadian soil dataset with the exception that the
distance matrices were calculated based on the SILVA-based alignment.

\textbf{\emph{Analysis of reference database.}} We utilized the 97\%
OTUs greengenes reference sequence and taxonomy data (v.13.8) that
accompanies the QIIME installation. Because the greengenes reference
alignment does a poor job of representing the secondary structure of the
16S rRNA gene (Schloss, 2010), we realigned the FASTA sequences to a
SILVA reference alignment to identify the V4 region of the sequences.

\textbf{\emph{Calculation of Matthew's Correlation Coefficient (MCC).}}
The MCC was calculated by two approaches in this study using only the
dereplicated sequence lists. First, we calculated the MCC to determine
the stability of OTU assignments following the approach of the He study.
We assumed that the clusters obtained from the 30 randomized full
datasets were correct. We counted the number of sequence pairs that were
in the same OTU for the subsetted dataset and the full dataset
(i.e.~true positives; TP), that were in different OTUs for the subsetted
dataset and the full dataset (i.e.~true negatives; TN), that were in the
same OTU for the subsetted dataset and different OTUs in the full
dataset (i.e.~false positives; FP), and that were in different OTUs for
the subsetted dataset and the same OTU in the full dataset (i.e.~false
negatives; FN). For each set of 30 random subsamplings of the dataset,
we counted these parameters against the 30 randomizations of the full
dataset. This gave 900 comparisons for each fraction of sequences being
used in the analysis. The Matthew's correlation coefficient was then
calculated as:

\[
MCC = \frac{TP \times TN - FP \times FN}{\sqrt{(TP +FP)(TP+FN)(TN+FP)(TN+FN)} }
\]

Second, we calculated the MCC to determine the quality of the
clusterings as previously described (Schloss \& Westcott, 2011).
Briefly, we compared the OTU assignments for pairs of sequences to the
distance matrix that was calculated between all pairs of aligned
sequences. For each dataset that was clustered, those pairs of sequences
that were in the same OTU and had a distance less than 3\% were TPs,
those that were in different OTUs and had a distance greater than 3\%
were TNs, those that were in the same OTU and had a distance greater
than 3\% were FPs, and those that were in different OTUs and had a
distance less than 3\% were FNs. The MCC was counted for each dataset
using the formula above as implemented in the \texttt{sens.spec} command
in mothur. To judge the quality of the Swarm-generated OTU assignments
we calculated the MCC value using thresholds incremented by 1\% between
0 and 5\% and selected the threshold that provided the optimal MCC
value.

\textbf{\emph{Software availability.}} A reproducible workflow including
all scripts and this manuscript as a literate programming document are
available at
\url{https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015}. The
workflow utilized QIIME (v.1.9.1; Caporaso et al., 2010), mothur
(v.1.37.0; Schloss et al., 2009), USEARCH (v.6.1; Edgar, 2010), VSEARCH
(v.1.5.0; Rognes et al., 2015), Swarm (v.2.1.1; Mahé et al., 2014), and
R (v.3.2.0; R Core Team, 2015). The SL, AL, and CL methods are called
nearest neighbor (NN), average neighbor (AN), and furthest neighbor (FN)
in mothur; we have used the terminology from the He study to minimize
confusion. The knitr (v.1.10.5; Xie, 2013), Rcpp (v. 0.11.6;
Eddelbuettel, 2013), rentrez (v. 1.0.0; Winter, Chamberlain \&
Guangchun, 2015), and jsonlite (v. 0.9.16; Ooms, 2014) packages were
used within R.

\newpage

\subsection{Figures}\label{figures}

\textbf{Figure 1. Comparison of the stability (A, B) and quality (C, D)
of \emph{de novo} and reference-based clustering methods using the
Canadian soil dataset.} The average stability of the OTUs was determined
by calculating the MCC with respect to the OTU assignments for the full
dataset using varying sized subsamples. The quality of the OTUs was
determined by calculating the MCC with respect to the distances between
the sequences using varying sized subsamples. Thirty randomizations were
performed for each fraction of the dataset and the average and 95\%
confidence interval are presented when using 60\% of the data. The
vertical gray lines in A and C indicates the fraction of the dataset
represented in B and D, respectively. The color and shape of the
plotting symbol is the same between the different panels and is
described along the x-axis of panel D. The optimum threshold for the
Swarm-generated assignments was 3\%.

\textbf{Figure 2. The clustering methods varied in their ability to
generate the same number of OTUs using a subset of the data as were
observed when the full dataset was rarefied.} The subsetted data are
depicted by closed circles and the data from the rarefied full dataset
is depicted by the open circles.

\textbf{Figure 3. Comparison of the stability (A, B) and quality (C, D)
of \emph{de novo} and reference-based clustering methods using the
murine dataset.} The average stability of the OTUs was determined by
calculating the MCC with respect to the OTU assignments for the full
dataset using varying sized subsamples. The quality of the OTUs was
determined by calculating the MCC with respect to the distances between
the sequences using varying sized subsamples. Thirty randomizations were
performed for each fraction of the dataset and the average and 95\%
confidence interval are presented when using 60\% of the data. The
vertical gray lines in A and C indicates the fraction of the dataset
represented in B and D, respectively. The color and shape of the
plotting symbol is the same between the different panels and is
described along the x-axis of panel D. The optimum threshold for the
Swarm-generated assignments was 2\%.

\textbf{Figure 4. The stability and quality of USEARCH and VSEARCH OTUs
generated by the AGC and DGC methods were similar.} The stability of the
OTUs was determined by calculating the MCC for OTUs calculated using
60\% of the data relative to the OTU assignments for the full dataset.
The quality of the OTUs was determined by calculating the MCC of the
OTUs calculated using the full dataset with respect to the distances
between the sequences. The error bars represent the 95\% confidence
interval across the 30 randomizations.

\textbf{Figure 5. The number of closed-reference OTUs observed in the
murine dataset when using USEARCH, VSEARCH, and without a heuristic.} In
addition to the default ordering of the references provided with the
QIIME package, the reference sequences were randomized 30 times; the
order of the murine dataset was not randomized. Regardless of whether
the default or randomized ordering was used, the number of OTUs
generated using VSEARCH did not differ. The non-heuristic approach
calculated the exact distance between the murine sequences and the the
reference sequences and assigned the sequences to the reference with the
smallest distance.

\newpage

\includepdf{results/figures/figure_1.pdf}
\includepdf{results/figures/figure_2.pdf}
\includepdf{results/figures/figure_3.pdf}
\includepdf{results/figures/figure_4.pdf}

\newpage

\hyperdef{}{references}{\label{references}}
\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\hyperdef{}{ref-Anderson2001}{\label{ref-Anderson2001}}
Anderson MJ. 2001. A new method for non-parametric multivariate analysis
of variance. \emph{Austral Ecology} 26:32--46. DOI:
\url{http://doi.org/10.1111/j.1442-9993.2001.01070.pp.x}.

\hyperdef{}{ref-Barriuso2011}{\label{ref-Barriuso2011}}
Barriuso J., Valverde JR., Mellado RP. 2011. Estimation of bacterial
diversity using next generation sequencing of 16S rDNA: A comparison of
different workflows. \emph{BMC Bioinformatics} 12:473. DOI:
\url{http://doi.org/10.1186/1471-2105-12-473}.

\hyperdef{}{ref-Bonder2012}{\label{ref-Bonder2012}}
Bonder MJ., Abeln S., Zaura E., Brandt BW. 2012. Comparing clustering
and pre-processing in taxonomy analysis. \emph{Bioinformatics}
28:2891--2897. DOI: \url{http://doi.org/10.1093/bioinformatics/bts552}.

\hyperdef{}{ref-Cai2011}{\label{ref-Cai2011}}
Cai Y., Sun Y. 2011. ESPRIT-tree: Hierarchical clustering analysis of
millions of 16S rRNA pyrosequences in quasilinear computational time.
\emph{Nucleic Acids Research} 39:e95--e95. DOI:
\url{http://doi.org/10.1093/nar/gkr349}.

\hyperdef{}{ref-Caporaso2010}{\label{ref-Caporaso2010}}
Caporaso JG., Kuczynski J., Stombaugh J., Bittinger K., Bushman FD.,
Costello EK., Fierer N., Peña AG., Goodrich JK., Gordon JI., Huttley
GA., Kelley ST., Knights D., Koenig JE., Ley RE., Lozupone CA., McDonald
D., Muegge BD., Pirrung M., Reeder J., Sevinsky JR., Turnbaugh PJ.,
Walters WA., Widmann J., Yatsunenko T., Zaneveld J., Knight R. 2010.
QIIME allows analysis of high-throughput community sequencing data.
\emph{Nature Methods} 7:335--336. DOI:
\url{http://doi.org/10.1038/nmeth.f.303}.

\hyperdef{}{ref-Chen2013}{\label{ref-Chen2013}}
Chen W., Zhang CK., Cheng Y., Zhang S., Zhao H. 2013. A comparison of
methods for clustering 16S rRNA sequences into OTUs. \emph{PLoS ONE}
8:e70837. DOI: \url{http://doi.org/10.1371/journal.pone.0070837}.

\hyperdef{}{ref-eddelbuettel2013seamless}{\label{ref-eddelbuettel2013seamless}}
Eddelbuettel D. 2013. \emph{Seamless R and C++ integration with Rcpp}.
New York: Springer.

\hyperdef{}{ref-Edgar2010}{\label{ref-Edgar2010}}
Edgar RC. 2010. Search and clustering orders of magnitude faster than
BLAST. \emph{Bioinformatics} 26:2460--2461. DOI:
\url{http://doi.org/10.1093/bioinformatics/btq461}.

\hyperdef{}{ref-Edgar2011}{\label{ref-Edgar2011}}
Edgar RC., Haas BJ., Clemente JC., Quince C., Knight R. 2011. UCHIME
improves sensitivity and speed of chimera detection.
\emph{Bioinformatics} 27:2194--2200. DOI:
\url{http://doi.org/10.1093/bioinformatics/btr381}.

\hyperdef{}{ref-Edgar2013}{\label{ref-Edgar2013}}
Edgar RC. 2013. UPARSE: Highly accurate OTU sequences from microbial
amplicon reads. \emph{Nature Methods} 10:996--998. DOI:
\url{http://doi.org/10.1038/nmeth.2604}.

\hyperdef{}{ref-Gilbert2011}{\label{ref-Gilbert2011}}
Gilbert JA., Steele JA., Caporaso JG., Steinbrück L., Reeder J.,
Temperton B., Huse S., McHardy AC., Knight R., Joint I., Somerfield P.,
Fuhrman JA., Field D. 2011. Defining seasonal marine microbial community
dynamics. \emph{The ISME Journal} 6:298--308. DOI:
\url{http://doi.org/10.1038/ismej.2011.107}.

\hyperdef{}{ref-Hamady2009}{\label{ref-Hamady2009}}
Hamady M., Lozupone C., Knight R. 2009. Fast UniFrac: Facilitating
high-throughput phylogenetic analyses of microbial communities including
analysis of pyrosequencing and PhyloChip data. \emph{The ISME Journal}
4:17--27. DOI: \url{http://doi.org/10.1038/ismej.2009.97}.

\hyperdef{}{ref-He2015}{\label{ref-He2015}}
He Y., Caporaso JG., Jiang X-T., Sheng H-F., Huse SM., Rideout JR.,
Edgar RC., Kopylova E., Walters WA., Knight R., Zhou H-W. 2015.
Stability of operational taxonomic units: An important but neglected
property for analyzing microbial diversity. \emph{Microbiome} 3. DOI:
\url{http://doi.org/10.1186/s40168-015-0081-x}.

\hyperdef{}{ref-Huse2010}{\label{ref-Huse2010}}
Huse SM., Welch DM., Morrison HG., Sogin ML. 2010. Ironing out the
wrinkles in the rare biosphere through improved OTU clustering.
\emph{Environmental Microbiology} 12:1889--1898. DOI:
\url{http://doi.org/10.1111/j.1462-2920.2010.02193.x}.

\hyperdef{}{ref-Huttenhower2012}{\label{ref-Huttenhower2012}}
Huttenhower C., Gevers D., Knight R., Abubucker S., Badger JH.,
Chinwalla AT., Creasy HH., Earl AM., FitzGerald MG., Fulton RS., Giglio
MG., Hallsworth-Pepin K., Lobos EA., Madupu R., Magrini V., Martin JC.,
Mitreva M., Muzny DM., Sodergren EJ., Versalovic J., Wollam AM., Worley
KC., Wortman JR., Young SK., Zeng Q., Aagaard KM., Abolude OO.,
Allen-Vercoe E., Alm EJ., Alvarado L., Andersen GL., Anderson S.,
Appelbaum E., Arachchi HM., Armitage G., Arze CA., Ayvaz T., Baker CC.,
Begg L., Belachew T., Bhonagiri V., Bihan M., Blaser MJ., Bloom T.,
Bonazzi V., Brooks JP., Buck GA., Buhay CJ., Busam DA., Campbell JL.,
Canon SR., Cantarel BL., Chain PSG., Chen I-MA., Chen L., Chhibba S.,
Chu K., Ciulla DM., Clemente JC., Clifton SW., Conlan S., Crabtree J.,
Cutting MA., Davidovics NJ., Davis CC., DeSantis TZ., Deal C.,
Delehaunty KD., Dewhirst FE., Deych E., Ding Y., Dooling DJ., Dugan SP.,
Dunne WM., Durkin AS., Edgar RC., Erlich RL., Farmer CN., Farrell RM.,
Faust K., Feldgarden M., Felix VM., Fisher S., Fodor AA., Forney LJ.,
Foster L., Francesco VD., Friedman J., Friedrich DC., Fronick CC.,
Fulton LL., Gao H., Garcia N., Giannoukos G., Giblin C., Giovanni MY.,
Goldberg JM., Goll J., Gonzalez A., Griggs A., Gujja S., Haake SK., Haas
BJ., Hamilton HA., Harris EL., Hepburn TA., Herter B., Hoffmann DE.,
Holder ME., Howarth C., Huang KH., Huse SM., Izard J., Jansson JK.,
Jiang H., Jordan C., Joshi V., Katancik JA., Keitel WA., Kelley ST.,
Kells C., King NB., Knights D., Kong HH., Koren O., Koren S., Kota KC.,
Kovar CL., Kyrpides NC., Rosa PSL., Lee SL., Lemon KP., Lennon N., Lewis
CM., Lewis L., Ley RE., Li K., Liolios K., Liu B., Liu Y., Lo C-C.,
Lozupone CA., Lunsford RD., Madden T., Mahurkar AA., Mannon PJ., Mardis
ER., Markowitz VM., Mavromatis K., McCorrison JM., McDonald D., McEwen
J., McGuire AL., McInnes P., Mehta T., Mihindukulasuriya KA., Miller
JR., Minx PJ., Newsham I., Nusbaum C., O'Laughlin M., Orvis J., Pagani
I., Palaniappan K., Patel SM., Pearson M., Peterson J., Podar M., Pohl
C., Pollard KS., Pop M., Priest ME., Proctor LM., Qin X., Raes J., Ravel
J., Reid JG., Rho M., Rhodes R., Riehle KP., Rivera MC.,
Rodriguez-Mueller B., Rogers Y-H., Ross MC., Russ C., Sanka RK., Sankar
P., Sathirapongsasuti JF., Schloss JA., Schloss PD., Schmidt TM., Scholz
M., Schriml L., Schubert AM., Segata N., Segre JA., Shannon WD., Sharp
RR., Sharpton TJ., Shenoy N., Sheth NU., Simone GA., Singh I., Smillie
CS., Sobel JD., Sommer DD., Spicer P., Sutton GG., Sykes SM., Tabbaa
DG., Thiagarajan M., Tomlinson CM., Torralba M., Treangen TJ., Truty
RM., Vishnivetskaya TA., Walker J., Wang L., Wang Z., Ward DV., Warren
W., Watson MA., Wellington C., Wetterstrand KA., White JR.,
Wilczek-Boney K., Wu Y., Wylie KM., Wylie T., Yandava C., Ye L., Ye Y.,
Yooseph S., Youmans BP., Zhang L., Zhou Y., Zhu Y., Zoloth L., Zucker
JD., Birren BW., Gibbs RA., Highlander SK., Methé BA., Nelson KE.,
Petrosino JF., Weinstock GM., Wilson RK., White O. 2012. Structure,
function and diversity of the healthy human microbiome. \emph{Nature}
486:207--214. DOI: \url{http://doi.org/10.1038/nature11234}.

\hyperdef{}{ref-Kim2011}{\label{ref-Kim2011}}
Kim M., Morrison M., Yu Z. 2011. Evaluation of different partial 16S
rRNA gene sequence regions for phylogenetic analysis of microbiomes.
\emph{Journal of Microbiological Methods} 84:81--87. DOI:
\url{http://doi.org/10.1016/j.mimet.2010.10.020}.

\hyperdef{}{ref-Koeppel2013}{\label{ref-Koeppel2013}}
Koeppel AF., Wu M. 2013. Surprisingly extensive mixed phylogenetic and
ecological signals among bacterial operational taxonomic units.
\emph{Nucleic Acids Research} 41:5175--5188. DOI:
\url{http://doi.org/10.1093/nar/gkt241}.

\hyperdef{}{ref-Kozich2013}{\label{ref-Kozich2013}}
Kozich JJ., Westcott SL., Baxter NT., Highlander SK., Schloss PD. 2013.
Development of a dual-index sequencing strategy and curation pipeline
for analyzing amplicon sequence data on the MiSeq Illumina sequencing
platform. \emph{Applied and Environmental Microbiology} 79:5112--5120.
DOI: \url{http://doi.org/10.1128/aem.01043-13}.

\hyperdef{}{ref-Langille2013}{\label{ref-Langille2013}}
Langille MGI., Zaneveld J., Caporaso JG., McDonald D., Knights D., Reyes
JA., Clemente JC., Burkepile DE., Thurber RLV., Knight R., Beiko RG.,
Huttenhower C. 2013. Predictive functional profiling of microbial
communities using 16S rRNA marker gene sequences. \emph{Nat Biotechnol}
31:814--821. DOI: \url{http://doi.org/10.1038/nbt.2676}.

\hyperdef{}{ref-Mah2014}{\label{ref-Mah2014}}
Mahé F., Rognes T., Quince C., Vargas C de., Dunthorn M. 2014. Swarm:
Robust and fast clustering method for amplicon-based studies.
\emph{PeerJ} 2:e593. DOI: \url{http://doi.org/10.7717/peerj.593}.

\hyperdef{}{ref-Matthews1975}{\label{ref-Matthews1975}}
Matthews B. 1975. Comparison of the predicted and observed secondary
structure of t4 phage lysozyme. \emph{Biochimica et Biophysica Acta
(BBA) - Protein Structure} 405:442--451. DOI:
\url{http://doi.org/10.1016/0005-2795(75)90109-9}.

\hyperdef{}{ref-May2014}{\label{ref-May2014}}
May A., Abeln S., Crielaard W., Heringa J., Brandt BW. 2014. Unraveling
the outcome of 16S rDNA-based taxonomy analysis through mock data and
simulations. \emph{Bioinformatics} 30:1530--1538. DOI:
\url{http://doi.org/10.1093/bioinformatics/btu085}.

\hyperdef{}{ref-NavasMolina2013}{\label{ref-NavasMolina2013}}
Navas-Molina JA., Peralta-Sánchez JM., González A., McMurdie PJ.,
Vázquez-Baeza Y., Xu Z., Ursell LK., Lauber C., Zhou H., Song SJ.,
Huntley J., Ackermann GL., Berg-Lyons D., Holmes S., Caporaso JG.,
Knight R. 2013. Advancing our understanding of the human microbiome
using QIIME. In: \emph{Methods in enzymology}. Elsevier BV, 371--444.
DOI: \url{http://doi.org/10.1016/b978-0-12-407863-5.00019-8}.

\hyperdef{}{ref-ooms2014jsonlite}{\label{ref-ooms2014jsonlite}}
Ooms J. 2014. The jsonlite package: A practical and consistent mapping
between JSON data and R objects. \emph{arXiv:1403.2805 {[}stat.CO{]}}.

\hyperdef{}{ref-Preheim2013}{\label{ref-Preheim2013}}
Preheim SP., Perrotta AR., Martin-Platero AM., Gupta A., Alm EJ. 2013.
Distribution-based clustering: Using ecology to refine the operational
taxonomic unit. \emph{Applied and Environmental Microbiology}
79:6593--6603. DOI: \url{http://doi.org/10.1128/aem.00342-13}.

\hyperdef{}{ref-Pruesse2007}{\label{ref-Pruesse2007}}
Pruesse E., Quast C., Knittel K., Fuchs BM., Ludwig W., Peplies J.,
Glockner FO. 2007. SILVA: A comprehensive online resource for quality
checked and aligned ribosomal RNA sequence data compatible with ARB.
\emph{Nucleic Acids Research} 35:7188--7196. DOI:
\url{http://doi.org/10.1093/nar/gkm864}.

\hyperdef{}{ref-language2015}{\label{ref-language2015}}
R Core Team. 2015. R: A language and environment for statistical
computing.

\hyperdef{}{ref-Rideout2014}{\label{ref-Rideout2014}}
Rideout JR., He Y., Navas-Molina JA., Walters WA., Ursell LK., Gibbons
SM., Chase J., McDonald D., Gonzalez A., Robbins-Pianka A., Clemente
JC., Gilbert JA., Huse SM., Zhou H-W., Knight R., Caporaso JG. 2014.
Subsampled open-reference clustering creates consistent, comprehensive
OTU definitions and scales to billions of sequences. \emph{PeerJ}
2:e545. DOI: \url{http://doi.org/10.7717/peerj.545}.

\hyperdef{}{ref-Roesch2007}{\label{ref-Roesch2007}}
Roesch LFW., Fulthorpe RR., Riva A., Casella G., Hadwin AKM., Kent AD.,
Daroub SH., Camargo FAO., Farmerie WG., Triplett EW. 2007.
Pyrosequencing enumerates and contrasts soil microbial diversity.
\emph{The ISME Journal}. DOI:
\url{http://doi.org/10.1038/ismej.2007.53}.

\hyperdef{}{ref-Rognes2015}{\label{ref-Rognes2015}}
Rognes T., Mahé F., Flouri T., McDonald; D. 2015. Vsearch: VSEARCH
1.4.0. DOI: \url{http://doi.org/10.5281/zenodo.31443}.

\hyperdef{}{ref-Schloss2009}{\label{ref-Schloss2009}}
Schloss PD., Westcott SL., Ryabin T., Hall JR., Hartmann M., Hollister
EB., Lesniewski RA., Oakley BB., Parks DH., Robinson CJ., Sahl JW.,
Stres B., Thallinger GG., Horn DJV., Weber CF. 2009. Introducing mothur:
Open-source, platform-independent, community-supported software for
describing and comparing microbial communities. \emph{Applied and
Environmental Microbiology} 75:7537--7541. DOI:
\url{http://doi.org/10.1128/aem.01541-09}.

\hyperdef{}{ref-Schloss2010}{\label{ref-Schloss2010}}
Schloss PD. 2010. The effects of alignment quality, distance calculation
method, sequence filtering, and region on the analysis of 16S rRNA
gene-based studies. \emph{PLoS Comput Biol} 6:e1000844. DOI:
\url{http://doi.org/10.1371/journal.pcbi.1000844}.

\hyperdef{}{ref-Schloss2012Stabilization}{\label{ref-Schloss2012Stabilization}}
Schloss PD., Schubert AM., Zackular JP., Iverson KD., Young VB.,
Petrosino JF. 2012. Stabilization of the murine gut microbiome following
weaning. \emph{Gut Microbes} 3:383--393. DOI:
\url{http://doi.org/10.4161/gmic.21008}.

\hyperdef{}{ref-Schloss2012Secondary}{\label{ref-Schloss2012Secondary}}
Schloss PD. 2012. Secondary structure improves OTU assignments of 16S
rRNA gene sequences. \emph{The ISME Journal} 7:457--460. DOI:
\url{http://doi.org/10.1038/ismej.2012.102}.

\hyperdef{}{ref-Schloss2011Assessing}{\label{ref-Schloss2011Assessing}}
Schloss PD., Westcott SL. 2011. Assessing and improving methods used in
operational taxonomic unit-based approaches for 16S rRNA gene sequence
analysis. \emph{Applied and Environmental Microbiology} 77:3219--3226.
DOI: \url{http://doi.org/10.1128/aem.02810-10}.

\hyperdef{}{ref-Schloss2011Reducing}{\label{ref-Schloss2011Reducing}}
Schloss PD., Gevers D., Westcott SL. 2011. Reducing the effects of PCR
amplification and sequencing artifacts on 16S rRNA-based studies.
\emph{PLoS ONE} 6:e27310. DOI:
\url{http://doi.org/10.1371/journal.pone.0027310}.

\hyperdef{}{ref-Schmidt2014Ecological}{\label{ref-Schmidt2014Ecological}}
Schmidt TSB., Rodrigues JFM., Mering C von. 2014a. Ecological
consistency of SSU rRNA-based operational taxonomic units at a global
scale. \emph{PLoS Comput Biol} 10:e1003594. DOI:
\url{http://doi.org/10.1371/journal.pcbi.1003594}.

\hyperdef{}{ref-Schmidt2014Limits}{\label{ref-Schmidt2014Limits}}
Schmidt TSB., Rodrigues JFM., Mering C von. 2014b. Limits to robustness
and reproducibility in the demarcation of operational taxonomic units.
\emph{Environ Microbiol} 17:1689--1706. DOI:
\url{http://doi.org/10.1111/1462-2920.12610}.

\hyperdef{}{ref-Schubert2015}{\label{ref-Schubert2015}}
Schubert AM., Sinani H., Schloss PD. 2015. Antibiotic-induced
alterations of the murine gut microbiota and subsequent effects on
colonization resistance against *clostridium difficile*. \emph{mBio}
6:e00974--15. DOI: \url{http://doi.org/10.1128/mbio.00974-15}.

\hyperdef{}{ref-Shade2013}{\label{ref-Shade2013}}
Shade A., Klimowicz AK., Spear RN., Linske M., Donato JJ., Hogan CS.,
McManus PS., Handelsman J. 2013. Streptomycin application has no
detectable effect on bacterial community structure in apple orchard
soil. \emph{Applied and Environmental Microbiology} 79:6617--6625. DOI:
\url{http://doi.org/10.1128/aem.02017-13}.

\hyperdef{}{ref-Sun2009}{\label{ref-Sun2009}}
Sun Y., Cai Y., Liu L., Yu F., Farrell ML., McKendree W., Farmerie W.
2009. ESPRIT: Estimating species richness using large collections of 16S
rRNA pyrosequences. \emph{Nucleic Acids Research} 37:e76--e76. DOI:
\url{http://doi.org/10.1093/nar/gkp285}.

\hyperdef{}{ref-Sun2011}{\label{ref-Sun2011}}
Sun Y., Cai Y., Huse SM., Knight R., Farmerie WG., Wang X., Mai V. 2011.
A large-scale benchmark study of existing algorithms for
taxonomy-independent microbial community analysis. \emph{Briefings in
Bioinformatics} 13:107--121. DOI:
\url{http://doi.org/10.1093/bib/bbr009}.

\hyperdef{}{ref-Wang2007}{\label{ref-Wang2007}}
Wang Q., Garrity GM., Tiedje JM., Cole JR. 2007. Naive bayesian
classifier for rapid assignment of rRNA sequences into the new bacterial
taxonomy. \emph{Applied and Environmental Microbiology} 73:5261--5267.
DOI: \url{http://doi.org/10.1128/aem.00062-07}.

\hyperdef{}{ref-White2010}{\label{ref-White2010}}
White JR., Navlakha S., Nagarajan N., Ghodsi M-R., Kingsford C., Pop M.
2010. Alignment and clustering of phylogenetic markers - implications
for microbial diversity studies. \emph{BMC Bioinformatics} 11:152. DOI:
\url{http://doi.org/10.1186/1471-2105-11-152}.

\hyperdef{}{ref-winter2015rentrez}{\label{ref-winter2015rentrez}}
Winter D., Chamberlain S., Guangchun H. 2015. Rentrez 1.0.0. DOI:
\url{http://doi.org/10.5281/zenodo.32420}.

\hyperdef{}{ref-xie2013dynamic}{\label{ref-xie2013dynamic}}
Xie Y. 2013. \emph{Dynamic documents with R and knitr}. Boca Raton,
Florida: Chapman; Hall/CRC.

\hyperdef{}{ref-Zackular2015}{\label{ref-Zackular2015}}
Zackular JP., Baxter NT., Chen GY., Schloss PD. 2015. Manipulation of
the gut microbiota reveals role in colon tumorigenesis. \emph{mSphere}
1:e00001--15. DOI: \url{http://doi.org/10.1128/mSphere.00001-15}.

\end{document}

### 2015-11-14 ResponseToReviewers.md
Reorganize section order

@@
 
-Thank you for the suggestion; we have moved this section to the paragraph in the Results & Discussion section describing the quality of the clustering methods.
+Thank you for the suggestion. Because we needed to reorganize the text, the details have been removed from the Introduction and are now in the Methods, which precedes the Results & Discussion.
 

### 2015-11-14 Schloss_Cluster_PeerJ_2015.Rmd
Remove figures from body of paper
Put figures back in
Reorganize section order

@@
 
+## Methods
+
+***454 FLX-generated Roesch Canadian soil dataset.***
+After obtaining the 16S rRNA gene fragments from GenBank (accessions EF308591-EF361836), we followed the methods outlined by the He study by removing any sequence that contained an ambiguous base, was identified as being a chimera, and fell outside a defined sequence length.
+Although they reported observing a total of 50,542 sequences that were represented by 13,293 unique sequences, we obtained a total of `r length(scan(file="data/he/canada_soil.good.unique.pick.redundant.fasta", what="", sep="\n", quiet=T))/2` sequences that were represented by `r length(scan(file="data/he/canada_soil.good.unique.pick.fasta", what="", sep="\n", quiet=T))/2` unique sequences.
+Similar to the He study, we randomly sampled, without replacement, 20, 40, 60, and 80% of the sequences from the full data set.
+The random sampling was repeated 30 times.
+The order of the sequences in the full dataset was randomly permuted without replacement to generate an additional 30 datasets.
+To perform the hierarchical clustering methods and to generate a distance matrix we followed the approach of the He study by calculating distances based on pairwise global alignments using the `pairwise.dist` command in mothur using the default Needleman-Wunsch alignment method and parameters.
+It should be noted that this approach has been strongly discouraged [@Schloss2012Secondary].
+Execution of the hierarchical clustering methods was performed as described in the original He study using mothur (v.1.37) and using the QIIME (v.1.9.1) parameter profiles provided in the supplementary material from the He study for the greedy and reference-based clustering methods.
+
+***MiSeq-generated Murine gut microbiota dataset.***
+The murine 16S rRNA gene sequence data generated from the V4 region using an Illumina MiSeq was obtained from http:/www.mothur.org/MiSeqDevelopmentData/StabilityNoMetaG.tar and was processed as outlined in the original study [@Kozich2013].
+Briefly, 250-nt read pairs were assembled into contigs by aligning the reads and correcting discordant base calls by requiring one of the base calls to have a Phred quality score at least 6 points higher than the other.
+Sequences where it was not possible to resolve the disagreement were culled from the dataset.
+The sequences were then aligned to a SILVA reference alignment [@Pruesse2007] and any reads that aligned outside of the V4 region were removed from the dataset.
+Sequences were pre-clustered by combining the abundances of sequences that differed by 2 or fewer nucleotides of a more abundant sequence.
+Each of the samples was then screened for chimeric sequences using the default parameters in UCHIME [@Edgar2011].
+The resulting sequences were processed in the same manner as the Canadian soil dataset with the exception that the distance matrices were calculated based on the SILVA-based alignment.
+
+***Analysis of reference database.***
+We utilized the 97% OTUs greengenes reference sequence and taxonomy data (v.13.8) that accompanies the QIIME installation.
+Because the greengenes reference alignment does a poor job of representing the secondary structure of the 16S rRNA gene [@Schloss2010], we realigned the FASTA sequences to a SILVA reference alignment to identify the V4 region of the sequences.
+
+***Calculation of Matthew's Correlation Coefficient (MCC).***
+The MCC was calculated by two approaches in this study using only the dereplicated sequence lists.
+First, we calculated the MCC to determine the stability of OTU assignments following the approach of the He study.
+We assumed that the clusters obtained from the 30 randomized full datasets were correct.
+We counted the number of sequence pairs that were in the same OTU for the subsetted dataset and the full dataset (i.e.
+true positives; TP), that were in different OTUs for the subsetted dataset and the full dataset (i.e.
+true negatives; TN), that were in the same OTU for the subsetted dataset and different OTUs in the full dataset (i.e.
+false positives; FP), and that were in different OTUs for the subsetted dataset and the same OTU in the full dataset (i.e.
+false negatives; FN).
+For each set of 30 random subsamplings of the dataset, we counted these parameters against the 30 randomizations of the full dataset.
+This gave 900 comparisons for each fraction of sequences being used in the analysis.
+The Matthew's correlation coefficient was then calculated as:
+
+$$
+MCC = \frac{TP \times TN - FP \times FN}{\sqrt{(TP +FP)(TP+FN)(TN+FP)(TN+FN)} }
+$$
+
+Second, we calculated the MCC to determine the quality of the clusterings as previously described [@Schloss2011Assessing].
+Briefly, we compared the OTU assignments for pairs of sequences to the distance matrix that was calculated between all pairs of aligned sequences.
+For each dataset that was clustered, those pairs of sequences that were in the same OTU and had a distance less than 3% were TPs, those that were in different OTUs and had a distance greater than 3% were TNs, those that were in the same OTU and had a distance greater than 3% were FPs, and those that were in different OTUs and had a distance less than 3% were FNs.
+The MCC was counted for each dataset using the formula above as implemented in the `sens.spec` command in mothur.
+To judge the quality of the Swarm-generated OTU assignments we calculated the MCC value using thresholds incremented by 1% between 0 and 5% and selected the threshold that provided the optimal MCC value.
+
+***Software availability.***
+A reproducible workflow including all scripts and this manuscript as a literate programming document are available at https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015.
+The workflow utilized QIIME [v.1.9.1; @Caporaso2010], mothur [v.1.37.0; @Schloss2009], USEARCH [v.6.1; @Edgar2010], VSEARCH [v.1.5.0; @Rognes2015], Swarm [v.2.1.1; @Mah2014], and R [v.`r paste(version$major, version$minor, sep='.')`; @language2015].
+The SL, AL, and CL methods are called nearest neighbor (NN), average neighbor (AN), and furthest neighbor (FN) in mothur; we have used the terminology from the He study to minimize confusion.
+The knitr [v.`r packageDescription("knitr")$Version`; @xie2013dynamic], Rcpp [v.
+`r packageDescription("Rcpp")$Version`; @eddelbuettel2013seamless], rentrez [v.
+`r packageDescription("rentrez")$Version`; @winter2015rentrez], and jsonlite [v.
+`r packageDescription("jsonlite")$Version`; @ooms2014jsonlite] packages were used within R.
+
 ## Results and Discussion
@@
 To assess the quality of OTU assignments by the various methods, we made use of the pairwise distance between the unique sequences to count the number of true positives and negatives and the number of false positives and negatives for each method and sampling depth.
-Those pairs of sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives.
-Meanwhile, those pairs of sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives.
 Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and negatives using the MCC.
@@
 
-## Methods
-
-***454 FLX-generated Roesch Canadian soil dataset.***
-After obtaining the 16S rRNA gene fragments from GenBank (accessions EF308591-EF361836), we followed the methods outlined by the He study by removing any sequence that contained an ambiguous base, was identified as being a chimera, and fell outside a defined sequence length.
-Although they reported observing a total of 50,542 sequences that were represented by 13,293 unique sequences, we obtained a total of `r length(scan(file="data/he/canada_soil.good.unique.pick.redundant.fasta", what="", sep="\n", quiet=T))/2` sequences that were represented by `r length(scan(file="data/he/canada_soil.good.unique.pick.fasta", what="", sep="\n", quiet=T))/2` unique sequences.
-Similar to the He study, we randomly sampled, without replacement, 20, 40, 60, and 80% of the sequences from the full data set.
-The random sampling was repeated 30 times.
-The order of the sequences in the full dataset was randomly permuted without replacement to generate an additional 30 datasets.
-To perform the hierarchical clustering methods and to generate a distance matrix we followed the approach of the He study by calculating distances based on pairwise global alignments using the `pairwise.dist` command in mothur using the default Needleman-Wunsch alignment method and parameters.
-It should be noted that this approach has been strongly discouraged [@Schloss2012Secondary].
-Execution of the hierarchical clustering methods was performed as described in the original He study using mothur (v.1.37) and using the QIIME (v.1.9.1) parameter profiles provided in the supplementary material from the He study for the greedy and reference-based clustering methods.
-
-***MiSeq-generated Murine gut microbiota dataset.***
-The murine 16S rRNA gene sequence data generated from the V4 region using an Illumina MiSeq was obtained from http:/www.mothur.org/MiSeqDevelopmentData/StabilityNoMetaG.tar and was processed as outlined in the original study [@Kozich2013].
-Briefly, 250-nt read pairs were assembled into contigs by aligning the reads and correcting discordant base calls by requiring one of the base calls to have a Phred quality score at least 6 points higher than the other.
-Sequences where it was not possible to resolve the disagreement were culled from the dataset.
-The sequences were then aligned to a SILVA reference alignment [@Pruesse2007] and any reads that aligned outside of the V4 region were removed from the dataset.
-Sequences were pre-clustered by combining the abundances of sequences that differed by 2 or fewer nucleotides of a more abundant sequence.
-Each of the samples was then screened for chimeric sequences using the default parameters in UCHIME [@Edgar2011].
-The resulting sequences were processed in the same manner as the Canadian soil dataset with the exception that the distance matrices were calculated based on the SILVA-based alignment.
-
-***Analysis of reference database.***
-We utilized the 97% OTUs greengenes reference sequence and taxonomy data (v.13.8) that accompanies the QIIME installation.
-Because the greengenes reference alignment does a poor job of representing the secondary structure of the 16S rRNA gene [@Schloss2010], we realigned the FASTA sequences to a SILVA reference alignment to identify the V4 region of the sequences.
-
-***Calculation of Matthew's Correlation Coefficient (MCC).***
-The MCC was calculated by two approaches in this study using only the dereplicated sequence lists.
-First, we calculated the MCC to determine the stability of OTU assignments following the approach of the He study.
-We assumed that the clusters obtained from the 30 randomized full datasets were correct.
-We counted the number of sequence pairs that were in the same OTU for the subsetted dataset and the full dataset (i.e.
-true positives; TP), that were in different OTUs for the subsetted dataset and the full dataset (i.e.
-true negatives; TN), that were in the same OTU for the subsetted dataset and different OTUs in the full dataset (i.e.
-false positives; FP), and that were in different OTUs for the subsetted dataset and the same OTU in the full dataset (i.e.
-false negatives; FN).
-For each set of 30 random subsamplings of the dataset, we counted these parameters against the 30 randomizations of the full dataset.
-This gave 900 comparisons for each fraction of sequences being used in the analysis.
-The Matthew's correlation coefficient was then calculated as:
-
-$$
-MCC = \frac{TP \times TN - FP \times FN}{\sqrt{(TP +FP)(TP+FN)(TN+FP)(TN+FN)} }
-$$
-
-Second, we calculated the MCC to determine the quality of the clusterings as previously described [@Schloss2011Assessing].
-Briefly, we compared the OTU assignments for pairs of sequences to the distance matrix that was calculated between all pairs of aligned sequences.
-For each dataset that was clustered, those pairs of sequences that were in the same OTU and had a distance less than 3% were TPs, those that were in different OTUs and had a distance greater than 3% were TNs, those that were in the same OTU and had a distance greater than 3% were FPs, and those that were in different OTUs and had a distance less than 3% were FNs.
-The MCC was counted for each dataset using the formula above as implemented in the `sens.spec` command in mothur.
-To judge the quality of the Swarm-generated OTU assignments we calculated the MCC value using thresholds incremented by 1% between 0 and 5% and selected the threshold that provided the optimal MCC value.
-
-***Software availability.***
-A reproducible workflow including all scripts and this manuscript as a literate programming document are available at https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015.
-The workflow utilized QIIME [v.1.9.1; @Caporaso2010], mothur [v.1.37.0; @Schloss2009], USEARCH [v.6.1; @Edgar2010], VSEARCH [v.1.5.0; @Rognes2015], Swarm [v.2.1.1; @Mah2014], and R [v.`r paste(version$major, version$minor, sep='.')`; @language2015].
-The SL, AL, and CL methods are called nearest neighbor (NN), average neighbor (AN), and furthest neighbor (FN) in mothur; we have used the terminology from the He study to minimize confusion.
-The knitr [v.`r packageDescription("knitr")$Version`; @xie2013dynamic], Rcpp [v.
-`r packageDescription("Rcpp")$Version`; @eddelbuettel2013seamless], rentrez [v.
-`r packageDescription("rentrez")$Version`; @winter2015rentrez], and jsonlite [v.
-`r packageDescription("jsonlite")$Version`; @ooms2014jsonlite] packages were used within R.
-
 \newpage
@@
 \includepdf{results/figures/figure_4.pdf}
-\includepdf{results/figures/figure_5.pdf}
 

### 2015-12-03 code/build_figure2.R
Change shapes of plotting symbols

@@ -16,6 +16,8 @@ build_figure2 <- function(dataset, output_file_name, position='bottomright'){
 
 	orig_par <- par()
 
+	pch <- c(22,21,24,23,25)
+
 	pdf(file=output_file_name, width=4, height=4)
 	par(mar=c(5,5,0.5,0.5))
 	plot(NA, xlim=c(0.8,length(methods)+0.2), ylim=c(0,max(rare_methods$rare_uci)), axes=F, xlab="", ylab="")
@@ -27,12 +29,12 @@ build_figure2 <- function(dataset, output_file_name, position='bottomright'){
 	#	the 95% confidence intervals were smaller or similar in size to the cex=1 plotting symbol
 	#	arrows(x0=m-offset, y0=subset[,"sobs"], y1=subset[,"sobs_lci"], col=clrs, length=0.05, angle=90)
 	#	arrows(x0=m-offset, y0=subset[,"sobs"], y1=subset[,"sobs_uci"], col=clrs, length=0.05, angle=90)
-		points(rep(m-offset, n_fractions), subset[,"sobs"], col=clrs, pch=19, cex=1.0)
+		points(rep(m-offset, n_fractions), subset[,"sobs"], col=clrs, pch=pch, cex=1.0, bg=clrs)
 
 	#	the 95% confidence intervals were smaller or similar in size to the cex=1 plotting symbol
 	#	arrows(x0=m+offset, y0=subset[,"rarefied"], y1=subset[,"rare_lci"], col=clrs, length=0.05, angle=90)
 	#	arrows(x0=m+offset, y0=subset[,"rarefied"], y1=subset[,"rare_uci"], col=clrs, length=0.05, angle=90)
-		points(rep(m+offset, n_fractions), subset[,"rarefied"], col=clrs, pch=21, cex=1.0)
+		points(rep(m+offset, n_fractions), subset[,"rarefied"], col=clrs, pch=pch, cex=1.0)
 	}
 
 	method_labels <- c("CL", "SL", "AL", "DGC", "AGC", "Closed-ref", "Open-ref", "Swarm")
@@ -42,11 +44,15 @@ build_figure2 <- function(dataset, output_file_name, position='bottomright'){
 
 	title(ylab="Number of OTUs", line=3.5)
 
-#	pos <- legend(x=6.2, y=1400, legend=c("20%", "40%", "60%", "80%", "100%"),pch=19, col="white", cex=0.7)
[…]

### 2016-02-22 papers/peerj_2015/Schloss_Cluster_PeerJ_2015.tex
Move PeerJ paper files to papers/peerj_2015

--- 2015-11-13 Schloss_Cluster_PeerJ_2015.tex
@@
 
+\subsection{Methods}\label{methods}
+
+\textbf{\emph{454 FLX-generated Roesch Canadian soil dataset.}} After
+obtaining the 16S rRNA gene fragments from GenBank (accessions
+EF308591-EF361836), we followed the methods outlined by the He study by
+removing any sequence that contained an ambiguous base, was identified
+as being a chimera, and fell outside a defined sequence length. Although
+they reported observing a total of 50,542 sequences that were
+represented by 13,293 unique sequences, we obtained a total of 50,946
+sequences that were represented by 13,393 unique sequences. Similar to
+the He study, we randomly sampled, without replacement, 20, 40, 60, and
+80\% of the sequences from the full data set. The random sampling was
+repeated 30 times. The order of the sequences in the full dataset was
+randomly permuted without replacement to generate an additional 30
+datasets. To perform the hierarchical clustering methods and to generate
+a distance matrix we followed the approach of the He study by
+calculating distances based on pairwise global alignments using the
+\texttt{pairwise.dist} command in mothur using the default
+Needleman-Wunsch alignment method and parameters. It should be noted
+that this approach has been strongly discouraged (Schloss, 2012).
+Execution of the hierarchical clustering methods was performed as
+described in the original He study using mothur (v.1.37) and using the
+QIIME (v.1.9.1) parameter profiles provided in the supplementary
+material from the He study for the greedy and reference-based clustering
+methods.
+
+\textbf{\emph{MiSeq-generated Murine gut microbiota dataset.}} The
+murine 16S rRNA gene sequence data generated from the V4 region using an
+Illumina MiSeq was obtained from
+\url{http:/www.mothur.org/MiSeqDevelopmentData/StabilityNoMetaG.tar} and
+was processed as outlined in the original study (Kozich et al., 2013).
+Briefly, 250-nt read pairs were assembled into contigs by aligning the
+reads and correcting discordant base calls by requiring one of the base
+calls to have a Phred quality score at least 6 points higher than the
+other. Sequences where it was not possible to resolve the disagreement
+were culled from the dataset. The sequences were then aligned to a SILVA
+reference alignment (Pruesse et al., 2007) and any reads that aligned
+outside of the V4 region were removed from the dataset. Sequences were
+pre-clustered by combining the abundances of sequences that differed by
+2 or fewer nucleotides of a more abundant sequence. Each of the samples
+was then screened for chimeric sequences using the default parameters in
+UCHIME (Edgar et al., 2011). The resulting sequences were processed in
+the same manner as the Canadian soil dataset with the exception that the
+distance matrices were calculated based on the SILVA-based alignment.
+
+\textbf{\emph{Analysis of reference database.}} We utilized the 97\%
+OTUs greengenes reference sequence and taxonomy data (v.13.8) that
+accompanies the QIIME installation. Because the greengenes reference
+alignment does a poor job of representing the secondary structure of the
+16S rRNA gene (Schloss, 2010), we realigned the FASTA sequences to a
+SILVA reference alignment to identify the V4 region of the sequences.
+
+\textbf{\emph{Calculation of Matthew's Correlation Coefficient (MCC).}}
+The MCC was calculated by two approaches in this study using only the
+dereplicated sequence lists. First, we calculated the MCC to determine
+the stability of OTU assignments following the approach of the He study.
+We assumed that the clusters obtained from the 30 randomized full
+datasets were correct. We counted the number of sequence pairs that were
+in the same OTU for the subsetted dataset and the full dataset
+(i.e.~true positives; TP), that were in different OTUs for the subsetted
+dataset and the full dataset (i.e.~true negatives; TN), that were in the
+same OTU for the subsetted dataset and different OTUs in the full
+dataset (i.e.~false positives; FP), and that were in different OTUs for
+the subsetted dataset and the same OTU in the full dataset (i.e.~false
+negatives; FN). For each set of 30 random subsamplings of the dataset,
+we counted these parameters against the 30 randomizations of the full
+dataset. This gave 900 comparisons for each fraction of sequences being
+used in the analysis. The Matthew's correlation coefficient was then
+calculated as:
+
+\[
+MCC = \frac{TP \times TN - FP \times FN}{\sqrt{(TP +FP)(TP+FN)(TN+FP)(TN+FN)} }
+\]
+
+Second, we calculated the MCC to determine the quality of the
+clusterings as previously described (Schloss \& Westcott, 2011).
+Briefly, we compared the OTU assignments for pairs of sequences to the
+distance matrix that was calculated between all pairs of aligned
+sequences. For each dataset that was clustered, those pairs of sequences
+that were in the same OTU and had a distance less than 3\% were TPs,
+those that were in different OTUs and had a distance greater than 3\%
+were TNs, those that were in the same OTU and had a distance greater
+than 3\% were FPs, and those that were in different OTUs and had a
+distance less than 3\% were FNs. The MCC was counted for each dataset
+using the formula above as implemented in the \texttt{sens.spec} command
+in mothur. To judge the quality of the Swarm-generated OTU assignments
+we calculated the MCC value using thresholds incremented by 1\% between
+0 and 5\% and selected the threshold that provided the optimal MCC
+value.
+
+\textbf{\emph{Software availability.}} A reproducible workflow including
+all scripts and this manuscript as a literate programming document are
+available at
+\url{https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015}. The
+workflow utilized QIIME (v.1.9.1; Caporaso et al., 2010), mothur
+(v.1.37.0; Schloss et al., 2009), USEARCH (v.6.1; Edgar, 2010), VSEARCH
+(v.1.5.0; Rognes et al., 2015), Swarm (v.2.1.1; Mahé et al., 2014), and
+R (v.3.2.0; R Core Team, 2015). The SL, AL, and CL methods are called
+nearest neighbor (NN), average neighbor (AN), and furthest neighbor (FN)
+in mothur; we have used the terminology from the He study to minimize
+confusion. The knitr (v.1.10.5; Xie, 2013), Rcpp (v. 0.11.6;
+Eddelbuettel, 2013), rentrez (v. 1.0.0; Winter, Chamberlain \&
+Guangchun, 2015), and jsonlite (v. 0.9.16; Ooms, 2014) packages were
+used within R.
+
 \subsection{Results and Discussion}\label{results-and-discussion}
@@
 number of false positives and negatives for each method and sampling
-depth. Those pairs of sequences that were similar to each other and
-found in the same OTU were called true positives while those that were
-similar and found in different OTUs were called false negatives.
-Meanwhile, those pairs of sequences that were different from each other
-and found in the same OTU were called false positives and those that
-were dissimilar and found in different OTUs were called true negatives.
-Counting the frequency of these different classes allowed us to judge
-how each method balanced the ratio of true positives and negatives to
-false positives and negatives using the MCC. We used the average MCC
+depth. Counting the frequency of these different classes allowed us to
+judge how each method balanced the ratio of true positives and negatives
+to false positives and negatives using the MCC. We used the average MCC
 value as a measure of a method's quality and its variation as a measure
@@
 clustering.
-
-\subsection{Methods}\label{methods}
-
-\textbf{\emph{454 FLX-generated Roesch Canadian soil dataset.}} After
-obtaining the 16S rRNA gene fragments from GenBank (accessions
-EF308591-EF361836), we followed the methods outlined by the He study by
-removing any sequence that contained an ambiguous base, was identified
-as being a chimera, and fell outside a defined sequence length. Although
-they reported observing a total of 50,542 sequences that were
-represented by 13,293 unique sequences, we obtained a total of 50,946
-sequences that were represented by 13,393 unique sequences. Similar to
-the He study, we randomly sampled, without replacement, 20, 40, 60, and
-80\% of the sequences from the full data set. The random sampling was
-repeated 30 times. The order of the sequences in the full dataset was
-randomly permuted without replacement to generate an additional 30
-datasets. To perform the hierarchical clustering methods and to generate
-a distance matrix we followed the approach of the He study by
-calculating distances based on pairwise global alignments using the
-\texttt{pairwise.dist} command in mothur using the default
-Needleman-Wunsch alignment method and parameters. It should be noted
-that this approach has been strongly discouraged (Schloss, 2012).
-Execution of the hierarchical clustering methods was performed as
-described in the original He study using mothur (v.1.37) and using the
-QIIME (v.1.9.1) parameter profiles provided in the supplementary
-material from the He study for the greedy and reference-based clustering
-methods.
-
-\textbf{\emph{MiSeq-generated Murine gut microbiota dataset.}} The
-murine 16S rRNA gene sequence data generated from the V4 region using an
-Illumina MiSeq was obtained from
-\url{http:/www.mothur.org/MiSeqDevelopmentData/StabilityNoMetaG.tar} and
-was processed as outlined in the original study (Kozich et al., 2013).
-Briefly, 250-nt read pairs were assembled into contigs by aligning the
-reads and correcting discordant base calls by requiring one of the base
-calls to have a Phred quality score at least 6 points higher than the
-other. Sequences where it was not possible to resolve the disagreement
-were culled from the dataset. The sequences were then aligned to a SILVA
-reference alignment (Pruesse et al., 2007) and any reads that aligned
-outside of the V4 region were removed from the dataset. Sequences were
-pre-clustered by combining the abundances of sequences that differed by
-2 or fewer nucleotides of a more abundant sequence. Each of the samples
-was then screened for chimeric sequences using the default parameters in
-UCHIME (Edgar et al., 2011). The resulting sequences were processed in
-the same manner as the Canadian soil dataset with the exception that the
-distance matrices were calculated based on the SILVA-based alignment.
-
-\textbf{\emph{Analysis of reference database.}} We utilized the 97\%
-OTUs greengenes reference sequence and taxonomy data (v.13.8) that
-accompanies the QIIME installation. Because the greengenes reference
-alignment does a poor job of representing the secondary structure of the
-16S rRNA gene (Schloss, 2010), we realigned the FASTA sequences to a
-SILVA reference alignment to identify the V4 region of the sequences.
-
-\textbf{\emph{Calculation of Matthew's Correlation Coefficient (MCC).}}
-The MCC was calculated by two approaches in this study using only the
-dereplicated sequence lists. First, we calculated the MCC to determine
-the stability of OTU assignments following the approach of the He study.
-We assumed that the clusters obtained from the 30 randomized full
-datasets were correct. We counted the number of sequence pairs that were
-in the same OTU for the subsetted dataset and the full dataset
-(i.e.~true positives; TP), that were in different OTUs for the subsetted
-dataset and the full dataset (i.e.~true negatives; TN), that were in the
-same OTU for the subsetted dataset and different OTUs in the full
-dataset (i.e.~false positives; FP), and that were in different OTUs for
-the subsetted dataset and the same OTU in the full dataset (i.e.~false
-negatives; FN). For each set of 30 random subsamplings of the dataset,
-we counted these parameters against the 30 randomizations of the full
-dataset. This gave 900 comparisons for each fraction of sequences being
-used in the analysis. The Matthew's correlation coefficient was then
-calculated as:
-
-\[
-MCC = \frac{TP \times TN - FP \times FN}{\sqrt{(TP +FP)(TP+FN)(TN+FP)(TN+FN)} }
-\]
-
-Second, we calculated the MCC to determine the quality of the
-clusterings as previously described (Schloss \& Westcott, 2011).
-Briefly, we compared the OTU assignments for pairs of sequences to the
-distance matrix that was calculated between all pairs of aligned
-sequences. For each dataset that was clustered, those pairs of sequences
-that were in the same OTU and had a distance less than 3\% were TPs,
-those that were in different OTUs and had a distance greater than 3\%
-were TNs, those that were in the same OTU and had a distance greater
-than 3\% were FPs, and those that were in different OTUs and had a
-distance less than 3\% were FNs. The MCC was counted for each dataset
-using the formula above as implemented in the \texttt{sens.spec} command
-in mothur. To judge the quality of the Swarm-generated OTU assignments
-we calculated the MCC value using thresholds incremented by 1\% between
-0 and 5\% and selected the threshold that provided the optimal MCC
-value.
-
-\textbf{\emph{Software availability.}} A reproducible workflow including
-all scripts and this manuscript as a literate programming document are
-available at
-\url{https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015}. The
-workflow utilized QIIME (v.1.9.1; Caporaso et al., 2010), mothur
-(v.1.37.0; Schloss et al., 2009), USEARCH (v.6.1; Edgar, 2010), VSEARCH
-(v.1.5.0; Rognes et al., 2015), Swarm (v.2.1.1; Mahé et al., 2014), and
-R (v.3.2.0; R Core Team, 2015). The SL, AL, and CL methods are called
-nearest neighbor (NN), average neighbor (AN), and furthest neighbor (FN)
-in mothur; we have used the terminology from the He study to minimize
-confusion. The knitr (v.1.10.5; Xie, 2013), Rcpp (v. 0.11.6;
-Eddelbuettel, 2013), rentrez (v. 1.0.0; Winter, Chamberlain \&
-Guangchun, 2015), and jsonlite (v. 0.9.16; Ooms, 2014) packages were
-used within R.
 

### 2016-02-22 papers/peerj_2015/Schloss_Cluster_PeerJ_2015.Rmd
Not sure what's going on

--- 2015-11-14 Schloss_Cluster_PeerJ_2015.Rmd
@@
     includes:
-      in_header: header.tex
+      in_header: ../header.tex
 csl: peerj.csl
@@
 
-\includepdf{results/figures/figure_1.pdf}
-\includepdf{results/figures/figure_2.pdf}
-\includepdf{results/figures/figure_3.pdf}
-\includepdf{results/figures/figure_4.pdf}
+\includepdf{../../results/figures/figure_1.pdf}
+\includepdf{../../results/figures/figure_2.pdf}
+\includepdf{../../results/figures/figure_3.pdf}
+\includepdf{../../results/figures/figure_4.pdf}
+\includepdf{../../results/figures/figure_5.pdf}
 

### 2016-02-22 code/build_simulated_names.R
Rough in simulated rules

@@ -1,0 +1,30 @@
+fasta_file <- 'data/staggered/staggered.fasta'
+
+get_seq_names <- function(fasta_file){
+	fasta_data <- scan(fasta_file, what=character(), sep='\n', quiet=T)
+	is_seq_name <- grepl("^>", fasta_data)
+	gsub('>', '', fasta_data[is_seq_name])
+}
+
+make_list <- function(name, freq){
+	redundant <- paste(name, (1:freq), sep="_")
+	redundant[1] <- name
+	redundant <- paste(redundant, collapse=",")
+	paste(name, redundant, sep='\t')
+}
+
+even <- function(fasta_file, n_seqs=100){
+	seq_names <- get_seq_names(fasta_file)
+	names_data <- sapply(seq_names, make_list, freq=n_seqs)
+	names_file <- gsub('fasta', 'names', fasta_file)
+	write(names_data, names_file)
+}
+
+staggered <- function(fasta_file, max_n_seqs=200){
+	set.seed(1)
+	seq_names <- get_seq_names(fasta_file)
+	freqs <- floor(runif(length(seq_names), 1, max_n_seqs))
+	names_data <- mapply(make_list, seq_names, freqs)
+	names_file <- gsub('fasta', 'names', fasta_file)
+	write(names_data, names_file)
+}

### 2016-02-23 code/otuclust_to_list.R
Add OTUClust rules and code

@@ -1,0 +1,12 @@
+# the otuclust format puts otus on separate rows and the sequences within each
+# otu on the same line separated by tabs
+
+otuclust_to_list <- function(otuclust_file_name){
+	otuclust_file_name <- "he_0.2_01.otuclust.clust"
+	otuclust_data <- scan(otuclust_file_name, what="", sep="\n", quiet=TRUE)
+	otuclust_data <- gsub('\t', ',', otuclust_data)
+	n_otus <- length(otuclust_data)
+	list_data <- paste("userLabel", n_otus, otuclust_data, collapse='\t')
+	list_file_name <- gsub("\\.clust", ".list", otuclust_file_name)
+	write(list_data, list_file_name)
+}

### 2016-02-23 code/run_otuclust.sh

@@ -1,0 +1,15 @@
+# This script implements the OTUClust algorithm. Because the output is weird
+# we change it into a mothur list file. The input is a degapped and redundant
+# fasta file and the output is a list file where otuclust is used as the method
+# tag. We'll assume that OTUClust is in the user's path. We will also assign
+# sequences to OTUs based on 97% similarity (-s 0.97) and because we want to
+# include all of the sequences we'll keep the singletons (-m 1)
+
+FASTA=$1
+OTUCLUST_CLUST=$(echo $FASTA | sed 's/fasta/otuclust.clust/')
+OTUCLUST_REP=$(echo $FASTA | sed 's/fasta/otuclust.rep/')
+
+otuclust -f fasta $FASTA --out-clust $OTUCLUST_CLUST --out-rep OTUCLUST_REP -s 0.97 -m 1
+
+R -e "source('code/otuclust_to_list.R'); otuclust_to_list('$OTUCLUST_CLUST')"
+rm $OTUCLUST_CLUST $OTUCLUST_REP

### 2016-02-23 code/otuclust_to_list.R
Getting OTUClust to run on axiom

@@ -2,7 +2,6 @@
 # otu on the same line separated by tabs
 
 otuclust_to_list <- function(otuclust_file_name){
-	otuclust_file_name <- "he_0.2_01.otuclust.clust"
 	otuclust_data <- scan(otuclust_file_name, what="", sep="\n", quiet=TRUE)
 	otuclust_data <- gsub('\t', ',', otuclust_data)
 	n_otus <- length(otuclust_data)

### 2016-02-23 code/otuclust_to_list.R
Debugging OTUClust execution

@@ -5,7 +5,8 @@ otuclust_to_list <- function(otuclust_file_name){
 	otuclust_data <- scan(otuclust_file_name, what="", sep="\n", quiet=TRUE)
 	otuclust_data <- gsub('\t', ',', otuclust_data)
 	n_otus <- length(otuclust_data)
-	list_data <- paste("userLabel", n_otus, otuclust_data, collapse='\t')
+	list_data <- paste(c("userLabel", n_otus, otuclust_data), collapse='\t')
 	list_file_name <- gsub("\\.clust", ".list", otuclust_file_name)
+	list_file_name <- gsub("\\.ng", "", list_file_name)
 	write(list_data, list_file_name)
 }

### 2016-02-23 code/run_otuclust.sh

@@ -9,7 +9,7 @@ FASTA=$1
 OTUCLUST_CLUST=$(echo $FASTA | sed 's/fasta/otuclust.clust/')
 OTUCLUST_REP=$(echo $FASTA | sed 's/fasta/otuclust.rep/')
 
-otuclust -f fasta $FASTA --out-clust $OTUCLUST_CLUST --out-rep OTUCLUST_REP -s 0.97 -m 1
+otuclust -f fasta $FASTA --out-clust $OTUCLUST_CLUST --out-rep $OTUCLUST_REP -s 0.97 -m 1
 
 R -e "source('code/otuclust_to_list.R'); otuclust_to_list('$OTUCLUST_CLUST')"
 rm $OTUCLUST_CLUST $OTUCLUST_REP

### 2016-02-23 code/run_sumaclust.sh
Bring in sumaclust

@@ -1,0 +1,15 @@
+# This script implements the sumaclust algorithm. Because the output is weird
+# we change it into a mothur list file. The input is a degapped and redundant
+# fasta file and the output is a list file where sumaclust is used as the method
+# tag. We'll assume that sumaclust is in code/sumaclust_v1.0.20. We will also
+# assign sequences to OTUs based on 97% similarity (-t 0.97) and because we want
+# to eventually generate a list file we want to output the mapping file with the
+# -O flag
+
+FASTA=$1
+SUMACLUST_CLUST=$(echo $FASTA | sed 's/fasta/sumaclust.clust/')
+
+code/sumaclust_v1.0.20/sumaclust -t 0.97 data/he/he_0.2_01.fasta -O $SUMACLUST_CLUST >/dev/null
+
+R -e "source('code/sumaclust_to_list.R'); sumaclust_to_list('$SUMACLUST_CLUST')"
+rm $SUMACLUST_CLUST

### 2016-02-23 code/sumaclust_to_list.R

@@ -1,0 +1,14 @@
+# the sumaclust format puts otus on separate rows and the sequences within each
+# otu on the same line separated by tabs. the representative sequence name is
+# repeated in the first two columns
+
+sumaclust_to_list <- function(sumaclust_file_name){
+	sumaclust_data <- scan(sumaclust_file_name, what="", sep="\n", quiet=TRUE)
+	sumaclust_data <- gsub("^[^\t]\t", "", sumaclust_data)
+	sumaclust_data <- gsub('\t', ',', sumaclust_data)
+	n_otus <- length(sumaclust_data)
+	list_data <- paste(c("userLabel", n_otus, sumaclust_data), collapse='\t')
+	list_file_name <- gsub("\\.clust", ".list", sumaclust_file_name)
+	list_file_name <- gsub("\\.ng", "", list_file_name)
+	write(list_data, list_file_name)
+}

### 2016-02-23 code/run_sumaclust.sh
Use FASTA variable in sumaclust

@@ -9,7 +9,7 @@
 FASTA=$1
 SUMACLUST_CLUST=$(echo $FASTA | sed 's/fasta/sumaclust.clust/')
 
-code/sumaclust_v1.0.20/sumaclust -t 0.97 data/he/he_0.2_01.fasta -O $SUMACLUST_CLUST >/dev/null
+code/sumaclust_v1.0.20/sumaclust -t 0.97 $FASTA -O $SUMACLUST_CLUST >/dev/null
 
 R -e "source('code/sumaclust_to_list.R'); sumaclust_to_list('$SUMACLUST_CLUST')"
 rm $SUMACLUST_CLUST

### 2016-02-23 code/sortmerna_to_list.R
Add sortmerna commands

@@ -1,0 +1,15 @@
+# the sortmrna format puts otus on separate rows and the sequences within each
+# otu on the same line separated by tabs. the reference sequence name is in the
+# first column and can be ignored.
+
+sortmerna_to_list <- function(sortmerna_folder_name){
+	sortmerna_file <- paste0(sortmerna_folder_name, "/seqs_otus.txt")
+	sortmerna_data <- scan(sortmerna_file, what="", sep="\n", quiet=TRUE)
+	sortmerna_data <- gsub("^[^\t]*\t", "", sortmerna_data)
+	sortmerna_data <- gsub('\t', ',', sortmerna_data)
+	n_otus <- length(sortmerna_data)
+	list_data <- paste(c("userLabel", n_otus, sortmerna_data), collapse='\t')
+	list_file_name <- paste0(sortmerna_folder_name, ".list")
+	list_file_name <- gsub("\\.ng", "", list_file_name)
+	write(list_data, list_file_name)
+}

### 2016-02-23 code/sortmerna_to_list.R
Get sortmerna scripts to work

@@ -2,8 +2,12 @@
 # otu on the same line separated by tabs. the reference sequence name is in the
 # first column and can be ignored.
 
+
+#he_1.0_01_otus.txt
+
 sortmerna_to_list <- function(sortmerna_folder_name){
-	sortmerna_file <- paste0(sortmerna_folder_name, "/seqs_otus.txt")
+	stub <- gsub(".sortmerna", "_otus.txt", sortmerna_folder_name)
+	sortmerna_file <- paste0(sortmerna_folder_name, "/", stub)
 	sortmerna_data <- scan(sortmerna_file, what="", sep="\n", quiet=TRUE)
 	sortmerna_data <- gsub("^[^\t]*\t", "", sortmerna_data)
 	sortmerna_data <- gsub('\t', ',', sortmerna_data)

### 2016-02-24 papers/msystems_2016/Schloss_Commentary_mSystems_2016.Rmd
Rough in coarse commentary structure

--- 2015-11-14 Schloss_Cluster_PeerJ_2015.Rmd
@@
 ---
-title: "***De novo* clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units**"
+title: "** INSERT TITLE **"
 bibliography: references.bib
@@
     includes:
-      in_header: header.tex
-csl: peerj.csl
+      in_header: ../header.tex
+csl: msystems.csl
 fontsize: 11pt
@@
 \vspace{25mm}
-Sarah L. Westcott and Patrick D. Schloss${^\dagger}$
+Patrick D. Schloss${^\dagger}$
 
@@
 
-**Background.** 16S rRNA gene sequences are routinely assigned to operational taxonomic units (OTUs) that are then used to analyze complex microbial communities.
-A number of methods have been employed to carry out the assignment of 16S rRNA gene sequences to OTUs leading to confusion over which method is optimal.
-A recent study suggested that a clustering method should be selected based on its ability to generate stable OTU assignments that do not change as additional sequences are added to the dataset.
-In contrast, we contend that the quality of the OTU assignments, the ability of the method to properly represent the distances between the sequences, is more important.
+\newpage
 
-**Methods.** Our analysis implemented six *de novo* clustering algorithms including the single linkage, complete linkage, average linkage, abundance-based greedy clustering, distance-based greedy clustering, and Swarm and the open and closed-reference methods.
-Using two previously published datasets we used the Matthew’s Correlation Coefficient (MCC) to assess the stability and quality of OTU assignments.
-
-**Results.** The stability of OTU assignments did not reflect the quality of the assignments.
-Depending on the dataset being analyzed, the average linkage and the distance and abundance-based greedy clustering methods generated OTUs that were more likely to represent the actual distances between sequences than the open and closed-reference methods.
-We also demonstrated that for the greedy algorithms VSEARCH produced assignments that were comparable to those produced by USEARCH making VSEARCH a viable free and open source alternative to USEARCH.
-Further interrogation of the reference-based methods indicated that when USEARCH or VSEARCH were used to identify the closest reference, the OTU assignments were sensitive to the order of the reference sequences because the reference sequences can be identical over the region being considered.
-More troubling was the observation that while both USEARCH and VSEARCH have a high level of sensitivity to detect reference sequences, the specificity of those matches was poor relative to the true best match.
-
-**Discussion.** Our analysis calls into question the quality and stability of OTU assignments generated by the open and closed-reference methods as implemented in current version of QIIME.
-This study demonstrates that *de novo* methods are the optimal method of assigning sequences into OTUs and that the quality of these assignments needs to be assessed for multiple methods to identify the optimal clustering method for a particular dataset.
+INSERT TEXT HERE
 
@@
 
-## Introduction
-
-The ability to affordably generate millions of 16S rRNA gene sequences has allowed microbial ecologists to thoroughly characterize the microbial community composition of hundreds of samples.
-To simplify the complexity of these large datasets, it is helpful to cluster sequences into meaningful bins.
-These bins, commonly known as operational taxonomic units (OTUs), are  used to compare the biodiversity contained within and between different samples [@Schloss2011Assessing].
-Such comparisons have enabled researchers to characterize the microbiota associated with the human body [e.g.
-@Huttenhower2012], soil [e.g.
-@Shade2013], aquatic ecosystems [e.g.
-@Gilbert2011], and numerous other environments.
-Within the field of microbial ecology, a convention has emerged where sequences are clustered into OTUs using a threshold of 97% similarity or a distance of 3%.
-One advantage of the OTU-based approach is that the definition of the bins is operational and can be changed to suit the needs of the particular project.
-However, with the dissemination of clustering methods within software such as mothur [@Schloss2009], QIIME [@Caporaso2010], and other tools [@Edgar2010; @Sun2009; @Mah2014; @Edgar2013; @Cai2011], it is important to understand how different clustering methods implement this conventional OTU threshold.
-Furthermore, it is necessary to understand how the selected method affects the precision and accuracy of assigning sequences to OTUs.
-Broadly speaking, three approaches have been developed to assign sequences to OTUs:
-
-The first approach has been referred to as phylotyping [@Schloss2011Assessing] or closed reference clustering [@NavasMolina2013].
-This approach involves comparing sequences to a curated database and then clustering sequences into the same OTU that are similar to the same reference sequence.
-Reference-based clustering methods suffer when the reference does not adequately reflect the biodiversity of the community.
-If a large fraction of sequences are novel, then they cannot be assigned to an OTU.
-In addition, the reference sequences are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene [@Schloss2010; @Kim2011].
-Thus, a sequence representing a fragment of the gene may be more than 97% similar to multiple reference sequences.
-Defining OTUs in the closed-reference approach is problematic because two sequences might be 97% similar to the same reference sequence, but they may only be 94% similar to each other.
-Alternatively, a sequence may be equally similar to two or more reference sequences.
-A alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@Schloss2011Assessing].
-The strengths of the reference-based methods include their speed, potential for trivial parallelization, ability to compare OTU assignments across studies, and the hope that as databases improve, the OTU assignments will also improve.
-
-The second approach has been referred to as distance-based [@Schloss2011Assessing] or *de novo* clustering [@NavasMolina2013].
-In this approach, the distance between sequences is used to cluster sequences into OTUs rather than the distance to a reference database.
-In contrast to the efficiency of closed-reference clustering, the computational cost of hierarchical *de novo* clustering methods scales quadratically with the number of unique sequences.
-The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences.
-If error rates can be reduced through stringent quality control measures, then these problems can be overcome [@Kozich2013].
-As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@Edgar2010; @Sun2009; @Mah2014].
-Two related heuristics implemented in USEARCH were recently described: distance-based greedy clustering (DGC) and abundance-based greedy clustering (AGC) [@Edgar2010; @He2015].
-These greedy methods cluster sequences within a defined similarity threshold of an index sequence or creates a new index sequence.
-If a sequence is more similar than the defined threshold, it is assigned to the closest centroid based (i.e.
-DGC) or the most abundant centroid (i.e.
-AGC).
-One critique of *de novo* approaches is that OTU assignments are sensitive to the input order of the sequences [@He2015; @Mah2014].
-Whether the differences in assignments is meaningful is unclear and the variation in results could represent equally valid clustering of the data.
-The strength of *de novo* clustering is its independence of references for carrying out the clustering step.
-For this reason, *de novo* clustering has been preferred across the field.
-After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU [@Schloss2011Assessing].
-
-The third approach, open-reference clustering, is a hybrid of the closed-reference and *de novo* approaches [@NavasMolina2013; @Rideout2014].
-Open-reference clustering involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference.
-In theory, this method should exploit the strengths of both closed-reference and *de novo* clustering; however, the different OTU definitions employed by commonly used closed-reference and *de novo* clustering implementations pose a possible problem when the methods are combined.
-An alternative to this approach has been to classify sequences to a bacterial family or genus and then assign those sequences to OTUs within those taxonomic groups using the average linkage method [@Schloss2011Assessing].
-For example, all sequences classified as belonging to the *Porphyromonadaceae* would then be assigned to OTUs using the average linkage method using a 3% distance threshold.
-Those sequences that did not classify to a known family would also be clustered using the average linkage method.
-An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group is seen as being independent and can be processed separately.
-Such an approach would overcome the difficulty of mixing OTU definitions between the closed-reference and *de novo* approaches; however, it would still suffer from the problems associated with database quality and classification error.
-
-The growth in options for assigning sequences using each of these three broad approaches has been considerable.
-It has been difficult to objectively assess the quality of OTU assignments.
-Some have focused on the time and memory required to process a dataset [@Sun2009; @Cai2011; @Rideout2014; @Mah2014].
-These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the OTU assignments.
-Others have attempted to judge the quality of a method by its ability to generate data that parallels classification data [@Cai2011; @Sun2011; @White2010].
-This approach is problematic because bacterial taxonomy often reflects historical biases amongst bacterial systematicists.
-Furthermore, it is well known that the rates of evolution across lineages are not the same [@Wang2007; @Schloss2010].
-A related approach has used clustering of mock community data to evaluate methods [@Edgar2013; @Mah2014; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014].
-Yet these approaches ignore the effects of sequencing errors that tend to accumulate with sequencing depth and represent highly idealized communities that lack the phylogenetic diversity of real microbial communities [@Kozich2013; @Schloss2011Reducing].
-Others have assessed the quality of clustering based on their ability to generate the same OTUs generated by other methods [@Schmidt2014Limits; @Rideout2014].
-This is problematic because it does not solve the fundamental question of which method is optimal.
-The concept of ecological consistency as a metric of quality asserts that sequences that cluster into the same OTU should share similar ecological affiliations [@Koeppel2013; @Schmidt2014Ecological; @Preheim2013].
-Although this is an intriguing approach and is a quantitative metric, it is unclear how the metric would be objectively validated.
-We recently proposed an approach for evaluating OTU assignments using the distances between pairs of sequences [@Schloss2011Assessing].
-We were able to synthesize the relationship between OTU assignments and the distances between sequences using the Matthew's correlation coefficient [MCC; @Matthews1975].
-MCC is can be interpreted as representing the correlation between the observed and expected classifications and can vary between -1.0 and 1.0.
-The strength of the MCC, as implemented by Schloss et al.
-[-@Schloss2011Assessing], is that it is an objective approach to assessing the quality of the OTU assignments that can be calculated for any set of OTU assignments where there is a distance matrix and a specific threshold without relying on an external reference.
-
-A recent analysis by He and colleagues [-@He2015] characterized the three general clustering approaches by focusing on what they called stability.
-They defined stability as the ability of a method to provide the same clustering on a subset of the data as was found in the full dataset.
-Their concept of stability did not account for the quality of the OTU assignments and instead focused on the precision of the assignments.
-A method may be very stable, but of poor quality.
-In the current analysis, we assessed the quality and stability of the various clustering methods.
-Building on our previous analysis of clustering methods, our hypothesis was that the methods praised by the He study for their stability actually suffered a lack of quality.
-In addition, we assess these parameters in light of sequence quality using the original 454 dataset and a larger and more modern dataset generated using the MiSeq platform.
-
-## Methods
-
-***454 FLX-generated Roesch Canadian soil dataset.***
-After obtaining the 16S rRNA gene fragments from GenBank (accessions EF308591-EF361836), we followed the methods outlined by the He study by removing any sequence that contained an ambiguous base, was identified as being a chimera, and fell outside a defined sequence length.
-Although they reported observing a total of 50,542 sequences that were represented by 13,293 unique sequences, we obtained a total of `r length(scan(file="data/he/canada_soil.good.unique.pick.redundant.fasta", what="", sep="\n", quiet=T))/2` sequences that were represented by `r length(scan(file="data/he/canada_soil.good.unique.pick.fasta", what="", sep="\n", quiet=T))/2` unique sequences.
-Similar to the He study, we randomly sampled, without replacement, 20, 40, 60, and 80% of the sequences from the full data set.
-The random sampling was repeated 30 times.
-The order of the sequences in the full dataset was randomly permuted without replacement to generate an additional 30 datasets.
-To perform the hierarchical clustering methods and to generate a distance matrix we followed the approach of the He study by calculating distances based on pairwise global alignments using the `pairwise.dist` command in mothur using the default Needleman-Wunsch alignment method and parameters.
-It should be noted that this approach has been strongly discouraged [@Schloss2012Secondary].
-Execution of the hierarchical clustering methods was performed as described in the original He study using mothur (v.1.37) and using the QIIME (v.1.9.1) parameter profiles provided in the supplementary material from the He study for the greedy and reference-based clustering methods.
-
-***MiSeq-generated Murine gut microbiota dataset.***
-The murine 16S rRNA gene sequence data generated from the V4 region using an Illumina MiSeq was obtained from http:/www.mothur.org/MiSeqDevelopmentData/StabilityNoMetaG.tar and was processed as outlined in the original study [@Kozich2013].
-Briefly, 250-nt read pairs were assembled into contigs by aligning the reads and correcting discordant base calls by requiring one of the base calls to have a Phred quality score at least 6 points higher than the other.
-Sequences where it was not possible to resolve the disagreement were culled from the dataset.
-The sequences were then aligned to a SILVA reference alignment [@Pruesse2007] and any reads that aligned outside of the V4 region were removed from the dataset.
-Sequences were pre-clustered by combining the abundances of sequences that differed by 2 or fewer nucleotides of a more abundant sequence.
-Each of the samples was then screened for chimeric sequences using the default parameters in UCHIME [@Edgar2011].
-The resulting sequences were processed in the same manner as the Canadian soil dataset with the exception that the distance matrices were calculated based on the SILVA-based alignment.
-
-***Analysis of reference database.***
-We utilized the 97% OTUs greengenes reference sequence and taxonomy data (v.13.8) that accompanies the QIIME installation.
-Because the greengenes reference alignment does a poor job of representing the secondary structure of the 16S rRNA gene [@Schloss2010], we realigned the FASTA sequences to a SILVA reference alignment to identify the V4 region of the sequences.
-
-***Calculation of Matthew's Correlation Coefficient (MCC).***
-The MCC was calculated by two approaches in this study using only the dereplicated sequence lists.
-First, we calculated the MCC to determine the stability of OTU assignments following the approach of the He study.
-We assumed that the clusters obtained from the 30 randomized full datasets were correct.
-We counted the number of sequence pairs that were in the same OTU for the subsetted dataset and the full dataset (i.e.
-true positives; TP), that were in different OTUs for the subsetted dataset and the full dataset (i.e.
-true negatives; TN), that were in the same OTU for the subsetted dataset and different OTUs in the full dataset (i.e.
-false positives; FP), and that were in different OTUs for the subsetted dataset and the same OTU in the full dataset (i.e.
-false negatives; FN).
-For each set of 30 random subsamplings of the dataset, we counted these parameters against the 30 randomizations of the full dataset.
-This gave 900 comparisons for each fraction of sequences being used in the analysis.
-The Matthew's correlation coefficient was then calculated as:
-
-$$
-MCC = \frac{TP \times TN - FP \times FN}{\sqrt{(TP +FP)(TP+FN)(TN+FP)(TN+FN)} }
-$$
-
-Second, we calculated the MCC to determine the quality of the clusterings as previously described [@Schloss2011Assessing].
-Briefly, we compared the OTU assignments for pairs of sequences to the distance matrix that was calculated between all pairs of aligned sequences.
-For each dataset that was clustered, those pairs of sequences that were in the same OTU and had a distance less than 3% were TPs, those that were in different OTUs and had a distance greater than 3% were TNs, those that were in the same OTU and had a distance greater than 3% were FPs, and those that were in different OTUs and had a distance less than 3% were FNs.
-The MCC was counted for each dataset using the formula above as implemented in the `sens.spec` command in mothur.
-To judge the quality of the Swarm-generated OTU assignments we calculated the MCC value using thresholds incremented by 1% between 0 and 5% and selected the threshold that provided the optimal MCC value.
-
-***Software availability.***
-A reproducible workflow including all scripts and this manuscript as a literate programming document are available at https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015.
-The workflow utilized QIIME [v.1.9.1; @Caporaso2010], mothur [v.1.37.0; @Schloss2009], USEARCH [v.6.1; @Edgar2010], VSEARCH [v.1.5.0; @Rognes2015], Swarm [v.2.1.1; @Mah2014], and R [v.`r paste(version$major, version$minor, sep='.')`; @language2015].
-The SL, AL, and CL methods are called nearest neighbor (NN), average neighbor (AN), and furthest neighbor (FN) in mothur; we have used the terminology from the He study to minimize confusion.
-The knitr [v.`r packageDescription("knitr")$Version`; @xie2013dynamic], Rcpp [v.
-`r packageDescription("Rcpp")$Version`; @eddelbuettel2013seamless], rentrez [v.
-`r packageDescription("rentrez")$Version`; @winter2015rentrez], and jsonlite [v.
-`r packageDescription("jsonlite")$Version`; @ooms2014jsonlite] packages were used within R.
-
-## Results and Discussion
-```{r he_stability_analysis}
-he_stability <- read.table(file="data/process/he.mcc_ref.summary", header=T)
-```
-
-***Summary and replication of He study.*** We obtained the Canadian soil dataset from Roesch et al.
-[-@Roesch2007] and processed the sequences as described by He and colleagues.
-Using these data, we reconsidered three of the more critical analyses performed in the He study.
-
-First, we sought to quantify whether the OTU assignments observed for a subset of the data represented the same assignments that were found with the full dataset.
-The He study used the MCC to quantify the degree to which pairs of sequences were in the same OTUs in subsampled and full datasets.
-A more robust approach would utilize metrics that quantify the mutual information held between two sets of clusterings and has been applied to assess inter-method variation in OTU composition [@Schmidt2014Limits].
-To maintain consistency with the original He study, we also calculated the MCC value as they described.
-The He study found that when they used the open and closed-reference methods the OTUs formed using the subsetted data most closely resembled those of the full dataset.
-Among the *de novo* methods they observed that the AGC method generated the most stable OTUs followed by the single linkage (SL), DGC, complete linkage (CL), and average linkage (AL) methods.
-We first calculated the MCC for the OTU assignments generated by each of the clustering methods using 20, 40, 60, and 80% of the sequences relative to the OTU composition formed by the methods using the full dataset (see Methods for description; Figure 1A).
-Similar to the He study, we replicated each method and subsampled to the desired fraction of the dataset 30 times.
-Multiple subsamplings were necessary because a random number generator is used in some of the methods to break ties where pairs of sequences have the same distance between them.
-Across these sequencing depths, we observed that the stability of the OTUs generated by the SL and CL methods were highly sensitive to sampling effort relative to the OTUs generated by the AL, AGC, and DGC methods (Figure 1A).
-Our results (Figure 1B) largely confirmed those of Figure 4C in the He study with one notable exception.
-The He study observed a broad range of MCC values among their AL replicates when analyzing OTUs generated using 60% of the data.
-This result appeared out of character and was not explained by the authors.
-They observed a mean MCC value of approximately 0.63 (95% confidence interval between approximately 0.15 and 0.75).
-In contrast, we observed a mean value of `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "mean"], 2)` (95% confidence interval between `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "lci"], 2)` and `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "uci"], 2)`).
-This result indicates that the AL assignments were far more stable than indicated in the He study.
-Regardless, although the assignments are quite stable, it does support the assertion that the OTU assignments observed for the subset of the data do not perfectly match the assignments that were found with the full dataset as they did with the reference-based methods; however, the significance of these differences is unclear.
-
-Second, the He study and the original Roesch study showed that rarefaction curves calculated using CL-generated OTU assignments obtained using a subsample of the sequencing data did not overlap with rarefaction curves generated using OTU assignments generated from the full dataset.
-The He and Roesch studies both found that the CL method produced fewer OTUs in the subset than in the rarefied data.
-In addition, the He study found that the SL method produced more OTUs, the AGC produced fewer, and the other methods produced similar numbers of OTUs than expected when comparing the subsetted data to the rarefied data.
-Our results support those of these previous studies (Figure 2).
-It was clear that inter-method differences were generally more pronounced than the differences observed between rarefying from the full dataset and from clustering the subsetted data.
-The number of OTUs observed was largest using the CL method, followed by the open-reference method.
-The AL, AGC, and DGC methods all provided comparable numbers of OTUs.
-Finally, the closed-reference and SL methods generated the fewest number of OTUs.
-
-Third, the authors attempted to describe the effects of the OTU assignment instability on comparisons of communities.
-They used Adonis to test whether the community structure represented in subsetted communities resembled that of the full dataset when only using the unstable OTUs [@Anderson2001].
-Although they were able to detect significant p-values, they appeared to be of marginal biological significance.
-Adonis R statistics close to zero indicate the community structures from the full and subsetted datasets overlapped while values of one indicate the communities are completely different.
-The He study observed adonis R statistics of 0.02 (closed-reference), 0.03 (open-reference), 0.07 (CL, AGC, DGC), and 0.16 (SL and AL).
-Regardless of the statistical or biological significance of these results, the analysis was tautological since, by definition, representing communities based on their unstable OTUs would yield differences.
-Furthermore, the *de novo* and open-reference approaches do not consistently label the OTUs that sequences belong to when the clustering methods are run multiple times with different random number seeds.
-To overcome this, the authors selected representative sequences from each OTU and used those representative sequences to link OTU assignments between the different sized sequence sets.
-It was not surprising that the only analysis that did not provide a significant p-value was for the closed-reference analysis, which is the only analysis that provides consistent OTU labels.
-Finally, the authors built off of this analysis to count the number of OTUs that were differentially represented between the subsetted and full datasets by each method.
-This entire analysis assumed that the OTUs generated using the full dataset were correct, which was an unsubstantiated assumption since the authors did not assess the quality of the OTU assignments.
-
-This re-analysis of the He study raised five complementary questions.
-First, how do the various methods vary in the quality of their OTU assignments?
-Second, how generalizable are these results to modern datasets generated using a large number of sequences that were deeply sequenced?
-Third, how does the stability and quality of OTU assignments generated by new methods compare to those analyzed in the He study?
-Fourth, are there open-source alternatives to USEARCH that perform just as well?
-Finally, although the stability of reference-based methods did not appear to be impacted by the input order of the sequences to be assigned to OTUs, is the stability of reference-based methods impacted by the order of the reference sequences?
-In the remainder of the Results and Discussion section we address each of these questions.
-
-***How do the various methods vary in the quality of their OTU assignments?*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs.
-A method can generate highly stable OTUs, but the OTU assignments may be meaningless if they poorly represent the specified cutoff and the actual distance between the sequences.
-To assess the quality of OTU assignments by the various methods, we made use of the pairwise distance between the unique sequences to count the number of true positives and negatives and the number of false positives and negatives for each method and sampling depth.
-Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and negatives using the MCC.
-We used the average MCC value as a measure of a method's quality and its variation as a measure of its consistency.
-We made three important observations.
-First, each of the *de novo* methods varied in how sensitive their MCC values were to additional sequences (Figure 1C).
-The SL and CL methods were the most sensitive; however, the quality of the OTU assignments did not meaningfully differ when 80 or 100% of the data were assigned to OTUs using the *de novo* methods.
-Second, the AL method had higher MCC values than the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL (Figure 1D).
-Third, with the possible exception of the CL method, the MCC values for each of the methods only demonstrated a small amount of variation between runs of the method with a different ordering of the input sequences.
-This indicates that although there may be variation between executions of the same method, they produced OTU assignments that were of equal quality.
-Revisiting the concept of stability, we question the value of obtaining stable OTUs when the full dataset is not optimally assigned to OTUs.
-Our analysis indicates that the most optimal method for assigning the Canadian soils sequences to OTUs using a 97% threshold was the AL method.
-
-```{r region_overlap}
-n_overlap <- read.table(file="data/gg_13_8/gg_13_8_97.overlap.count", row.names=1)
-n_v4_overlap <- n_overlap["v4", 1]
-n_v9_overlap <- n_overlap["v9", 1]
-n_v19_total <- n_overlap["v19", 1]
-```
-
-```{r miseq_analysis}
-miseq_info <- read.table(file="data/miseq/miseq.seq.info", row.names=1)
-n_groups <- miseq_info["n_groups",1]
-n_miseq_seqs <- miseq_info["total_seqs",1]
-n_unique_miseq_seqs <- miseq_info["unique_seqs",1]
-
-miseq_stability <- read.table(file="data/process/miseq.mcc_ref.summary", header=T)
-miseq_mcc <- read.table(file="data/process/miseq.mcc.summary", header=T)
-```
-
-***How generalizable are these results to modern datasets generated using a large number of sequences that were deeply sequenced?*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering methods.
-First, it was one of the earliest 16S rRNA gene sequence datasets published using the 454 FLX platform.
-Developments in sequencing technology now permit the sequencing of millions of sequences for a study.
-In addition, because the original Phred quality scores and flowgram data are not available, it was not possible for us to adequately remove sequencing errors [@Schloss2011Reducing].
-The large number of sequencing errors that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering methods.
-Second, the dataset used in the He study covered the V9 region of the 16S rRNA gene.
-For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference methods.
-Of the `r n_v19_total` sequences in the default QIIME database, only `r n_v9_overlap` fully cover the V9 region.
-In contrast, `r n_v4_overlap` of the sequences fully covered the V4 region.
-Inadequate coverage of the V9 region would adversely affect the ability of the reference-based methods to assign sequences to OTUs.
-Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene [@Schloss2010].
-With these points in mind, we compared the clustering assignment for each of these methods using a time series experiment that was obtained using mouse feces [@Schloss2012Stabilization; @Kozich2013].
-The MiSeq platform was used to generate `r n_miseq_seqs` sequences from the V4 region of the 16S rRNA gene of `r n_groups` samples.
-Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02% [@Kozich2013].
-Although no dataset is perfect for exhaustively testing these clustering methods, this dataset was useful for demonstrating several points.
-First, when using 60% of the data, the stability relationships amongst the different methods were similar to what we observed using the Canadian soil dataset (Figure 3AB).
-With the exception for the clusters generated using CL, the methods all performed very well with stabilities greater than `r round(min(miseq_stability[miseq_stability$fraction=="0.6" & miseq_stability$method != "fn","mean"]), 2)`.
-Second, the MCC values calculated relative to the distances between sequences were generally higher than was observed for the Canadian soil dataset for all of the methods except the CL and SL methods.
-Surprisingly, the MCC values for the DGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "dgc","mean"],2)`)  and AGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "agc","mean"],2)`) methods were comparable to the AL method (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "an","mean"],2)`; Figure 3CD).
-This result suggests that the optimal method is likely to be database-dependent.
-
-Finally, as was observed with the Canadian soil dataset, there was little variation in MCC values observed among the 30 randomizations.
-Therefore, although the methods have a stochastic component, the OTU assignments did not vary meaningfully between runs.
-The results from both the Canadian soil and murine microbiota datasets demonstrate that the *de novo* methods can generate stable OTU assignments and that the the overall quality of the assignments were consistent.
-Most importantly, these analyses demonstrate that the OTU assignments using the AL, AGC, and DGC *de novo* methods were consistently better than either of the reference-based methods.
-
-***How does the stability and quality of OTU assignments generated by new methods compare to those analyzed in the He study?*** The Swarm algorithm is a recently proposed *de novo* method for assigning sequences to OTUs that uses user-defined parameters to break up chains generated by SL clustering [@Mah2014].
-Swarm was originally validated by comparing the results against the expected clusters formed based on the taxonomic composition of a mock community.
-Similar to the authors of the He study, the Swarm developers suggest that methods are needed that are insensitive to input order.
-Use of Swarm on the Canadian soil and murine datasets demonstrated that similar to the other *de novo* methods, Swarm's OTU assignments changed as sequences were added (Figures 1A and 3A).
-When we compared the OTU assignments for both datasets when using all of the sequence data, the variation in MCC values across the 30 randomizations were not meaningfully different (Figures 1D and 3D).
-Most importantly, when we selected the distance threshold that optimized the MCC value, the quality of the OTU assignments was close to that of the AL assignments when using the Canadian soil dataset and considerably worse than that of the murine dataset (Figures 1D and 3D).
-Interestingly, the distance thresholds that resulted in the largest MCC values were 3 and 2% for the Canadian soil and murine datasets, respectively.
-This suggests that distance-based OTU definitions are not consistent across datasets when using the Swarm algorithm, although they do appear to be within the neighborhood of 3%.
-Finally, the Swarm developers indicated that hierarchical *de novo* algorithms were too impractical to use on large MiSeq-generated datasets.
-Our ability to apply AL to the large mouse dataset and even larger datasets suggests that it is not necessary to sacrifice OTU assignment quality for speed [e.g.
-@Schubert2015; @Zackular2015].
-
-***Are there open-source alternatives to USEARCH that perform just as well?*** For some datasets the AGC and DGC methods appear to perform as well or better than the hierarchical clustering methods.
-As originally described in the He study, the AGC and DGC methods utilized the USEARCH program and the DGC method is used for clustering in UPARSE [@Edgar2010; @Edgar2013].
-The source code for USEARCH is not publicly available and only the 32-bit executables are available for free to academic users.
-Access for non-academic users and those needing the 64-bit version is available commercially from the developer.
-An alternative to USEARCH is VSEARCH, which is being developed in parallel to USEARCH as an open-source alternative.
-One subtle difference between the two programs is that USEARCH employs a heuristic to generate candidate alignments whereas VSEARCH generates the actual global alignments.
-The VSEARCH developers claim that this difference enhances the sensitivity of VSEARCH relative to USEARCH.
-Using the two datasets, we determined whether the AGC and DGC methods, as implemented by the two programs, yielded OTU assignments of similar quality.
-In general the overall trends that we observed with the USEARCH-version of AGC and DGC were also observed with the VSEARCH-version of the methods (Figure 4).
-When we compared the two implementations of the AGC and DGC methods, the OTUs generated by the VSEARCH-version of the methods were as stable or more stable than the USEARCH-version when using 60% of the datasets.
-In addition, the MCC values for the entire datasets, calculated relative to the distance matrix, were virtually indistinguishable.
-These results are a strong indication that VSEARCH is a suitable and possibly better option for executing the AGC and DGC methods.
-
-```{r rand_ref_analysis}
-rand_ref_u <- read.table(file="data/rand_ref/hits.uclosed.summary", row.names=1)
-rand_ref_v <- read.table(file="data/rand_ref/hits.vclosed.summary", row.names=1)
-
-uclosed <- read.table(file="data/rand_ref/closed_ref.usearch.sensspec", header=T, row.names=1)[2:31,]
-vclosed <- read.table(file="data/rand_ref/closed_ref.vsearch.sensspec", header=T, row.names=1)[2:31,]
-
-miseq_closed_ref <- read.table(file="data/rand_ref/closed_ref.redundancy.analysis", header=T, row.names=1)
-```
-
-***Is the stability of reference-based methods impacted by the order of the reference sequences?*** The He study and our replication attempt validated that the closed-reference method generated perfectly stable OTUs.
-This was unsurprising since, by definition, the method is designed to return one-to-one mapping of reads to a reference.
-Furthermore, because it treats the input sequences independently the input order or use of a random number generator is not an issue.
-An important test that was not performed in the He study was to determine whether the clustering was sensitive to the order of the sequences in the database.
-The default database used in QIIME, which was also used in the He study, contains full-length sequences that are at most 97% similar to each other.
-We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference method with the full murine dataset, which contained `r n_unique_miseq_seqs` unique sequences (Figure 5).
-Surprisingly, we observed that the number of OTUs generated was not the same in each of the randomizations.
-On average there were `r round(rand_ref_u["mean", 1])` sequences that mapped to a reference OTU per randomization (range from `r rand_ref_u["min", 1]` to `r rand_ref_u["max", 1]`).
-The original ordering of the reference resulted in `r rand_ref_u["original", 1]` sequences being mapped, less than the minimum observed number of mapped sequences when the references were randomized.
-This surprising result was likely due to the performance of the USEARCH heuristic.
-To test this further, we substituted VSEARCH for USEARCH in the closed-reference method.
-When we used VSEARCH the original ordering of the reference sequences and all randomizations were able to map `r rand_ref_v["original", 1]` sequences to reference OTUs.
-When we calculated the true distance between each of the murine sequences and the references, we were able to map `r nrow(miseq_closed_ref)` of the murine sequences to the reference sequences when using a 97% similarity threshold without the use of a heuristic.
-This result indicates that the closed reference approach, whether using USEARCH or VSEARCH, does not exhaustively or accurately map reads to the closest reference.
-To quantify this further, we calculated the MCC for the USEARCH and VSEARCH assignments relative to the assignments using the non-heuristic approach.
-Using USEARCH the average MCC was `r mean(uclosed$mcc)` (range: `r min(uclosed$mcc)` to `r max(uclosed$mcc)`) and using VSEARCH the average MCC was `r mean(vclosed$mcc)` (range: `r min(vclosed$mcc)` to `r max(vclosed$mcc)`).
-The two methods had similar sensitivities (USEARCH: `r mean(uclosed$sensitivity)` and VSEARCH: `r mean(vclosed$sensitivity)`), but the USEARCH specificity (`r mean(uclosed$specificity)`) was considerably higher than VSEARCH (`r mean(vclosed$specificity)`).
-Overall, these results indicate that although heuristic approaches may be fast, they do a poor job of mapping reads to the correct reference sequence relative to non-heuristic approaches.
-
-```{r database_analysis}
-small_dists <- read.table(file="data/gg_13_8/gg_13_8_97.small_dist.count", row.names=1)
-n_small_v4_dist <- small_dists["v4",1]
-n_small_v19_dist <- small_dists["v19",1]
-
-n_v4_unique <- length(scan(file="data/gg_13_8/gg_13_8_97.v4_ref.names", what="", sep="\n", quiet=T))
-
-u_overlap <- read.table(file="data/rand_ref/overlap.uclosed.summary")
-v_overlap <- read.table(file="data/rand_ref/overlap.vclosed.summary")
-
-dups <- read.table(file="data/gg_13_8/duplicate.analysis")
-n_true_mappings <- nrow(dups)
-max_taxa <- dups[dups$V3 == max(dups$V3),]
-```
-
-We also observed that regardless of whether we used USEARCH or VSEARCH, the reference OTU labels that were assigned to each OTU differed between randomizations.
-When we used USEARCH to perform closed-reference clustering, an average of `r round(100*u_overlap["orig", "mean"], 2)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*u_overlap["orig", "min"], 2)` to `r round(100*u_overlap["orig", "max"], 2)`%).
-If we instead used VSEARCH an average of `r round(100*v_overlap["orig", "mean"], 2)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*v_overlap["orig", "min"], 2)` to `r round(100*v_overlap["orig", "max"], 2)`%).
-To better understand this result, we further analyzed QIIME's reference database.
-We hypothesized that within a given region there would be sequences that were more than 97% similar and possibly identical to each other.
-When a sequence was used to search the randomized databases, it would encounter a different reference sequence as the first match with each randomization.
-Among the `r n_v4_overlap` reference sequences that fully overlap the V4 region, there were `r n_small_v19_dist` pairs of sequences that were more than 97% similar to each other over the full length of the 16S rRNA gene.
-When the extracted V4 sequences were dereplicated, we identified `r n_v4_unique` unique sequences.
-Among these dereplicated V4 sequences there were `r n_small_v4_dist` pairs of sequences that were more than 97% similar to each other.
-The presence of duplicate and highly similar V4 reference sequences explains the lack of labeling stability when using either USEARCH or VSEARCH to carry out the closed-reference method.
-We suspect that the reference database was designed to only include sequences that were at most 97% similar to each other as a way to overcome the limitations of the USEARCH search heuristic.
-
-```{r }
-n_actual_map <- nrow(miseq_closed_ref)
-
-n_map_more_than_one_expanded <- sum(miseq_closed_ref$n_dups_expanded > 1)
-n_map_one_uniqued <- sum(miseq_closed_ref$n_dups_expanded>1 & miseq_closed_ref$n_dups_uniqued==1)
-n_map_more_than_one_uniqued <- sum(miseq_closed_ref$n_dups_expanded>1 & miseq_closed_ref$n_dups_uniqued>1)
-n_mismatch_taxonomy <- sum(table(miseq_closed_ref$n_taxa_expanded, miseq_closed_ref$n_dups_expanded)[-1,-1])
-```
-
-Beyond comparing the abundance of specific OTUs across samples, the reference database is used in the open and closed-reference methods to generate OTU labels that can be used in several downstream applications.
-It is commonly used to extract information from a reference phylogenetic tree to carrying out UniFrac-based analyses [@Hamady2009] and to identify reference genomes for performing analyses such as PICRUSt [@Langille2013].
-Because these downstream applications depend on the correct and unique labeling of the OTUs, the lack of label stability is problematic.
-As one illustration of the effects that incorrect labels would have on an analysis, we asked whether the duplicate sequences had the same taxonomies.
-Among the `r sum(dups$V2==2)` V4 reference sequences that had one duplicate, `r sum(dups$V2==2 & dups$V3==2)` had discordant taxonomies.
-Furthermore, among those `r sum(dups$V2 > 2)` V4 reference sequences with two or more duplicates, `r sum(dups$V2>2 & dups$V3>=2)` had discordant taxonomies.
-Two V4 reference sequences mapped to `r max_taxa[1,2]-1` and `r max_taxa[2,2]-1` duplicate sequences and both contained `r max_taxa[1,3]` different taxonomies.
-Among the V4 sequences within the database, there was also a sequence that had `r max(dups[,2])-1` duplicates and represented `r dups[which.max(dups[,2]),3]` different taxonomies.
-When we analyzed the `r n_actual_map` sequences that mapped to the V4 reference sequences using a non-heuristic approach, we observed that `r n_map_more_than_one_expanded` of the sequences mapped to more than one reference sequence.
-Of these sequences, `r n_map_one_uniqued` (`r 100*n_map_one_uniqued/n_map_more_than_one_expanded`%) mapped to references that were identical over the V4 region and `r n_map_more_than_one_uniqued` (`r 100*n_map_more_than_one_uniqued/n_map_more_than_one_expanded`%) mapped equally well to two or more references that were not identical over the V4 region.
-Among the combined `r n_map_more_than_one_expanded` sequences that mapped to multiple reference sequences, the taxonomy of the multiple reference sequences conflicted for `r n_mismatch_taxonomy` (`r 100*n_mismatch_taxonomy/n_map_more_than_one_expanded`%).
-Together, these results demonstrate some of the considerable problems with the reference-based clustering of sequences.
-
-## Conclusions
-It is worth noting that the analysis from the Roesch study that motivated the He study is not typical of microbial ecology studies.
-First, their analysis was based on a single soil sample.
-Researchers generally have dozens or hundreds of samples that are pooled and clustered together to enable comparison across samples.
-Second, all of the sequence data from these datasets is pooled for a single analysis.
-Rarely would a researcher rarefy their data prior to clustering since it can be more efficiently done after all of the data are assigned to OTUs.
-Third, the CL method used in the original Roesch study has since been shown to not generate optimal OTUs [@Schloss2011Assessing].
-As for the approach used in the He study, the value of identifying stable OTUs is unclear.
-Although there is concern that running the methods multiple times yields different clusterings, we have shown that there is little variation in their quality.
-This suggests that the different clusterings by the same method are equally good.
-Greater emphasis should be placed on obtaining an optimal balance between splitting similar sequences into separate OTUs and merging disparate sequences into the same OTU.
-
-The approach of the current study quantified the effects of merging and splitting by using an objective metric.
-Through the use of the pairwise distances between sequences, we were able to use the MCC to demonstrate that, in general, the AL method was consistently the optimal method for each dataset, but that Swarm, AGC, and DGC sometimes perform as well as AL.
-At least for the murine dataset, Swarm also could be among the methods that performed poorly.
-It is impossible to obtain a clustering with no false positives or false negatives and the optimal method may vary by dataset.
-With this in mind, researchers are encouraged to calculate and report their MCC values and to use these values to justify using methods other than the AL.
-As an alternative to the He study's method of measuring stability, we propose using the variation in the quality of the clustering of the full dataset.
-Given the tight 95% confidence intervals shown in Figures 1D and 3D, with the exception of CL, it is clear that this variation is quite small.
-This indicates that although the order of the sequences being clustered can affect the actual cluster assignments, the quality of those different clusterings is not meaningfully different.
-
-Our analysis of those methods that implemented USEARCH as a method for clustering sequences revealed that its heuristic limited its specificity.
-When we replaced USEARCH with VSEARCH, the clustering quality was as good or better.
-Although there may be parameters in USEARCH that can be tuned to improve the heuristic, these parameters are likely dataset dependent.
-Based on the data presented in this study, its availability as an open source, and free program, VSEARCH should replace USEARCH in the *de novo* clustering methods; however, USEARCH performed better than VSEARCH for closed-reference clustering.
-Furthermore, although not tested in our study, VSEARCH can be parallelized leading to potentially significant improvements in speed.
-Although USEARCH and VSEARCH do not utilize aligned sequences, it is important to note that a sequence curation pipeline including denoising, alignment, trimming to a consistent region of the 16S rRNA gene, and chimera checking are critical to making proper inferences [@Schloss2011Reducing; @Kozich2013; @Schloss2012Secondary].
-
-We have assessed the ability of reference-based clustering methods to capture the actual distance between the sequences in a dataset in parallel with *de novo* methods.
-Several studies have lauded both the open and closed-reference approaches for generating reproducible clusterings [@He2015; @Rideout2014; @NavasMolina2013], yet we have shown that both reference-based approaches did a poor job of representing the distance between the sequences compared to the *de novo* approaches.
-Although the OTU assignments are reproducible and stable across a range of library sizes, the reference-based OTU assignments are a poor representation of the data.
-We also observed that the assignments were not actually reproducible when the order of the reference sequences was randomized.
-When USEARCH was used, the actual number of sequences that mapped to the reference changed depended on the order of the reference.
-Perhaps most alarming was that the default order of the database provided the worst MCC of any of the randomizations we attempted.
-This has the potential to introduce systematic a bias rather than a random error.
-Even when we used VSEARCH to perform closed-reference clustering and were able to obtain a consistent clusterings, we observed that the labels on the OTUs differed between randomizations.
-Because the OTU labels are frequently used to identify representative sequences for those OTUs, variation in labels, often representing different taxonomic groups, will have a detrimental effect on the interpretation of downstream analyses.
-
-Because the open-reference method is a hybrid of the closed-reference and DGC methods, it is also negatively affected by the various problems using USEARCH.
-An added problem with the open-reference method is that the two phases of the method employ different thresholds to define its OTUs.
-In the closed-reference step, sequences must be within a threshold of a reference to be in the same OTU.
-This means that in the worst case scenario two sequences that are 97% similar to a reference and are joined into the same OTU, may only be 94% similar to each other.
-In the DGC step, the goal is to approximate the AL method which requires that, on average, the sequences within an OTU are, on average, 97% similar to each other.
-The end result of the open-reference approach is that sequences that are similar to previously observed sequences are clustered with one threshold while those that are not similar to previously observed sequences are clustered with a different threshold.
-
-As the throughput of sequencing technologies have improved, development of clustering algorithms must continue to keep pace.
-*De novo* clustering methods are considerably slower and more computationally intensive than reference-based methods and the greedy *de novo* methods are faster than the hierarchical methods.
-In our experience [@Kozich2013], the most significant detriment to execution speed of the *de novo* methods has been the inadequate removal of sequencing error and chimeras.
-As the rate of sequencing error increases so do the number of unique sequences that must be clustered.
-The speed of the *de novo* methods scales approximately quadratically, so that doubling the number of sequences results in a four-fold increase in the time required to execute the method.
-The rapid expansion in sequencing throughput has been likened to the Red Queen in Lewis Carroll's, *Through the Looking-Glass* who must run in place to keep up to her changing surroundings [@Schloss2009].
-Microbial ecologists must continue to refine clustering methods to better handle the size of the datasets, but they must also take steps to improve the quality of the underlying data.
-Ultimately, objective standards must be applied to assess the quality of the data and the quality of OTU clustering.
+**Figure 1.
+INSERT LEGEND HERE** Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
 
@@
 
-## Figures
-
-**Figure 1.
-Comparison of the stability (A, B) and quality (C, D) of *de novo* and reference-based clustering methods using the Canadian soil dataset.** The average stability of the OTUs was determined by calculating the MCC with respect to the OTU assignments for the full dataset using varying sized subsamples.
-The quality of the OTUs was determined by calculating the MCC with respect to the distances between the sequences using varying sized subsamples.
-Thirty randomizations were performed for each fraction of the dataset and the average and 95% confidence interval are presented when using 60% of the data.
-The vertical gray lines in A and C indicates the fraction of the dataset represented in B and D, respectively.
-The color and shape of the plotting symbol is the same between the different panels and is described along the x-axis of panel D.
-The optimum threshold for the Swarm-generated assignments was 3%.
-
-**Figure 2.
-The clustering methods varied in their ability to generate the same number of OTUs using a subset of the data as were observed when the full dataset was rarefied.** The subsetted data are depicted by closed circles and the data from the rarefied full dataset is depicted by the open circles.
-
-**Figure 3.
-Comparison of the stability (A, B) and quality (C, D) of *de novo* and reference-based clustering methods using the murine dataset.** The average stability of the OTUs was determined by calculating the MCC with respect to the OTU assignments for the full dataset using varying sized subsamples.
-The quality of the OTUs was determined by calculating the MCC with respect to the distances between the sequences using varying sized subsamples.
-Thirty randomizations were performed for each fraction of the dataset and the average and 95% confidence interval are presented when using 60% of the data.
-The vertical gray lines in A and C indicates the fraction of the dataset represented in B and D, respectively.
-The color and shape of the plotting symbol is the same between the different panels and is described along the x-axis of panel D.
-The optimum threshold for the Swarm-generated assignments was 2%.
-
-**Figure 4.
-The stability and quality of USEARCH and VSEARCH OTUs generated by the AGC and DGC methods were similar.**  The stability of the OTUs was determined by calculating the MCC for OTUs calculated using 60% of the data relative to the OTU assignments for the full dataset.
-The quality of the OTUs was determined by calculating the MCC of the OTUs calculated using the full dataset with respect to the distances between the sequences.
-The error bars represent the 95% confidence interval across the 30 randomizations.
-
-**Figure 5.
-The number of closed-reference OTUs observed in the murine dataset when using USEARCH, VSEARCH, and without a heuristic.** In addition to the default ordering of the references provided with the QIIME package, the reference sequences were randomized 30 times; the order of the murine dataset was not randomized.
-Regardless of whether the default or randomized ordering was used, the number of OTUs generated using VSEARCH did not differ.
-The non-heuristic approach calculated the exact distance between the murine sequences and the the reference sequences and assigned the sequences to the reference with the smallest distance.
-
-\newpage
-
-\includepdf{results/figures/figure_1.pdf}
-\includepdf{results/figures/figure_2.pdf}
-\includepdf{results/figures/figure_3.pdf}
-\includepdf{results/figures/figure_4.pdf}
+\begin{comment}
+	\includepdf{../../results/figures/commentary_figure_1.pdf}
+\end{comment}
 

### 2016-02-24 papers/msystems_2016/Schloss_Commentary_mSystems_2016.Rmd
First draft of commentary

@@
 ## Abstract
+150 words
 
@@
 
-INSERT TEXT HERE
+* Add new Swarm reference
+
+The ability to assign 16S rRNA gene sequences to operational taxonomic units (OTUs) allows microbial ecologists to overcome the inconsistencies and biases within bacterial taxonomy and provides a strategy for clustering similar sequences that do not have representatives in a reference database.
+These sequences are clustered into OTUs based on their distance from (or similarity to) each other.
+Numerous algorithms for solving this seemingly simple problem have blossomed in recent years and were recently the subject of benchmarking studies performed by Westcott and myself [ref x 2], He et al.
+[ref], and Kopylova et al [ref].
+These studies provide a thorough review of the sequencing clustering landscape, which can be divided into three general approaches: (i) *de novo* clustering where sequences are clustered without first mapping sequences to a reference database, (ii) closed-reference clustering where sequences are clustered based on the references that the sequences map to, and (iii) open reference clustering where sequences that do not map adequately to the reference are then clustered using a *de novo* approach.
+My studies have highlighted a persistent problem in the development of clustering algorithms, which is assessing the quality of the clustering assignments.
+
+The most recent analysis of Kopylova et al [@ref] repeats many of the benchmarking strategies employed by previous researchers.
+First, they and others have compared the time and memory required to cluster sequences in a dataset [@Sun2009; @Cai2011; @Rideout2014; @Mah2014].
+These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the OTU assignments.
+In the extreme, we could easily develop a toy a clustering algorithm could randomly assign sequences to a predetermined number of OTUs and be efficient, but also poorly reflect the genetic diversity within the community.
+Second, they and others have compared the number of OTUs generated by various methods for a common dataset [@refs].
+Again, the toy clustering algorithm could cluster to a target number of OTUs, but the clusterings would likely be meaningless.
+Third, because many of the methods are sensitive to the initial order of the sequences, a metric of OTU stability has been proposed as a way to assess algorithms [@He2015].
+Although it is important that the methods generate reproducible OTU assignments, this approach ignores the possability that the variation in assignments may be equally robust or that the assignemnts by a highly reproducible algorithm may be quite poor.
+Fourth, the method that Kopylova et al.
+[@ref] relied upon the most was to cluster sequences from simulated data or data from synthetic communities of cultured organisms and quantify how well the OTU assignments matched the organisms' taxonomy [@Edgar2013; @Mah2014; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014; @Cai2011; @Sun2011; @White2010].
+This approach is flawed because bacterial taxonomy often reflects those inconsistencies and biases within bacterial taxonomy that OTU-based methods strive to overcome.
+Furthermore, this benchmarking strategy can only be applied when the actual taxonomy of the organisms are known and so it is unclear how the methods scale to sequences from the novel organisms we are likely to encounter in deep sequencing surveys.
+In a final approach, Kopylova and others have assessed the quality of clustering based on their ability to generate the same OTUs generated by other methods [@Schmidt2014Limits; @Rideout2014].
+Unfortunately, without the ability to ground truth any method, such comparisons are tenuous.
+Westcott and I have proposed an unbiased and objective method for assessing the quality of OTU assignments that can be applied to any collection of sequences [@Schloss2011Assessing; @Westcott2015].
+
+Our approach uses the observed dissimilarity between pairs of sequences and information about whether sequences were clustered together to quantify how well similar sequences are clustered together and dissimilar sequences are clustered apart.
+To quantify the correlation between the observed and expected OTU assignments, we synthesize the relationship between OTU assignments and the distances between sequences using the Matthew's correlation coefficient [MCC; @Matthews1975].
+In the most recent application of this approach [@Westcott2015], we found that closed-reference clustering algorithms could be sensitive to the order of the sequences in the reference database and frequently clustered sequences together that were more than 3% different from each other.
+We also discussed that because open-reference clustering was dependent on closed-reference clustering it had sensitivity to the order of the database.
+Furthermore, we described how it had a nebulous threshold for OTUs since sequences are clustered based on a radius of 3% under the closed-reference phase and a diameter of 3% under the open reference phase.
+Finally, we showed that *de novo* clustering algorithms generated the most robust OTU assignments and confirmed our previous analysis that the average neighbor algorithm consistently performed the best [@Schloss2011Assessing].
+Given the observation that the best algorithm may vary by dataset we concluded that researchers should quantify the MCC for several *do novo* algorithms before selecting an algorithm.
+
+To revisit these results, I have expanded the analysis to incorporate additional *de novo* and closed-reference algorithms and two simulated datasets following the approach described by Kopylova et al [@Kopylova2016].
+I evaluated three heirarchical and XXXXX greedy *de novo* algorithms, one open-reference clustering algorithm, and three closed-reference algorithms (Figure 1).
+To test these approaches I applied each of them to four datasets from soil [@Roesch2007], mouse feces [@Schloss2012Stabilization], and two simulated communities.
+The simulated communities were generated by randomly selecting 10,000 16S rRNA sequences that were unique within the V4 region from the SILVA non-redundant database [@ref].
+Next, an even community was generated by specifying that each sequence had a frequency of 100 reads and a staggered community was generated by specifying that the abundance of each sequence was a randomly drawn a uniform distribution between 1 and 200.
+The benchmarking approach is the same as the approach used previously [@Westcott2015].
+A reproducible version of this manuscript and analysis has been added to the repository available at https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015.
+
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+Although the goal of Kopylova et al.
+[-@Kopylova2016] was to compare various clustering algorithms, they also studied these algorithms in the broader context of raw sequence processing, screening for chimeras, and removal of singletons.
+Each of these are critical decisions in a comprehensive pipeline; however, they confound the analysis of how best to cluster sequences into OTUs that reflect a specific distance threshold.
+Through the use of objective criteria that measure the quality of the clusterings, independent of taxonomy or database, researchers will be able to evaluate which clustering algorithm is the best fit for their data.
 
@@
 **Figure 1.
-INSERT LEGEND HERE** Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+Comparison of OTU quality generated by multiple algorithms applied to four datasets.** The nearest, average, and furthest neighbor clustering algorithms were used as implemented in mothur (v.1.37).
+Abundance (AGC) and Distance-based greedy clustering (DGC) were implemented using USEARCH (v.6.1) and VSEARCH (v.1.5.0).
+Other *de novo* clustering algorithms included Swarm (v.2.1.1), OTUCLUST (v.0.1), and SUMACLUST (v.1.0.20).
+The MCC values for Swarm were determined by selecting the distance threshold that generated the maximum MCC value for each dataset.
+The USEARCH and SortMeRNA (v.2.0) closed-reference clusterings were performed using QIIME (v.1.9.1) and the VSEARCH closed-reference clusterings were performed substituting VSEARCH for USEARCH in the QIIME workflows.
+The order of the sequences in each dataset was randomized thirty times and the intra-method range in MCC values was smaller than the plotting symbol.
+MCC values were calculated in mothur.
+
+Nearest neighbor
+Average neighbor
+Furthest neighbor
+USEARCH AGC
+VSEARCH AGC
+USEARCH DGC
+VSEARCH DGC
+Swarm
+OTUClust
+SumaClust
+Open-reference
+USEARCH closed-reference
+VSEARCH closed-reference
+SortMeRNA closed-reference
 

### 2016-02-24 papers/msystems_2016/Schloss_Commentary_mSystems_2016.Rmd
Edit legend to remove VSEARCH from closed-reference clustering

@@
 The MCC values for Swarm were determined by selecting the distance threshold that generated the maximum MCC value for each dataset.
-The USEARCH and SortMeRNA (v.2.0) closed-reference clusterings were performed using QIIME (v.1.9.1) and the VSEARCH closed-reference clusterings were performed substituting VSEARCH for USEARCH in the QIIME workflows.
+The USEARCH and SortMeRNA (v.2.0) closed-reference clusterings were performed using QIIME (v.1.9.1).
 The order of the sequences in each dataset was randomized thirty times and the intra-method range in MCC values was smaller than the plotting symbol.

### 2016-02-24 papers/msystems_2016/Schloss_Commentary_mSystems_2016.Rmd

@@
 To revisit these results, I have expanded the analysis to incorporate additional *de novo* and closed-reference algorithms and two simulated datasets following the approach described by Kopylova et al [@Kopylova2016].
-I evaluated three heirarchical and XXXXX greedy *de novo* algorithms, one open-reference clustering algorithm, and three closed-reference algorithms (Figure 1).
+I evaluated three heirarchical and seven greedy *de novo* algorithms, one open-reference clustering algorithm, and two closed-reference algorithms (Figure 1).
 To test these approaches I applied each of them to four datasets from soil [@Roesch2007], mouse feces [@Schloss2012Stabilization], and two simulated communities.

### 2016-02-25 code/run_rand_sref.sh
Got rand_ref version of SortMeRNA to run

@@ -1,17 +1,32 @@
 REF=$1
-#REF=data/rand_ref/rand_ref_1.0_03.fasta
+#REF=data/rand_ref/rand_ref_1.0_04.fasta
 
 FASTA=data/rand_ref/miseq.fasta
-CLOSED_PATH=$(echo $REF | sed 's/fasta/sclosed/')
 
-IDX_FILE=$(echo $REF | sed 's/fasta/idx/')
 
-rm -rf $CLOSED_PATH/
+# For whatever reason, the reference idx files cannot be in the same linear
+# path as the folder that the output goees to. here we set up a folder for the
+# idx files that are next to the output folder. if we don't do this the computer
+# just spins its wheels for a long time (forever?). Note that we generate new
+# idx files for each randomization of the reference. Not sure whether this is
+# necessary, but better safe than sorry and it makes sense that it be done.
 
+IDX_FOLDER=$(echo $REF | sed 's/fasta/idx/')
+IDX_FILE=$(echo $IDX_FOLDER/reference.idx)
+
+mkdir -p $IDX_FOLDER
 indexdb_rna --ref $REF,$IDX_FILE -v
+
+
+# The clustering...
+
+CLOSED_PATH=$(echo $REF | sed 's/fasta/sclosed/')
+rm -rf $CLOSED_PATH/
 
 parallel_pick_otus_sortmerna.py -i $FASTA -o $CLOSED_PATH -r $REF -T --jobs_to_start 1 --threads 10 --sortmerna_db $IDX_FILE
 
+
+# Cleaning up
+
 mv $CLOSED_PATH/miseq_otus.txt $CLOSED_PATH.sc
-rm -rf $CLOSED_PATH
-rm $IDX_FILE*
+rm -rf $CLOSED_PATH $IDX_FOLDER

### 2016-02-25 code/sortmerna_to_list.R
Fix output file names

@@ -6,7 +6,7 @@
 #he_1.0_01_otus.txt
 
 sortmerna_to_list <- function(sortmerna_folder_name){
-	stub <- gsub(".sortmerna", "_otus.txt", sortmerna_folder_name)
+	stub <- gsub(".*/(.*).sortmerna", "\\1_otus.txt", sortmerna_folder_name)
 	sortmerna_file <- paste0(sortmerna_folder_name, "/", stub)
 	sortmerna_data <- scan(sortmerna_file, what="", sep="\n", quiet=TRUE)
 	sortmerna_data <- gsub("^[^\t]*\t", "", sortmerna_data)

### 2016-02-25 code/summarize_rand_ref.R
Expand functionality to use rand_ref SortMeRNA data

@@ -8,10 +8,26 @@ summary_stats <- function(x){
 	c(min=min, max=max, median=median, mean=mean, sd=sd)
 }
 
+parse_sc_line <- function(line){
+	split_line <- unlist(strsplit(line, "\t"))
+	refrence <- split_line[1]
+	split_line <- split_line[-1]
+	n_repeats <- length(split_line)
+	rep(refrence, n_repeats)
+}
 
-get_hits <- function(cluster_file){
-	cluster_data <- read.table(file=cluster_file, stringsAsFactors=FALSE)
-	cluster_data[cluster_data$V1 == "H", 10]
+get_hits <- function(cluster_file, method){
+	hits <- as.numeric()
+
+	if(method == 's'){
+		cluster_data <- scan(file=cluster_file, what=character(), sep='\n', quiet=T)
+		listing <- lapply(cluster_data, parse_sc_line)
+		hits <- unlist(listing)
+	} else {
+		cluster_data <- read.table(file=cluster_file, stringsAsFactors=FALSE)
+		hists <- cluster_data[cluster_data$V1 == "H", 10]
+	}
+	return(hits)
 }
 
 
@@ -23,10 +39,8 @@ get_hits_summary <- function(rand_hits, orig_hits, method){
 	output_filename <- paste0("data/rand_ref/hits.", method, "closed.summary")
 	write.table(file=output_filename, x=output_data, row.names=TRUE, col.names=FALSE, quote=FALSE, sep="\t")
 
-       output_filename <- paste0("data/rand_ref/hits.", method, "closed.counts")
-       write.table(file=output_filename, c(rand_n_hits, orig_n_hits), row.names=c(1:30, "original"), col.names=FALSE, sep="\t", quote=FALSE)
-
-
+	output_filename <- paste0("data/rand_ref/hits.", method, "closed.counts")
+	write.table(file=output_filename, c(rand_n_hits, orig_n_hits), row.names=c(1:30, "original"), col.names=FALSE, sep="\t", quote=FALSE)
 }
 
 
@@ -68,7 +82,7 @@ get_overlap_summary <- function(rand_hits, orig_hits, method){
 
 	describe_overlap <- aggregate(overlap, by=list(rand_orig), summary_stats)
 	summarize_overlap <- describe_overlap$x
-	rownames(summarize_overlap) <- describe_overlap$Group.1 
+	rownames(summarize_overlap) <- describe_overlap$Group.1
 
[…]

### 2016-02-25 code/run_rand_sref.sh
Fix typo in otu file name

@@ -1,7 +1,6 @@
 REF=$1
-#REF=data/rand_ref/rand_ref_1.0_04.fasta
 
-FASTA=data/rand_ref/miseq.fasta
+FASTA=data/rand_ref/miseq.unique.fasta
 
 
 # For whatever reason, the reference idx files cannot be in the same linear
@@ -28,5 +27,5 @@ parallel_pick_otus_sortmerna.py -i $FASTA -o $CLOSED_PATH -r $REF -T --jobs_to_s
 
 # Cleaning up
 
-mv $CLOSED_PATH/miseq_otus.txt $CLOSED_PATH.sc
+mv $CLOSED_PATH/miseq*_otus.txt $CLOSED_PATH.sc
 rm -rf $CLOSED_PATH $IDX_FOLDER

### 2016-02-25 code/summarize_rand_ref.R
Get summarization of rand_ref analysis to run

@@ -25,7 +25,7 @@ get_hits <- function(cluster_file, method){
 		hits <- unlist(listing)
 	} else {
 		cluster_data <- read.table(file=cluster_file, stringsAsFactors=FALSE)
-		hists <- cluster_data[cluster_data$V1 == "H", 10]
+		hits <- cluster_data[cluster_data$V1 == "H", 10]
 	}
 	return(hits)
 }
@@ -66,7 +66,7 @@ get_overlap_summary <- function(rand_hits, orig_hits, method){
 	rand_orig <- rep("rand", counter-1)
 
 	A <- unique(orig_hits)
-	n_A <- length(A)	#[1] 1307
+	n_A <- length(A)
 
 	for(i in 1:length(rand_hits)){
 		B <- unique(rand_hits[[i]])

### 2016-02-25 code/cvsearch_to_list.R
Rough in usearch-based closed reference clustering

@@ -1,0 +1,11 @@
+form_otu <- function(line){
+	unlist(strplit(line, "\t"))[-1]
+}
+
+cvsearch_to_list <- function(clustered_file_name){
+	clustered_file_name <- "data/he/he_1.0_01.vclosed/vsearch/ref_clustered.uc"
+	clustered_data <- scan(clustered_file_name, what=character(), sep='\n', quiet=T)
+
+
+
+}

### 2016-02-25 code/run_cvsearch.sh

@@ -1,0 +1,15 @@
+FASTA=$1
+FASTA=data/he/he_1.0_01.fasta
+REF=data/references/97_otus.fasta
+
+CLOSED_PATH=$(echo $FASTA | sed 's/fasta/vclosed/')
+rm -rf $CLOSED_PATH/
+mkdir $CLOSED_PATH
+mkdir $CLOSED_PATH/vsearch
+
+code/vsearch/vsearch --sizeout --derep_fulllength $FASTA --minseqlength 30 --threads 1 --uc $CLOSED_PATH/vsearch/abundance_sorted.uc --output $CLOSED_PATH/vsearch/abundance_sorted.fna --strand both --log $CLOSED_PATH/vsearch/abundance_sorted.log
+code/vsearch/vsearch --maxaccepts 16 --id 0.97 --minseqlength 30 --threads 1 --wordlength 8 --uc $CLOSED_PATH/vsearch/ref_clustered.uc --maxrejects 64 --strand both --db $REF --log $CLOSED_PATH/vsearch/ref_clustered.log --usearch_global $CLOSED_PATH/vsearch/abundance_sorted.fna
+
+R -e "source(code/cvsearch_to_list.R); cvsearch_to_list.R('$CLOSED_PATH/vsearch/ref_clustered.uc')"
+
+rm -rf $CLOSED_PATH/

### 2016-02-25 code/cvsearch_to_list.R
Write scripts to process closed-reference vsearch clustering

@@ -1,11 +1,38 @@
-form_otu <- function(line){
-	unlist(strplit(line, "\t"))[-1]
-}
-
-cvsearch_to_list <- function(clustered_file_name){
+cvsearch_to_list <- function(unique_file_name, clustered_file_name){
+	unique_file_name <- "data/he/he_1.0_01.vclosed/vsearch/abundance_sorted.uc"
 	clustered_file_name <- "data/he/he_1.0_01.vclosed/vsearch/ref_clustered.uc"
-	clustered_data <- scan(clustered_file_name, what=character(), sep='\n', quiet=T)
 
+	uniqued <- read.table(file=unique_file_name, stringsAsFactors=FALSE)
 
+	names_first_column <- uniqued[uniqued$V1=="S", "V9"]
+	names_second_column <- names_first_column
 
+	hits <- uniqued[uniqued$V1=="H", ]
+
+	for(i in 0:(length(names_first_column)-1)){
+		dups <- paste(hits[hits$V2==i, "V9"], collapse=",")
+		names_second_column[i+1] <- paste(names_second_column[i+1], dups, sep=",")
+	}
+	names_second_column <- gsub(",$", "", names_second_column)
+	names(names_second_column) <- names_first_column
+
+	clustered <- read.table(file=clustered_file_name, stringsAsFactors=FALSE)
+	unique_sequence <- gsub(";size=.*", "", clustered$V9)
+	duplicate_sequences <- names_second_column[unique_sequence]
+
+	ref_sequence <- as.character(clustered$V10)
+
+	otu_list <- character()
+
+	for(i in 1:length(ref_sequence)){
+		if(is.na(otu_list[ref_sequence[i]])){
+			otu_list[ref_sequence[i]] <- duplicate_sequences[i]
+		} else {
+			otu_list[ref_sequence[i]] <- paste(otu_list[ref_sequence[i]], duplicate_sequences[i], sep=',')
+		}
+	}
+
+	list_file_data <- paste(c("userLabel", length(otu_list), otu_list), collapse='\t')
+	list_file_name <- gsub("/vsearch/abundance_sorted.uc", ".list", unique_file_name)
+	write(list_file_data, list_file_name)
 }

### 2016-02-25 code/run_cvsearch.sh

@@ -10,6 +10,9 @@ mkdir $CLOSED_PATH/vsearch
 code/vsearch/vsearch --sizeout --derep_fulllength $FASTA --minseqlength 30 --threads 1 --uc $CLOSED_PATH/vsearch/abundance_sorted.uc --output $CLOSED_PATH/vsearch/abundance_sorted.fna --strand both --log $CLOSED_PATH/vsearch/abundance_sorted.log
 code/vsearch/vsearch --maxaccepts 16 --id 0.97 --minseqlength 30 --threads 1 --wordlength 8 --uc $CLOSED_PATH/vsearch/ref_clustered.uc --maxrejects 64 --strand both --db $REF --log $CLOSED_PATH/vsearch/ref_clustered.log --usearch_global $CLOSED_PATH/vsearch/abundance_sorted.fna
 
-R -e "source(code/cvsearch_to_list.R); cvsearch_to_list.R('$CLOSED_PATH/vsearch/ref_clustered.uc')"
+	unique_file_name <- "data/he/he_1.0_01.vclosed/vsearch/abundance_sorted.uc"
+	clustered_file_name <- "data/he/he_1.0_01.vclosed/vsearch/ref_clustered.uc"
+
+R -e "source(code/cvsearch_to_list.R); cvsearch_to_list.R('$CLOSED_PATH/vsearch/abundance_sorted.uc', '$CLOSED_PATH/vsearch/ref_clustered.uc')"
 
 rm -rf $CLOSED_PATH/

### 2016-02-25 code/cvsearch_to_list.R
Remove names of test files from cvsearch analysis

@@ -1,6 +1,4 @@
 cvsearch_to_list <- function(unique_file_name, clustered_file_name){
-	unique_file_name <- "data/he/he_1.0_01.vclosed/vsearch/abundance_sorted.uc"
-	clustered_file_name <- "data/he/he_1.0_01.vclosed/vsearch/ref_clustered.uc"
 
 	uniqued <- read.table(file=unique_file_name, stringsAsFactors=FALSE)
 
@@ -33,6 +31,7 @@ cvsearch_to_list <- function(unique_file_name, clustered_file_name){
 	}
 
 	list_file_data <- paste(c("userLabel", length(otu_list), otu_list), collapse='\t')
-	list_file_name <- gsub("/vsearch/abundance_sorted.uc", ".list", unique_file_name)
+	list_file_name <- gsub("vclosed/vsearch/abundance_sorted.uc", "cvsearch.list", unique_file_name)
+	list_file_name <- gsub("ng.", "", list_file_name)
 	write(list_file_data, list_file_name)
 }

### 2016-02-25 code/run_cvsearch.sh

@@ -1,5 +1,4 @@
 FASTA=$1
-FASTA=data/he/he_1.0_01.fasta
 REF=data/references/97_otus.fasta
 
 CLOSED_PATH=$(echo $FASTA | sed 's/fasta/vclosed/')
@@ -10,9 +9,6 @@ mkdir $CLOSED_PATH/vsearch
 code/vsearch/vsearch --sizeout --derep_fulllength $FASTA --minseqlength 30 --threads 1 --uc $CLOSED_PATH/vsearch/abundance_sorted.uc --output $CLOSED_PATH/vsearch/abundance_sorted.fna --strand both --log $CLOSED_PATH/vsearch/abundance_sorted.log
 code/vsearch/vsearch --maxaccepts 16 --id 0.97 --minseqlength 30 --threads 1 --wordlength 8 --uc $CLOSED_PATH/vsearch/ref_clustered.uc --maxrejects 64 --strand both --db $REF --log $CLOSED_PATH/vsearch/ref_clustered.log --usearch_global $CLOSED_PATH/vsearch/abundance_sorted.fna
 
-	unique_file_name <- "data/he/he_1.0_01.vclosed/vsearch/abundance_sorted.uc"
-	clustered_file_name <- "data/he/he_1.0_01.vclosed/vsearch/ref_clustered.uc"
-
-R -e "source(code/cvsearch_to_list.R); cvsearch_to_list.R('$CLOSED_PATH/vsearch/abundance_sorted.uc', '$CLOSED_PATH/vsearch/ref_clustered.uc')"
+R -e "source('code/cvsearch_to_list.R'); cvsearch_to_list('$CLOSED_PATH/vsearch/abundance_sorted.uc', '$CLOSED_PATH/vsearch/ref_clustered.uc')"
 
 rm -rf $CLOSED_PATH/

### 2016-02-26 code/ninja_to_list.R
Add NINJA-OPS rules

@@ -1,0 +1,23 @@
+# Using the LOG options in NINJA-OPS creates a ninja_pass.log file where the
+# first column is the sequence name and the second column is the references
+# sequence that it mapped to. We need to convert this to a list file...
+
+library(dplyr)
+
+ninja_to_list <- function(folder_name){
+
+	folder_name <- "even_1.0_01.ng.ninja"
+	input_file_name <- paste0(folder_name, '/ninja_pass.log')
+	output_file_name <- paste0(folder_name, '.list')
+	output_file_name <- gsub("ng.", "", output_file_name)
+
+	mapping <- read.table(file=input_file_name, stringsAsFactors=FALSE)
+	mapping$V2 <- as.character(mapping$V2)
+
+	list_data <- mapping %>% group_by(factor(V2)) %>% summarise(otu=paste(V1, collapse=','))
+
+	list_output <- paste(c("userLabel", length(list_data), list_data), collapse='\t')
+
+	write(list_output, output_file_name)
+
+}

### 2016-02-26 code/run_ninja.sh

@@ -1,0 +1,16 @@
+FASTA=$1
+DB_STUB=code/NINJA-OPS/databases/greengenes97/greengenes97
+
+NINJA_FOLDER=$(echo $FASTA | sed 's/fasta/ninja/')
+
+rm -rf $NINJA_FOLDER
+mkdir $NINJA_FOLDER
+code/NINJA-OPS/bin/ninja_filter_linux $FASTA $NINJA_FOLDER/ninja D 1 LOG
+
+bowtie2-align-s --no-head -x $DB_STUB -S $NINJA_FOLDER/alignments.txt --np 0 --mp 1,1 --rdg 0,1 --rfg 0,1 --score-min L,0,-0.03 --norc -f $NINJA_FOLDER/ninja_filt.fa -p 4 -k 1 --very-sensitive
+
+code/NINJA-OPS/bin/ninja_parse_filtered_linux $NINJA_FOLDER/ninja $NINJA_FOLDER/alignments.txt $DB_STUB.db $DB_STUB.taxonomy --legacy LOG
+
+R -e "source('code/ninja_to_list.R');ninja_to_list('$NINJA_FOLDER')"
+
+rm -rf $NINJA_FOLDER

### 2016-02-26 code/ninja_to_list.R
Remove typos to get NINJA-OPS rules running

@@ -2,11 +2,10 @@
 # first column is the sequence name and the second column is the references
 # sequence that it mapped to. We need to convert this to a list file...
 
-library(dplyr)
+library(dplyr, warn.conflicts=F, quietly=T)
 
 ninja_to_list <- function(folder_name){
 
-	folder_name <- "even_1.0_01.ng.ninja"
 	input_file_name <- paste0(folder_name, '/ninja_pass.log')
 	output_file_name <- paste0(folder_name, '.list')
 	output_file_name <- gsub("ng.", "", output_file_name)
@@ -16,7 +15,7 @@ ninja_to_list <- function(folder_name){
 
 	list_data <- mapping %>% group_by(factor(V2)) %>% summarise(otu=paste(V1, collapse=','))
 
-	list_output <- paste(c("userLabel", length(list_data), list_data), collapse='\t')
+	list_output <- paste(c("userLabel", length(list_data$otu), list_data$otu), collapse='\t')
 
 	write(list_output, output_file_name)
 

### 2016-02-27 code/run_ninja.sh
Remove -k 1 option to reflect new behavior of v.1.3.2

@@ -7,7 +7,7 @@ rm -rf $NINJA_FOLDER
 mkdir $NINJA_FOLDER
 code/NINJA-OPS/bin/ninja_filter_linux $FASTA $NINJA_FOLDER/ninja D 1 LOG
 
-bowtie2-align-s --no-head -x $DB_STUB -S $NINJA_FOLDER/alignments.txt --np 0 --mp 1,1 --rdg 0,1 --rfg 0,1 --score-min L,0,-0.03 --norc -f $NINJA_FOLDER/ninja_filt.fa -p 4 -k 1 --very-sensitive
+bowtie2-align-s --no-head -x $DB_STUB -S $NINJA_FOLDER/alignments.txt --np 0 --mp 1,1 --rdg 0,1 --rfg 0,1 --score-min L,0,-0.03 --norc -f $NINJA_FOLDER/ninja_filt.fa -p 4 --very-sensitive
 
 code/NINJA-OPS/bin/ninja_parse_filtered_linux $NINJA_FOLDER/ninja $NINJA_FOLDER/alignments.txt $DB_STUB.db $DB_STUB.taxonomy --legacy LOG
 

### 2016-02-27 code/run_rand_nref.sh
Create script to run ninja-ops on rand_ref datasets

@@ -1,0 +1,35 @@
+REF=$1
+REF=data/rand_ref/rand_ref_1.0_01.fasta
+FASTA=data/rand_ref/miseq.unique.fasta
+
+
+DB_FOLDER=$(echo $REF | sed 's/fasta/ninja_db/')
+DB_NAME=$(echo $DB_FOLDER | sed 's/data\/rand_ref\///')
+
+rm -rf $DB_FOLDER
+mkdir -p $DB_FOLDER
+
+cp data/references/97_otus.taxonomy $DB_FOLDER/$DB_NAME.taxonomy
+
+code/NINJA-OPS/bin/ninja_prep_linux $REF $DB_FOLDER/out_preDB.fa $DB_FOLDER/$DB_NAME.db
+bowtie2-build $DB_FOLDER/out_preDB.fa $DB_FOLDER/$DB_NAME
+
+
+
+# The clustering...
+
+NINJA_FOLDER=$(echo $REF | sed 's/fasta/ninja/')
+
+rm -rf $NINJA_FOLDER
+mkdir $NINJA_FOLDER
+
+code/NINJA-OPS/bin/ninja_filter_linux $FASTA $NINJA_FOLDER/ninja D 1 LOG
+
+bowtie2-align-s --no-head -x $DB_FOLDER/$DB_NAME -S $NINJA_FOLDER/alignments.txt --np 0 --mp 1,1 --rdg 0,1 --rfg 0,1 --score-min L,0,-0.03 --norc -f $NINJA_FOLDER/ninja_filt.fa -p 4 --very-sensitive
+
+code/NINJA-OPS/bin/ninja_parse_filtered_linux $NINJA_FOLDER/ninja $NINJA_FOLDER/alignments.txt $DB_FOLDER/$DB_NAME.db $DB_FOLDER/$DB_NAME.taxonomy --legacy
+
+mv $NINJA_FOLDER/ninja_pass.log $NINJA_FOLDER.nc
+
+# Cleaning up
+rm -rf $DB_FOLDER $NINJA_FOLDER

### 2016-02-27 code/run_rand_nref.sh
Remove hardcoded REF and use LOG option

@@ -1,5 +1,4 @@
 REF=$1
-REF=data/rand_ref/rand_ref_1.0_01.fasta
 FASTA=data/rand_ref/miseq.unique.fasta
 
 
@@ -27,7 +26,7 @@ code/NINJA-OPS/bin/ninja_filter_linux $FASTA $NINJA_FOLDER/ninja D 1 LOG
 
 bowtie2-align-s --no-head -x $DB_FOLDER/$DB_NAME -S $NINJA_FOLDER/alignments.txt --np 0 --mp 1,1 --rdg 0,1 --rfg 0,1 --score-min L,0,-0.03 --norc -f $NINJA_FOLDER/ninja_filt.fa -p 4 --very-sensitive
 
-code/NINJA-OPS/bin/ninja_parse_filtered_linux $NINJA_FOLDER/ninja $NINJA_FOLDER/alignments.txt $DB_FOLDER/$DB_NAME.db $DB_FOLDER/$DB_NAME.taxonomy --legacy
+code/NINJA-OPS/bin/ninja_parse_filtered_linux $NINJA_FOLDER/ninja $NINJA_FOLDER/alignments.txt $DB_FOLDER/$DB_NAME.db $DB_FOLDER/$DB_NAME.taxonomy --legacy LOG
 
 mv $NINJA_FOLDER/ninja_pass.log $NINJA_FOLDER.nc
 

### 2016-02-27 code/summarize_rand_ref.R
Add nclosed processing to summarize_rand_ref

@@ -23,6 +23,9 @@ get_hits <- function(cluster_file, method){
 		cluster_data <- scan(file=cluster_file, what=character(), sep='\n', quiet=T)
 		listing <- lapply(cluster_data, parse_sc_line)
 		hits <- unlist(listing)
+	} else if(method == 'n'){
+		cluster_data <- read.table(file=cluster_file, stringsAsFactors=FALSE)
+		hits <- cluster_data$V2		
 	} else {
 		cluster_data <- read.table(file=cluster_file, stringsAsFactors=FALSE)
 		hits <- cluster_data[cluster_data$V1 == "H", 10]

### 2016-03-03 code/extract_opt_swarm.R
Extract data from cutoff for optimal swarm MCC value

@@ -1,0 +1,18 @@
+library(plyr)
+
+extract_opt_swarm <- function(opt_file_name="data/he/he.swarm.opt.sensspec"){
+	pool_file <- gsub("opt\\.", "pool_", opt_file_name)
+
+	opt_file <- read.table(file=opt_file_name, header=T)
+	opt_data <- ddply(.data=opt_file,
+						.variables="label",
+						.fun=function(x)x[which.max(x$mcc),])
+
+	fraction <- gsub(".*_(\\d\\.\\d)_.*", "\\1", opt_data$label)
+	replicate <- gsub(".*_(\\d\\d)\\..*", "\\1", opt_data$label)
+	label <- "userLabel"
+
+	pool <- data.frame(fraction, replicate, label, opt_data[,-1])
+
+	write.table(file=pool_file, pool, quote=F, row.names=F, sep='\t')
+}

### 2016-03-03 code/summarize_mcc.R
Get summarize_mcc to play nicely with swarm data

@@ -8,14 +8,7 @@ summarize_mcc <- function(dataset){
 	write(x="mean\tlci\tuci\tfraction\tmethod", file=output_file_name)
 
 	for(m in methods){
-		file_name <- ""
-
-		if(m != 'swarm')
-			file_name <- paste0("data/", dataset, "/", dataset, ".", m, ".pool_sensspec")
-		} else {
-			file_name <- paste0("data/", dataset, "/", dataset, ".swarm.opt.sensspec")
-		}
-
+		file_name <- paste0("data/", dataset, "/", dataset, ".", m, ".pool_sensspec")
 		mcc_data <-read.table(file=file_name, header=T)
 		mcc_stats_table <- aggregate(mcc_data$mcc, by=list(mcc_data$fraction), function(x)c(mean=mean(x), quantile(x, probs=c(0.025, 0.975))))
 

### 2016-03-03 code/build_all_methods_compare_plot.R
Add code and rules to build figure

@@ -1,0 +1,91 @@
+library("ggplot2")
+library("plyr")
+
+pool_summaries <- function(df, dataset){
+	summary_file_name <- paste0("data/process/", dataset, ".mcc.summary")
+
+	data <- read.table(file=summary_file_name, header=T)
+	data <- data[data$fraction == 1.0,]
+	data$dataset <- dataset
+
+	if(is.null(df)){
+		df <- data
+	} else {
+		df <- rbind(df, data)
+	}
+	return(df)
+}
+
+is_denovo <- function(method){
+	method %in% c("an", "nn", "fn", "agc", "dgc", "vagc", "vdgc", "swarm",
+								"sumaclust", "otuclust")
+}
+
+pooled_data <- NULL
+pooled_data <- pool_summaries(pooled_data, "he")
+#pooled_data <- pool_summaries(NULL, "miseq")
+pooled_data <- pool_summaries(pooled_data, "even")
+pooled_data <- pool_summaries(pooled_data, "staggered")
+
+ordering <- daply(
+	.data = pooled_data[is_denovo(pooled_data$method),],
+	.variables = "method",
+	.fun = function(x) mean(x$mean)
+)
+ordering <- ordering[!is.na(ordering)]
+de_novo_labels <- names(ordering)[order(ordering, decreasing=TRUE)]
+
+ordering <- daply(
+	.data = pooled_data[!is_denovo(pooled_data$method),],
+	.variables = "method",
+	.fun = function(x) mean(x$mean)
+)
+ordering <- ordering[!is.na(ordering)]
+reference_labels <- names(ordering)[order(ordering, decreasing=TRUE)]
+
+pooled_data$method <- factor(pooled_data$method, levels = c(de_novo_labels,reference_labels))
+
+pretty_methods <- c(an = "AN", sumaclust = "Sumaclust", vdgc="V-DGC", 
+										vagc="V-AGC", dgc = "U-DGC",
+										agc = "U-AGC", fn  = "FN", otuclust = "OTUCLUST",
+										swarm = "Swarm",     nn = "NN", cvsearch="V-Closed",
+										sortmerna = "SortMeRNA", open = "Open",
+										closed = "U-Closed", ninja = "NINJA-OPS")
+
+ggplot(pooled_data, aes(method, mean, col=dataset, shape=dataset)) +
+	
+	geom_rect(aes(xmin = -Inf, ymin = 0.9, xmax= Inf, ymax = Inf), col="lightgray",
+						fill="lightgray") +
[…]

### 2016-03-03 code/build_all_methods_compare_plot.R
Add preliminary mouse data to plot

@@ -23,7 +23,7 @@ is_denovo <- function(method){
 
 pooled_data <- NULL
 pooled_data <- pool_summaries(pooled_data, "he")
-#pooled_data <- pool_summaries(NULL, "miseq")
+pooled_data <- pool_summaries(pooled_data, "miseq")
 pooled_data <- pool_summaries(pooled_data, "even")
 pooled_data <- pool_summaries(pooled_data, "staggered")
 
@@ -45,7 +45,7 @@ reference_labels <- names(ordering)[order(ordering, decreasing=TRUE)]
 
 pooled_data$method <- factor(pooled_data$method, levels = c(de_novo_labels,reference_labels))
 
-pretty_methods <- c(an = "AN", sumaclust = "Sumaclust", vdgc="V-DGC", 
+pretty_methods <- c(an = "AN", sumaclust = "Sumaclust", vdgc="V-DGC",
 										vagc="V-AGC", dgc = "U-DGC",
 										agc = "U-AGC", fn  = "FN", otuclust = "OTUCLUST",
 										swarm = "Swarm",     nn = "NN", cvsearch="V-Closed",
@@ -53,39 +53,42 @@ pretty_methods <- c(an = "AN", sumaclust = "Sumaclust", vdgc="V-DGC",
 										closed = "U-Closed", ninja = "NINJA-OPS")
 
 ggplot(pooled_data, aes(method, mean, col=dataset, shape=dataset)) +
-	
-	geom_rect(aes(xmin = -Inf, ymin = 0.9, xmax= Inf, ymax = Inf), col="lightgray",
-						fill="lightgray") +
-	geom_text(aes(x = 5.5, y = 0.98, label = "De novo"), size = 4, color="black", fontface="plain") +
-	geom_text(aes(x = 13.00, y = 0.98, label = "Reference-based"), size = 4, color="black", fontface="plain") +
+
+	geom_rect(aes(xmin = -Inf, ymin = 0.9, xmax= Inf, ymax = Inf),
+					col="lightgray", fill="lightgray") +
+	geom_text(aes(x = 5.5, y = 0.98, label = "De novo"), size = 4,
+					color="black", fontface="plain") +
+	geom_text(aes(x = 13.00, y = 0.98, label = "Reference-based"), size = 4,
+	 				color="black", fontface="plain") +
 	geom_segment(aes( x=10.5, xend=10.5, y=-Inf, yend =Inf), color="black") +
-	
+
[…]

### 2016-03-03 code/summaraize_rand_ref.R
Summarize rand_ref data and output to data/process/

@@ -1,0 +1,43 @@
+method <- c(u="usearch", v="vsearch", n="ninja", s="sortmerna")
+
+summarize_closed_ref_sensspec <- function(df, file_name){
+
+	data <- read.table(file=file_name, header=T)
+	method <- gsub("data/rand_ref/closed_ref.(.*).sensspec", "\\1", file_name)
+
+	sensspec_summary <- data.frame(method=method, t(apply(data, 2, mean)))
+
+	if(is.null(df)){
+		df <- sensspec_summary
+	} else {
+		df <- rbind(df, sensspec_summary)
+	}
+
+	return(df)
+}
+
+summarize_hits <- function(df, file_name){
+
+	data <- read.table(file=file_name, header=F, row.names=1)
+	m <- gsub(".*hits.(.*)closed.summary", "\\1", file_name)
+
+	df[df$method==method[m], c("min_hits")] <- data["min", 1]
+	df[df$method==method[m], c("mean_hits")] <- data["mean", 1]
+	df[df$method==method[m], c("max_hits")] <- data["max", 1]
+
+	df
+}
+
+sensspec <- NULL
+sensspec <- summarize_closed_ref_sensspec(sensspec, "data/rand_ref/closed_ref.vsearch.sensspec")
+sensspec <- summarize_closed_ref_sensspec(sensspec, "data/rand_ref/closed_ref.usearch.sensspec")
+sensspec <- summarize_closed_ref_sensspec(sensspec, "data/rand_ref/closed_ref.sortmerna.sensspec")
+sensspec <- summarize_closed_ref_sensspec(sensspec, "data/rand_ref/closed_ref.ninja.sensspec")
+
+sensspec <- summarize_hits(sensspec, "data/rand_ref/hits.vclosed.summary")
+sensspec <- summarize_hits(sensspec, "data/rand_ref/hits.uclosed.summary")
+sensspec <- summarize_hits(sensspec, "data/rand_ref/hits.sclosed.summary")
+sensspec <- summarize_hits(sensspec, "data/rand_ref/hits.nclosed.summary")
+
+write.table(sensspec, file="data/process/closed_ref_sensspec.summary", quote=F,
+		row.names=F, sep='\t')

### 2016-03-03 papers/msystems_2016/Schloss_Commentary_mSystems_2016.Rmd

@@
 
-* Add new Swarm reference
-
 The ability to assign 16S rRNA gene sequences to operational taxonomic units (OTUs) allows microbial ecologists to overcome the inconsistencies and biases within bacterial taxonomy and provides a strategy for clustering similar sequences that do not have representatives in a reference database.
 These sequences are clustered into OTUs based on their distance from (or similarity to) each other.
-Numerous algorithms for solving this seemingly simple problem have blossomed in recent years and were recently the subject of benchmarking studies performed by Westcott and myself [ref x 2], He et al.
-[ref], and Kopylova et al [ref].
+Numerous algorithms for solving this seemingly simple problem have blossomed in recent years and were recently the subject of benchmarking studies performed by Westcott and myself [-@Schloss2011Assessing; -@Westcott2015], He et al.
+[-@He2015], and Kopylova et al [-@Kopylova2016].
 These studies provide a thorough review of the sequencing clustering landscape, which can be divided into three general approaches: (i) *de novo* clustering where sequences are clustered without first mapping sequences to a reference database, (ii) closed-reference clustering where sequences are clustered based on the references that the sequences map to, and (iii) open reference clustering where sequences that do not map adequately to the reference are then clustered using a *de novo* approach.
@@
 
-The most recent analysis of Kopylova et al [@ref] repeats many of the benchmarking strategies employed by previous researchers.
+The most recent analysis of Kopylova et al [-@Kopylova2016] repeats many of the benchmarking strategies employed by previous researchers.
 First, they and others have compared the time and memory required to cluster sequences in a dataset [@Sun2009; @Cai2011; @Rideout2014; @Mah2014].
 These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the OTU assignments.
-In the extreme, we could easily develop a toy a clustering algorithm could randomly assign sequences to a predetermined number of OTUs and be efficient, but also poorly reflect the genetic diversity within the community.
-Second, they and others have compared the number of OTUs generated by various methods for a common dataset [@refs].
-Again, the toy clustering algorithm could cluster to a target number of OTUs, but the clusterings would likely be meaningless.
+In the extreme, we could easily develop a "toy" a clustering algorithm could randomly assign sequences to a predetermined number of OTUs and be efficient, but also poorly reflect the genetic diversity within the community.
+Second, they and others have compared the number of OTUs generated by various methods for a common dataset [@Edgar2013; @Kopylova2016].
+Again, a toy clustering algorithm could cluster to a target number of OTUs, but the clusterings would likely be meaningless.
 Third, because many of the methods are sensitive to the initial order of the sequences, a metric of OTU stability has been proposed as a way to assess algorithms [@He2015].
@@
 Fourth, the method that Kopylova et al.
-[@ref] relied upon the most was to cluster sequences from simulated data or data from synthetic communities of cultured organisms and quantify how well the OTU assignments matched the organisms' taxonomy [@Edgar2013; @Mah2014; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014; @Cai2011; @Sun2011; @White2010].
-This approach is flawed because bacterial taxonomy often reflects those inconsistencies and biases within bacterial taxonomy that OTU-based methods strive to overcome.
+[-@Kopylova2016] relied upon the most was to cluster sequences from simulated data or data from synthetic communities of cultured organisms and quantify how well the OTU assignments matched the organisms' taxonomy [@Edgar2013; @Mah2014; @Mah2015; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014; @Cai2011; @Sun2011; @White2010; @AlGhalith2016].
+This commonly used approach is flawed because bacterial taxonomy often reflects those inconsistencies and biases within bacterial taxonomy that OTU-based methods strive to overcome.
 Furthermore, this benchmarking strategy can only be applied when the actual taxonomy of the organisms are known and so it is unclear how the methods scale to sequences from the novel organisms we are likely to encounter in deep sequencing surveys.
@@
 To revisit these results, I have expanded the analysis to incorporate additional *de novo* and closed-reference algorithms and two simulated datasets following the approach described by Kopylova et al [@Kopylova2016].
-I evaluated three heirarchical and seven greedy *de novo* algorithms, one open-reference clustering algorithm, and two closed-reference algorithms (Figure 1).
+I evaluated three heirarchical and seven greedy *de novo* algorithms, one open-reference clustering algorithm, and four closed-reference algorithms (Figure 1).
 To test these approaches I applied each of them to four datasets from soil [@Roesch2007], mouse feces [@Schloss2012Stabilization], and two simulated communities.
-The simulated communities were generated by randomly selecting 10,000 16S rRNA sequences that were unique within the V4 region from the SILVA non-redundant database [@ref].
+The simulated communities were generated by randomly selecting 10,000 16S rRNA sequences that were unique within the V4 region from the SILVA non-redundant database [@Kopylova2016; @Pruesse2007].
 Next, an even community was generated by specifying that each sequence had a frequency of 100 reads and a staggered community was generated by specifying that the abundance of each sequence was a randomly drawn a uniform distribution between 1 and 200.
@@
 Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Lorem ipsum dolor sit amet, consectetur.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
 Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
@@
 Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
 Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

### 2016-03-04 code/build_all_methods_compare_plot.R
Write out method names

@@ -45,12 +45,13 @@ reference_labels <- names(ordering)[order(ordering, decreasing=TRUE)]
 
 pooled_data$method <- factor(pooled_data$method, levels = c(de_novo_labels,reference_labels))
 
-pretty_methods <- c(an = "AN", sumaclust = "Sumaclust", vdgc="V-DGC",
-										vagc="V-AGC", dgc = "U-DGC",
-										agc = "U-AGC", fn  = "FN", otuclust = "OTUCLUST",
-										swarm = "Swarm",     nn = "NN", cvsearch="V-Closed",
-										sortmerna = "SortMeRNA", open = "Open",
-										closed = "U-Closed", ninja = "NINJA-OPS")
+pretty_methods <- c(an = "Average\nNeighbor", sumaclust = "Sumaclust",
+					vdgc="DGC with\nVSEARCH", vagc="AGC with\nVSEARCH",
+					dgc = "DGC with\nUSEARCH", agc = "AGC with\nUSEARCH",
+					fn  = "Furthest\nNeighbor", otuclust = "OTUCLUST",
+					swarm = "Swarm", nn = "Nearest\nNeighbor", cvsearch="VSEARCH",
+					sortmerna = "SortMeRNA", open = "Open",
+					closed = "USEARCH", ninja = "NINJA-OPS")
 
 ggplot(pooled_data, aes(method, mean, col=dataset, shape=dataset)) +
 

### 2016-03-07 papers/msystems_2016/Schloss_Commentary_mSystems_2016.Rmd
First full draft of manuscript

@@
 ---
-title: "** INSERT TITLE **"
+title: "Application of database-independent methods to assess the quality of OTU picking methods"
 bibliography: references.bib
@@
     includes:
-      in_header: ../header.tex
+      in_header: header.tex
 csl: msystems.csl
@@
 
-To revisit these results, I have expanded the analysis to incorporate additional *de novo* and closed-reference algorithms and two simulated datasets following the approach described by Kopylova et al [@Kopylova2016].
-I evaluated three heirarchical and seven greedy *de novo* algorithms, one open-reference clustering algorithm, and four closed-reference algorithms (Figure 1).
-To test these approaches I applied each of them to four datasets from soil [@Roesch2007], mouse feces [@Schloss2012Stabilization], and two simulated communities.
+To revisit these results, I have expanded the analysis to evaluate three hierarchical and seven greedy *de novo* algorithms, one open-reference clustering algorithm, and four closed-reference algorithms (Figure 1).
+To test these approaches I applied each of them to datasets from soil [@Roesch2007], mouse feces [@Schloss2012Stabilization], and two simulated datasets.
 The simulated communities were generated by randomly selecting 10,000 16S rRNA sequences that were unique within the V4 region from the SILVA non-redundant database [@Kopylova2016; @Pruesse2007].
 Next, an even community was generated by specifying that each sequence had a frequency of 100 reads and a staggered community was generated by specifying that the abundance of each sequence was a randomly drawn a uniform distribution between 1 and 200.
-The benchmarking approach is the same as the approach used previously [@Westcott2015].
 A reproducible version of this manuscript and analysis has been added to the repository available at https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015.
 
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-Lorem ipsum dolor sit amet, consectetur.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+```{r de_novo}
+get_de_novo <- function(dataset){
+	de_novo <- read.table(file=paste0("data/process/", dataset, ".mcc.summary"), header=T)
+	de_novo <- de_novo[de_novo$fraction == 1.0,]
 
-adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+	output <- de_novo$mean
+	names(output) <- de_novo$method
+
+	return(output)
+}
+
+mouse <- get_de_novo("miseq")
+soil <- get_de_novo("he")
+even <- get_de_novo("even")
+staggered <- get_de_novo("staggered")
+```
+
+I replicated the benchmarking approach that I have used previously to assess the ability of an algorithm to correctly group sequences that are similar to each other and split sequences that are dissimilar to each other using the MCC [@Westcott2015; @Schloss2011Assessing].
+When I compared the MCC values calculated using the ten *de novo* algorithms with the four datasets, the average neighbor algorithm reliably performed as well or better than the other methods (Figure 1).
+The MCC values for the VSEARCH (AGC: `r mouse["vagc"]` and DGC: `r mouse["vdgc"]`) and USEARCH-based (AGC: `r mouse["agc"]` and DGC: `r mouse["dgc"]`) algorithms, Sumaclust (`r mouse["sumaclust"]`), and average neighbor (`r mouse["an"]`) were similarly high for the murine dataset.
+For each of the other datasets, the MCC value for the average neighbor algorithm was at least 5% higher than the next best method.
+Swarm does not use a traditional distance-based criteria to cluster sequences into OTUs and instead looks for natural subnetworks in the data.
+When I used the distance threshold that gave the best MCC value for the Swarm data, the MCC values were generally not as high as they were using the average neighbor algorithm.
+The one exception was for the soil dataset.
+Among the reference-based methods, all of the MCC values suffer because when sequences that are at least 97% similar to a reference are pooled, the sequences within an OTU could be as much as 6% different from each other.
+The effect of this is observed in the MCC values that were calculated for the OTUs assigned by these methods generally being lower than those observed using the *de novo* approaches (Figure 1).
+It is also important to note that the MCC values are somewhat inflated because sequences were not clustered into OTUs if there was not a reference sequence that was more than 97% similar to the sequence.
+Given the consistent quality of the clusterings formed by the average neighbor algorithm, these results confirm the conclusion from the previous analysis that researchers should use the average neighbor algorithm or calculate MCC values for several methods and use the clustering that gives the best MCC value.
+
+```{r closed_reference}
+closed <- read.table(file="data/process/closed_ref_sensspec.summary", header=T, row.names=1)
+```
+
+Next, I investigated the ability of the reference-based methods to properly assign sequences to OTUs.
+The full-length 16S rRNA gene sequences in the default reference taxonomy that accompanies QIIME are not more than 97% similar to each other, but within the V4 region many of the sequences were more similar to each other and even identical to each other.
+As a result, we previously found that For USEARCH and VSEARCH there was a dependence between the ordering of sequences in the reference database and the OTU assignments.
+To explore this further, we analyzed the 32,106 unique sequences from the murine dataset with randomized databases.
+VSEARCH always found matches for `r closed["vsearch","mean_hits"]` murine sequences, the reference matched to those sequences differed between randomizations.
+For USEARCH there were between `r paste(closed['usearch', c('min_hits', 'max_hits')], collapse=" and ")` matches depending on the order of the reference.
+In the updated analysis we found that SortMeRNA resulted in between `r paste(closed['sortmerna', c('min_hits', 'max_hits')], collapse=" and ")` matches.
+Using NINJA-OPS with different orderings of the reference sequences generated the same `r closed['ninja', 'min_hits']` matches.
+These results point to an additional problem with closed-reference clustering, which is the inability for the method to assign sequences to OTUs when a similar reference sequence does not exist in the database.
+For the well-characterized murine microbiota, NINJA-OPS did the best by finding relatives for `r 100*closed['ninja', 'min_hits']/32106`% of the sequences.
+As indicated by the variation in the number of sequences that matched a reference sequence, these methods varied in their sensitivity and specificity to find the best reference sequence.
+Of the closed-reference methods, NINJA-OPS had the best sensitivity (`r 100*closed["ninja", "sensitivity"]`%) and specificity (`r 100*closed["ninja", "specificity"]`%) while SortMeRNA had the worst sensitivity (`r 100*closed["sortmerna", "sensitivity"]`%) and VSEARCH had the worst specificity (`r 100*closed["vsearch", "specificity"]`%).
+Reference-based clustering algorithms are much faster than *de novo* approaches, but do not generate OTUs that are as robust.
 
@@
 The USEARCH and SortMeRNA (v.2.0) closed-reference clusterings were performed using QIIME (v.1.9.1).
+Closed-reference clustering was also performed using VSEARCH (v.1.5.0) and NINJA-OPS (v.1.3.2).
 The order of the sequences in each dataset was randomized thirty times and the intra-method range in MCC values was smaller than the plotting symbol.
-MCC values were calculated in mothur.
-
-Nearest neighbor
-Average neighbor
-Furthest neighbor
-USEARCH AGC
-VSEARCH AGC
-USEARCH DGC
-VSEARCH DGC
-Swarm
-OTUClust
-SumaClust
-Open-reference
-USEARCH closed-reference
-VSEARCH closed-reference
-SortMeRNA closed-reference
-
-\newpage
-
-\begin{comment}
-	\includepdf{../../results/figures/commentary_figure_1.pdf}
-\end{comment}
+MCC values were calculated using mothur.
 

### 2016-03-07 papers/msystems_2016/Schloss_Commentary_mSystems_2016.Rmd
Editing

@@
 ## Abstract
-150 words
+Assigning 16S rRNA gene sequences to operational taxonomic units (OTUs) allows microbial ecologists to overcome the inconsistencies and biases within bacterial taxonomy and provides a strategy for clustering similar sequences that do not have representatives in a reference database.
+I have applied the Matthew's correlation coefficient to assess the ability of 15 reference-independent and -dependent clustering algorithms to assign sequences to OTUs.
+This metric quantifies the ability of an algorithm to reflect the relationships between sequences without the use of a reference and can be applied to any dataset or method.
+Overall, the most consistent method was the average neighbor algorithm, but that for some datasets other algorithms can match its performance.
 
@@
 
-The ability to assign 16S rRNA gene sequences to operational taxonomic units (OTUs) allows microbial ecologists to overcome the inconsistencies and biases within bacterial taxonomy and provides a strategy for clustering similar sequences that do not have representatives in a reference database.
-These sequences are clustered into OTUs based on their distance from (or similarity to) each other.
-Numerous algorithms for solving this seemingly simple problem have blossomed in recent years and were recently the subject of benchmarking studies performed by Westcott and myself [-@Schloss2011Assessing; -@Westcott2015], He et al.
-[-@He2015], and Kopylova et al [-@Kopylova2016].
+Numerous algorithms have been developed for solving the seemingly simple problem of assigning 16S rRNA gene sequences to operational taxonomic units (OTUs).
+These algorithms were recently the subject of benchmarking studies performed by Westcott and myself [-@Schloss2011Assessing; -@Westcott2015], He et al [-@He2015], and Kopylova et al [-@Kopylova2016].
 These studies provide a thorough review of the sequencing clustering landscape, which can be divided into three general approaches: (i) *de novo* clustering where sequences are clustered without first mapping sequences to a reference database, (ii) closed-reference clustering where sequences are clustered based on the references that the sequences map to, and (iii) open reference clustering where sequences that do not map adequately to the reference are then clustered using a *de novo* approach.
-My studies have highlighted a persistent problem in the development of clustering algorithms, which is assessing the quality of the clustering assignments.
+Assessing the quality of the clustering assignments has been a persistent problem in the development of clustering algorithms.
 
-The most recent analysis of Kopylova et al [-@Kopylova2016] repeats many of the benchmarking strategies employed by previous researchers.
-First, they and others have compared the time and memory required to cluster sequences in a dataset [@Sun2009; @Cai2011; @Rideout2014; @Mah2014].
-These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the OTU assignments.
-In the extreme, we could easily develop a "toy" a clustering algorithm could randomly assign sequences to a predetermined number of OTUs and be efficient, but also poorly reflect the genetic diversity within the community.
-Second, they and others have compared the number of OTUs generated by various methods for a common dataset [@Edgar2013; @Kopylova2016].
-Again, a toy clustering algorithm could cluster to a target number of OTUs, but the clusterings would likely be meaningless.
-Third, because many of the methods are sensitive to the initial order of the sequences, a metric of OTU stability has been proposed as a way to assess algorithms [@He2015].
-Although it is important that the methods generate reproducible OTU assignments, this approach ignores the possability that the variation in assignments may be equally robust or that the assignemnts by a highly reproducible algorithm may be quite poor.
-Fourth, the method that Kopylova et al.
-[-@Kopylova2016] relied upon the most was to cluster sequences from simulated data or data from synthetic communities of cultured organisms and quantify how well the OTU assignments matched the organisms' taxonomy [@Edgar2013; @Mah2014; @Mah2015; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014; @Cai2011; @Sun2011; @White2010; @AlGhalith2016].
-This commonly used approach is flawed because bacterial taxonomy often reflects those inconsistencies and biases within bacterial taxonomy that OTU-based methods strive to overcome.
-Furthermore, this benchmarking strategy can only be applied when the actual taxonomy of the organisms are known and so it is unclear how the methods scale to sequences from the novel organisms we are likely to encounter in deep sequencing surveys.
-In a final approach, Kopylova and others have assessed the quality of clustering based on their ability to generate the same OTUs generated by other methods [@Schmidt2014Limits; @Rideout2014].
+The recent analysis of Kopylova et al [-@Kopylova2016] repeats many of the benchmarking strategies employed by previous researchers.
+Many algorithm developers have clustered sequences from simulated data or data from synthetic communities of cultured organisms and quantify how well the OTU assignments matched the organisms' taxonomy [@Edgar2013; @Mah2014; @Mah2015; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014; @Cai2011; @Sun2011; @White2010; @AlGhalith2016].
+This approach is flawed because bacterial taxonomy often reflects those inconsistencies and biases within bacterial taxonomy that OTU-based methods strive to overcome.
+Furthermore, it is unclear how the methods scale to sequences from the novel organisms we are likely to encounter in deep sequencing surveys.
+In a second approach, developers have compared the time and memory required to cluster sequences in a dataset [@Sun2009; @Cai2011; @Rideout2014; @Mah2014].
+These are valid parameters to assess when judging a clustering method, but indicate little regarding the quality of the OTU assignments.
+One could randomly assign sequences to a predetermined number of OTUs.
+This would be efficient.
+It would also poorly reflect the genetic diversity within the community.
+In a third approach, developers have compared the number of OTUs generated by various methods for a common dataset [@Edgar2013; @Kopylova2016].
+Again, one could randomly cluster sequences into a target number of OTUs, but the clusterings would likely be meaningless.
+In a fourth approach a metric of OTU stability has been proposed as a way to assess algorithms [@He2015].
+Although it is important that the methods generate reproducible OTU assignments when the initial order of the sequences is randomized.
+This metric ignores the possibility that the variation in assignments may be equally robust or that the assignments by a highly reproducible algorithm may be quite poor.
+In a final approach, some developers have assessed the quality of clustering based on the method's ability to generate the same OTUs generated by other methods [@Schmidt2014Limits; @Rideout2014].
 Unfortunately, without the ability to ground truth any method, such comparisons are tenuous.
+There is a need for an objective metric to assess the quality of OTU assignments.
+
 Westcott and I have proposed an unbiased and objective method for assessing the quality of OTU assignments that can be applied to any collection of sequences [@Schloss2011Assessing; @Westcott2015].
-
 Our approach uses the observed dissimilarity between pairs of sequences and information about whether sequences were clustered together to quantify how well similar sequences are clustered together and dissimilar sequences are clustered apart.
 To quantify the correlation between the observed and expected OTU assignments, we synthesize the relationship between OTU assignments and the distances between sequences using the Matthew's correlation coefficient [MCC; @Matthews1975].
-In the most recent application of this approach [@Westcott2015], we found that closed-reference clustering algorithms could be sensitive to the order of the sequences in the reference database and frequently clustered sequences together that were more than 3% different from each other.
-We also discussed that because open-reference clustering was dependent on closed-reference clustering it had sensitivity to the order of the database.
-Furthermore, we described how it had a nebulous threshold for OTUs since sequences are clustered based on a radius of 3% under the closed-reference phase and a diameter of 3% under the open reference phase.
-Finally, we showed that *de novo* clustering algorithms generated the most robust OTU assignments and confirmed our previous analysis that the average neighbor algorithm consistently performed the best [@Schloss2011Assessing].
-Given the observation that the best algorithm may vary by dataset we concluded that researchers should quantify the MCC for several *do novo* algorithms before selecting an algorithm.
-
-To revisit these results, I have expanded the analysis to evaluate three hierarchical and seven greedy *de novo* algorithms, one open-reference clustering algorithm, and four closed-reference algorithms (Figure 1).
+I have expanded the analysis to evaluate three hierarchical and seven greedy *de novo* algorithms, one open-reference clustering algorithm, and four closed-reference algorithms (Figure 1).
 To test these approaches I applied each of them to datasets from soil [@Roesch2007], mouse feces [@Schloss2012Stabilization], and two simulated datasets.
@@
 The effect of this is observed in the MCC values that were calculated for the OTUs assigned by these methods generally being lower than those observed using the *de novo* approaches (Figure 1).
-It is also important to note that the MCC values are somewhat inflated because sequences were not clustered into OTUs if there was not a reference sequence that was more than 97% similar to the sequence.
+It is also important to note that the MCC values for the closed-reference OTUs are inflated because sequences were removed from the analysis if there was not a reference sequence that was more than 97% similar to the sequence.
 Given the consistent quality of the clusterings formed by the average neighbor algorithm, these results confirm the conclusion from the previous analysis that researchers should use the average neighbor algorithm or calculate MCC values for several methods and use the clustering that gives the best MCC value.
@@
 The full-length 16S rRNA gene sequences in the default reference taxonomy that accompanies QIIME are not more than 97% similar to each other, but within the V4 region many of the sequences were more similar to each other and even identical to each other.
-As a result, we previously found that For USEARCH and VSEARCH there was a dependence between the ordering of sequences in the reference database and the OTU assignments.
+As a result, we previously found that there was a dependence between the ordering of sequences in the reference database and the OTU assignments with USEARCH and VSEARCH.
 To explore this further, we analyzed the 32,106 unique sequences from the murine dataset with randomized databases.
 VSEARCH always found matches for `r closed["vsearch","mean_hits"]` murine sequences, the reference matched to those sequences differed between randomizations.
-For USEARCH there were between `r paste(closed['usearch', c('min_hits', 'max_hits')], collapse=" and ")` matches depending on the order of the reference.
-In the updated analysis we found that SortMeRNA resulted in between `r paste(closed['sortmerna', c('min_hits', 'max_hits')], collapse=" and ")` matches.
+For USEARCH there were between `r closed['usearch', 'min_hits']` and `r closed['usearch', 'max_hits']` matches depending on the order of the reference.
+In the updated analysis we found that SortMeRNA resulted in between `r closed['sortmerna', 'min_hits']` and `r closed['sortmerna', 'max_hits']` matches.
 Using NINJA-OPS with different orderings of the reference sequences generated the same `r closed['ninja', 'min_hits']` matches.
 These results point to an additional problem with closed-reference clustering, which is the inability for the method to assign sequences to OTUs when a similar reference sequence does not exist in the database.
-For the well-characterized murine microbiota, NINJA-OPS did the best by finding relatives for `r 100*closed['ninja', 'min_hits']/32106`% of the sequences.
+For the well-characterized murine microbiota, NINJA-OPS did the best by finding relatives for `r format(100*closed['ninja', 'min_hits']/32106, digits=3)`% of the sequences.
 As indicated by the variation in the number of sequences that matched a reference sequence, these methods varied in their sensitivity and specificity to find the best reference sequence.
-Of the closed-reference methods, NINJA-OPS had the best sensitivity (`r 100*closed["ninja", "sensitivity"]`%) and specificity (`r 100*closed["ninja", "specificity"]`%) while SortMeRNA had the worst sensitivity (`r 100*closed["sortmerna", "sensitivity"]`%) and VSEARCH had the worst specificity (`r 100*closed["vsearch", "specificity"]`%).
+Of the closed-reference methods, NINJA-OPS had the best sensitivity (`r format(100*closed["ninja", "sensitivity"], digits=3)`%) and specificity (`r format(100*closed["ninja", "specificity"], digits=3)`%) while SortMeRNA had the worst sensitivity (`r format(100*closed["sortmerna", "sensitivity"], digits=3)`%) and VSEARCH had the worst specificity (`r format(100*closed["vsearch", "specificity"], digits=3)`%).
 Reference-based clustering algorithms are much faster than *de novo* approaches, but do not generate OTUs that are as robust.
 
-Although the goal of Kopylova et al.
-[-@Kopylova2016] was to compare various clustering algorithms, they also studied these algorithms in the broader context of raw sequence processing, screening for chimeras, and removal of singletons.
-Each of these are critical decisions in a comprehensive pipeline; however, they confound the analysis of how best to cluster sequences into OTUs that reflect a specific distance threshold.
+Although the goal of Kopylova et al [-@Kopylova2016] was to compare various clustering algorithms, they also studied these algorithms in the broader context of raw sequence processing, screening for chimeras, and removal of singletons.
+Each of these are critical decisions in a comprehensive pipeline; however, by including these steps, they confounded the analysis of how best to cluster sequences into OTUs.
+The effect of differences in MCC values on one's ability to draw inferences is unclear and admittedly may be relatively minor for some datasets.
+Because of this uncertainty, researchers should use the most reliable methods available in case the differences in clustering does effect the conclusions they can draw from their particular dataset.
 Through the use of objective criteria that measure the quality of the clusterings, independent of taxonomy or database, researchers will be able to evaluate which clustering algorithm is the best fit for their data.

### 2016-03-08 papers/msystems_2016/Schloss_Commentary_mSystems_2016.Rmd
Final editing and addition of references to Fig. 1 caption

@@
 ---
-title: "Application of database-independent methods to assess the quality of OTU picking methods"
+title: "**Application of database-independent methods to assess the quality of OTU picking methods**"
 bibliography: references.bib
@@
 This metric quantifies the ability of an algorithm to reflect the relationships between sequences without the use of a reference and can be applied to any dataset or method.
-Overall, the most consistent method was the average neighbor algorithm, but that for some datasets other algorithms can match its performance.
+The most consistently robust method was the average neighbor algorithm; however, for some datasets other algorithms matched its performance.
 
@@
 
-The recent analysis of Kopylova et al [-@Kopylova2016] repeats many of the benchmarking strategies employed by previous researchers.
-Many algorithm developers have clustered sequences from simulated data or data from synthetic communities of cultured organisms and quantify how well the OTU assignments matched the organisms' taxonomy [@Edgar2013; @Mah2014; @Mah2015; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014; @Cai2011; @Sun2011; @White2010; @AlGhalith2016].
+The recent analysis of Kopylova et al [-@Kopylova2016] repeated many of the benchmarking strategies employed by previous researchers.
+Many algorithm developers have clustered sequences from simulated communities or sequencing data from synthetic communities of cultured organisms and quantify how well the OTU assignments matched the organisms' taxonomy [@Edgar2013; @Mah2014; @Mah2015; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014; @Cai2011; @Sun2011; @White2010; @AlGhalith2016].
 This approach is flawed because bacterial taxonomy often reflects those inconsistencies and biases within bacterial taxonomy that OTU-based methods strive to overcome.
@@
 In a fourth approach a metric of OTU stability has been proposed as a way to assess algorithms [@He2015].
-Although it is important that the methods generate reproducible OTU assignments when the initial order of the sequences is randomized.
-This metric ignores the possibility that the variation in assignments may be equally robust or that the assignments by a highly reproducible algorithm may be quite poor.
+Although it is important that the methods generate reproducible OTU assignments when the initial order of the sequences is randomized, this metric ignores the possibility that the variation in assignments may be equally robust or that the assignments by a highly reproducible algorithm may be quite poor.
 In a final approach, some developers have assessed the quality of clustering based on the method's ability to generate the same OTUs generated by other methods [@Schmidt2014Limits; @Rideout2014].
@@
 To quantify the correlation between the observed and expected OTU assignments, we synthesize the relationship between OTU assignments and the distances between sequences using the Matthew's correlation coefficient [MCC; @Matthews1975].
-I have expanded the analysis to evaluate three hierarchical and seven greedy *de novo* algorithms, one open-reference clustering algorithm, and four closed-reference algorithms (Figure 1).
+I have expanded our previous analysis to evaluate three hierarchical and seven greedy *de novo* algorithms, one open-reference clustering algorithm, and four closed-reference algorithms (Figure 1).
 To test these approaches I applied each of them to datasets from soil [@Roesch2007], mouse feces [@Schloss2012Stabilization], and two simulated datasets.
@@
 When I compared the MCC values calculated using the ten *de novo* algorithms with the four datasets, the average neighbor algorithm reliably performed as well or better than the other methods (Figure 1).
-The MCC values for the VSEARCH (AGC: `r mouse["vagc"]` and DGC: `r mouse["vdgc"]`) and USEARCH-based (AGC: `r mouse["agc"]` and DGC: `r mouse["dgc"]`) algorithms, Sumaclust (`r mouse["sumaclust"]`), and average neighbor (`r mouse["an"]`) were similarly high for the murine dataset.
+For the murine dataset, the MCC values for the VSEARCH (AGC: `r mouse["vagc"]` and DGC: `r mouse["vdgc"]`) and USEARCH-based (AGC: `r mouse["agc"]` and DGC: `r mouse["dgc"]`) algorithms, Sumaclust (`r mouse["sumaclust"]`), and average neighbor (`r mouse["an"]`) were similarly high.
 For each of the other datasets, the MCC value for the average neighbor algorithm was at least 5% higher than the next best method.
@@
 It is also important to note that the MCC values for the closed-reference OTUs are inflated because sequences were removed from the analysis if there was not a reference sequence that was more than 97% similar to the sequence.
-Given the consistent quality of the clusterings formed by the average neighbor algorithm, these results confirm the conclusion from the previous analysis that researchers should use the average neighbor algorithm or calculate MCC values for several methods and use the clustering that gives the best MCC value.
+Given the consistent quality of the clusterings formed by the average neighbor algorithm, these results confirm the conclusion from the previous analysis that researchers should use the average neighbor algorithm or calculate MCC values for several methods and use the clustering that gives the best MCC value [@Westcott2015].
 
@@
 Next, I investigated the ability of the reference-based methods to properly assign sequences to OTUs.
-The full-length 16S rRNA gene sequences in the default reference taxonomy that accompanies QIIME are not more than 97% similar to each other, but within the V4 region many of the sequences were more similar to each other and even identical to each other.
-As a result, we previously found that there was a dependence between the ordering of sequences in the reference database and the OTU assignments with USEARCH and VSEARCH.
+The full-length 16S rRNA gene sequences in the default reference taxonomy that accompanies QIIME are less than 97% similar to each other.
+Within the V4 region, however, many of the sequences were more similar to each other and even identical to each other.
+As a result, we previously found that there was a dependence between the ordering of sequences in the reference database and the OTU assignments with USEARCH and VSEARCH [@Westcott2015].
 To explore this further, we analyzed the 32,106 unique sequences from the murine dataset with randomized databases.
@@
 These results point to an additional problem with closed-reference clustering, which is the inability for the method to assign sequences to OTUs when a similar reference sequence does not exist in the database.
-For the well-characterized murine microbiota, NINJA-OPS did the best by finding relatives for `r format(100*closed['ninja', 'min_hits']/32106, digits=3)`% of the sequences.
+For the well-characterized murine microbiota, NINJA-OPS did the best by finding relatives for `r format(100*closed['ninja', 'min_hits']/32106, digits=3)`% of the unique murine sequences.
 As indicated by the variation in the number of sequences that matched a reference sequence, these methods varied in their sensitivity and specificity to find the best reference sequence.
@@
 Although the goal of Kopylova et al [-@Kopylova2016] was to compare various clustering algorithms, they also studied these algorithms in the broader context of raw sequence processing, screening for chimeras, and removal of singletons.
-Each of these are critical decisions in a comprehensive pipeline; however, by including these steps, they confounded the analysis of how best to cluster sequences into OTUs.
+Each of these are critical decisions in a comprehensive pipeline.
+By including these steps, they confounded their analysis of how best to cluster sequences into OTUs.
 The effect of differences in MCC values on one's ability to draw inferences is unclear and admittedly may be relatively minor for some datasets.
-Because of this uncertainty, researchers should use the most reliable methods available in case the differences in clustering does effect the conclusions they can draw from their particular dataset.
+Because of this uncertainty, researchers should use the most reliable methods available in case the differences in clustering do effect the conclusions that can be drawn from a particular dataset.
 Through the use of objective criteria that measure the quality of the clusterings, independent of taxonomy or database, researchers will be able to evaluate which clustering algorithm is the best fit for their data.
@@
 **Figure 1.
-Comparison of OTU quality generated by multiple algorithms applied to four datasets.** The nearest, average, and furthest neighbor clustering algorithms were used as implemented in mothur (v.1.37).
-Abundance (AGC) and Distance-based greedy clustering (DGC) were implemented using USEARCH (v.6.1) and VSEARCH (v.1.5.0).
-Other *de novo* clustering algorithms included Swarm (v.2.1.1), OTUCLUST (v.0.1), and SUMACLUST (v.1.0.20).
+Comparison of OTU quality generated by multiple algorithms applied to four datasets.** The nearest, average, and furthest neighbor clustering algorithms were used as implemented in mothur (v.1.37)[@Schloss2009].
+Abundance (AGC) and Distance-based greedy clustering (DGC) were implemented using USEARCH (v.6.1) and VSEARCH (v.1.5.0)[@He2015; @Edgar2013; @Rognes2015].
+Other *de novo* clustering algorithms included Swarm (v.2.1.1)[@Mah2014; @Mah2015], OTUCLUST (v.0.1)[@Albanese2015], and Sumaclust (v.1.0.20).
 The MCC values for Swarm were determined by selecting the distance threshold that generated the maximum MCC value for each dataset.
-The USEARCH and SortMeRNA (v.2.0) closed-reference clusterings were performed using QIIME (v.1.9.1).
-Closed-reference clustering was also performed using VSEARCH (v.1.5.0) and NINJA-OPS (v.1.3.2).
+The USEARCH and SortMeRNA (v.2.0) closed-reference clusterings were performed using QIIME (v.1.9.1) [@Kopylova2012; @Caporaso2010].
+Closed-reference clustering was also performed using VSEARCH (v.1.5.0) and NINJA-OPS (v.1.3.2) [@AlGhalith2016].
 The order of the sequences in each dataset was randomized thirty times and the intra-method range in MCC values was smaller than the plotting symbol.

### 2016-03-08 papers/msystems_2016/Schloss_Commentary_mSystems_2016.Rmd
Fix title

@@
 ---
-title: "**Application of database-independent methods to assess the quality of OTU picking methods**"
+title: "**Application of database-independent approach to assess the quality of OTU picking methods**"
 bibliography: references.bib

### 2016-03-25 papers/msystems_2016/Schloss_Commentary_mSystems_2016.Rmd
Responses to reviewer comments

@@
 Many algorithm developers have clustered sequences from simulated communities or sequencing data from synthetic communities of cultured organisms and quantify how well the OTU assignments matched the organisms' taxonomy [@Edgar2013; @Mah2014; @Mah2015; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014; @Cai2011; @Sun2011; @White2010; @AlGhalith2016].
-This approach is flawed because bacterial taxonomy often reflects those inconsistencies and biases within bacterial taxonomy that OTU-based methods strive to overcome.
+Although an OTU definition would ideally match bacterial taxonomy, bacterial taxonomy has proven itself to be fluid and reflect the biases of various research interests.
 Furthermore, it is unclear how the methods scale to sequences from the novel organisms we are likely to encounter in deep sequencing surveys.
@@
 These are valid parameters to assess when judging a clustering method, but indicate little regarding the quality of the OTU assignments.
-One could randomly assign sequences to a predetermined number of OTUs.
-This would be efficient.
-It would also poorly reflect the genetic diversity within the community.
+For example, reference-based methods are very efficient, but do a poor job of reflecting the genetic diversity within the community when novel sequences are encountered [@Westcott2015].
 In a third approach, developers have compared the number of OTUs generated by various methods for a common dataset [@Edgar2013; @Kopylova2016].
-Again, one could randomly cluster sequences into a target number of OTUs, but the clusterings would likely be meaningless.
+Although methods need to guard against excessive splitting of sequences across OTUs, by focusing on minimizing the number of OTUs in a community developers excessively lumping sequences that do not belong together.
 In a fourth approach a metric of OTU stability has been proposed as a way to assess algorithms [@He2015].
@@
 It is also important to note that the MCC values for the closed-reference OTUs are inflated because sequences were removed from the analysis if there was not a reference sequence that was more than 97% similar to the sequence.
+By choosing to focus on the ability to regenerate taxonomic clusterings, minimizing the number of OTUs, and performance Kopylova et al [-@Kopylova2016] concluded that Swarm and SUMACLUST had the most consistent performance among the *de novo* methods.
+My objective MCC-based approach found that Sumaclust performed well, but was matched or outperformed by the average neighbor algorithm; using a 3% threshold, Swarm was actually one of the worst methods.
 Given the consistent quality of the clusterings formed by the average neighbor algorithm, these results confirm the conclusion from the previous analysis that researchers should use the average neighbor algorithm or calculate MCC values for several methods and use the clustering that gives the best MCC value [@Westcott2015].

### 2016-03-25 code/run_rand_nref.sh
Modify NINJA-OPS-related code to use v1.5

@@ -11,12 +11,13 @@ mkdir -p $DB_FOLDER
 
 cp data/references/97_otus.taxonomy $DB_FOLDER/$DB_NAME.taxonomy
 
-code/NINJA-OPS/bin/ninja_prep_linux $REF $DB_FOLDER/out_preDB.fa $DB_FOLDER/$DB_NAME.db
-bowtie2-build $DB_FOLDER/out_preDB.fa $DB_FOLDER/$DB_NAME
+#code/NINJA-OPS/bin/ninja_prep_linux $REF $DB_FOLDER/out_preDB $DB_FOLDER/$DB_NAME
+code/NINJA-OPS/bin/ninja_prep_linux $REF $DB_FOLDER/$DB_NAME
+bowtie2-build $DB_FOLDER/$DB_NAME.fa $DB_FOLDER/$DB_NAME
 
 
 # The clustering...
-NINJA_FOLDER=$(echo $REF | sed 's/fasta/ninja/')
+NINJA_FOLDER=$(echo $REF | sed 's/fasta/nclosed/')
 
 rm -rf $NINJA_FOLDER
 mkdir $NINJA_FOLDER
@@ -27,10 +28,9 @@ bowtie2-align-s --no-head -x $DB_FOLDER/$DB_NAME -S $NINJA_FOLDER/alignments.txt
 
 code/NINJA-OPS/bin/ninja_parse_filtered_linux $NINJA_FOLDER/ninja $NINJA_FOLDER/alignments.txt $DB_FOLDER/$DB_NAME.db $DB_FOLDER/$DB_NAME.taxonomy --legacy LOG
 
-# something is weird about the indices that NINJA-OPS is returning in the log
-# file. taxonomically they're way off and I can't get them to match with what is
-# in the alignments file. going to have to go this alone...
-R -e "source('code/fix_ninja_indexing.R');fix_ninja_indexing('$NINJA_FOLDER', '$DB_FOLDER')"
+
+mv $NINJA_FOLDER/ninja_otumap.txt $NINJA_FOLDER.nc
+
 
 # Cleaning up
 rm -rf $DB_FOLDER $NINJA_FOLDER

### 2016-03-25 code/summarize_rand_ref.R

@@ -19,13 +19,10 @@ parse_sc_line <- function(line){
 get_hits <- function(cluster_file, method){
 	hits <- as.numeric()
 
-	if(method == 's'){
+	if(method == 's' || method == 'n'){
 		cluster_data <- scan(file=cluster_file, what=character(), sep='\n', quiet=T)
 		listing <- lapply(cluster_data, parse_sc_line)
 		hits <- unlist(listing)
-	} else if(method == 'n'){
-		cluster_data <- read.table(file=cluster_file, stringsAsFactors=FALSE)
-		hits <- cluster_data$V2		
 	} else {
 		cluster_data <- read.table(file=cluster_file, stringsAsFactors=FALSE)
 		hits <- cluster_data[cluster_data$V1 == "H", 10]

### 2016-03-26 code/closed_ref_analysis.R
Modify NINJA-OPS-related code to use v1.5

@@ -41,13 +41,7 @@ get_sens_spec_data <- function(search_file, true_mapping, duplicate_lookup){
 		obs_map <- as.character(clusters$V10)
 		names(obs_map) <- gsub("-\\d*;size=.*;", "", clusters$V9)
 		names(obs_map) <- gsub('-', "_", names(obs_map))
-	} else if(tag == "nc") {
-		clusters <- read.table(file=search_file, stringsAsFactors=F)
-		obs_map <- as.character(clusters$V2)
-		names(obs_map) <- clusters$V1
-		names(obs_map) <- gsub('-', "_", names(obs_map))
-		names(obs_map) <- gsub('_[^_]*$', "", names(obs_map))
-	} else if(tag == "sc") {
+	} else if(tag == "sc" || tag == "nc") {
 		clusters <- scan(file=search_file, what=character(), sep='\n', quiet=T)
 		listing <- lapply(clusters, parse_sc_line)
 		obs_map <- unlist(listing)

### 2016-03-26 papers/msystems_2016/Schloss_Commentary_mSystems_2016.Rmd
Editing text

@@
 The recent analysis of Kopylova et al [-@Kopylova2016] repeated many of the benchmarking strategies employed by previous researchers.
-Many algorithm developers have clustered sequences from simulated communities or sequencing data from synthetic communities of cultured organisms and quantify how well the OTU assignments matched the organisms' taxonomy [@Edgar2013; @Mah2014; @Mah2015; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014; @Cai2011; @Sun2011; @White2010; @AlGhalith2016].
+Many algorithm developers have clustered sequences from simulated communities or sequencing data from synthetic communities of cultured organisms and quantified how well the OTU assignments matched the organisms' taxonomy [@Edgar2013; @Mah2014; @Mah2015; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014; @Cai2011; @Sun2011; @White2010; @AlGhalith2016].
 Although an OTU definition would ideally match bacterial taxonomy, bacterial taxonomy has proven itself to be fluid and reflect the biases of various research interests.
@@
 In a third approach, developers have compared the number of OTUs generated by various methods for a common dataset [@Edgar2013; @Kopylova2016].
-Although methods need to guard against excessive splitting of sequences across OTUs, by focusing on minimizing the number of OTUs in a community developers excessively lumping sequences that do not belong together.
-In a fourth approach a metric of OTU stability has been proposed as a way to assess algorithms [@He2015].
+Although methods need to guard against excessive splitting of sequences across OTUs, by focusing on minimizing the number of OTUs in a community developers risk excessively lumping sequences together that are not similar.
+In a fourth approach, a metric of OTU stability has been proposed as a way to assess algorithms [@He2015].
 Although it is important that the methods generate reproducible OTU assignments when the initial order of the sequences is randomized, this metric ignores the possibility that the variation in assignments may be equally robust or that the assignments by a highly reproducible algorithm may be quite poor.
@@
 The USEARCH and SortMeRNA (v.2.0) closed-reference clusterings were performed using QIIME (v.1.9.1) [@Kopylova2012; @Caporaso2010].
-Closed-reference clustering was also performed using VSEARCH (v.1.5.0) and NINJA-OPS (v.1.3.2) [@AlGhalith2016].
+Closed-reference clustering was also performed using VSEARCH (v.1.5.0) and NINJA-OPS (v.1.5.0) [@AlGhalith2016].
 The order of the sequences in each dataset was randomized thirty times and the intra-method range in MCC values was smaller than the plotting symbol.

### 2016-03-26 papers/msystems_2016/Schloss_Commentary_mSystems_2016.Rmd
Keep TeX output file to conversion to docx

@@
   pdf_document:
+    keep_tex: true
     includes:

### 2016-03-26 papers/msystems_2016/Schloss_Commentary_mSystems_2016.tex
Keep TeX output file to conversion to docx

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Application of database-independent approach to assess the quality of OTU picking methods},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{\textbf{Application of database-independent approach to assess the
quality of OTU picking methods}}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}
\usepackage{comment}

% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\begin{document}
\maketitle

\begin{center}
\vspace{25mm}
Patrick D. Schloss${^\dagger}$

\vspace{30mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
\end{center}

\newpage

\linenumbers

\subsection{Abstract}\label{abstract}

Assigning 16S rRNA gene sequences to operational taxonomic units (OTUs)
allows microbial ecologists to overcome the inconsistencies and biases
within bacterial taxonomy and provides a strategy for clustering similar
sequences that do not have representatives in a reference database. I
have applied the Matthew's correlation coefficient to assess the ability
of 15 reference-independent and -dependent clustering algorithms to
assign sequences to OTUs. This metric quantifies the ability of an
algorithm to reflect the relationships between sequences without the use
of a reference and can be applied to any dataset or method. The most
consistently robust method was the average neighbor algorithm; however,
for some datasets other algorithms matched its performance.

\newpage

Numerous algorithms have been developed for solving the seemingly simple
problem of assigning 16S rRNA gene sequences to operational taxonomic
units (OTUs). These algorithms were recently the subject of benchmarking
studies performed by Westcott and myself (1, 2), He et al (3), and
Kopylova et al (4). These studies provide a thorough review of the
sequencing clustering landscape, which can be divided into three general
approaches: (i) \emph{de novo} clustering where sequences are clustered
without first mapping sequences to a reference database, (ii)
closed-reference clustering where sequences are clustered based on the
references that the sequences map to, and (iii) open reference
clustering where sequences that do not map adequately to the reference
are then clustered using a \emph{de novo} approach. Assessing the
quality of the clustering assignments has been a persistent problem in
the development of clustering algorithms.

The recent analysis of Kopylova et al (4) repeated many of the
benchmarking strategies employed by previous researchers. Many algorithm
developers have clustered sequences from simulated communities or
sequencing data from synthetic communities of cultured organisms and
quantified how well the OTU assignments matched the organisms' taxonomy
(5--16). Although an OTU definition would ideally match bacterial
taxonomy, bacterial taxonomy has proven itself to be fluid and reflect
the biases of various research interests. Furthermore, it is unclear how
the methods scale to sequences from the novel organisms we are likely to
encounter in deep sequencing surveys. In a second approach, developers
have compared the time and memory required to cluster sequences in a
dataset (6, 13, 17, 18). These are valid parameters to assess when
judging a clustering method, but indicate little regarding the quality
of the OTU assignments. For example, reference-based methods are very
efficient, but do a poor job of reflecting the genetic diversity within
the community when novel sequences are encountered (2). In a third
approach, developers have compared the number of OTUs generated by
various methods for a common dataset (4, 5). Although methods need to
guard against excessive splitting of sequences across OTUs, by focusing
on minimizing the number of OTUs in a community developers risk
excessively lumping sequences together that are not similar. In a fourth
approach, a metric of OTU stability has been proposed as a way to assess
algorithms (3). Although it is important that the methods generate
reproducible OTU assignments when the initial order of the sequences is
randomized, this metric ignores the possibility that the variation in
assignments may be equally robust or that the assignments by a highly
reproducible algorithm may be quite poor. In a final approach, some
developers have assessed the quality of clustering based on the method's
ability to generate the same OTUs generated by other methods (18, 19).
Unfortunately, without the ability to ground truth any method, such
comparisons are tenuous. There is a need for an objective metric to
assess the quality of OTU assignments.

Westcott and I have proposed an unbiased and objective method for
assessing the quality of OTU assignments that can be applied to any
collection of sequences (1, 2). Our approach uses the observed
dissimilarity between pairs of sequences and information about whether
sequences were clustered together to quantify how well similar sequences
are clustered together and dissimilar sequences are clustered apart. To
quantify the correlation between the observed and expected OTU
assignments, we synthesize the relationship between OTU assignments and
the distances between sequences using the Matthew's correlation
coefficient (20). I have expanded our previous analysis to evaluate
three hierarchical and seven greedy \emph{de novo} algorithms, one
open-reference clustering algorithm, and four closed-reference
algorithms (Figure 1). To test these approaches I applied each of them
to datasets from soil (21), mouse feces (22), and two simulated
datasets. The simulated communities were generated by randomly selecting
10,000 16S rRNA sequences that were unique within the V4 region from the
SILVA non-redundant database (4, 23). Next, an even community was
generated by specifying that each sequence had a frequency of 100 reads
and a staggered community was generated by specifying that the abundance
of each sequence was a randomly drawn a uniform distribution between 1
and 200. A reproducible version of this manuscript and analysis has been
added to the repository available at
\url{https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015}.

I replicated the benchmarking approach that I have used previously to
assess the ability of an algorithm to correctly group sequences that are
similar to each other and split sequences that are dissimilar to each
other using the MCC (1, 2). When I compared the MCC values calculated
using the ten \emph{de novo} algorithms with the four datasets, the
average neighbor algorithm reliably performed as well or better than the
other methods (Figure 1). For the murine dataset, the MCC values for the
VSEARCH (AGC: 0.76 and DGC: 0.78) and USEARCH-based (AGC: 0.76 and DGC:
0.77) algorithms, Sumaclust (0.76), and average neighbor (0.76) were
similarly high. For each of the other datasets, the MCC value for the
average neighbor algorithm was at least 5\% higher than the next best
method. Swarm does not use a traditional distance-based criteria to
cluster sequences into OTUs and instead looks for natural subnetworks in
the data. When I used the distance threshold that gave the best MCC
value for the Swarm data, the MCC values were generally not as high as
they were using the average neighbor algorithm. The one exception was
for the soil dataset. Among the reference-based methods, all of the MCC
values suffer because when sequences that are at least 97\% similar to a
reference are pooled, the sequences within an OTU could be as much as
6\% different from each other. The effect of this is observed in the MCC
values that were calculated for the OTUs assigned by these methods
generally being lower than those observed using the \emph{de novo}
approaches (Figure 1). It is also important to note that the MCC values
for the closed-reference OTUs are inflated because sequences were
removed from the analysis if there was not a reference sequence that was
more than 97\% similar to the sequence. By choosing to focus on the
ability to regenerate taxonomic clusterings, minimizing the number of
OTUs, and performance Kopylova et al (4) concluded that Swarm and
SUMACLUST had the most consistent performance among the \emph{de novo}
methods. My objective MCC-based approach found that Sumaclust performed
well, but was matched or outperformed by the average neighbor algorithm;
using a 3\% threshold, Swarm was actually one of the worst methods.
Given the consistent quality of the clusterings formed by the average
neighbor algorithm, these results confirm the conclusion from the
previous analysis that researchers should use the average neighbor
algorithm or calculate MCC values for several methods and use the
clustering that gives the best MCC value (2).

Next, I investigated the ability of the reference-based methods to
properly assign sequences to OTUs. The full-length 16S rRNA gene
sequences in the default reference taxonomy that accompanies QIIME are
less than 97\% similar to each other. Within the V4 region, however,
many of the sequences were more similar to each other and even identical
to each other. As a result, we previously found that there was a
dependence between the ordering of sequences in the reference database
and the OTU assignments with USEARCH and VSEARCH (2). To explore this
further, we analyzed the 32,106 unique sequences from the murine dataset
with randomized databases. VSEARCH always found matches for 27,737
murine sequences, the reference matched to those sequences differed
between randomizations. For USEARCH there were between 28,007 and 28,111
matches depending on the order of the reference. In the updated analysis
we found that SortMeRNA resulted in between 23,912 and 28,464 matches.
Using NINJA-OPS with different orderings of the reference sequences
generated the same 28,499 matches. These results point to an additional
problem with closed-reference clustering, which is the inability for the
method to assign sequences to OTUs when a similar reference sequence
does not exist in the database. For the well-characterized murine
microbiota, NINJA-OPS did the best by finding relatives for 88.8\% of
the unique murine sequences. As indicated by the variation in the number
of sequences that matched a reference sequence, these methods varied in
their sensitivity and specificity to find the best reference sequence.
Of the closed-reference methods, NINJA-OPS had the best sensitivity
(99.7\%) and specificity (79.7\%) while SortMeRNA had the worst
sensitivity (95.7\%) and VSEARCH had the worst specificity (60.3\%).
Reference-based clustering algorithms are much faster than \emph{de
novo} approaches, but do not generate OTUs that are as robust.

Although the goal of Kopylova et al (4) was to compare various
clustering algorithms, they also studied these algorithms in the broader
context of raw sequence processing, screening for chimeras, and removal
of singletons. Each of these are critical decisions in a comprehensive
pipeline. By including these steps, they confounded their analysis of
how best to cluster sequences into OTUs. The effect of differences in
MCC values on one's ability to draw inferences is unclear and admittedly
may be relatively minor for some datasets. Because of this uncertainty,
researchers should use the most reliable methods available in case the
differences in clustering do effect the conclusions that can be drawn
from a particular dataset. Through the use of objective criteria that
measure the quality of the clusterings, independent of taxonomy or
database, researchers will be able to evaluate which clustering
algorithm is the best fit for their data.

\newpage

\textbf{Figure 1. Comparison of OTU quality generated by multiple
algorithms applied to four datasets.} The nearest, average, and furthest
neighbor clustering algorithms were used as implemented in mothur
(v.1.37)(24). Abundance (AGC) and Distance-based greedy clustering (DGC)
were implemented using USEARCH (v.6.1) and VSEARCH (v.1.5.0)(3, 5, 25).
Other \emph{de novo} clustering algorithms included Swarm (v.2.1.1)(6,
7), OTUCLUST (v.0.1)(26), and Sumaclust (v.1.0.20). The MCC values for
Swarm were determined by selecting the distance threshold that generated
the maximum MCC value for each dataset. The USEARCH and SortMeRNA
(v.2.0) closed-reference clusterings were performed using QIIME
(v.1.9.1) (27, 28). Closed-reference clustering was also performed using
VSEARCH (v.1.5.0) and NINJA-OPS (v.1.5.0) (16). The order of the
sequences in each dataset was randomized thirty times and the
intra-method range in MCC values was smaller than the plotting symbol.
MCC values were calculated using mothur.

\newpage

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\hypertarget{ref-Schloss2011Assessing}{}
1. \textbf{Schloss PD}, \textbf{Westcott SL}. 2011. Assessing and
improving methods used in operational taxonomic unit-based approaches
for 16S rRNA gene sequence analysis. Applied and Environmental
Microbiology \textbf{77}:3219--3226.
doi:\href{https://doi.org/10.1128/aem.02810-10}{10.1128/aem.02810-10}.

\hypertarget{ref-Westcott2015}{}
2. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
methods outperform reference-based methods for assigning 16S rRNA gene
sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.

\hypertarget{ref-He2015}{}
3. \textbf{He Y}, \textbf{Caporaso JG}, \textbf{Jiang X-T},
\textbf{Sheng H-F}, \textbf{Huse SM}, \textbf{Rideout JR}, \textbf{Edgar
RC}, \textbf{Kopylova E}, \textbf{Walters WA}, \textbf{Knight R},
\textbf{Zhou H-W}. 2015. Stability of operational taxonomic units: An
important but neglected property for analyzing microbial diversity.
Microbiome \textbf{3}.
doi:\href{https://doi.org/10.1186/s40168-015-0081-x}{10.1186/s40168-015-0081-x}.

\hypertarget{ref-Kopylova2016}{}
4. \textbf{Kopylova E}, \textbf{Navas-Molina JA}, \textbf{Mercier C},
\textbf{Xu ZZ}, \textbf{Mahé F}, \textbf{He Y}, \textbf{Zhou H-W},
\textbf{Rognes T}, \textbf{Caporaso JG}, \textbf{Knight R}. 2016.
Open-source sequence clustering methods improve the state of the art.
mSystems \textbf{1}:e00003--15.
doi:\href{https://doi.org/10.1128/msystems.00003-15}{10.1128/msystems.00003-15}.

\hypertarget{ref-Edgar2013}{}
5. \textbf{Edgar RC}. 2013. UPARSE: Highly accurate OTU sequences from
microbial amplicon reads. Nature Methods \textbf{10}:996--998.
doi:\href{https://doi.org/10.1038/nmeth.2604}{10.1038/nmeth.2604}.

\hypertarget{ref-Mah2014}{}
6. \textbf{Mahé F}, \textbf{Rognes T}, \textbf{Quince C}, \textbf{Vargas
C de}, \textbf{Dunthorn M}. 2014. Swarm: Robust and fast clustering
method for amplicon-based studies. PeerJ \textbf{2}:e593.
doi:\href{https://doi.org/10.7717/peerj.593}{10.7717/peerj.593}.

\hypertarget{ref-Mah2015}{}
7. \textbf{Mahé F}, \textbf{Rognes T}, \textbf{Quince C}, \textbf{Vargas
C de}, \textbf{Dunthorn M}. 2015. Swarm v2: Highly-scalable and
high-resolution amplicon clustering. PeerJ \textbf{3}:e1420.
doi:\href{https://doi.org/10.7717/peerj.1420}{10.7717/peerj.1420}.

\hypertarget{ref-Barriuso2011}{}
8. \textbf{Barriuso J}, \textbf{Valverde JR}, \textbf{Mellado RP}. 2011.
Estimation of bacterial diversity using next generation sequencing of
16S rDNA: A comparison of different workflows. BMC Bioinformatics
\textbf{12}:473.
doi:\href{https://doi.org/10.1186/1471-2105-12-473}{10.1186/1471-2105-12-473}.

\hypertarget{ref-Bonder2012}{}
9. \textbf{Bonder MJ}, \textbf{Abeln S}, \textbf{Zaura E},
\textbf{Brandt BW}. 2012. Comparing clustering and pre-processing in
taxonomy analysis. Bioinformatics \textbf{28}:2891--2897.
doi:\href{https://doi.org/10.1093/bioinformatics/bts552}{10.1093/bioinformatics/bts552}.

\hypertarget{ref-Chen2013}{}
10. \textbf{Chen W}, \textbf{Zhang CK}, \textbf{Cheng Y}, \textbf{Zhang
S}, \textbf{Zhao H}. 2013. A comparison of methods for clustering 16S
rRNA sequences into OTUs. PLoS ONE \textbf{8}:e70837.
doi:\href{https://doi.org/10.1371/journal.pone.0070837}{10.1371/journal.pone.0070837}.

\hypertarget{ref-Huse2010}{}
11. \textbf{Huse SM}, \textbf{Welch DM}, \textbf{Morrison HG},
\textbf{Sogin ML}. 2010. Ironing out the wrinkles in the rare biosphere
through improved OTU clustering. Environmental Microbiology
\textbf{12}:1889--1898.
doi:\href{https://doi.org/10.1111/j.1462-2920.2010.02193.x}{10.1111/j.1462-2920.2010.02193.x}.

\hypertarget{ref-May2014}{}
12. \textbf{May A}, \textbf{Abeln S}, \textbf{Crielaard W},
\textbf{Heringa J}, \textbf{Brandt BW}. 2014. Unraveling the outcome of
16S rDNA-based taxonomy analysis through mock data and simulations.
Bioinformatics \textbf{30}:1530--1538.
doi:\href{https://doi.org/10.1093/bioinformatics/btu085}{10.1093/bioinformatics/btu085}.

\hypertarget{ref-Cai2011}{}
13. \textbf{Cai Y}, \textbf{Sun Y}. 2011. ESPRIT-tree: Hierarchical
clustering analysis of millions of 16S rRNA pyrosequences in quasilinear
computational time. Nucleic Acids Research \textbf{39}:e95--e95.
doi:\href{https://doi.org/10.1093/nar/gkr349}{10.1093/nar/gkr349}.

\hypertarget{ref-Sun2011}{}
14. \textbf{Sun Y}, \textbf{Cai Y}, \textbf{Huse SM}, \textbf{Knight R},
\textbf{Farmerie WG}, \textbf{Wang X}, \textbf{Mai V}. 2011. A
large-scale benchmark study of existing algorithms for
taxonomy-independent microbial community analysis. Briefings in
Bioinformatics \textbf{13}:107--121.
doi:\href{https://doi.org/10.1093/bib/bbr009}{10.1093/bib/bbr009}.

\hypertarget{ref-White2010}{}
15. \textbf{White JR}, \textbf{Navlakha S}, \textbf{Nagarajan N},
\textbf{Ghodsi M-R}, \textbf{Kingsford C}, \textbf{Pop M}. 2010.
Alignment and clustering of phylogenetic markers - implications for
microbial diversity studies. BMC Bioinformatics \textbf{11}:152.
doi:\href{https://doi.org/10.1186/1471-2105-11-152}{10.1186/1471-2105-11-152}.

\hypertarget{ref-AlGhalith2016}{}
16. \textbf{Al-Ghalith GA}, \textbf{Montassier E}, \textbf{Ward HN},
\textbf{Knights D}. 2016. NINJA-OPS: Fast accurate marker gene alignment
using concatenated ribosomes. PLOS Computational Biology
\textbf{12}:e1004658.
doi:\href{https://doi.org/10.1371/journal.pcbi.1004658}{10.1371/journal.pcbi.1004658}.

\hypertarget{ref-Sun2009}{}
17. \textbf{Sun Y}, \textbf{Cai Y}, \textbf{Liu L}, \textbf{Yu F},
\textbf{Farrell ML}, \textbf{McKendree W}, \textbf{Farmerie W}. 2009.
ESPRIT: Estimating species richness using large collections of 16S rRNA
pyrosequences. Nucleic Acids Research \textbf{37}:e76--e76.
doi:\href{https://doi.org/10.1093/nar/gkp285}{10.1093/nar/gkp285}.

\hypertarget{ref-Rideout2014}{}
18. \textbf{Rideout JR}, \textbf{He Y}, \textbf{Navas-Molina JA},
\textbf{Walters WA}, \textbf{Ursell LK}, \textbf{Gibbons SM},
\textbf{Chase J}, \textbf{McDonald D}, \textbf{Gonzalez A},
\textbf{Robbins-Pianka A}, \textbf{Clemente JC}, \textbf{Gilbert JA},
\textbf{Huse SM}, \textbf{Zhou H-W}, \textbf{Knight R}, \textbf{Caporaso
JG}. 2014. Subsampled open-reference clustering creates consistent,
comprehensive OTU definitions and scales to billions of sequences. PeerJ
\textbf{2}:e545.
doi:\href{https://doi.org/10.7717/peerj.545}{10.7717/peerj.545}.

\hypertarget{ref-Schmidt2014Limits}{}
19. \textbf{Schmidt TSB}, \textbf{Rodrigues JFM}, \textbf{Mering C von}.
2014. Limits to robustness and reproducibility in the demarcation of
operational taxonomic units. Environ Microbiol \textbf{17}:1689--1706.
doi:\href{https://doi.org/10.1111/1462-2920.12610}{10.1111/1462-2920.12610}.

\hypertarget{ref-Matthews1975}{}
20. \textbf{Matthews B}. 1975. Comparison of the predicted and observed
secondary structure of t4 phage lysozyme. Biochimica et Biophysica Acta
(BBA) - Protein Structure \textbf{405}:442--451.
doi:\href{https://doi.org/10.1016/0005-2795(75)90109-9}{10.1016/0005-2795(75)90109-9}.

\hypertarget{ref-Roesch2007}{}
21. \textbf{Roesch LFW}, \textbf{Fulthorpe RR}, \textbf{Riva A},
\textbf{Casella G}, \textbf{Hadwin AKM}, \textbf{Kent AD},
\textbf{Daroub SH}, \textbf{Camargo FAO}, \textbf{Farmerie WG},
\textbf{Triplett EW}. 2007. Pyrosequencing enumerates and contrasts soil
microbial diversity. The ISME Journal.
doi:\href{https://doi.org/10.1038/ismej.2007.53}{10.1038/ismej.2007.53}.

\hypertarget{ref-Schloss2012Stabilization}{}
22. \textbf{Schloss PD}, \textbf{Schubert AM}, \textbf{Zackular JP},
\textbf{Iverson KD}, \textbf{Young VB}, \textbf{Petrosino JF}. 2012.
Stabilization of the murine gut microbiome following weaning. Gut
Microbes \textbf{3}:383--393.
doi:\href{https://doi.org/10.4161/gmic.21008}{10.4161/gmic.21008}.

\hypertarget{ref-Pruesse2007}{}
23. \textbf{Pruesse E}, \textbf{Quast C}, \textbf{Knittel K},
\textbf{Fuchs BM}, \textbf{Ludwig W}, \textbf{Peplies J},
\textbf{Glockner FO}. 2007. SILVA: A comprehensive online resource for
quality checked and aligned ribosomal RNA sequence data compatible with
ARB. Nucleic Acids Research \textbf{35}:7188--7196.
doi:\href{https://doi.org/10.1093/nar/gkm864}{10.1093/nar/gkm864}.

\hypertarget{ref-Schloss2009}{}
24. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology
\textbf{75}:7537--7541.
doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.

\hypertarget{ref-Rognes2015}{}
25. \textbf{Rognes T}, \textbf{Mahé F}, \textbf{Flouri T},
\textbf{McDonald; D}. 2015. Vsearch: VSEARCH 1.4.0. Zenodo.

\hypertarget{ref-Albanese2015}{}
26. \textbf{Albanese D}, \textbf{Fontana P}, \textbf{Filippo CD},
\textbf{Cavalieri D}, \textbf{Donati C}. 2015. MICCA: A complete and
accurate software for taxonomic profiling of metagenomic data. Sci Rep
\textbf{5}:9743.
doi:\href{https://doi.org/10.1038/srep09743}{10.1038/srep09743}.

\hypertarget{ref-Kopylova2012}{}
27. \textbf{Kopylova E}, \textbf{Noe L}, \textbf{Touzet H}. 2012.
SortMeRNA: Fast and accurate filtering of ribosomal RNAs in
metatranscriptomic data. Bioinformatics \textbf{28}:3211--3217.
doi:\href{https://doi.org/10.1093/bioinformatics/bts611}{10.1093/bioinformatics/bts611}.

\hypertarget{ref-Caporaso2010}{}
28. \textbf{Caporaso JG}, \textbf{Kuczynski J}, \textbf{Stombaugh J},
\textbf{Bittinger K}, \textbf{Bushman FD}, \textbf{Costello EK},
\textbf{Fierer N}, \textbf{Peña AG}, \textbf{Goodrich JK},
\textbf{Gordon JI}, \textbf{Huttley GA}, \textbf{Kelley ST},
\textbf{Knights D}, \textbf{Koenig JE}, \textbf{Ley RE},
\textbf{Lozupone CA}, \textbf{McDonald D}, \textbf{Muegge BD},
\textbf{Pirrung M}, \textbf{Reeder J}, \textbf{Sevinsky JR},
\textbf{Turnbaugh PJ}, \textbf{Walters WA}, \textbf{Widmann J},
\textbf{Yatsunenko T}, \textbf{Zaneveld J}, \textbf{Knight R}. 2010.
QIIME allows analysis of high-throughput community sequencing data.
Nature Methods \textbf{7}:335--336.
doi:\href{https://doi.org/10.1038/nmeth.f.303}{10.1038/nmeth.f.303}.

\end{document}
